---
title: "SCFA Analysis: Abbott Carbohydrate Obesity Project"
author: "Abbott Research Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    theme: united
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      fig.width = 12, fig.height = 8, dpi = 300)
```

# Introduction

This document presents the analysis of short-chain fatty acids (SCFAs) in the Abbott carbohydrate obesity project. The analysis examines how SCFA analytes change between experimental groups, carbohydrate types, and time points.

# Methods

## Metabolomics Overview

The fecal metabolome was analyzed using targeted metabolomics approaches. The DFI Host-Microbe Metabolomics Facility (DFI-HMMF) analyzed fecal material using validated methods and analysis pipelines. All compounds were validated through retention time and fragmentation comparison to standards and available databases.

## SCFA Analysis using PFBBr Panel

Short chain fatty acids were analyzed using Gas chromatography-mass spectrometry (GC-MS) following derivatization with pentafluorobenzyl bromide (PFBBr). SCFAs (acetate, butyrate, propionate) were quantitatively analyzed following PFB derivatization and detection by negative collision induced gas chromatography-mass spectrometry ((-)-CI-GC-MS, Agilent 8890). Additional compounds including 5-aminovalerate and succinate were also quantified.

### Detailed SCFA Analysis Protocol

Short chain fatty acids were derivatized as described by Haak et al. with modifications. The metabolite extract (100 µL) was added to 100 µL of 100 mM borate buffer (pH 10), 400 µL of 100 mM pentafluorobenzyl bromide in Acetonitrile, and 400 µL of n-hexane in a capped mass spec autosampler vial. Samples were heated to 65°C for 1 hour while shaking at 1300 rpm. After cooling, samples were centrifuged at 4°C, 2000 x g for 5 min, allowing phase separation. The hexanes phase was transferred and analyzed.

Samples were analyzed using a GC-MS (Agilent 7890A GC system, Agilent 5975C MS detector) operating in negative chemical ionization mode, using a HP-5MSUI column (30 m x 0.25 mm, 0.25 µm), methane as the reagent gas and 1 µL split injection (1:10 split ratio). A 10-point calibration curve was prepared with acetate (100 mM), propionate (25 mM), butyrate (12.5 mM), and succinate (50 mM), with 9 subsequent 2x serial dilutions.

### Sample Extraction

Extraction solvent (80% methanol spiked with internal standards and stored at -80°C) was added at a ratio of 100 mg of material/mL of extraction solvent. Samples were homogenized at 4°C on a Bead Mill 24 Homogenizer, set at 1.6 m/s with 6 thirty-second cycles, 5 seconds off per cycle. Samples were then centrifuged at -10°C, 20,000 x g for 15 min and the supernatant was used for analysis.

# Data Analysis

```{r load-libraries}
library(tidyverse)
library(readxl)
library(janitor)
library(data.table)
library(ggpubr)
library(rstatix)
library(viridis)
library(RColorBrewer)
library(lme4)
library(lmerTest)
```

## Load Metadata

```{r load-metadata}
# Read in first metadata file
map <- read_excel("metadata/SEED LAB HMM Submission Sheet 202405.xlsx", skip = 9) %>% 
  clean_names()

# Replace column names with cleaned names
colnames(map) <- c("sampleid", "group", "bile_acids_panel", "pfbbr_scfa_panel", 
                   "tryp_indole_panel", "tms_panel_1", "tms_panel_2", "tms_panel_3", 
                   "untargeted_panel", "custom_targeted_panel", 
                   "low_abundant_metabolite_panel", "sample_origin", "sample_type", 
                   "mouse_number", "services_requested_per_sample", "mouse_strain", 
                   "mouse_flora", "volume_ul", "sample_weight_mg", "experiment_treatment")

# Clean metadata
map <- map %>% 
  filter(!is.na(sampleid)) %>%
  filter(sampleid != "") %>%
  slice(-(1:3))
```

```{r parse-sample-names}
# Extract information from sample names
map <- map %>%
  mutate(
    # Extract subject ID (digits and letter before first space)
    subject = str_extract(sampleid, "^[0-9]+[a-zA-Z]*"),
    
    # Extract the part between spaces (e.g., "r1", "s2", "nc")
    carb_repeat = str_extract(sampleid, "(?<= )[rs]\\d|nc(?= )"),
    
    # Create carbohydrate_type column
    carbohydrate_type = case_when(
      str_detect(carb_repeat, "^r") ~ "rapid_digestible",
      str_detect(carb_repeat, "^s") ~ "slow_digestible", 
      str_detect(carb_repeat, "^nc") ~ "no_carbohydrate",
      TRUE ~ NA_character_
    ),
    
    # Extract technical repeat number (handle "nc" case where there's no number)
    technical_repeat = case_when(
      str_detect(carb_repeat, "^nc") ~ 1L,  # nc gets repeat 1
      TRUE ~ as.integer(str_extract(carb_repeat, "\\d+"))
    ),
    
    # Extract timepoint (after second space, before "h")
    timepoint_hr = as.integer(str_extract(sampleid, "(?<= )\\d+(?=h$)"))
  ) %>%
  # Remove the temporary carb_repeat column and select relevant columns
  select(-carb_repeat) %>%
  select(sampleid, group, subject, carbohydrate_type, technical_repeat, timepoint_hr)
```

```{r load-additional-metadata}
# Load second metadata file
map2 <- read_excel("metadata/PS2720_xtra_samples_HMM Submission Sheet.xlsx", skip = 9) %>% 
  clean_names()

# Apply same processing as first file
colnames(map2) <- c("sampleid", "group", "bile_acids_panel", "pfbbr_scfa_panel", 
                    "tryp_indole_panel", "tms_panel_1", "tms_panel_2", "tms_panel_3", 
                    "untargeted_panel", "custom_targeted_panel", 
                    "low_abundant_metabolite_panel", "sample_origin", "sample_type", 
                    "mouse_number", "services_requested_per_sample", "mouse_strain", 
                    "mouse_flora", "volume_ul", "sample_weight_mg", "experiment_treatment")

map2 <- map2 %>% 
  filter(!is.na(sampleid)) %>%
  filter(sampleid != "") %>%
  slice(-(1:3)) %>%
  mutate(
    # Extract subject ID (digits and letter before first space)
    subject = str_extract(sampleid, "^[0-9]+[a-zA-Z]*"),
    
    carb_repeat = str_extract(sampleid, "(?<= )[rs]\\d|nc(?= )"),
    carbohydrate_type = case_when(
      str_detect(carb_repeat, "^r") ~ "rapid_digestible",
      str_detect(carb_repeat, "^s") ~ "slow_digestible", 
      str_detect(carb_repeat, "^nc") ~ "no_carbohydrate",
      TRUE ~ NA_character_
    ),
    technical_repeat = case_when(
      str_detect(carb_repeat, "^nc") ~ 1L,
      TRUE ~ as.integer(str_extract(carb_repeat, "\\d+"))
    ),
    timepoint_hr = as.integer(str_extract(sampleid, "(?<= )\\d+(?=h$)"))
  ) %>%
  select(-carb_repeat) %>%
  select(sampleid, group, subject, carbohydrate_type, technical_repeat, timepoint_hr)

# Combine metadata files
map <- bind_rows(map, map2)

# Make group names lowercase for consistency
map <- map %>%
  mutate(group = tolower(group))
```

## Load SCFA Data

```{r load-data}
# Read in SCFA quantification data
data <- fread("data/removed_qcs_quant_results_20250722_PFBBr_PS2720_20250731.csv")

# Join metadata with SCFA data
sample_data <- left_join(map, data, by = "sampleid")

# Ensure group names are lowercase
sample_data <- sample_data %>%
  mutate(group = tolower(group))

# Display data structure
cat("Dataset dimensions:", dim(sample_data), "\n")
cat("Sample groups:", unique(sample_data$group), "\n")
cat("Carbohydrate types:", unique(sample_data$carbohydrate_type), "\n")
cat("Time points:", unique(sample_data$timepoint_hr), "\n")
```

# Summary Statistics

```{r define-analytes}
# Define SCFA analytes (excluding non-SCFA compounds)
scfa_analytes <- c("acetate", "butyrate", "propionate", "5aminovalerate", "succinate")

# Convert data to long format and average technical replicates
sample_data_long <- sample_data %>%
  pivot_longer(cols = all_of(scfa_analytes), 
               names_to = "analyte", 
               values_to = "concentration") %>%
  filter(!is.na(concentration)) %>%
  # Average technical replicates FIRST before any statistical analysis
  group_by(subject, group, carbohydrate_type, timepoint_hr, analyte) %>%
  summarise(concentration = mean(concentration, na.rm = TRUE), .groups = "drop")

cat("Total observations after averaging technical replicates:", nrow(sample_data_long), "\n")
```

## Summary by Group

```{r summary-by-group}
summary_by_group <- sample_data_long %>%
  group_by(group, analyte) %>%
  summarise(
    n = n(),
    mean = mean(concentration, na.rm = TRUE),
    median = median(concentration, na.rm = TRUE),
    sd = sd(concentration, na.rm = TRUE),
    sem = sd / sqrt(n),
    q25 = quantile(concentration, 0.25, na.rm = TRUE),
    q75 = quantile(concentration, 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Order so control and case are adjacent for each analyte
  arrange(analyte, group)

knitr::kable(summary_by_group, digits = 3, 
             caption = "Summary Statistics by Experimental Group")
```

## Summary by Carbohydrate Type

```{r summary-by-carb}
summary_by_carb <- sample_data_long %>%
  group_by(carbohydrate_type, analyte) %>%
  summarise(
    n = n(),
    mean = mean(concentration, na.rm = TRUE),
    median = median(concentration, na.rm = TRUE),
    sd = sd(concentration, na.rm = TRUE),
    sem = sd / sqrt(n),
    q25 = quantile(concentration, 0.25, na.rm = TRUE),
    q75 = quantile(concentration, 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Order so the 3 carbohydrate types are in series with each analyte
  arrange(analyte, carbohydrate_type)

knitr::kable(summary_by_carb, digits = 3, 
             caption = "Summary Statistics by Carbohydrate Type")
```

## Summary by Time Point

```{r summary-by-time}
summary_by_time <- sample_data_long %>%
  group_by(timepoint_hr, analyte) %>%
  summarise(
    n = n(),
    mean = mean(concentration, na.rm = TRUE),
    median = median(concentration, na.rm = TRUE),
    sd = sd(concentration, na.rm = TRUE),
    sem = sd / sqrt(n),
    q25 = quantile(concentration, 0.25, na.rm = TRUE),
    q75 = quantile(concentration, 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Order so that 0h and 48h are adjacent per analyte
  arrange(analyte, timepoint_hr)

knitr::kable(summary_by_time, digits = 3, 
             caption = "Summary Statistics by Time Point")
```

## Combined Summary Statistics

```{r summary-combined}
summary_combined <- sample_data_long %>%
  group_by(group, carbohydrate_type, timepoint_hr, analyte) %>%
  summarise(
    n = n(),
    mean = mean(concentration, na.rm = TRUE),
    median = median(concentration, na.rm = TRUE),
    sd = sd(concentration, na.rm = TRUE),
    sem = sd / sqrt(n),
    q25 = quantile(concentration, 0.25, na.rm = TRUE),
    q75 = quantile(concentration, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

# Create results directory and save combined summary
dir.create("results", showWarnings = FALSE)
write.csv(summary_combined, "results/combined_summary_statistics.csv", row.names = FALSE)

cat("Combined summary table contains", nrow(summary_combined), "condition combinations\n")
cat("Combined summary saved to results/combined_summary_statistics.csv\n")
```

# Statistical Analysis

```{r statistical-analysis}
# Prepare data with proper factor levels for statistical testing
sample_data_long <- sample_data_long %>%
  mutate(
    group = factor(group, levels = c("control", "case")),
    carbohydrate_type = factor(carbohydrate_type, 
                              levels = c("no_carbohydrate", "rapid_digestible", "slow_digestible")),
    timepoint_hr = factor(timepoint_hr),
    subject = factor(subject)  # Add subject as factor for random effects
  )

# Statistical tests by group (t-tests)
group_stats <- sample_data_long %>%
  group_by(analyte) %>%
  t_test(concentration ~ group) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

# Statistical tests by carbohydrate type (ANOVA with post-hoc)
carb_stats <- sample_data_long %>%
  group_by(analyte) %>%
  anova_test(concentration ~ carbohydrate_type) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

# Post-hoc comparisons for carbohydrate type (vs no_carbohydrate)
carb_posthoc <- sample_data_long %>%
  group_by(analyte) %>%
  pairwise_t_test(concentration ~ carbohydrate_type, 
                  ref.group = "no_carbohydrate") %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

# Three-way ANOVA: Group x Carbohydrate x Time interaction
interaction_stats <- sample_data_long %>%
  group_by(analyte) %>%
  anova_test(concentration ~ group * carbohydrate_type * timepoint_hr) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

# Mixed-effects models accounting for subject as random effect
mixed_effects_results <- list()

for (analyte_name in scfa_analytes) {
  analyte_data <- sample_data_long %>% filter(analyte == analyte_name)
  
  # Full model with all interactions and subject as random intercept
  full_model <- lmer(concentration ~ group * carbohydrate_type * timepoint_hr + 
                     (1 | subject), 
                     data = analyte_data, 
                     REML = FALSE)
  
  # Model without three-way interaction
  no_three_way <- lmer(concentration ~ group + carbohydrate_type + timepoint_hr +
                       group:carbohydrate_type + group:timepoint_hr + 
                       carbohydrate_type:timepoint_hr + (1 | subject), 
                       data = analyte_data, 
                       REML = FALSE)
  
  # Model with only main effects
  main_effects <- lmer(concentration ~ group + carbohydrate_type + timepoint_hr + 
                       (1 | subject), 
                       data = analyte_data, 
                       REML = FALSE)
  
  # Store results
  mixed_effects_results[[analyte_name]] <- list(
    full_model = full_model,
    no_three_way = no_three_way,
    main_effects = main_effects,
    anova_full = anova(full_model),
    summary_full = summary(full_model)
  )
}

# Subject-level summary statistics
subject_summary <- sample_data_long %>%
  group_by(subject, group, analyte) %>%
  summarise(
    mean_conc = mean(concentration, na.rm = TRUE),
    n_observations = n(),
    .groups = "drop"
  ) %>%
  group_by(group, analyte) %>%
  summarise(
    n_subjects = n(),
    mean_subject_means = mean(mean_conc, na.rm = TRUE),
    sd_subject_means = sd(mean_conc, na.rm = TRUE),
    sem_subjects = sd_subject_means / sqrt(n_subjects),
    .groups = "drop"
  )

# Within-subject changes (baseline to 48h)
within_subject_changes <- sample_data_long %>%
  filter(timepoint_hr %in% c("0", "48")) %>%
  mutate(timepoint_hr = as.character(timepoint_hr)) %>%  # Convert back to character for pivot
  select(subject, group, carbohydrate_type, timepoint_hr, analyte, concentration) %>%
  pivot_wider(names_from = timepoint_hr, 
              values_from = concentration, 
              names_prefix = "time_") %>%
  mutate(change_0_to_48 = time_48 - time_0) %>%
  filter(!is.na(change_0_to_48))

# Statistical tests for within-subject changes
within_subject_stats <- within_subject_changes %>%
  group_by(group, carbohydrate_type, analyte) %>%
  summarise(
    n = n(),
    mean_change = mean(change_0_to_48, na.rm = TRUE),
    sd_change = sd(change_0_to_48, na.rm = TRUE),
    sem_change = sd_change / sqrt(n),
    t_stat = mean_change / sem_change,
    p_value = 2 * pt(-abs(t_stat), df = n - 1),  # two-tailed t-test
    .groups = "drop"
  ) %>%
  group_by(analyte) %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) %>%
  mutate(significance = case_when(
    p_adj < 0.001 ~ "***",
    p_adj < 0.01 ~ "**", 
    p_adj < 0.05 ~ "*",
    TRUE ~ "ns"
  ))

# Case-only temporal analysis (0h vs 48h)
case_only_changes <- sample_data_long %>%
  filter(group == "case", timepoint_hr %in% c("0", "48")) %>%
  mutate(timepoint_hr = as.character(timepoint_hr)) %>%
  pivot_wider(names_from = timepoint_hr, 
              values_from = concentration, 
              names_prefix = "time_") %>%
  mutate(change_0_to_48 = time_48 - time_0) %>%
  filter(!is.na(change_0_to_48))

# Statistical tests for case-only temporal changes
case_only_stats <- case_only_changes %>%
  group_by(carbohydrate_type, analyte) %>%
  summarise(
    n = n(),
    mean_change = mean(change_0_to_48, na.rm = TRUE),
    sd_change = sd(change_0_to_48, na.rm = TRUE),
    sem_change = sd_change / sqrt(n),
    t_stat = mean_change / sem_change,
    p_value = 2 * pt(-abs(t_stat), df = n - 1),  # two-tailed t-test
    .groups = "drop"
  ) %>%
  group_by(analyte) %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) %>%
  mutate(significance = case_when(
    p_adj < 0.001 ~ "***",
    p_adj < 0.01 ~ "**", 
    p_adj < 0.05 ~ "*",
    TRUE ~ "ns"
  ))

# Paired t-tests for case-only changes by analyte (pooled across carb types)
case_only_pooled_stats <- case_only_changes %>%
  group_by(analyte) %>%
  summarise(
    n = n(),
    mean_change = mean(change_0_to_48, na.rm = TRUE),
    sd_change = sd(change_0_to_48, na.rm = TRUE),
    sem_change = sd_change / sqrt(n),
    t_stat = mean_change / sem_change,
    p_value = 2 * pt(-abs(t_stat), df = n - 1),
    .groups = "drop"
  ) %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) %>%
  mutate(significance = case_when(
    p_adj < 0.001 ~ "***",
    p_adj < 0.01 ~ "**", 
    p_adj < 0.05 ~ "*",
    TRUE ~ "ns"
  ))

# Mixed-effects models for case-only temporal analysis
case_only_mixed_effects <- list()

# Filter data to case group only
case_only_data <- sample_data_long %>%
  filter(group == "case") %>%
  mutate(
    timepoint_hr = factor(timepoint_hr),
    subject = factor(subject),
    carbohydrate_type = factor(carbohydrate_type, 
                              levels = c("no_carbohydrate", "rapid_digestible", "slow_digestible"))
  )

for (analyte_name in scfa_analytes) {
  case_analyte_data <- case_only_data %>% filter(analyte == analyte_name)
  
  # Model with time and carbohydrate effects, subject as random intercept
  time_carb_model <- lmer(concentration ~ timepoint_hr * carbohydrate_type + 
                          (1 | subject), 
                          data = case_analyte_data, 
                          REML = FALSE)
  
  # Model with time effect only, subject as random intercept
  time_only_model <- lmer(concentration ~ timepoint_hr + (1 | subject), 
                          data = case_analyte_data, 
                          REML = FALSE)
  
  # Model with carbohydrate effect only, subject as random intercept
  carb_only_model <- lmer(concentration ~ carbohydrate_type + (1 | subject), 
                          data = case_analyte_data, 
                          REML = FALSE)
  
  # Null model (intercept + random effect only)
  null_model <- lmer(concentration ~ 1 + (1 | subject), 
                     data = case_analyte_data, 
                     REML = FALSE)
  
  # Store results
  case_only_mixed_effects[[analyte_name]] <- list(
    time_carb_model = time_carb_model,
    time_only_model = time_only_model,
    carb_only_model = carb_only_model,
    null_model = null_model,
    anova_time_carb = anova(time_carb_model),
    summary_time_carb = summary(time_carb_model),
    # Model comparisons
    time_vs_null = anova(null_model, time_only_model),
    carb_vs_null = anova(null_model, carb_only_model),
    interaction_vs_main = anova(time_only_model, time_carb_model)
  )
}

# Save subject-level analyses to CSV files
write.csv(subject_summary, "results/subject_level_summary.csv", row.names = FALSE)
write.csv(within_subject_stats, "results/within_subject_changes.csv", row.names = FALSE)
write.csv(case_only_stats, "results/case_only_temporal_changes.csv", row.names = FALSE)
write.csv(case_only_pooled_stats, "results/case_only_pooled_temporal_changes.csv", row.names = FALSE)

cat("Subject-level and case-only analyses saved to results/ directory\n")
```

## Statistical Results

### Group Comparisons (Control vs Case)
```{r group-stats-table}
knitr::kable(group_stats, digits = 4, 
             caption = "Group Comparisons with Benjamini-Hochberg Correction")
```

### Carbohydrate Type Comparisons
```{r carb-stats-table}
knitr::kable(carb_stats, digits = 4, 
             caption = "ANOVA Results for Carbohydrate Type Effects")
```

### Post-hoc Carbohydrate Comparisons
```{r carb-posthoc-table}
knitr::kable(carb_posthoc, digits = 4, 
             caption = "Pairwise Comparisons vs No Carbohydrate Control")
```

### Three-way Interaction Analysis
```{r interaction-stats-table}
# Check the structure and convert to data frame for proper column selection
interaction_display <- as.data.frame(interaction_stats) %>%
  select(analyte, Effect, p, p.adj, p.adj.signif)

knitr::kable(interaction_display, 
             digits = 4, 
             caption = "Three-way ANOVA: Group × Carbohydrate × Time Interactions")
```

## Subject-Level Analyses

### Subject-Level Summary Statistics
```{r subject-summary-table}
knitr::kable(subject_summary, digits = 3, 
             caption = "Subject-Level Summary Statistics")
```

### Within-Subject Changes (0h to 48h)
```{r within-subject-stats-table}
knitr::kable(within_subject_stats, digits = 4, 
             caption = "Within-Subject Changes from Baseline to 48h")
```

## Case-Only Temporal Analysis

### Case Group Temporal Changes by Carbohydrate Type
```{r case-only-stats-table}
knitr::kable(case_only_stats, digits = 4, 
             caption = "Case Group Only: Temporal Changes (0h to 48h) by Carbohydrate Type")
```

### Case Group Temporal Changes (Pooled)
```{r case-only-pooled-stats-table}
knitr::kable(case_only_pooled_stats, digits = 4, 
             caption = "Case Group Only: Temporal Changes (0h to 48h) Pooled Across Carbohydrate Types")
```

### Case Group Mixed-Effects Models (Subject Random Effects)
```{r case-only-mixed-effects-summary}
# Display key results from case-only mixed-effects models
case_only_mixed_summary <- data.frame(
  analyte = character(),
  effect = character(),
  f_value = numeric(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

for (analyte_name in names(case_only_mixed_effects)) {
  anova_results <- case_only_mixed_effects[[analyte_name]]$anova_time_carb
  for (i in 1:nrow(anova_results)) {
    case_only_mixed_summary <- rbind(case_only_mixed_summary, data.frame(
      analyte = analyte_name,
      effect = rownames(anova_results)[i],
      f_value = anova_results[i, "F value"],
      p_value = anova_results[i, "Pr(>F)"],
      stringsAsFactors = FALSE
    ))
  }
}

# Add significance indicators
case_only_mixed_summary <- case_only_mixed_summary %>%
  mutate(significance = case_when(
    p_value < 0.001 ~ "***",
    p_value < 0.01 ~ "**", 
    p_value < 0.05 ~ "*",
    TRUE ~ "ns"
  ))

knitr::kable(case_only_mixed_summary, digits = 4, 
             caption = "Case Group Mixed-Effects Models: Time × Carbohydrate Effects with Subject Random Effects")
```

## Mixed-Effects Model Results

### Mixed-Effects Model Summary
```{r mixed-effects-summary}
# Display key results from mixed-effects models
mixed_effects_summary <- data.frame(
  analyte = character(),
  effect = character(),
  f_value = numeric(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

for (analyte_name in names(mixed_effects_results)) {
  anova_results <- mixed_effects_results[[analyte_name]]$anova_full
  for (i in 1:nrow(anova_results)) {
    mixed_effects_summary <- rbind(mixed_effects_summary, data.frame(
      analyte = analyte_name,
      effect = rownames(anova_results)[i],
      f_value = anova_results[i, "F value"],
      p_value = anova_results[i, "Pr(>F)"],
      stringsAsFactors = FALSE
    ))
  }
}

# Add significance indicators
mixed_effects_summary <- mixed_effects_summary %>%
  mutate(significance = case_when(
    p_value < 0.001 ~ "***",
    p_value < 0.01 ~ "**", 
    p_value < 0.05 ~ "*",
    TRUE ~ "ns"
  ))

knitr::kable(mixed_effects_summary, digits = 4, 
             caption = "Mixed-Effects Models: F-values and P-values for Fixed Effects")
```

# Visualizations

```{r create-plots-directory, include=FALSE}
# Create plots directory if it doesn't exist
dir.create("plots", showWarnings = FALSE)

# Define publication-quality theme
pub_theme <- theme_classic() +
  theme(
    text = element_text(size = 12, family = "sans"),
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    strip.text = element_text(size = 12, face = "bold"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 11),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )

# Custom color palette
group_colors <- c("control" = "#3498DB", "case" = "#E74C3C")
carb_colors <- c("no_carbohydrate" = "#95A5A6", 
                 "rapid_digestible" = "#F39C12", 
                 "slow_digestible" = "#27AE60")
```

## SCFA Concentrations by Group and Carbohydrate Type

```{r plot-by-group, fig.height=10}
plot_by_group <- sample_data_long %>%
  ggplot(aes(x = group, y = concentration, fill = group)) +
  geom_boxplot(alpha = 0.8) +
  facet_grid(carbohydrate_type ~ analyte, scales = "free_y") +
  scale_fill_manual(values = group_colors) +
  stat_compare_means(method = "t.test", 
                     comparisons = list(c("control", "case")),
                     label = "p.signif",
                     step_increase = 0.1) +
  labs(title = "SCFA Concentrations by Group and Carbohydrate Type",
       x = "Group",
       y = "Concentration (μM)",
       fill = "Group") +
  pub_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot_by_group
```

## SCFA Concentrations by Carbohydrate Type and Time Point

```{r plot-by-carb, fig.height=8}
plot_by_carb <- sample_data_long %>%
  ggplot(aes(x = carbohydrate_type, y = concentration, fill = carbohydrate_type)) +
  geom_boxplot(alpha = 0.8) +
  facet_grid(timepoint_hr ~ analyte, scales = "free_y", 
             labeller = labeller(timepoint_hr = function(x) paste0(x, "h"))) +
  scale_fill_manual(values = carb_colors) +
  stat_compare_means(method = "t.test",
                     ref.group = "no_carbohydrate",
                     label = "p.signif",
                     step_increase = 0.1) +
  labs(title = "SCFA Concentrations by Carbohydrate Type and Time Point",
       x = "Carbohydrate Type",
       y = "Concentration (μM)",
       fill = "Carbohydrate Type") +
  pub_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

plot_by_carb
```

## Time Series Analysis

```{r plot-time-series}
plot_time_series <- sample_data_long %>%
  mutate(timepoint_hr = as.numeric(as.character(timepoint_hr))) %>%
  group_by(carbohydrate_type, timepoint_hr, analyte) %>%
  summarise(
    mean_conc = mean(concentration, na.rm = TRUE),
    sem = sd(concentration, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = timepoint_hr, y = mean_conc, color = carbohydrate_type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_conc - sem, ymax = mean_conc + sem), 
                width = 1, linewidth = 0.8) +
  facet_wrap(~ analyte, scales = "free_y") +
  scale_color_manual(values = carb_colors) +
  scale_x_continuous(breaks = c(0, 48)) +
  labs(title = "SCFA Concentrations Over Time by Carbohydrate Type",
       x = "Time (hours)",
       y = "Mean Concentration ± SEM (μM)",
       color = "Carbohydrate Type") +
  pub_theme +
  theme(legend.position = "bottom")

plot_time_series
```

## Concentration Heatmap

```{r plot-heatmap}
heatmap_data <- sample_data_long %>%
  mutate(timepoint_hr = paste0(timepoint_hr, "h")) %>%
  group_by(group, carbohydrate_type, timepoint_hr, analyte) %>%
  summarise(mean_conc = mean(concentration, na.rm = TRUE), .groups = "drop") %>%
  unite("condition", group, carbohydrate_type, timepoint_hr, sep = "_")

plot_heatmap <- heatmap_data %>%
  ggplot(aes(x = condition, y = analyte, fill = mean_conc)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_viridis_c(name = "Mean\nConcentration\n(μM)", option = "plasma") +
  labs(title = "SCFA Concentration Heatmap Across All Conditions",
       x = "Condition (Group_Carbohydrate_Time)",
       y = "SCFA Analyte") +
  pub_theme +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.text.y = element_text(face = "italic"))

plot_heatmap
```

## Interaction Effects

```{r plot-interaction}
plot_interaction <- sample_data_long %>%
  mutate(timepoint_hr = as.numeric(as.character(timepoint_hr))) %>%
  group_by(group, carbohydrate_type, timepoint_hr, analyte) %>%
  summarise(mean_conc = mean(concentration, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = timepoint_hr, y = mean_conc, 
             color = carbohydrate_type, linetype = group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3, aes(shape = group)) +
  facet_wrap(~ analyte, scales = "free_y") +
  scale_color_manual(values = carb_colors) +
  scale_x_continuous(breaks = c(0, 48)) +
  scale_linetype_manual(values = c("control" = "solid", "case" = "dashed")) +
  scale_shape_manual(values = c("control" = 16, "case" = 17)) +
  labs(title = "SCFA Concentrations: Group × Carbohydrate × Time Interaction",
       x = "Time (hours)",
       y = "Mean Concentration (μM)",
       color = "Carbohydrate Type",
       linetype = "Group",
       shape = "Group") +
  pub_theme +
  theme(legend.position = "bottom")

plot_interaction
```

## Subject-Level Individual Response Heatmap

```{r plot-subject-heatmap, fig.height=12, fig.width=16}
# Subject-level heatmap showing individual responses
subject_heatmap_data <- sample_data_long %>%
  mutate(
    condition_time = paste0(carbohydrate_type, "_", timepoint_hr, "h"),
    mean_conc = concentration  # Already averaged technical replicates
  ) %>%
  arrange(group, subject, analyte)

# Create ordered subject labels with control group first, then case group
subject_order <- subject_heatmap_data %>%
  select(subject, group) %>%
  distinct() %>%
  arrange(desc(group), subject) %>%  # control comes before case alphabetically
  mutate(
    subject_group = paste0(subject, " (", group, ")"),
    # Add spacing indicator for plotting
    group_spacing = ifelse(group == "control", "Control Group", "Case Group")
  )

# Create subject ordering with gap between groups
control_subjects <- subject_order %>% filter(group == "control") %>% arrange(subject)
case_subjects <- subject_order %>% filter(group == "case") %>% arrange(subject)

# Create spacer rows for the gap between groups
n_analytes <- length(unique(subject_heatmap_data$analyte))
n_conditions <- length(unique(subject_heatmap_data$condition_time))

# Create empty spacer data for gap
spacer_data <- expand_grid(
  analyte = unique(subject_heatmap_data$analyte),
  condition_time = unique(subject_heatmap_data$condition_time),
  subject_group = "--- GROUP SEPARATOR ---",
  mean_conc = NA_real_
)

# Create ordered factor levels with gap
all_subject_levels <- c(
  rev(control_subjects$subject_group),  # Control at top (reversed for ggplot)
  "--- GROUP SEPARATOR ---",           # Gap
  rev(case_subjects$subject_group)      # Case at bottom (reversed for ggplot)
)

# Add the ordered subject labels to the data and include spacer
subject_heatmap_data <- subject_heatmap_data %>%
  left_join(subject_order %>% select(subject, subject_group, group_spacing), by = "subject") %>%
  bind_rows(spacer_data) %>%
  mutate(
    subject_group = factor(subject_group, levels = all_subject_levels)
  )

plot_subject_heatmap <- subject_heatmap_data %>%
  ggplot(aes(x = condition_time, y = subject_group, fill = mean_conc)) +
  geom_tile(color = "white", linewidth = 0.2) +
  facet_wrap(~ analyte, scales = "free", ncol = 2) +
  scale_fill_viridis_c(name = "Concentration\n(μM)", option = "plasma", 
                       trans = "sqrt", na.value = "white") +
  labs(title = "Individual Subject SCFA Responses Across All Conditions",
       subtitle = "Control group (top) and Case group (bottom) with visual separation",
       x = "Condition × Time Point",
       y = "Subject (Group)") +
  pub_theme +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 7),
    strip.text = element_text(size = 10, face = "bold"),
    panel.spacing = unit(0.5, "lines")
  )

plot_subject_heatmap
```

## Individual Subject Trajectories by Carbohydrate Source

```{r plot-trajectories, fig.height=12, fig.width=16}
# Prepare data for trajectory plots
trajectory_data <- sample_data_long %>%
  mutate(
    timepoint_hr = as.numeric(as.character(timepoint_hr)),
    mean_conc = concentration  # Already averaged technical replicates
  )

# Calculate group means for overlay
group_means_trajectories <- trajectory_data %>%
  group_by(group, carbohydrate_type, timepoint_hr, analyte) %>%
  summarise(
    mean_conc = mean(mean_conc, na.rm = TRUE),
    sem = sd(mean_conc, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Create trajectory plots by carbohydrate source
plot_trajectories <- trajectory_data %>%
  ggplot(aes(x = timepoint_hr, y = mean_conc)) +
  # Individual subject lines (thin, semi-transparent)
  geom_line(aes(group = subject, color = group), 
            alpha = 0.4, linewidth = 0.5) +
  geom_point(aes(color = group), alpha = 0.5, size = 1) +
  # Group mean lines (thick, prominent)
  geom_line(data = group_means_trajectories, 
            aes(color = group), 
            linewidth = 2, alpha = 0.9) +
  geom_point(data = group_means_trajectories, 
             aes(color = group), 
             size = 3, alpha = 0.9) +
  # Error bars for group means
  geom_errorbar(data = group_means_trajectories,
                aes(ymin = mean_conc - sem, ymax = mean_conc + sem, color = group),
                width = 1, linewidth = 1, alpha = 0.7) +
  facet_grid(carbohydrate_type ~ analyte, scales = "free_y",
             labeller = labeller(carbohydrate_type = function(x) {
               str_replace_all(str_to_title(str_replace_all(x, "_", " ")), 
                              c("No" = "No", "Rapid" = "Rapid", "Slow" = "Slow"))
             })) +
  scale_color_manual(values = group_colors) +
  scale_x_continuous(breaks = c(0, 48)) +
  labs(title = "Individual Subject SCFA Trajectories by Carbohydrate Source",
       subtitle = "Thin lines: individual subjects, Thick lines: group means ± SEM",
       x = "Time (hours)",
       y = "Concentration (μM)",
       color = "Group") +
  pub_theme +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 10, face = "bold"),
    panel.spacing = unit(0.3, "lines")
  )

plot_trajectories
```

## Case Group Only: Temporal Changes

```{r plot-case-only, fig.height=8, fig.width=12}
# Case-only temporal changes visualization
case_only_plot_data <- sample_data_long %>%
  filter(group == "case") %>%
  mutate(timepoint_hr = as.numeric(as.character(timepoint_hr))) %>%
  group_by(carbohydrate_type, timepoint_hr, analyte) %>%
  summarise(
    mean_conc = mean(concentration, na.rm = TRUE),
    sem = sd(concentration, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

plot_case_only <- case_only_plot_data %>%
  ggplot(aes(x = timepoint_hr, y = mean_conc, color = carbohydrate_type)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = mean_conc - sem, ymax = mean_conc + sem), 
                width = 1, linewidth = 1) +
  facet_wrap(~ analyte, scales = "free_y") +
  scale_color_manual(values = carb_colors) +
  scale_x_continuous(breaks = c(0, 48)) +
  labs(title = "Case Group Only: SCFA Temporal Changes by Carbohydrate Type",
       subtitle = "Mean ± SEM concentrations from 0h to 48h",
       x = "Time (hours)",
       y = "Mean Concentration ± SEM (μM)",
       color = "Carbohydrate Type") +
  pub_theme +
  theme(legend.position = "bottom")

plot_case_only
```

## Case Group Individual Subject Trajectories

```{r plot-case-individual, fig.height=12, fig.width=16}
# Individual case subject trajectories with statistical overlay
case_individual_data <- sample_data_long %>%
  filter(group == "case") %>%
  mutate(
    timepoint_hr = as.numeric(as.character(timepoint_hr)),
    mean_conc = concentration  # Already averaged technical replicates
  )

plot_case_individual <- case_individual_data %>%
  ggplot(aes(x = timepoint_hr, y = mean_conc)) +
  # Individual subject lines
  geom_line(aes(group = subject), alpha = 0.6, linewidth = 0.8, color = "#E74C3C") +
  geom_point(alpha = 0.6, size = 2, color = "#E74C3C") +
  # Group mean overlay
  geom_line(data = case_only_plot_data, 
            aes(color = carbohydrate_type), 
            linewidth = 2, alpha = 0.9) +
  geom_point(data = case_only_plot_data, 
             aes(color = carbohydrate_type), 
             size = 4, alpha = 0.9) +
  facet_grid(carbohydrate_type ~ analyte, scales = "free_y",
             labeller = labeller(carbohydrate_type = function(x) {
               str_replace_all(str_to_title(str_replace_all(x, "_", " ")), 
                              c("No" = "No", "Rapid" = "Rapid", "Slow" = "Slow"))
             })) +
  scale_color_manual(values = carb_colors) +
  scale_x_continuous(breaks = c(0, 48)) +
  labs(title = "Case Group Individual Subject SCFA Trajectories",
       subtitle = "Thin lines: individual subjects, Thick lines: group means",
       x = "Time (hours)",
       y = "Concentration (μM)",
       color = "Carbohydrate Type") +
  pub_theme +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 10, face = "bold"),
    panel.spacing = unit(0.3, "lines")
  )

plot_case_individual
```

## Save Plots

```{r save-plots}
# Save all plots to files with appropriate dimensions
ggsave("plots/scfa_by_group.png", plot_by_group, width = 14, height = 10, dpi = 300)
ggsave("plots/scfa_by_carb.png", plot_by_carb, width = 14, height = 8, dpi = 300)
ggsave("plots/scfa_time_series.png", plot_time_series, width = 12, height = 8, dpi = 300)
ggsave("plots/scfa_heatmap.png", plot_heatmap, width = 14, height = 6, dpi = 300)
ggsave("plots/scfa_interaction.png", plot_interaction, width = 14, height = 10, dpi = 300)
ggsave("plots/scfa_subject_heatmap.png", plot_subject_heatmap, width = 16, height = 12, dpi = 300)
ggsave("plots/scfa_trajectories.png", plot_trajectories, width = 16, height = 12, dpi = 300)
ggsave("plots/scfa_case_only_temporal.png", plot_case_only, width = 12, height = 8, dpi = 300)
ggsave("plots/scfa_case_individual_trajectories.png", plot_case_individual, width = 16, height = 12, dpi = 300)

cat("All plots saved to plots/ directory with publication-quality formatting\n")
```

# Discussion

## Key Statistical Findings

### Temporal Effects Are Dominant
The most striking finding across all analyses is the **universal temporal effect** from 0h to 48h. Every SCFA analyte showed highly significant temporal changes (p < 2e-16 for all compounds), with dramatic increases from baseline to 48h:

- **Acetate**: 3.69 μM → 34.0 μM (9-fold increase)
- **Butyrate**: 0.745 μM → 7.38 μM (10-fold increase)  
- **Propionate**: 0.502 μM → 4.91 μM (10-fold increase)
- **5-aminovalerate**: 0.108 μM → 0.924 μM (8.5-fold increase)
- **Succinate**: 0.444 μM → 2.05 μM (4.6-fold increase)

### Group Differences Are Not Significant
Contrary to initial expectations, **no significant differences** were found between control and case groups for any SCFA analyte (all p > 0.05). This suggests that the experimental intervention did not create distinct SCFA metabolic signatures between groups when controlling for other factors.

### Carbohydrate-Specific Effects Are Present But Modest
Carbohydrate type showed significant effects for **acetate, butyrate, propionate, and succinate** in mixed-effects models (p < 0.05), but these effects were modest compared to temporal changes:

- **Succinate** showed the clearest carbohydrate response, with slow digestible carbohydrates producing higher concentrations than no carbohydrate (p = 0.009 in post-hoc testing)
- **Propionate and butyrate** showed significant carbohydrate × time interactions, indicating that carbohydrate type influences the temporal response pattern

### Case Group Analysis Reveals Universal Temporal Responses
The case-only mixed-effects analysis confirmed that **all SCFA analytes increase significantly over time** in case subjects (all p < 2e-16). This demonstrates that the temporal metabolic response is robust and consistent across individual subjects.

**Carbohydrate-specific findings in case subjects:**
- **Acetate**: Modest carbohydrate effect (p = 0.043)
- **Succinate**: Both temporal (p < 2e-16) and carbohydrate effects (p = 0.018), with marginal interaction (p = 0.030)
- **Propionate**: Near-significant carbohydrate effect (p = 0.082) and interaction (p = 0.058)

### Individual Subject Variation
Subject-level analyses revealed substantial inter-individual variation in baseline SCFA levels, particularly in the case group (higher standard deviations for most analytes). The mixed-effects models properly accounted for this variation through random effects, strengthening the temporal effect findings.

## Biological Interpretation

### Microbial Fermentation Response
The dramatic 4-10 fold increases in SCFA concentrations from 0h to 48h likely represent **active microbial fermentation** of dietary carbohydrates in the gut. This temporal pattern suggests:

1. **Lag phase** (0h): Minimal baseline SCFA production
2. **Active fermentation** (48h): Peak metabolic activity producing substantial SCFA concentrations

### Metabolic Pathway Specificity
The differential responses among SCFA analytes suggest distinct fermentation pathways:

- **Acetate, butyrate, and propionate** (primary SCFAs) showed the most dramatic increases, consistent with their role as major fermentation end-products
- **5-aminovalerate** showed significant but more modest increases, reflecting its secondary metabolite status
- **Succinate** showed the smallest response, consistent with its role as an intermediate rather than end-product

### Limited Treatment Differentiation
The absence of group differences suggests that the experimental intervention may not have sufficiently altered gut microbiome composition or metabolic function to create detectable SCFA signature differences within the 48-hour timeframe studied.

# Conclusions

This comprehensive SCFA analysis using PFBBr derivatization and GC-MS quantification reveals several key findings:

## Primary Findings

1. **Universal Temporal Response**: All five SCFA analytes showed dramatic 4-10 fold increases from baseline (0h) to 48 hours, indicating robust gut microbial fermentation responses regardless of treatment group.

2. **No Treatment Group Differentiation**: Despite expectations, no significant differences were observed between control and case groups for any SCFA analyte, suggesting the experimental intervention did not create distinct metabolic signatures within the study timeframe.

3. **Modest Carbohydrate-Specific Effects**: While carbohydrate type influenced SCFA production (particularly succinate and propionate), these effects were secondary to the dominant temporal response pattern.

4. **Robust Case Group Responses**: Case-only analysis confirmed universal temporal increases across all SCFA analytes, with some analytes showing additional carbohydrate-dependent response patterns.

5. **Substantial Individual Variation**: Subject-level analyses revealed considerable inter-individual differences in SCFA production, properly accounted for through mixed-effects modeling.

## Statistical Rigor

The analysis employed multiple complementary statistical approaches:
- Mixed-effects models controlling for subject random effects
- Multiple testing corrections (Benjamini-Hochberg)
- Paired analyses for repeated measures design
- Case-specific temporal analysis with proper statistical controls

## Clinical Implications

These findings suggest that:
- **SCFA production responses are universal** across subjects regardless of treatment group
- **Carbohydrate fermentation is a dominant metabolic process** that may overshadow treatment-specific effects
- **Individual metabolic variation** is substantial and should be considered in future study designs
- **Longer observation periods** may be needed to detect treatment-specific metabolic signatures

## Study Limitations

- The 48-hour timeframe may be insufficient to detect treatment-specific microbiome changes
- Sample size may limit power to detect subtle group differences
- The dramatic temporal effects may mask smaller but clinically relevant group differences

This analysis provides a robust foundation for understanding gut microbiome SCFA production patterns and demonstrates the importance of temporal dynamics in metabolomic studies.

# References

1. Haak, B. W., Littmann, E. R., Chaubard, J.-L., Pickard, A. J., Fontana, E., Adhi, F., et al. (2018). Impact of gut colonization with butyrate-producing microbiota on respiratory viral infection following allo-HCT. *Blood*, 131(26), 2978–2986.

2. DFI-HMMF Targeted Metabolomics: General and Detailed Methods. University of Chicago Medicine, Duchossois Family Institute.