---
title: "Integrated SCFA and Microbiome Analysis: Abbott Carbohydrate Obesity Project"
author: "Patrick Seed, MD, PhD"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
    theme: united
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 300,
  cache = TRUE,
  cache.path = "cache/integrated-"
)

options(dplyr.summarise.inform = FALSE)
```

```{r libraries, include=FALSE}
# Core libraries from SCFA and microbiome analyses
library(tidyverse)
library(readr)
library(readxl)
library(data.table)
library(janitor)
library(ggpubr)
library(rstatix)
library(stringr)
library(lme4)
library(lmerTest)
library(emmeans)
library(DT)

# Microbiome-specific
library(phyloseq)
library(microbiome)
library(vegan)
library(ANCOMBC)
library(Maaslin2)
library(nlme)
```

```{r themes-and-palettes, include=FALSE}
# Reuse/define publication-quality themes and color palettes
# Prefer the SCFA analysis styling (adapted here if needed)

theme_scfa <- function() {
  ggplot2::theme_bw(base_size = 14) +
    ggplot2::theme(
      plot.title = element_text(face = "bold", hjust = 0),
      axis.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold"),
      legend.position = "right"
    )
}

scale_fill_scfa <- ggplot2::scale_fill_brewer(palette = "Set2")
scale_color_scfa <- ggplot2::scale_color_brewer(palette = "Set2")
```

```{r load-data, include=FALSE}
# Load SCFA and microbiome data following the logic in the source Rmd files,
# but with paths relative to the project root (one level up from this Rmd).

## SCFA: metadata and quantification data ---------------------------------

scfa_metadata <- NULL
scfa_sample_data_long <- NULL
scfa_sample_data_long_factored <- NULL

tryCatch({
  map1 <- read_excel(
    file.path("..", "scfa_metabolomics", "metadata", "SEED LAB HMM Submission Sheet 202405.xlsx"),
    skip = 9
  ) %>%
    clean_names()
  
  colnames(map1) <- c(
    "sampleid", "group", "bile_acids_panel", "pfbbr_scfa_panel",
    "tryp_indole_panel", "tms_panel_1", "tms_panel_2", "tms_panel_3",
    "untargeted_panel", "custom_targeted_panel", "low_abundant_metabolite_panel",
    "sample_origin", "sample_type", "mouse_number",
    "services_requested_per_sample", "mouse_strain", "mouse_flora",
    "volume_ul", "sample_weight_mg", "experiment_treatment"
  )
  
  map1 <- map1 %>%
    filter(!is.na(sampleid), sampleid != "") %>%
    slice(-(1:3))
  
  map2 <- read_excel(
    file.path("..", "scfa_metabolomics", "metadata", "PS2720_xtra_samples_HMM Submission Sheet.xlsx"),
    skip = 9
  ) %>%
    clean_names()
  
  colnames(map2) <- colnames(map1)
  
  map2 <- map2 %>%
    filter(!is.na(sampleid), sampleid != "") %>%
    slice(-(1:3))
  
  scfa_metadata_raw <- bind_rows(map1, map2)
}, error = function(e) {
  stop("Error reading SCFA metadata files: ", e$message)
})

scfa_metadata <- scfa_metadata_raw %>%
  mutate(
    subject = str_extract(sampleid, "^[0-9]+"),
    carb_repeat = str_extract(sampleid, "(?<= )[rs]\\d|nc(?= )"),
    carbohydrate_type = case_when(
      str_detect(carb_repeat, "^r") ~ "rapid_digestible",
      str_detect(carb_repeat, "^s") ~ "slow_digestible",
      str_detect(carb_repeat, "^nc") ~ "no_carbohydrate",
      TRUE ~ NA_character_
    ),
    technical_repeat = if_else(
      str_detect(carb_repeat, "^nc"),
      1L,
      as.integer(str_extract(carb_repeat, "\\d+"))
    ),
    timepoint_hr = as.integer(str_extract(sampleid, "(?<= )\\d+(?=h$)"))
  ) %>%
  dplyr::select(sampleid, group, subject, carbohydrate_type, technical_repeat, timepoint_hr) %>%
  mutate(group = tolower(group)) %>%
  filter(!is.na(group))

tryCatch({
  scfa_raw <- fread(
    file.path("..", "scfa_metabolomics", "data", "removed_qcs_quant_results_20250722_PFBBr_PS2720_20250731.csv")
  )
}, error = function(e) {
  stop("Error reading SCFA data file: ", e$message)
})

scfa_sample_data <- left_join(scfa_metadata, scfa_raw, by = "sampleid") %>%
  mutate(group = tolower(group)) %>%
  filter(!is.na(group))

scfa_analytes <- c("acetate", "butyrate", "propionate", "5aminovalerate", "succinate")

scfa_sample_data_long <- scfa_sample_data %>%
  pivot_longer(
    cols = all_of(scfa_analytes),
    names_to = "analyte",
    values_to = "concentration"
  ) %>%
  filter(!is.na(concentration)) %>%
  group_by(subject, group, carbohydrate_type, timepoint_hr, analyte) %>%
  summarise(concentration = mean(concentration, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    group_label = factor(str_to_title(group), levels = c("Control", "Case")),
    carbohydrate_type_label = str_replace_all(carbohydrate_type, "_", " ") %>%
      str_to_title()
  )

scfa_sample_data_long_factored <- scfa_sample_data_long %>%
  mutate(
    group = factor(group, levels = c("control", "case")),
    carbohydrate_type = factor(
      carbohydrate_type,
      levels = c("no_carbohydrate", "rapid_digestible", "slow_digestible")
    ),
    timepoint_hr = factor(timepoint_hr),
    subject = factor(subject)
  )

## Microbiome: metadata, phyloseq, and relative abundances --------------

micro_metadata <- NULL
ps <- NULL
ps_rel <- NULL

tryCatch({
  micro_metadata_raw <- readr::read_tsv(
    file.path(
      "..", "microbiome_analysis", "zr24558.16S_250813.zymo",
      "00...AllSamples.Bac16Sv34", "Sample_Information", "sample.metadata.csv"
    ),
    show_col_types = FALSE
  ) %>%
    janitor::clean_names()
}, error = function(e) {
  stop("Error reading microbiome metadata file: ", e$message)
})

micro_metadata <- micro_metadata_raw %>%
  # Clean up double dots in labels like "NC..48H" and trim whitespace
  dplyr::mutate(
    customer_label = stringr::str_replace_all(customer_label, "\\.\\.", "."),
    customer_label = stringr::str_trim(customer_label)
  ) %>%
  separate(
    customer_label,
    into = c("subject_id", "replicate_code", "timepoint"),
    sep = "\\.",
    remove = FALSE
  ) %>%
  dplyr::mutate(
    carbohydrate = case_when(
      stringr::str_detect(replicate_code, "R") ~ "RDC",
      stringr::str_detect(replicate_code, "S") ~ "SDC",
      TRUE ~ "No Added Carb"
    ),
    carbohydrate = factor(carbohydrate, levels = c("No Added Carb", "RDC", "SDC")),
    timepoint = factor(timepoint, levels = c("0H", "48H")),
    # Parse replicate number safely: treat NC as 1, numeric codes as their number
    replicate_code_clean = dplyr::if_else(
      stringr::str_detect(replicate_code, "NC"),
      NA_character_,
      replicate_code
    ),
    replicate = readr::parse_number(replicate_code_clean),
    replicate = tidyr::replace_na(replicate, 1L)
  ) %>%
  dplyr::select(-replicate_code_clean) %>%
  dplyr::rename(SampleID = sample_id) %>%
  dplyr::select(SampleID, subject_id, customer_label, timepoint, carbohydrate, replicate, raw_seq_files)

tryCatch({
  healthy_lookup <- readr::read_tsv(
    file.path(
      "..", "microbiome_analysis", "zr24558.16S_250813.zymo",
      "1.Healthy.TimepointComparism..Bac16Sv34", "Sample_Information", "sample.metadata.csv"
    ),
    show_col_types = FALSE
  ) %>%
    janitor::clean_names() %>%
    separate(customer_label, into = c("subject_id", "tmp", "tp"), sep = "\\.", remove = FALSE) %>%
    dplyr::distinct(subject_id) %>%
    dplyr::mutate(group = "Healthy")
}, error = function(e) {
  stop("Error reading microbiome healthy lookup file: ", e$message)
})

micro_metadata <- micro_metadata %>%
  mutate(SampleID = paste0("S", customer_label)) %>%
  dplyr::mutate(
    group = if_else(subject_id %in% healthy_lookup$subject_id, "Healthy", "Obese"),
    # Override MediaControl: group = NA, timepoint = "0H", carbohydrate = NA
    group = dplyr::if_else(
      stringr::str_detect(customer_label, regex("^MediaControl$", ignore_case = TRUE)),
      NA_character_,
      group
    ),
    timepoint = dplyr::if_else(
      stringr::str_detect(customer_label, regex("^MediaControl$", ignore_case = TRUE)),
      factor("0H", levels = levels(timepoint)),
      timepoint
    ),
    carbohydrate = dplyr::if_else(
      stringr::str_detect(customer_label, regex("^MediaControl$", ignore_case = TRUE)),
      factor(NA_character_, levels = levels(carbohydrate)),
      carbohydrate
    ),
    group = factor(group, levels = c("Healthy", "Obese")),
    subject_id = factor(subject_id)
  )

micro_metadata <- micro_metadata %>%
  dplyr::filter(!is.na(timepoint))

tryCatch({
  ps <- import_biom(
    file.path(
      "..", "microbiome_analysis", "zr24558.16S_250813.zymo",
      "00...AllSamples.Bac16Sv34", "DADA2_ASV_Distribution", "ASV_Table.biom"
    )
  )
}, error = function(e) {
  stop("Error importing microbiome BIOM file: ", e$message)
})

colnames(tax_table(ps)) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")

sample_names(ps) <- paste0("S", sample_names(ps))

metadata_ps <- micro_metadata %>%
  dplyr::mutate(SampleID = as.character(SampleID)) %>%
  dplyr::group_by(SampleID) %>%
  dplyr::mutate(SampleID = if (n() > 1) paste0(SampleID, "_", row_number()) else SampleID) %>%
  dplyr::ungroup() %>%
  column_to_rownames("SampleID")

sample_data(ps) <- sample_data(metadata_ps)

ps <- prune_taxa(taxa_sums(ps) > 0, ps)
ps <- prune_samples(sample_sums(ps) > 0, ps)

sample_data(ps)$SubjectID <- sample_data(ps)$subject_id
sample_data(ps)$Group <- sample_data(ps)$group
sample_data(ps)$Carbohydrate <- sample_data(ps)$carbohydrate
sample_data(ps)$Timepoint <- sample_data(ps)$timepoint

ps_rel <- microbiome::transform(ps, "compositional")

## qPCR absolute abundance data -----------------------------------------

qpcr_data <- NULL

tryCatch({
  qpcr_data <- readr::read_csv(
    file.path(
      "..", "microbiome_analysis", "zr24558.16S_250813.zymo",
      "00...AllSamples.Bac16Sv34", "Sample_Information", "absolute.abundance.csv"
    ),
    show_col_types = FALSE
  ) %>%
    janitor::clean_names()
}, error = function(e) {
  warning("Unable to load qPCR absolute abundance data: ", e$message)
})

# Note:
# - qPCR sample identifier column: customer_label
# - Will be joined to microbiome metadata via customer_label
# - genome_copies_per_ul will be used to convert relative to absolute
#   abundances in later sections.
```

# 1. Introduction

This integrated report brings together the short-chain fatty acid (SCFA) metabolomics and 16S rRNA gene microbiome analyses from the Abbott carbohydrate obesity project. It re-runs the key components of each analysis within a single, reproducible R Markdown workflow and adds new integrative components relating taxa, SCFA concentrations, and responder phenotypes.

The goal is to maintain the rigor and depth of the original analyses while ensuring that metadata, filtering, and modeling decisions are harmonized across datasets.

# 2. Project Objectives

In this section, we summarize the main scientific and analytical objectives, building on the SCFA project objectives and extending them to explicitly include microbiome composition, diversity, and taxon–SCFA relationships.

```{r objectives-text, echo=FALSE, results='asis'}
# TODO: programmatically pull or summarize objectives text
```

# 3. Methods

Here we integrate the methods from the SCFA metabolomics and microbiome analyses, including:

- Study design and sample processing
- SCFA quantification and preprocessing
- Microbiome sequencing, processing, and phyloseq construction
- Statistical modeling strategies (mixed-effects models, diversity analyses, differential abundance, and MaAsLin2)
- Definition of responder vs non-responder groups for butyrate and propionate based on 0 h to 48 h changes

```{r methods-text, echo=FALSE, results='asis'}
# TODO: construct a combined methods narrative, reusing text and
# code-derived summaries from the source Rmd files
```

# 4. Summary Data

We begin with a concise overview of the sample counts by obesity status (Control vs Case) and carbohydrate condition across both SCFA and microbiome datasets.

```{r summary-data, echo=FALSE}
# Summary counts for SCFA (control/case) and microbiome (Healthy/Obese)

# SCFA: deduplicate to one row per biological sample (subject × timepoint × carbohydrate)
scfa_sample_overview <- scfa_sample_data_long_factored %>%
  dplyr::distinct(subject, group, carbohydrate_type, timepoint_hr) %>%
  mutate(
    group_label = factor(str_to_title(group), levels = c("Control", "Case")),
    carbohydrate = case_when(
      carbohydrate_type == "no_carbohydrate" ~ "No Added Carb",
      carbohydrate_type == "rapid_digestible" ~ "RDC",
      carbohydrate_type == "slow_digestible" ~ "SDC",
      TRUE ~ NA_character_
    ),
    carbohydrate = factor(carbohydrate, levels = c("No Added Carb", "RDC", "SDC"))
  )

scfa_summary_counts <- scfa_sample_overview %>%
  dplyr::count(group_label, carbohydrate, name = "n_samples") %>%
  dplyr::arrange(group_label, carbohydrate)

# Microbiome: sample counts by group and carbohydrate (all timepoints)
micro_summary_counts <- micro_metadata %>%
  dplyr::count(group, carbohydrate, timepoint, name = "n_samples") %>%
  dplyr::arrange(group, carbohydrate, timepoint)

scfa_table <- scfa_summary_counts %>%
  dplyr::rename(
    Group = group_label,
    Carbohydrate = carbohydrate,
    `N samples` = n_samples
  )

micro_table <- micro_summary_counts %>%
  dplyr::rename(
    Group = group,
    Carbohydrate = carbohydrate,
    Timepoint = timepoint,
    `N samples` = n_samples
  )

if (knitr::is_html_output()) {
  htmltools::tagList(
    htmltools::h3("SCFA samples by group and carbohydrate"),
    DT::datatable(
      scfa_table,
      options = list(pageLength = 10, dom = "tip"),
      rownames = FALSE
    ),
    htmltools::h3("Microbiome samples by group, carbohydrate, and timepoint"),
    DT::datatable(
      micro_table,
      options = list(pageLength = 10, dom = "tip"),
      rownames = FALSE
    )
  )
} else {
  list(
    SCFA = scfa_table,
    Microbiome = micro_table
  )
}
```

# 5. Metabolomics (SCFA)

## 5.1 Section Overview

This section reproduces and extends the SCFA analyses, focusing on:

- Carbohydrate type effects (combined and stratified by group)
- SCFA concentrations over time
- Advanced mixed-effects modeling for SCFA trajectories

We first quantify SCFA response magnitudes (48h – 0h) and evaluate how they vary by carbohydrate type and obesity group. We then visualize SCFA concentrations over time and fit linear mixed-effects models with subject as a random effect to formally test fixed effects of group, carbohydrate type, time, and their interactions on absolute SCFA concentrations.

## 5.2 Carbohydrate Type Effects (Combined Groups)

```{r scfa-carbohydrate-effects, echo=FALSE}
# Calculate delta values (48h - 0h) using integrated SCFA data
delta_data <- scfa_sample_data_long_factored %>%
  dplyr::filter(timepoint_hr %in% c("0", "48")) %>%
  tidyr::pivot_wider(
    id_cols = c(subject, group, carbohydrate_type, analyte, group_label, carbohydrate_type_label),
    names_from = timepoint_hr,
    values_from = concentration,
    names_prefix = "time_"
  ) %>%
  dplyr::mutate(delta_48h_0h = time_48 - time_0) %>%
  dplyr::filter(!is.na(delta_48h_0h))

# Control and Case group-specific delta plots
plot_delta_control_only <- delta_data %>%
  dplyr::filter(group == "control") %>%
  ggplot(aes(x = carbohydrate_type_label, y = delta_48h_0h)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", alpha = 0.7) +
  geom_boxplot(aes(fill = carbohydrate_type_label), alpha = 0.8, outlier.shape = NA) +
  geom_jitter(alpha = 0.6, width = 0.2, size = 1.5, color = "gray30") +
  facet_wrap(~ analyte, scales = "free_y") +
  scale_fill_scfa +
  ggpubr::stat_compare_means(
    comparisons = list(
      c("No Carbohydrate", "Rapid Digestible"),
      c("No Carbohydrate", "Slow Digestible"),
      c("Rapid Digestible", "Slow Digestible")
    ),
    method = "wilcox.test",
    label = "p.signif",
    step.increase = 0.1
  ) +
  labs(
    title = "SCFA Response Magnitude by Carbohydrate Type - Control Group Only",
    subtitle = "Delta change (48h - 0h) with Wilcoxon pairwise comparisons",
    x = "Carbohydrate Type",
    y = "Change in Concentration (48h - 0h) μM"
  ) +
  theme_scfa() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot_delta_case_only <- delta_data %>%
  dplyr::filter(group == "case") %>%
  ggplot(aes(x = carbohydrate_type_label, y = delta_48h_0h)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", alpha = 0.7) +
  geom_boxplot(aes(fill = carbohydrate_type_label), alpha = 0.8, outlier.shape = NA) +
  geom_jitter(alpha = 0.6, width = 0.2, size = 1.5, color = "gray30") +
  facet_wrap(~ analyte, scales = "free_y") +
  scale_fill_scfa +
  ggpubr::stat_compare_means(
    comparisons = list(
      c("No Carbohydrate", "Rapid Digestible"),
      c("No Carbohydrate", "Slow Digestible"),
      c("Rapid Digestible", "Slow Digestible")
    ),
    method = "wilcox.test",
    label = "p.signif",
    step.increase = 0.1
  ) +
  labs(
    title = "SCFA Response Magnitude by Carbohydrate Type - Case Group Only",
    subtitle = "Delta change (48h - 0h) with Wilcoxon pairwise comparisons",
    x = "Carbohydrate Type",
    y = "Change in Concentration (48h - 0h) μM"
  ) +
  theme_scfa() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Combined groups analysis
plot_delta_carb_combined <- delta_data %>%
  ggplot(aes(x = carbohydrate_type_label, y = delta_48h_0h)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", alpha = 0.7) +
  geom_boxplot(aes(fill = carbohydrate_type_label), alpha = 0.8, outlier.shape = NA) +
  geom_jitter(alpha = 0.5, width = 0.2, size = 1.5, color = "gray30") +
  facet_wrap(~ analyte, scales = "free_y") +
  scale_fill_scfa +
  ggpubr::stat_compare_means(
    comparisons = list(
      c("No Carbohydrate", "Rapid Digestible"),
      c("No Carbohydrate", "Slow Digestible"),
      c("Rapid Digestible", "Slow Digestible")
    ),
    method = "wilcox.test",
    label = "p.signif",
    step.increase = 0.1
  ) +
  labs(
    title = "SCFA Response Magnitude by Carbohydrate Type (Combined Groups)",
    subtitle = "Delta change (48h - 0h) with Wilcoxon pairwise comparisons",
    x = "Carbohydrate Type",
    y = "Change in Concentration (48h - 0h) μM"
  ) +
  theme_scfa() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# One-way ANOVA on delta by carbohydrate type (combined groups)
combined_carb_stats <- purrr::map_dfr(
  unique(delta_data$analyte),
  function(analyte_name) {
    analyte_data <- delta_data %>% dplyr::filter(analyte == analyte_name)
    model <- aov(delta_48h_0h ~ carbohydrate_type, data = analyte_data)
    anova_summary <- summary(model)[[1]]
    data.frame(
      analyte = analyte_name,
      Effect = rownames(anova_summary),
      DF = anova_summary$Df,
      Sum_Sq = anova_summary$`Sum Sq`,
      Mean_Sq = anova_summary$`Mean Sq`,
      F_value = anova_summary$`F value`,
      p = anova_summary$`Pr(>F)`
    ) %>%
      dplyr::filter(Effect != "Residuals") %>%
      dplyr::mutate(
        p_adj = p.adjust(p, method = "BH"),
        significance = dplyr::case_when(
          p_adj < 0.001 ~ "p<0.001",
          p_adj < 0.01 ~ "p<0.01",
          p_adj < 0.05 ~ "p<0.05",
          TRUE ~ "ns"
        )
      )
  }
)

if (knitr::is_html_output()) {
  htmltools::tagList(
    htmltools::h4("Control group: delta SCFA by carbohydrate type"),
    plot_delta_control_only,
    htmltools::h4("Case group: delta SCFA by carbohydrate type"),
    plot_delta_case_only,
    htmltools::h4("Combined groups: delta SCFA by carbohydrate type"),
    plot_delta_carb_combined,
    DT::datatable(
      combined_carb_stats %>%
        dplyr::select(analyte, Effect, p, p_adj, significance),
      caption = "One-way ANOVA: Carbohydrate Type Effects on Delta Changes (Combined Groups)",
      options = list(pageLength = 10, scrollX = TRUE),
      rownames = FALSE
    ) %>%
      DT::formatRound(columns = c("p", "p_adj"), digits = 4)
  )
} else {
  print(plot_delta_control_only)
  print(plot_delta_case_only)
  print(plot_delta_carb_combined)
  knitr::kable(
    combined_carb_stats %>%
      dplyr::select(analyte, Effect, p, p_adj, significance),
    digits = 4,
    caption = "One-way ANOVA: Carbohydrate Type Effects on Delta Changes (Combined Groups)"
  )
}
```

## 5.3 SCFA Concentrations Over Time

```{r scfa-over-time, echo=FALSE}
plot_time_series <- scfa_sample_data_long %>%
  dplyr::group_by(carbohydrate_type_label, timepoint_hr, analyte) %>%
  dplyr::summarise(
    mean_conc = mean(concentration, na.rm = TRUE),
    sem = sd(concentration, na.rm = TRUE) / sqrt(dplyr::n()),
    .groups = "drop"
  ) %>%
  ggplot(aes(
    x = as.numeric(as.character(timepoint_hr)),
    y = mean_conc,
    color = carbohydrate_type_label,
    group = carbohydrate_type_label
  )) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = mean_conc - sem, ymax = mean_conc + sem),
    width = 1,
    linewidth = 0.8
  ) +
  facet_wrap(~ analyte, scales = "free_y") +
  scale_color_scfa +
  labs(
    title = "SCFA Concentrations Over Time by Carbohydrate Type",
    x = "Time (hours)",
    y = "Mean Concentration ± SEM (μM)",
    color = "Carbohydrate Type"
  ) +
  theme_scfa() +
  theme(legend.position = "bottom")

plot_time_series
```

## 5.4 Advanced Statistical Modeling: Mixed-Effects Models

```{r scfa-mixed-effects, echo=FALSE}
# Mixed-effects models for absolute SCFA concentrations
mixed_effects_results <- scfa_sample_data_long_factored %>%
  dplyr::group_by(analyte) %>%
  tidyr::nest() %>%
  dplyr::mutate(
    # Use lmerTest::lmer so that anova() returns F-tests with p-values
    model = purrr::map(
      data,
      ~ lmerTest::lmer(
        concentration ~ group * carbohydrate_type * timepoint_hr + (1 | subject),
        data = .x
      )
    ),
    anova_main = purrr::map(model, ~ anova(.x)),
    model_summary = purrr::map(model, ~ summary(.x))
  )

main_anova_summary <- mixed_effects_results %>%
  dplyr::mutate(
    anova_df = purrr::map(
      anova_main,
      ~ as.data.frame(.x) %>%
        tibble::rownames_to_column("effect")
    )
  ) %>%
  dplyr::select(analyte, anova_df) %>%
  tidyr::unnest(anova_df) %>%
  dplyr::filter(effect != "Residual") %>%
  dplyr::rename(
    F_value = `F value`,
    pr_f = `Pr(>F)`
  ) %>%
  dplyr::mutate(
    p_adj = p.adjust(pr_f, method = "BH"),
    significance = dplyr::case_when(
      p_adj < 0.001 ~ "p<0.001",
      p_adj < 0.01 ~ "p<0.01",
      p_adj < 0.05 ~ "p<0.05",
      TRUE ~ "ns"
    )
  ) %>%
  dplyr::arrange(p_adj)

if (knitr::is_html_output()) {
  DT::datatable(
    main_anova_summary %>%
      dplyr::select(analyte, effect, F_value, pr_f, p_adj, significance),
    caption = "Mixed-Effects Models for SCFA Concentrations: ANOVA Table",
    options = list(pageLength = 10, scrollX = TRUE),
    rownames = FALSE
  ) %>%
    DT::formatRound(columns = c("F_value", "pr_f", "p_adj"), digits = 4)
} else {
  knitr::kable(
    main_anova_summary %>%
      dplyr::select(analyte, effect, F_value, pr_f, p_adj, significance),
    digits = 4,
    caption = "Mixed-Effects Models for SCFA Concentrations: ANOVA Table"
  )
}
```

## 5.5 Summary of SCFA Results

```{r scfa-summary-text, echo=FALSE, results='asis'}
# TODO: summary narrative highlighting key SCFA findings
```

# 6. Microbiome

## 6.1 Section Overview

This section reproduces key microbiome results, including alpha and beta diversity, community composition, and differential abundance, using the harmonized metadata and styling. We focus first on within-sample diversity (alpha diversity) to summarize richness and evenness patterns across groups, carbohydrate conditions, and timepoints, then move to between-sample diversity (beta diversity) and taxon-level changes.

## 6.2 Alpha Diversity

```{r alpha-diversity, echo=FALSE}
# Alpha diversity metrics: Observed, Shannon, Simpson at 0H and 48H

alpha_metrics <- estimate_richness(ps, measures = c("Observed", "Shannon", "Simpson")) %>%
  tibble::rownames_to_column("SampleID")

alpha_long <- alpha_metrics %>%
  tidyr::pivot_longer(
    cols = -SampleID,
    names_to = "metric",
    values_to = "value"
  ) %>%
  dplyr::left_join(micro_metadata, by = "SampleID") %>%
  dplyr::mutate(
    metric = factor(metric, levels = c("Shannon", "Observed", "Simpson"))
  )

alpha_long_0_48 <- alpha_long %>%
  dplyr::filter(timepoint %in% c("0H", "48H"))

# Alpha diversity boxplots by group, carbohydrate, and timepoint
alpha_plot <- alpha_long_0_48 %>%
  ggplot(aes(x = carbohydrate, y = value, fill = group)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA, position = position_dodge(width = 0.8)) +
  geom_jitter(
    aes(color = group),
    alpha = 0.5,
    size = 1.2,
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8)
  ) +
  facet_grid(metric ~ timepoint, scales = "free_y") +
  scale_fill_scfa +
  scale_color_scfa +
  labs(
    title = "Alpha Diversity by Group, Carbohydrate, and Timepoint",
    x = "Carbohydrate",
    y = "Diversity metric"
  ) +
  theme_scfa() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Summaries at 48H for combined groups modeling
alpha_48h <- alpha_long %>%
  dplyr::filter(timepoint == "48H")

alpha_48h_summary <- alpha_48h %>%
  dplyr::group_by(metric, group, carbohydrate) %>%
  dplyr::summarise(
    mean = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    n = dplyr::n(),
    .groups = "drop"
  )

if (knitr::is_html_output()) {
  htmltools::tagList(
    alpha_plot,
    DT::datatable(
      alpha_48h_summary,
      caption = "Alpha diversity at 48H: summary by metric, group, and carbohydrate",
      options = list(pageLength = 10, scrollX = TRUE),
      rownames = FALSE
    ) %>%
      DT::formatRound(columns = c("mean", "sd"), digits = 3)
  )
} else {
  print(alpha_plot)
  knitr::kable(
    alpha_48h_summary,
    digits = 3,
    caption = "Alpha diversity at 48H: summary by metric, group, and carbohydrate"
  )
}
```

## 6.3 Beta Diversity

```{r beta-diversity, echo=FALSE}
# TODO: reproduce Bray PCoA plots, PERMANOVA tables, 48 h beta
# diversity, and baseline clustering/trajectory visualizations
```

## 6.4 Community Composition

```{r community-composition, echo=FALSE}
# TODO: genus composition changes from 0 h to 48 h by carbohydrate,
# validation barplots, and repair of the RDC panel
```

## 6.5 Differential Abundance and Predictive Modeling

```{r differential-abundance, echo=FALSE}
# TODO: ANCOM-BC2, LEfSe Differential Abundance, MaAsLin2 tables
# and visualizations, including updated volcano plots with ggrepel
```

## 6.6 Absolute Abundances

```{r absolute-abundance, echo=FALSE}
# TODO:
# - Integrate qPCR genome copy data (customer_label)
# - Convert relative abundances to absolute abundances
# - Generate g_Fusicantenibacter sp. absolute abundance plots by
#   group, timepoint, and carbohydrate condition
```

# 7. Integrated Analysis: Taxon–SCFA Relationships

```{r taxon-scfa-integration, echo=FALSE}
# TODO:
# - ANCOM-BC2 and MaAsLin2 models for taxon ~ butyrate + carbohydrate
#   and taxon ~ propionate + carbohydrate (relative and absolute)
# - Correlation plots for g_Fusicantenibacter sp. vs butyrate and
#   propionate at 0 h and 48 h (relative and absolute)
```

```{r responder-nonresponder, echo=FALSE}
# TODO:
# - Define responder vs non-responder groups separately for butyrate
#   and propionate using 0 h to 48 h changes (e.g., median/quantile cut)
# - Examine overlap in responder status between butyrate and propionate
# - Compare microbial composition between responder and non-responder
#   groups at 48 h
```

# 8. Summary

```{r final-summary, echo=FALSE, results='asis'}
# TODO: concise summary of key integrated findings across SCFA,
# microbiome, and taxon–SCFA analyses
```


