---
title: "Integrated SCFA and Microbiome Analysis: Abbott Carbohydrate Obesity Project"
author: "Patrick Seed, MD, PhD"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: hide
    theme: united
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 300,
  cache = TRUE,
  cache.path = "cache/integrated-"
)

options(dplyr.summarise.inform = FALSE)
```

```{r libraries, include=FALSE}
# Core libraries from SCFA and microbiome analyses
library(tidyverse)
library(readr)
library(readxl)
library(data.table)
library(janitor)
library(ggpubr)
library(rstatix)
library(stringr)
library(lme4)
library(lmerTest)
library(emmeans)
library(DT)

# Microbiome-specific
library(phyloseq)
library(microbiome)
library(vegan)
library(ANCOMBC)
library(Maaslin2)
library(nlme)
library(ggalluvial)
library(scico)
library(patchwork)
library(ggrepel)
library(purrr)
```

```{r themes-and-palettes, include=FALSE}
# Reuse/define publication-quality themes and color palettes
# Prefer the SCFA analysis styling (adapted here if needed)

theme_scfa <- function() {
  ggplot2::theme_bw(base_size = 14) +
    ggplot2::theme(
      plot.title = element_text(face = "bold", hjust = 0),
      axis.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold"),
      legend.position = "right"
    )
}

# Define as functions to avoid serialization issues when rendering to HTML
scale_fill_scfa <- function(...) ggplot2::scale_fill_brewer(palette = "Set2", ...)
scale_color_scfa <- function(...) ggplot2::scale_color_brewer(palette = "Set2", ...)
```

```{r load-data, include=FALSE}
# Load SCFA and microbiome data following the logic in the source Rmd files,
# but with paths relative to the project root (one level up from this Rmd).

## SCFA: metadata and quantification data ---------------------------------

scfa_metadata <- NULL
scfa_sample_data_long <- NULL
scfa_sample_data_long_factored <- NULL

tryCatch({
  map1 <- read_excel(
    file.path("..", "scfa_metabolomics", "metadata", "SEED LAB HMM Submission Sheet 202405.xlsx"),
    skip = 9
  ) %>%
    clean_names()
  
  colnames(map1) <- c(
    "sampleid", "group", "bile_acids_panel", "pfbbr_scfa_panel",
    "tryp_indole_panel", "tms_panel_1", "tms_panel_2", "tms_panel_3",
    "untargeted_panel", "custom_targeted_panel", "low_abundant_metabolite_panel",
    "sample_origin", "sample_type", "mouse_number",
    "services_requested_per_sample", "mouse_strain", "mouse_flora",
    "volume_ul", "sample_weight_mg", "experiment_treatment"
  )
  
  map1 <- map1 %>%
    filter(!is.na(sampleid), sampleid != "") %>%
    slice(-(1:3))
  
  map2 <- read_excel(
    file.path("..", "scfa_metabolomics", "metadata", "PS2720_xtra_samples_HMM Submission Sheet.xlsx"),
    skip = 9
  ) %>%
    clean_names()
  
  colnames(map2) <- colnames(map1)
  
  map2 <- map2 %>%
    filter(!is.na(sampleid), sampleid != "") %>%
    slice(-(1:3))
  
  scfa_metadata_raw <- bind_rows(map1, map2)
}, error = function(e) {
  stop("Error reading SCFA metadata files: ", e$message)
})

scfa_metadata <- scfa_metadata_raw %>%
  mutate(
    subject = str_extract(sampleid, "^[0-9]+"),
    carb_repeat = str_extract(sampleid, "(?<= )[rs]\\d|nc(?= )"),
    carbohydrate_type = case_when(
      str_detect(carb_repeat, "^r") ~ "rapid_digestible",
      str_detect(carb_repeat, "^s") ~ "slow_digestible",
      str_detect(carb_repeat, "^nc") ~ "no_carbohydrate",
      TRUE ~ NA_character_
    ),
    technical_repeat = if_else(
      str_detect(carb_repeat, "^nc"),
      1L,
      as.integer(str_extract(carb_repeat, "\\d+"))
    ),
    timepoint_hr = as.integer(str_extract(sampleid, "(?<= )\\d+(?=h$)"))
  ) %>%
  dplyr::select(sampleid, group, subject, carbohydrate_type, technical_repeat, timepoint_hr) %>%
  mutate(group = tolower(group)) %>%
  filter(!is.na(group))

tryCatch({
  scfa_raw <- fread(
    file.path("..", "scfa_metabolomics", "data", "removed_qcs_quant_results_20250722_PFBBr_PS2720_20250731.csv")
  )
}, error = function(e) {
  stop("Error reading SCFA data file: ", e$message)
})

scfa_sample_data <- left_join(scfa_metadata, scfa_raw, by = "sampleid") %>%
  mutate(group = tolower(group)) %>%
  filter(!is.na(group))

scfa_analytes <- c("acetate", "butyrate", "propionate", "5aminovalerate", "succinate")

scfa_sample_data_long <- scfa_sample_data %>%
  pivot_longer(
    cols = all_of(scfa_analytes),
    names_to = "analyte",
    values_to = "concentration"
  ) %>%
  filter(!is.na(concentration)) %>%
  group_by(subject, group, carbohydrate_type, timepoint_hr, analyte) %>%
  summarise(concentration = mean(concentration, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    group_label = factor(str_to_title(group), levels = c("Control", "Case")),
    carbohydrate_type_label = str_replace_all(carbohydrate_type, "_", " ") %>%
      str_to_title()
  )

scfa_sample_data_long_factored <- scfa_sample_data_long %>%
  mutate(
    group = factor(group, levels = c("control", "case")),
    carbohydrate_type = factor(
      carbohydrate_type,
      levels = c("no_carbohydrate", "rapid_digestible", "slow_digestible")
    ),
    timepoint_hr = factor(timepoint_hr),
    subject = factor(subject)
  )

## Microbiome: metadata, phyloseq, and relative abundances --------------

micro_metadata <- NULL
ps <- NULL
ps_rel <- NULL

tryCatch({
  micro_metadata_raw <- readr::read_tsv(
    file.path(
      "..", "microbiome_analysis", "zr24558.16S_250813.zymo",
      "00...AllSamples.Bac16Sv34", "Sample_Information", "sample.metadata.csv"
    ),
    show_col_types = FALSE
  ) %>%
    janitor::clean_names()
}, error = function(e) {
  stop("Error reading microbiome metadata file: ", e$message)
})

micro_metadata <- micro_metadata_raw %>%
  # Clean up double dots in labels like "NC..48H" and trim whitespace
  dplyr::mutate(
    customer_label = stringr::str_replace_all(customer_label, "\\.\\.", "."),
    customer_label = stringr::str_trim(customer_label)
  ) %>%
  separate(
    customer_label,
    into = c("subject_id", "replicate_code", "timepoint"),
    sep = "\\.",
    remove = FALSE
  ) %>%
  dplyr::mutate(
    carbohydrate = case_when(
      stringr::str_detect(replicate_code, "R") ~ "RDC",
      stringr::str_detect(replicate_code, "S") ~ "SDC",
      TRUE ~ "No Added Carb"
    ),
    carbohydrate = factor(carbohydrate, levels = c("No Added Carb", "RDC", "SDC")),
    timepoint = factor(timepoint, levels = c("0H", "48H")),
    # Parse replicate number safely: treat NC as 1, numeric codes as their number
    replicate_code_clean = dplyr::if_else(
      stringr::str_detect(replicate_code, "NC"),
      NA_character_,
      replicate_code
    ),
    replicate = readr::parse_number(replicate_code_clean),
    replicate = tidyr::replace_na(replicate, 1L)
  ) %>%
  dplyr::select(-replicate_code_clean) %>%
  dplyr::rename(SampleID = sample_id) %>%
  dplyr::select(SampleID, subject_id, customer_label, timepoint, carbohydrate, replicate, raw_seq_files)

tryCatch({
  healthy_lookup <- readr::read_tsv(
    file.path(
      "..", "microbiome_analysis", "zr24558.16S_250813.zymo",
      "1.Healthy.TimepointComparism..Bac16Sv34", "Sample_Information", "sample.metadata.csv"
    ),
    show_col_types = FALSE
  ) %>%
    janitor::clean_names() %>%
    separate(customer_label, into = c("subject_id", "tmp", "tp"), sep = "\\.", remove = FALSE) %>%
    dplyr::distinct(subject_id) %>%
    dplyr::mutate(group = "Healthy")
}, error = function(e) {
  stop("Error reading microbiome healthy lookup file: ", e$message)
})

micro_metadata <- micro_metadata %>%
  mutate(SampleID = paste0("S", customer_label)) %>%
  dplyr::mutate(
    group = if_else(subject_id %in% healthy_lookup$subject_id, "Healthy", "Obese"),
    # Override MediaControl: group = NA, timepoint = "0H", carbohydrate = NA
    group = dplyr::if_else(
      stringr::str_detect(customer_label, regex("^MediaControl$", ignore_case = TRUE)),
      NA_character_,
      group
    ),
    timepoint = dplyr::if_else(
      stringr::str_detect(customer_label, regex("^MediaControl$", ignore_case = TRUE)),
      factor("0H", levels = levels(timepoint)),
      timepoint
    ),
    carbohydrate = dplyr::if_else(
      stringr::str_detect(customer_label, regex("^MediaControl$", ignore_case = TRUE)),
      factor(NA_character_, levels = levels(carbohydrate)),
      carbohydrate
    ),
    group = factor(group, levels = c("Healthy", "Obese")),
    subject_id = factor(subject_id)
  )

micro_metadata <- micro_metadata %>%
  dplyr::filter(!is.na(timepoint))

tryCatch({
  ps <- import_biom(
    file.path(
      "..", "microbiome_analysis", "zr24558.16S_250813.zymo",
      "00...AllSamples.Bac16Sv34", "DADA2_ASV_Distribution", "ASV_Table.biom"
    )
  )
}, error = function(e) {
  stop("Error importing microbiome BIOM file: ", e$message)
})

colnames(tax_table(ps)) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")

sample_names(ps) <- paste0("S", sample_names(ps))

metadata_ps <- micro_metadata %>%
  dplyr::mutate(SampleID = as.character(SampleID)) %>%
  dplyr::group_by(SampleID) %>%
  dplyr::mutate(SampleID = if (n() > 1) paste0(SampleID, "_", row_number()) else SampleID) %>%
  dplyr::ungroup() %>%
  column_to_rownames("SampleID")

sample_data(ps) <- sample_data(metadata_ps)

ps <- prune_taxa(taxa_sums(ps) > 0, ps)
ps <- prune_samples(sample_sums(ps) > 0, ps)

sample_data(ps)$SubjectID <- sample_data(ps)$subject_id
sample_data(ps)$Group <- sample_data(ps)$group
sample_data(ps)$Carbohydrate <- sample_data(ps)$carbohydrate
sample_data(ps)$Timepoint <- sample_data(ps)$timepoint

ps_rel <- microbiome::transform(ps, "compositional")

## qPCR absolute abundance data -----------------------------------------

qpcr_data <- NULL

tryCatch({
  qpcr_data <- readr::read_csv(
    file.path(
      "..", "microbiome_analysis", "zr24558.16S_250813.zymo",
      "00...AllSamples.Bac16Sv34", "Sample_Information", "absolute.abundance.csv"
    ),
    show_col_types = FALSE
  ) %>%
    janitor::clean_names()
}, error = function(e) {
  warning("Unable to load qPCR absolute abundance data: ", e$message)
})

# Note:
# - qPCR sample identifier column: customer_label
# - Will be joined to microbiome metadata via customer_label
# - genome_copies_per_ul will be used to convert relative to absolute
#   abundances in later sections.
```

# 1. Introduction

This integrated report brings together the short-chain fatty acid (SCFA) metabolomics and 16S rRNA gene microbiome analyses from the Abbott carbohydrate obesity project. It re-runs the key components of each analysis within a single, reproducible R Markdown workflow and adds new integrative components relating taxa, SCFA concentrations, and responder phenotypes.

The goal is to maintain the rigor and depth of the original analyses while ensuring that metadata, filtering, and modeling decisions are harmonized across datasets.

# 2. Project Objectives

In this section, we summarize the main scientific and analytical objectives, building on the SCFA project objectives and extending them to explicitly include microbiome composition, diversity, and taxon–SCFA relationships.

```{r objectives-text, echo=FALSE, results='asis'}
# TODO: programmatically pull or summarize objectives text
```

# 3. Methods

Here we integrate the methods from the SCFA metabolomics and microbiome analyses, including:

- Study design and sample processing
- SCFA quantification and preprocessing
- Microbiome sequencing, processing, and phyloseq construction
- Statistical modeling strategies (mixed-effects models, diversity analyses, differential abundance, and MaAsLin2)
- Definition of responder vs non-responder groups for butyrate and propionate based on 0 h to 48 h changes

```{r methods-text, echo=FALSE, results='asis'}
# TODO: construct a combined methods narrative, reusing text and
# code-derived summaries from the source Rmd files
```

# 4. Summary Data

We begin with a concise overview of the sample counts by obesity status (Control vs Case) and carbohydrate condition across both SCFA and microbiome datasets.

```{r summary-data, echo=FALSE}
# Summary counts for SCFA (control/case) and microbiome (Healthy/Obese)

# SCFA: deduplicate to one row per biological sample (subject × timepoint × carbohydrate)
scfa_sample_overview <- scfa_sample_data_long_factored %>%
  dplyr::distinct(subject, group, carbohydrate_type, timepoint_hr) %>%
  mutate(
    group_label = factor(str_to_title(group), levels = c("Control", "Case")),
    carbohydrate = case_when(
      carbohydrate_type == "no_carbohydrate" ~ "No Added Carb",
      carbohydrate_type == "rapid_digestible" ~ "RDC",
      carbohydrate_type == "slow_digestible" ~ "SDC",
      TRUE ~ NA_character_
    ),
    carbohydrate = factor(carbohydrate, levels = c("No Added Carb", "RDC", "SDC"))
  )

scfa_summary_counts <- scfa_sample_overview %>%
  dplyr::count(group_label, carbohydrate, name = "n_samples") %>%
  dplyr::arrange(group_label, carbohydrate)

# Microbiome: sample counts by group and carbohydrate (all timepoints)
micro_summary_counts <- micro_metadata %>%
  dplyr::count(group, carbohydrate, timepoint, name = "n_samples") %>%
  dplyr::arrange(group, carbohydrate, timepoint)

scfa_table <- scfa_summary_counts %>%
  dplyr::rename(
    Group = group_label,
    Carbohydrate = carbohydrate,
    `N samples` = n_samples
  )

micro_table <- micro_summary_counts %>%
  dplyr::rename(
    Group = group,
    Carbohydrate = carbohydrate,
    Timepoint = timepoint,
    `N samples` = n_samples
  )

if (knitr::is_html_output()) {
  htmltools::tagList(
    htmltools::h3("SCFA samples by group and carbohydrate"),
    DT::datatable(
      scfa_table,
      options = list(pageLength = 10, dom = "tip"),
      rownames = FALSE
    ),
    htmltools::h3("Microbiome samples by group, carbohydrate, and timepoint"),
    DT::datatable(
      micro_table,
      options = list(pageLength = 10, dom = "tip"),
      rownames = FALSE
    )
  )
} else {
  list(
    SCFA = scfa_table,
    Microbiome = micro_table
  )
}
```

# 5. Metabolomics (SCFA)

## 5.1 Section Overview

This section reproduces and extends the SCFA analyses, focusing on:

- Carbohydrate type effects (combined and stratified by group)
- SCFA concentrations over time
- Advanced mixed-effects modeling for SCFA trajectories

We first quantify SCFA response magnitudes (48h – 0h) and evaluate how they vary by carbohydrate type and obesity group. We then visualize SCFA concentrations over time and fit linear mixed-effects models with subject as a random effect to formally test fixed effects of group, carbohydrate type, time, and their interactions on absolute SCFA concentrations.

## 5.2 Carbohydrate Type Effects (Combined Groups)

```{r scfa-carbohydrate-effects, echo=FALSE}
# Calculate delta values (48h - 0h) using integrated SCFA data
delta_data <- scfa_sample_data_long_factored %>%
  dplyr::filter(timepoint_hr %in% c("0", "48")) %>%
  tidyr::pivot_wider(
    id_cols = c(subject, group, carbohydrate_type, analyte, group_label, carbohydrate_type_label),
    names_from = timepoint_hr,
    values_from = concentration,
    names_prefix = "time_"
  ) %>%
  dplyr::mutate(delta_48h_0h = time_48 - time_0) %>%
  dplyr::filter(!is.na(delta_48h_0h))

# Control and Case group-specific delta plots
plot_delta_control_only <- delta_data %>%
  dplyr::filter(group == "control") %>%
  ggplot(aes(x = carbohydrate_type_label, y = delta_48h_0h)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", alpha = 0.7) +
  geom_boxplot(aes(fill = carbohydrate_type_label), alpha = 0.8, outlier.shape = NA) +
  geom_jitter(alpha = 0.6, width = 0.2, size = 1.5, color = "gray30") +
  facet_wrap(~ analyte, scales = "free_y") +
  scale_fill_scfa() +
  ggpubr::stat_compare_means(
    comparisons = list(
      c("No Carbohydrate", "Rapid Digestible"),
      c("No Carbohydrate", "Slow Digestible"),
      c("Rapid Digestible", "Slow Digestible")
    ),
    method = "wilcox.test",
    label = "p.signif",
    step.increase = 0.1
  ) +
  labs(
    title = "SCFA Response Magnitude by Carbohydrate Type - Control Group Only",
    subtitle = "Delta change (48h - 0h) with Wilcoxon pairwise comparisons",
    x = "Carbohydrate Type",
    y = "Change in Concentration (48h - 0h) μM"
  ) +
  theme_scfa() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot_delta_case_only <- delta_data %>%
  dplyr::filter(group == "case") %>%
  ggplot(aes(x = carbohydrate_type_label, y = delta_48h_0h)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", alpha = 0.7) +
  geom_boxplot(aes(fill = carbohydrate_type_label), alpha = 0.8, outlier.shape = NA) +
  geom_jitter(alpha = 0.6, width = 0.2, size = 1.5, color = "gray30") +
  facet_wrap(~ analyte, scales = "free_y") +
  scale_fill_scfa() +
  ggpubr::stat_compare_means(
    comparisons = list(
      c("No Carbohydrate", "Rapid Digestible"),
      c("No Carbohydrate", "Slow Digestible"),
      c("Rapid Digestible", "Slow Digestible")
    ),
    method = "wilcox.test",
    label = "p.signif",
    step.increase = 0.1
  ) +
  labs(
    title = "SCFA Response Magnitude by Carbohydrate Type - Case Group Only",
    subtitle = "Delta change (48h - 0h) with Wilcoxon pairwise comparisons",
    x = "Carbohydrate Type",
    y = "Change in Concentration (48h - 0h) μM"
  ) +
  theme_scfa() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Combined groups analysis
plot_delta_carb_combined <- delta_data %>%
  ggplot(aes(x = carbohydrate_type_label, y = delta_48h_0h)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", alpha = 0.7) +
  geom_boxplot(aes(fill = carbohydrate_type_label), alpha = 0.8, outlier.shape = NA) +
  geom_jitter(alpha = 0.5, width = 0.2, size = 1.5, color = "gray30") +
  facet_wrap(~ analyte, scales = "free_y") +
  scale_fill_scfa() +
  ggpubr::stat_compare_means(
    comparisons = list(
      c("No Carbohydrate", "Rapid Digestible"),
      c("No Carbohydrate", "Slow Digestible"),
      c("Rapid Digestible", "Slow Digestible")
    ),
    method = "wilcox.test",
    label = "p.signif",
    step.increase = 0.1
  ) +
  labs(
    title = "SCFA Response Magnitude by Carbohydrate Type (Combined Groups)",
    subtitle = "Delta change (48h - 0h) with Wilcoxon pairwise comparisons",
    x = "Carbohydrate Type",
    y = "Change in Concentration (48h - 0h) μM"
  ) +
  theme_scfa() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# One-way ANOVA on delta by carbohydrate type (combined groups)
combined_carb_stats <- purrr::map_dfr(
  unique(delta_data$analyte),
  function(analyte_name) {
    analyte_data <- delta_data %>% dplyr::filter(analyte == analyte_name)
    model <- aov(delta_48h_0h ~ carbohydrate_type, data = analyte_data)
    anova_summary <- summary(model)[[1]]
    data.frame(
      analyte = analyte_name,
      Effect = rownames(anova_summary),
      DF = anova_summary$Df,
      Sum_Sq = anova_summary$`Sum Sq`,
      Mean_Sq = anova_summary$`Mean Sq`,
      F_value = anova_summary$`F value`,
      p = anova_summary$`Pr(>F)`
    ) %>%
      dplyr::filter(Effect != "Residuals") %>%
      dplyr::mutate(
        p_adj = p.adjust(p, method = "BH"),
        significance = dplyr::case_when(
          p_adj < 0.001 ~ "p<0.001",
          p_adj < 0.01 ~ "p<0.01",
          p_adj < 0.05 ~ "p<0.05",
          TRUE ~ "ns"
        )
      )
  }
)

# Print plots and table - ggplot objects render automatically in HTML
cat("#### Control group: delta SCFA by carbohydrate type\n\n")
print(plot_delta_control_only)

cat("\n\n#### Case group: delta SCFA by carbohydrate type\n\n")
print(plot_delta_case_only)

cat("\n\n#### Combined groups: delta SCFA by carbohydrate type\n\n")
print(plot_delta_carb_combined)

# Display statistics table
if (knitr::is_html_output()) {
  DT::datatable(
    combined_carb_stats %>%
      dplyr::select(analyte, Effect, p, p_adj, significance),
    caption = "One-way ANOVA: Carbohydrate Type Effects on Delta Changes (Combined Groups)",
    options = list(pageLength = 10, scrollX = TRUE),
    rownames = FALSE
  ) %>%
    DT::formatRound(columns = c("p", "p_adj"), digits = 4)
} else {
  knitr::kable(
    combined_carb_stats %>%
      dplyr::select(analyte, Effect, p, p_adj, significance),
    digits = 4,
    caption = "One-way ANOVA: Carbohydrate Type Effects on Delta Changes (Combined Groups)"
  )
}
```

## 5.3 SCFA Concentrations Over Time

```{r scfa-over-time, echo=FALSE}
plot_time_series <- scfa_sample_data_long %>%
  dplyr::group_by(carbohydrate_type_label, timepoint_hr, analyte) %>%
  dplyr::summarise(
    mean_conc = mean(concentration, na.rm = TRUE),
    sem = sd(concentration, na.rm = TRUE) / sqrt(dplyr::n()),
    .groups = "drop"
  ) %>%
  ggplot(aes(
    x = as.numeric(as.character(timepoint_hr)),
    y = mean_conc,
    color = carbohydrate_type_label,
    group = carbohydrate_type_label
  )) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = mean_conc - sem, ymax = mean_conc + sem),
    width = 1,
    linewidth = 0.8
  ) +
  facet_wrap(~ analyte, scales = "free_y") +
  scale_color_scfa() +
  labs(
    title = "SCFA Concentrations Over Time by Carbohydrate Type",
    x = "Time (hours)",
    y = "Mean Concentration ± SEM (μM)",
    color = "Carbohydrate Type"
  ) +
  theme_scfa() +
  theme(legend.position = "bottom")

plot_time_series
```

## 5.4 Advanced Statistical Modeling: Mixed-Effects Models

```{r scfa-mixed-effects, echo=FALSE}
# Mixed-effects models for absolute SCFA concentrations
mixed_effects_results <- scfa_sample_data_long_factored %>%
  dplyr::group_by(analyte) %>%
  tidyr::nest() %>%
  dplyr::mutate(
    # Use lmerTest::lmer so that anova() returns F-tests with p-values
    model = purrr::map(
      data,
      ~ lmerTest::lmer(
        concentration ~ group * carbohydrate_type * timepoint_hr + (1 | subject),
        data = .x
      )
    ),
    anova_main = purrr::map(model, ~ anova(.x)),
    model_summary = purrr::map(model, ~ summary(.x))
  )

main_anova_summary <- mixed_effects_results %>%
  dplyr::mutate(
    anova_df = purrr::map(
      anova_main,
      ~ as.data.frame(.x) %>%
        tibble::rownames_to_column("effect")
    )
  ) %>%
  dplyr::select(analyte, anova_df) %>%
  tidyr::unnest(anova_df) %>%
  dplyr::filter(effect != "Residual") %>%
  dplyr::rename(
    F_value = `F value`,
    pr_f = `Pr(>F)`
  ) %>%
  dplyr::mutate(
    p_adj = p.adjust(pr_f, method = "BH"),
    significance = dplyr::case_when(
      p_adj < 0.001 ~ "p<0.001",
      p_adj < 0.01 ~ "p<0.01",
      p_adj < 0.05 ~ "p<0.05",
      TRUE ~ "ns"
    )
  ) %>%
  dplyr::arrange(p_adj)

if (knitr::is_html_output()) {
  DT::datatable(
    main_anova_summary %>%
      dplyr::select(analyte, effect, F_value, pr_f, p_adj, significance),
    caption = "Mixed-Effects Models for SCFA Concentrations: ANOVA Table",
    options = list(pageLength = 10, scrollX = TRUE),
    rownames = FALSE
  ) %>%
    DT::formatRound(columns = c("F_value", "pr_f", "p_adj"), digits = 4)
} else {
  knitr::kable(
    main_anova_summary %>%
      dplyr::select(analyte, effect, F_value, pr_f, p_adj, significance),
    digits = 4,
    caption = "Mixed-Effects Models for SCFA Concentrations: ANOVA Table"
  )
}
```

## 5.5 Summary of SCFA Results

```{r scfa-summary-text, echo=FALSE, results='asis'}
# TODO: summary narrative highlighting key SCFA findings
```

# 6. Microbiome

## 6.1 Section Overview

This section reproduces key microbiome results, including alpha and beta diversity, community composition, and differential abundance, using the harmonized metadata and styling. We focus first on within-sample diversity (alpha diversity) to summarize richness and evenness patterns across groups, carbohydrate conditions, and timepoints, then move to between-sample diversity (beta diversity) and taxon-level changes.

## 6.2 Alpha Diversity

```{r alpha-diversity, echo=FALSE}
# Alpha diversity metrics: Observed, Shannon, Simpson at 0H and 48H

alpha_metrics <- estimate_richness(ps, measures = c("Observed", "Shannon", "Simpson")) %>%
  tibble::rownames_to_column("SampleID")

alpha_long <- alpha_metrics %>%
  tidyr::pivot_longer(
    cols = -SampleID,
    names_to = "metric",
    values_to = "value"
  ) %>%
  dplyr::left_join(micro_metadata, by = "SampleID") %>%
  dplyr::mutate(
    metric = factor(metric, levels = c("Shannon", "Observed", "Simpson"))
  )

alpha_long_0_48 <- alpha_long %>%
  dplyr::filter(timepoint %in% c("0H", "48H"))

# Alpha diversity boxplots by group, carbohydrate, and timepoint
# with per-facet Healthy vs Obese Wilcoxon p-values
alpha_plot <- alpha_long_0_48 %>%
  ggplot(aes(x = carbohydrate, y = value, fill = group)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA, position = position_dodge(width = 0.8)) +
  geom_jitter(
    aes(color = group),
    alpha = 0.5,
    size = 1.2,
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8)
  ) +
  ggpubr::stat_compare_means(
    aes(group = group),
    method = "wilcox.test",
    label = "p.format",
    label.y = Inf,
    vjust = 1.5
  ) +
  facet_grid(metric ~ timepoint, scales = "free_y") +
  scale_fill_scfa() +
  scale_color_scfa() +
  labs(
    title = "Alpha Diversity by Group, Carbohydrate, and Timepoint",
    x = "Carbohydrate",
    y = "Diversity metric"
  ) +
  theme_scfa() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Summaries at 48H for combined groups modeling
alpha_48h <- alpha_long %>%
  dplyr::filter(timepoint == "48H")

alpha_48h_summary <- alpha_48h %>%
  dplyr::group_by(metric, group, carbohydrate) %>%
  dplyr::summarise(
    mean = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    n = dplyr::n(),
    .groups = "drop"
  )

# Wilcoxon tests: Healthy vs Obese within carbohydrate × timepoint for each alpha metric
alpha_wilcox_group <- alpha_long_0_48 %>%
  dplyr::group_by(metric, carbohydrate, timepoint) %>%
  dplyr::filter(dplyr::n_distinct(group) == 2) %>%  # Only test if both groups present
  rstatix::wilcox_test(value ~ group, p.adjust.method = "BH") %>%
  rstatix::add_significance() %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    comparison = paste0(group1, " vs ", group2),
    metric_carb_time = paste(metric, carbohydrate, timepoint, sep = " | ")
  ) %>%
  dplyr::select(metric, carbohydrate, timepoint, comparison, p, p.adj, p.adj.signif)

# Wilcoxon tests: Carbohydrate comparisons within group × timepoint for each alpha metric
alpha_wilcox_carb <- alpha_long_0_48 %>%
  dplyr::group_by(metric, group, timepoint) %>%
  dplyr::filter(dplyr::n_distinct(carbohydrate) >= 2) %>%  # Only test if multiple carb levels present
  rstatix::wilcox_test(value ~ carbohydrate, p.adjust.method = "BH") %>%
  rstatix::add_significance() %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    comparison = paste0(group1, " vs ", group2),
    metric_group_time = paste(metric, group, timepoint, sep = " | ")
  ) %>%
  dplyr::select(metric, group, timepoint, comparison, p, p.adj, p.adj.signif)

# Print plot and tables - ggplot objects render automatically in HTML
print(alpha_plot)

# Display summary table
cat("\n\n#### Alpha diversity at 48H: summary by metric, group, and carbohydrate\n\n")
if (knitr::is_html_output()) {
  DT::datatable(
    alpha_48h_summary,
    caption = "Alpha diversity at 48H: summary by metric, group, and carbohydrate",
    options = list(pageLength = 10, scrollX = TRUE),
    rownames = FALSE
  ) %>%
    DT::formatRound(columns = c("mean", "sd"), digits = 3)
} else {
  knitr::kable(
    alpha_48h_summary,
    digits = 3,
    caption = "Alpha diversity at 48H: summary by metric, group, and carbohydrate"
  )
}

# Display Wilcoxon group comparison table
cat("\n\n#### Wilcoxon tests: Healthy vs Obese within carbohydrate × timepoint\n\n")
if (knitr::is_html_output()) {
  DT::datatable(
    alpha_wilcox_group %>%
      dplyr::select(metric, carbohydrate, timepoint, comparison, p, p.adj, p.adj.signif),
    caption = "Wilcoxon tests: Healthy vs Obese within carbohydrate × timepoint for each alpha metric",
    options = list(pageLength = 10, scrollX = TRUE),
    rownames = FALSE
  ) %>%
    DT::formatRound(columns = c("p", "p.adj"), digits = 4)
} else {
  knitr::kable(
    alpha_wilcox_group %>%
      dplyr::select(metric, carbohydrate, timepoint, comparison, p, p.adj, p.adj.signif),
    digits = 4
  )
}

# Display Wilcoxon carbohydrate comparison table
cat("\n\n#### Wilcoxon tests: Carbohydrate comparisons within group × timepoint\n\n")
if (knitr::is_html_output()) {
  DT::datatable(
    alpha_wilcox_carb %>%
      dplyr::select(metric, group, timepoint, comparison, p, p.adj, p.adj.signif),
    caption = "Wilcoxon tests: Carbohydrate comparisons within group × timepoint for each alpha metric",
    options = list(pageLength = 10, scrollX = TRUE),
    rownames = FALSE
  ) %>%
    DT::formatRound(columns = c("p", "p.adj"), digits = 4)
} else {
  knitr::kable(
    alpha_wilcox_carb %>%
      dplyr::select(metric, group, timepoint, comparison, p, p.adj, p.adj.signif),
    digits = 4
  )
}
```

## 6.3 Beta Diversity

```{r beta-diversity, echo=FALSE}
# Beta diversity and community structure using Bray-Curtis distance

# Helper: PCoA data frame
make_pcoa_df <- function(ps_obj, distance_method) {
  ord <- ordinate(ps_obj, method = "PCoA", distance = distance_method)
  vecs <- as_tibble(ord$vectors[, 1:2], rownames = "SampleID")
  colnames(vecs)[2:3] <- c("PC1", "PC2")
  vecs <- vecs %>%
    dplyr::mutate(
      axis1_var = ord$values$Relative_eig[1],
      axis2_var = ord$values$Relative_eig[2]
    ) %>%
    dplyr::left_join(
      as_tibble(micro_metadata, .name_repair = "minimal"),
      by = "SampleID"
    )
  vecs
}

# Bray-Curtis PCoA across all samples
pcoa_bray <- make_pcoa_df(ps_rel, "bray")

# PERMANOVA: group × carbohydrate + timepoint with subject-level strata
dist_bray <- phyloseq::distance(ps_rel, method = "bray")
df_beta <- as(sample_data(ps_rel), "data.frame")

permanova_bray <- vegan::adonis2(
  dist_bray ~ group * carbohydrate + timepoint,
  data = df_beta,
  permutations = 999,
  strata = df_beta$subject_id
)

permanova_bray_tbl <- as.data.frame(permanova_bray) %>%
  rownames_to_column("term") %>%
  janitor::clean_names()

# Try to standardize column names for downstream use
permanova_bray_p_col <- intersect(
  c("pr_greater_f_", "pr_greater_f", "pr_f", "p_value"),
  names(permanova_bray_tbl)
)
permanova_bray_f_col <- intersect(
  c("f", "f_model", "pseudo_f"),
  names(permanova_bray_tbl)
)
permanova_bray_r2_col <- intersect(
  c("r2", "r_squared", "r2_adj"),
  names(permanova_bray_tbl)
)

permanova_bray_tbl <- permanova_bray_tbl %>%
  dplyr::mutate(
    f = .data[[permanova_bray_f_col[1]]],
    r2 = .data[[permanova_bray_r2_col[1]]],
    p_value = .data[[permanova_bray_p_col[1]]]
  )

# PCoA plot with PERMANOVA annotation (R² and p for group:carbohydrate)
permanova_gc <- permanova_bray_tbl %>%
  dplyr::filter(term == "group:carbohydrate")

annotation_text <- NULL
if (nrow(permanova_gc) > 0) {
  annotation_text <- glue::glue(
    "R² = {round(permanova_gc$r2[1], 3)}, p = {signif(permanova_gc$p_value[1], 3)}"
  )
}

beta_pcoa_plot <- {
  p <- pcoa_bray %>%
    ggplot(aes(x = PC1, y = PC2, color = carbohydrate, shape = group)) +
    geom_point(size = 2.5, alpha = 0.85) +
    stat_ellipse(
      aes(group = interaction(carbohydrate, timepoint), color = carbohydrate),
      linewidth = 0.4,
      linetype = "dashed",
      show.legend = FALSE
    ) +
    facet_wrap(~timepoint) +
    scale_color_scfa() +
    scale_shape_manual(values = c(16, 17)) +
    labs(
      x = glue::glue("PC1 ({scales::percent(round(pcoa_bray$axis1_var[1], 3))})"),
      y = glue::glue("PC2 ({scales::percent(round(pcoa_bray$axis2_var[1], 3))})"),
      title = "PCoA – Bray-Curtis distance"
    ) +
    theme_scfa()

  if (!is.null(annotation_text)) {
    p <- p +
      annotate(
        "label",
        x = -Inf,
        y = Inf,
        hjust = -0.05,
        vjust = 1.1,
        label = annotation_text,
        size = 3,
        fontface = "bold",
        fill = "white",
        alpha = 0.8
      )
  }

  p
}

# 48H-only Bray-Curtis PCoA and PERMANOVA
ps_48h <- phyloseq::subset_samples(ps_rel, Timepoint == "48H")
pcoa_48h <- make_pcoa_df(ps_48h, "bray")

dist_48h <- phyloseq::distance(ps_48h, method = "bray")
metadata_48h <- as(sample_data(ps_48h), "data.frame")

permanova_48h <- vegan::adonis2(
  dist_48h ~ group * carbohydrate,
  data = metadata_48h,
  permutations = 999,
  strata = metadata_48h$subject_id
)

permanova_48h_tbl <- as.data.frame(permanova_48h) %>%
  rownames_to_column("term") %>%
  janitor::clean_names()

permanova_48h_p_col <- intersect(
  c("pr_greater_f_", "pr_greater_f", "pr_f", "p_value"),
  names(permanova_48h_tbl)
)
permanova_48h_f_col <- intersect(
  c("f", "f_model", "pseudo_f"),
  names(permanova_48h_tbl)
)
permanova_48h_r2_col <- intersect(
  c("r2", "r_squared", "r2_adj"),
  names(permanova_48h_tbl)
)

permanova_48h_tbl <- permanova_48h_tbl %>%
  dplyr::mutate(
    f = .data[[permanova_48h_f_col[1]]],
    r2 = .data[[permanova_48h_r2_col[1]]],
    p_value = .data[[permanova_48h_p_col[1]]]
  )

beta_pcoa_48h_plot <- pcoa_48h %>%
  ggplot(aes(x = PC1, y = PC2, color = carbohydrate, shape = group)) +
  geom_point(size = 3, alpha = 0.9) +
  stat_ellipse(aes(group = carbohydrate, color = carbohydrate),
               linewidth = 0.8, linetype = "dashed", alpha = 0.8) +
  scale_color_scfa() +
  scale_shape_manual(values = c(16, 17)) +
  labs(
    x = glue::glue("PC1 ({scales::percent(round(pcoa_48h$axis1_var[1], 3))})"),
    y = glue::glue("PC2 ({scales::percent(round(pcoa_48h$axis2_var[1], 3))})"),
    title = "PCoA – Bray-Curtis distance (48H samples)"
  ) +
  theme_scfa()

if (knitr::is_html_output()) {
  htmltools::tagList(
    beta_pcoa_plot,
    DT::datatable(
      permanova_bray_tbl %>%
        dplyr::select(term, f, r2, p_value),
      caption = "PERMANOVA results (Bray-Curtis; all timepoints, subject-level strata)",
      options = list(pageLength = 10, scrollX = TRUE),
      rownames = FALSE
    ) %>%
      DT::formatRound(columns = c("f", "r2", "p_value"), digits = 3),
    beta_pcoa_48h_plot,
    DT::datatable(
      permanova_48h_tbl %>%
        dplyr::select(term, f, r2, p_value),
      caption = "PERMANOVA results (Bray-Curtis; 48H only, subject-level strata)",
      options = list(pageLength = 10, scrollX = TRUE),
      rownames = FALSE
    ) %>%
      DT::formatRound(columns = c("f", "r2", "p_value"), digits = 3)
  )
} else {
  print(beta_pcoa_plot)
  print(beta_pcoa_48h_plot)
  knitr::kable(
    permanova_bray_tbl %>%
      dplyr::select(term, f, r2, p_value),
    digits = 3,
    caption = "PERMANOVA results (Bray-Curtis; all timepoints, subject-level strata)"
  )
  knitr::kable(
    permanova_48h_tbl %>%
      dplyr::select(term, f, r2, p_value),
    digits = 3,
    caption = "PERMANOVA results (Bray-Curtis; 48H only, subject-level strata)"
  )
}
```

## 6.4 Community Composition


### 6.4.1 Genus-Level Aggregation and Overview

```{r genus-aggregation, echo=FALSE}
# Create genus-level aggregation using tax_glom
ps_genus <- tax_glom(ps, taxrank = "Genus", NArm = FALSE)
ps_genus <- prune_taxa(taxa_sums(ps_genus) > 0, ps_genus)

# Keep only top 20 most abundant genera for visualization
genus_abundance <- sort(taxa_sums(ps_genus), decreasing = TRUE)
top_genera <- names(genus_abundance)[1:min(20, length(genus_abundance))]
ps_genus <- prune_taxa(top_genera, ps_genus)

# Genus-level bar plots by group, carbohydrate, and timepoint
ps_genus_rel <- transform_sample_counts(ps_genus, function(x) x / sum(x))

genus_barplot_all <- psmelt(ps_genus_rel) %>%
  ggplot(aes(x = carbohydrate, y = Abundance, fill = Genus)) +
  geom_col(position = "fill") +
  facet_grid(group ~ timepoint) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = scico::scico(21, palette = "batlow")) +
  labs(
    x = "Carbohydrate",
    y = "Relative abundance",
    title = "Genus-level composition by group, carbohydrate, and timepoint",
    fill = "Genus"
  ) +
  theme_scfa() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Genus-level bar plot at 48H only
ps_48h_genus <- phyloseq::subset_samples(ps_genus, timepoint == "48H")

genus_barplot_48h <- psmelt(transform_sample_counts(ps_48h_genus, function(x) x / sum(x))) %>%
  ggplot(aes(x = carbohydrate, y = Abundance, fill = Genus)) +
  geom_col(position = "fill") +
  facet_wrap(~ group) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = scico::scico(21, palette = "batlow")) +
  labs(
    x = "Carbohydrate",
    y = "Relative abundance",
    title = "Genus-level composition at 48H",
    fill = "Genus"
  ) +
  theme_scfa() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

if (knitr::is_html_output()) {
  htmltools::tagList(
    genus_barplot_all,
    genus_barplot_48h
  )
} else {
  print(genus_barplot_all)
  print(genus_barplot_48h)
}
```

### 6.4.2 Genus Composition Changes from 0H to 48H by Carbohydrate Type

```{r genus-alluvial, echo=FALSE, fig.width=14, fig.height=10}
# Prepare data for alluvial diagrams showing flow from 0H to 48H
ps_genus_rel <- transform_sample_counts(ps_genus, function(x) x / sum(x))

# Extract data
otu_mat <- as.matrix(otu_table(ps_genus_rel))
if (!taxa_are_rows(ps_genus_rel)) {
  otu_mat <- t(otu_mat)
}

tax_df <- as.data.frame(tax_table(ps_genus_rel))
meta_df <- data.frame(sample_data(ps_genus_rel)) %>%
  tibble::rownames_to_column("SampleID")

# Create long format data
alluvial_data_raw <- otu_mat %>%
  as.data.frame() %>%
  tibble::rownames_to_column("ASV") %>%
  tidyr::pivot_longer(-ASV, names_to = "SampleID", values_to = "Abundance") %>%
  dplyr::left_join(
    tax_df %>% tibble::rownames_to_column("ASV") %>% dplyr::select(ASV, Genus),
    by = "ASV"
  ) %>%
  dplyr::left_join(
    meta_df %>% dplyr::select(SampleID, timepoint, carbohydrate),
    by = "SampleID"
  ) %>%
  dplyr::filter(timepoint %in% c("0H", "48H"))

# Remove NA values and filter to only valid combinations
alluvial_data_raw_clean <- alluvial_data_raw %>%
  dplyr::filter(
    !is.na(Abundance),
    !is.na(Genus),
    !is.na(timepoint),
    !is.na(carbohydrate)
  )

# Group by Genus, timepoint, carbohydrate to get mean abundance
# Use na.rm = TRUE to handle any remaining edge cases
alluvial_data_grouped <- alluvial_data_raw_clean %>%
  dplyr::group_by(Genus, timepoint, carbohydrate) %>%
  dplyr::summarise(mean_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")

# Get top 12 genera for visualization (use max abundance across all timepoints/carbs)
# But don't filter the data yet - we'll use all data and then aggregate
top_genera_by_abundance <- alluvial_data_grouped %>%
  dplyr::group_by(Genus) %>%
  dplyr::summarise(max_abund = max(mean_abundance), .groups = "drop") %>%
  dplyr::arrange(desc(max_abund)) %>%
  head(12) %>%
  dplyr::pull(Genus)

# Define genera that are important from DA analysis and should be included
# Check with single underscore "g_", double underscore "g__", and without prefix
# since taxonomy formats may vary (tax_table uses "g__" format)
da_important_genera <- c(
  "g_Fusicatenibacter",
  "g__Fusicatenibacter",
  "Fusicatenibacter",
  "g_Fusicatenibacters",
  "g__Fusicatenibacters",
  "Fusicatenibacters",
  "g_Hungatella",
  "g__Hungatella",
  "Hungatella",
  "g_Eggerthella",
  "g__Eggerthella",
  "Eggerthella",
  "g_Peptoniphilus",
  "g__Peptoniphilus",
  "Peptoniphilus"
)

# Check which of these genera exist in the data
# IMPORTANT: Get available_genera from the original tax_table, not from grouped data
# This ensures we find all genera in the taxonomy, even if they have zero/low abundance
available_genera_from_tax <- unique(tax_df$Genus)
available_genera_grouped <- unique(alluvial_data_grouped$Genus)

# Check exact matches first
da_genera_in_data <- da_important_genera[da_important_genera %in% available_genera_from_tax]

# Also check for variations in the full taxonomy (e.g., "g__Fusicatenibacter sp." might be stored differently)
# This will catch any genus containing these names, regardless of prefix or suffix
da_genera_variations <- available_genera_from_tax[
  stringr::str_detect(available_genera_from_tax, regex("Fusicatenibacter|Hungatella|Eggerthella|Peptoniphilus", ignore_case = TRUE))
]

# Combine exact matches and variations
da_genera_found <- unique(c(da_genera_in_data, da_genera_variations))

# Combine top 12 by abundance with DA-important genera
# Remove duplicates and keep up to 16 genera total
top_genera_alluvial <- unique(c(top_genera_by_abundance, da_genera_found))
if (length(top_genera_alluvial) > 16) {
  # If we have more than 16, prioritize top 12 by abundance, then add DA genera
  top_genera_alluvial <- c(
    top_genera_by_abundance,
    setdiff(da_genera_found, top_genera_by_abundance)
  )[1:min(16, length(top_genera_alluvial))]
}

# Filter to top genera and add "Other" category
# Use the grouped data (which has mean abundance per Genus × timepoint × carbohydrate)
# IMPORTANT: Use ALL data from alluvial_data_grouped (not filtered), so we preserve
# low-abundance combinations like RDC 0H
# Replace NaN with 0 before aggregating
alluvial_plot_data_pre_complete <- alluvial_data_grouped %>%
  dplyr::mutate(
    mean_abundance = ifelse(is.nan(mean_abundance) | is.na(mean_abundance), 0, mean_abundance),
    Genus_grouped = ifelse(Genus %in% top_genera_alluvial, as.character(Genus), "Other")
  ) %>%
  dplyr::group_by(Genus_grouped, timepoint, carbohydrate) %>%
  dplyr::summarise(Abundance = sum(mean_abundance, na.rm = TRUE), .groups = "drop") %>%
  dplyr::mutate(
    Abundance = ifelse(is.nan(Abundance) | is.na(Abundance), 0, Abundance)
  )

# Ensure we have data for both timepoints for each carbohydrate AND all Genus_grouped values
alluvial_plot_data <- alluvial_plot_data_pre_complete %>%
  tidyr::complete(
    Genus_grouped = c(top_genera_alluvial, "Other"),
    timepoint = c("0H", "48H"),
    carbohydrate = c("No Added Carb", "RDC", "SDC"),
    fill = list(Abundance = 0)
  )

# Create color palette
genus_colors <- c(
  scico::scico(length(top_genera_alluvial), palette = "batlow"),
  "Other" = "grey80"
)
names(genus_colors)[1:length(top_genera_alluvial)] <- top_genera_alluvial

# Create alluvial plots for each carbohydrate
plot_list_alluvial <- list()

for (carb in c("No Added Carb", "RDC", "SDC")) {
  plot_data <- alluvial_plot_data %>%
    dplyr::filter(carbohydrate == carb)
  
  # Ensure both timepoints exist, even if one has zero abundance (fix for RDC panel)
  # Get all unique Genus_grouped values first
  all_genera <- unique(plot_data$Genus_grouped)
  plot_data <- plot_data %>%
    tidyr::complete(
      Genus_grouped = all_genera,
      timepoint = c("0H", "48H"),
      fill = list(Abundance = 0, carbohydrate = carb)
    ) %>%
    dplyr::mutate(
      Genus_grouped = factor(
        Genus_grouped,
        levels = c(top_genera_alluvial[top_genera_alluvial %in% unique(Genus_grouped)], "Other")
      ),
      timepoint = factor(timepoint, levels = c("0H", "48H")),
      carbohydrate = carb
    )
  
  p <- ggplot(
    plot_data,
    aes(
      x = timepoint,
      y = Abundance,
      alluvium = Genus_grouped,
      stratum = Genus_grouped,
      fill = Genus_grouped
    )
  ) +
    geom_alluvium(alpha = 0.7, width = 0.3) +
    geom_stratum(width = 0.3, alpha = 0.9) +
    stat_stratum(
      geom = "text",
      aes(label = Genus_grouped),
      size = 2.3,
      fontface = "bold"
    ) +
    scale_fill_manual(values = genus_colors) +
    scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
    scale_x_discrete(expand = expansion(add = c(0.3, 0.3))) +
    labs(
      title = carb,
      x = "Timepoint",
      y = "Relative Abundance",
      fill = "Genus"
    ) +
    theme_scfa() +
    theme(
      legend.position = "none",
      panel.grid.major.x = element_blank()
    )
  
  plot_list_alluvial[[carb]] <- p
}

# Combine plots using patchwork
combined_alluvial <- wrap_plots(plot_list_alluvial, ncol = 3) +
  plot_annotation(
    title = "Genus Composition Changes from 0H to 48H by Carbohydrate Type",
    subtitle = paste0("Flow diagram showing relative abundance changes (top ", length(top_genera_alluvial), " genera including DA-important taxa)"),
    theme = theme(plot.title = element_text(size = 16, face = "bold"))
  )

print(combined_alluvial)
```

### 6.4.3 Validation: Bar Plots of Genus Abundances by Timepoint and Carbohydrate

```{r genus-validation, echo=FALSE, fig.width=14, fig.height=10}
# Extract genus-level data for validation bar plots
ps_genus_rel <- transform_sample_counts(ps_genus, function(x) x / sum(x))

otu_mat <- as.matrix(otu_table(ps_genus_rel))
if (!taxa_are_rows(ps_genus_rel)) {
  otu_mat <- t(otu_mat)
}
otu_df <- as.data.frame(t(otu_mat))

tax_df <- as.data.frame(tax_table(ps_genus_rel))
meta_df <- data.frame(sample_data(ps_genus_rel)) %>%
  tibble::rownames_to_column("SampleID")

# Combine data
genus_long <- otu_df %>%
  tibble::rownames_to_column("SampleID") %>%
  tidyr::pivot_longer(-SampleID, names_to = "ASV", values_to = "Abundance") %>%
  dplyr::left_join(
    tax_df %>% tibble::rownames_to_column("ASV") %>% dplyr::select(ASV, Genus),
    by = "ASV"
  ) %>%
  dplyr::left_join(
    meta_df %>% dplyr::select(SampleID, timepoint, carbohydrate),
    by = "SampleID"
  ) %>%
  dplyr::filter(timepoint %in% c("0H", "48H"))

# Remove NA values to prevent NaN in calculations
genus_long_clean <- genus_long %>%
  dplyr::filter(
    !is.na(Abundance),
    !is.na(Genus),
    !is.na(timepoint),
    !is.na(carbohydrate)
  )

# Calculate mean abundance per genus, timepoint, and carbohydrate
# Use ALL data first (don't filter yet) to preserve low-abundance combinations like RDC 0H
genus_summary_all <- genus_long_clean %>%
  dplyr::group_by(Genus, timepoint, carbohydrate) %>%
  dplyr::summarise(
    mean_abundance = mean(Abundance, na.rm = TRUE),
    se_abundance = sd(Abundance, na.rm = TRUE) / sqrt(dplyr::n()),
    n_samples = dplyr::n(),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    # Replace any NaN or NA with 0
    mean_abundance = ifelse(is.nan(mean_abundance) | is.na(mean_abundance), 0, mean_abundance),
    se_abundance = ifelse(is.nan(se_abundance) | is.na(se_abundance), 0, se_abundance)
  )

# Get top 15 genera overall (based on max abundance across all conditions)
top_genera_by_abundance_val <- genus_summary_all %>%
  dplyr::group_by(Genus) %>%
  dplyr::summarise(max_abundance = max(mean_abundance, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(desc(max_abundance)) %>%
  head(15) %>%
  dplyr::pull(Genus)

# Define genera that are important from DA analysis and should be included
# Check with single underscore "g_", double underscore "g__", and without prefix
# since taxonomy formats may vary (tax_table uses "g__" format)
da_important_genera <- c(
  "g_Fusicatenibacter",
  "g__Fusicatenibacter",
  "Fusicatenibacter",
  "g_Fusicatenibacters",
  "g__Fusicatenibacters",
  "Fusicatenibacters",
  "g_Hungatella",
  "g__Hungatella",
  "Hungatella",
  "g_Eggerthella",
  "g__Eggerthella",
  "Eggerthella",
  "g_Peptoniphilus",
  "g__Peptoniphilus",
  "Peptoniphilus"
)

# Check which of these genera exist in the data
# IMPORTANT: Get available_genera from the original tax_table, not from grouped data
# This ensures we find all genera in the taxonomy, even if they have zero/low abundance
available_genera_from_tax <- unique(tax_df$Genus)
available_genera_grouped <- unique(genus_summary_all$Genus)

# Check exact matches first
da_genera_in_data <- da_important_genera[da_important_genera %in% available_genera_from_tax]

# Also check for variations in the full taxonomy (e.g., "g__Fusicatenibacter sp." might be stored differently)
# This will catch any genus containing these names, regardless of prefix or suffix
da_genera_variations <- available_genera_from_tax[
  stringr::str_detect(available_genera_from_tax, regex("Fusicatenibacter|Hungatella|Eggerthella|Peptoniphilus", ignore_case = TRUE))
]

# Combine exact matches and variations
da_genera_found <- unique(c(da_genera_in_data, da_genera_variations))

# Combine top 15 by abundance with DA-important genera
# Remove duplicates and keep up to 19 genera total (15 + up to 4 DA genera)
top_genera_val <- unique(c(top_genera_by_abundance_val, da_genera_found))
if (length(top_genera_val) > 19) {
  # If we have more than 19, prioritize top 15 by abundance, then add DA genera
  top_genera_val <- c(
    top_genera_by_abundance_val,
    setdiff(da_genera_found, top_genera_by_abundance_val)
  )[1:min(19, length(top_genera_val))]
}

# Filter to top genera BUT keep ALL timepoint/carbohydrate combinations
# IMPORTANT: Don't filter by abundance threshold here - we want to see all combinations
# for the top genera, including low-abundance ones like RDC 0H
genus_summary_top <- genus_summary_all %>%
  dplyr::filter(Genus %in% top_genera_val)

# Create faceted bar plot
genus_validation_plot <- ggplot(
  genus_summary_top,
  aes(x = timepoint, y = mean_abundance, fill = timepoint)
) +
  geom_col(position = "dodge", width = 0.7) +
  geom_errorbar(
    aes(ymin = mean_abundance - se_abundance, ymax = mean_abundance + se_abundance),
    width = 0.2,
    position = position_dodge(0.7)
  ) +
  facet_grid(Genus ~ carbohydrate, scales = "free_y") +
  scale_fill_manual(values = c("0H" = "#E69F00", "48H" = "#56B4E9")) +
  labs(
    title = "Validation: Genus Abundance by Timepoint and Carbohydrate Type",
    subtitle = paste0("Mean relative abundance ± SE for top ", length(top_genera_val), " genera (including DA-important taxa)"),
    x = "Timepoint",
    y = "Mean Relative Abundance",
    fill = "Timepoint"
  ) +
  theme_scfa() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    strip.text.y = element_text(angle = 0, hjust = 0),
    panel.spacing = unit(0.5, "lines")
  )

print(genus_validation_plot)
```

## 6.5 Differential Abundance and Predictive Modeling

Differential abundance analysis identifies microbial taxa that differ significantly between experimental conditions. We applied multiple complementary methods to ensure robust identification of differentially abundant taxa: ANCOM-BC2 (Analysis of Compositions of Microbiomes with Bias Correction), which accounts for compositional bias and uses mixed-effects models; MaAsLin2 (Microbiome Multivariable Associations with Linear Models), which handles multivariate associations with normalization and transformations; and LEfSe (Linear Discriminant Analysis Effect Size), which identifies biomarker taxa between groups.

### 6.5.1 ANCOM-BC2 Analysis

ANCOM-BC2 models microbial abundance using a generalized linear mixed-effects model with bias correction for compositional data. We tested the main effects of group (Healthy vs Obese), carbohydrate type (RDC vs No Added Carb, SDC vs No Added Carb), timepoint (48H vs 0H), and their interactions.

```{r ancombc2-analysis, echo=FALSE, cache=TRUE, cache.path="cache/integrated-"}
# Prepare phyloseq object for ANCOM-BC2 (needs count data, not relative)
# Use ps (counts) and ensure we have genus-level data
ps_ancombc2 <- ps_genus

# Remove samples with too many zeros that cause estimation issues
zero_prop <- apply(otu_table(ps_ancombc2), 2, function(x) sum(x == 0) / length(x))
problematic_samples <- names(zero_prop[zero_prop > 0.95])

if (length(problematic_samples) > 0) {
  ps_ancombc2_clean <- subset_samples(ps_ancombc2, !sample_names(ps_ancombc2) %in% problematic_samples)
} else {
  ps_ancombc2_clean <- ps_ancombc2
}

# Check if we have enough samples
if (nsamples(ps_ancombc2_clean) < 10) {
  cat("WARNING: Insufficient samples for ANCOM-BC2 analysis\n")
  ancombc2_res <- NULL
} else {
  # Run ANCOM-BC2 with mixed-effects model
  ancombc2_res <- tryCatch({
    ANCOMBC::ancombc2(
      data = ps_ancombc2_clean,
      tax_level = "Genus",
      fix_formula = "group + carbohydrate * timepoint",
      rand_formula = "(1|subject_id)",
      p_adj_method = "BH",
      prv_cut = 0.10,
      lib_cut = 0,
      s0_perc = 0.05,
      group = "carbohydrate",
      struc_zero = TRUE,
      neg_lb = TRUE,
      alpha = 0.05,
      global = TRUE
    )
  }, error = function(e) {
    cat("ANCOM-BC2 failed:", e$message, "\n")
    NULL
  })
}

# Extract results if available
if (!is.null(ancombc2_res) && !is.null(ancombc2_res$res)) {
  res_data <- ancombc2_res$res
  
  # Get the effect names
  effect_names <- c("groupObese", "carbohydrateRDC", "carbohydrateSDC", "timepoint48H", 
                    "carbohydrateRDC:timepoint48H", "carbohydrateSDC:timepoint48H")
  
  # Create a data frame from the results for each effect
  ancombc2_tbl <- purrr::map_dfr(effect_names, function(effect) {
    lfc_col <- paste0("lfc_", effect)
    se_col <- paste0("se_", effect)
    W_col <- paste0("W_", effect)
    p_col <- paste0("p_", effect)
    q_col <- paste0("q_", effect)
    diff_col <- paste0("diff_", effect)
    
    if (all(c(lfc_col, se_col, W_col, p_col, q_col, diff_col) %in% colnames(res_data))) {
      tibble(
        taxon = res_data$taxon,
        effect = effect,
        beta = res_data[[lfc_col]],
        se = res_data[[se_col]],
        W = res_data[[W_col]],
        p_value = res_data[[p_col]],
        q_value = res_data[[q_col]],
        diff_abn = res_data[[diff_col]]
      )
    } else {
      tibble()
    }
  }) %>% dplyr::arrange(q_value)
} else {
  ancombc2_tbl <- tibble(
    taxon = character(0),
    effect = character(0),
    beta = numeric(0),
    se = numeric(0),
    W = numeric(0),
    p_value = numeric(0),
    q_value = numeric(0),
    diff_abn = logical(0)
  )
  cat("No ANCOM-BC2 results available - analysis may have failed\n")
}

# Show summary of significant results
if (nrow(ancombc2_tbl) > 0 && "q_value" %in% names(ancombc2_tbl)) {
  ancombc2_sig <- ancombc2_tbl %>% dplyr::filter(q_value < 0.1)
  cat("ANCOM-BC2 significant results (q < 0.1):", nrow(ancombc2_sig), "taxon-effect combinations\n")
}
```

### 6.5.2 ANCOM-BC2 Forest Plots

```{r ancombc2-forest-plot, echo=FALSE, fig.width=16, fig.height=12}
# Prepare data for forest plot
if (nrow(ancombc2_tbl) > 0 && "q_value" %in% names(ancombc2_tbl)) {
  forest_data <- ancombc2_tbl %>%
    dplyr::filter(q_value < 0.1) %>%
    dplyr::mutate(
      # Calculate confidence intervals
      ci_lower = beta - 1.96 * se,
      ci_upper = beta + 1.96 * se,
      # Create effect labels
      effect_label = dplyr::case_when(
        effect == "groupObese" ~ "Group: Obese vs Healthy",
        effect == "carbohydrateRDC" ~ "Carbohydrate: RDC vs No Added Carb",
        effect == "carbohydrateSDC" ~ "Carbohydrate: SDC vs No Added Carb", 
        effect == "timepoint48H" ~ "Timepoint: 48H vs 0H",
        effect == "carbohydrateRDC:timepoint48H" ~ "RDC × 48H Interaction",
        effect == "carbohydrateSDC:timepoint48H" ~ "SDC × 48H Interaction",
        TRUE ~ effect
      ),
      # Create taxon labels (remove g__ prefix for cleaner display)
      taxon_label = stringr::str_replace(taxon, "^g__", ""),
      # Color by significance level
      significance = dplyr::case_when(
        q_value < 0.001 ~ "q < 0.001",
        q_value < 0.01 ~ "q < 0.01", 
        q_value < 0.05 ~ "q < 0.05",
        q_value < 0.1 ~ "q < 0.1",
        TRUE ~ "q ≥ 0.1"
      )
    ) %>%
    # Order by effect size (absolute value)
    dplyr::arrange(desc(abs(beta)))
  
  if (nrow(forest_data) > 0) {
    # Create forest plot
    ancombc2_forest_plot <- ggplot(forest_data, aes(x = beta, y = reorder(taxon_label, beta))) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
      geom_point(aes(color = significance), size = 2) +
      geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper, color = significance), 
                     height = 0.2, alpha = 0.8) +
      facet_wrap(~ effect_label, scales = "free_x", ncol = 2) +
      scale_color_manual(
        values = c("q < 0.001" = "#d73027", 
                   "q < 0.01" = "#f46d43",
                   "q < 0.05" = "#fdae61", 
                   "q < 0.1" = "#fee08b"),
        name = "FDR-adjusted\np-value"
      ) +
      labs(
        title = "ANCOM-BC2 Results: Forest Plot of Significant Taxa",
        subtitle = "Effect sizes (log-fold changes) with 95% confidence intervals",
        x = "Log-fold change (β)",
        y = "Genus"
      ) +
      theme_scfa() +
      theme(
        axis.text.y = element_text(size = 8),
        axis.text.x = element_text(size = 9),
        strip.text = element_text(size = 10, face = "bold"),
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 11),
        legend.position = "bottom",
        panel.grid.minor = element_blank()
      )
    
    print(ancombc2_forest_plot)
  } else {
    cat("No significant ANCOM-BC2 results found (q < 0.1) for forest plot\n")
  }
} else {
  cat("No ANCOM-BC2 results available for forest plot - analysis may have failed\n")
}
```

### 6.5.3 ANCOM-BC2 Side-by-Side Comparison: Carbohydrate Effects

```{r ancombc2-carb-comparison, echo=FALSE, fig.width=14, fig.height=10}
# Compare RDC vs No Carb and SDC vs No Carb effects side by side
if (nrow(ancombc2_tbl) > 0 && "effect" %in% names(ancombc2_tbl)) {
  comparison_data <- ancombc2_tbl %>%
    dplyr::filter(effect %in% c("carbohydrateRDC", "carbohydrateSDC")) %>%
    dplyr::select(taxon, effect, beta, se, p_value, q_value, diff_abn) %>%
    tidyr::pivot_wider(
      names_from = effect,
      values_from = c(beta, se, p_value, q_value, diff_abn),
      names_sep = "_"
    ) %>%
    dplyr::mutate(
      # Calculate difference between RDC and SDC effects
      effect_difference = beta_carbohydrateRDC - beta_carbohydrateSDC,
      # Determine which treatment has stronger effect
      stronger_effect = dplyr::case_when(
        abs(beta_carbohydrateRDC) > abs(beta_carbohydrateSDC) ~ "RDC",
        abs(beta_carbohydrateSDC) > abs(beta_carbohydrateRDC) ~ "SDC",
        TRUE ~ "Equal"
      ),
      # Check if effects are opposite (different signs)
      opposite_effects = sign(beta_carbohydrateRDC) != sign(beta_carbohydrateSDC),
      # Check if either effect is significant
      either_significant = (q_value_carbohydrateRDC < 0.1) | (q_value_carbohydrateSDC < 0.1),
      # Clean taxon names
      taxon_clean = stringr::str_replace(taxon, "^g__", "")
    ) %>%
    dplyr::filter(either_significant | opposite_effects) %>%
    dplyr::arrange(desc(abs(effect_difference)))
  
  if (nrow(comparison_data) > 0) {
    # Prepare long format for plotting
    comparison_plot_data <- comparison_data %>%
      dplyr::select(taxon_clean, beta_carbohydrateRDC, beta_carbohydrateSDC, 
                    se_carbohydrateRDC, se_carbohydrateSDC,
                    q_value_carbohydrateRDC, q_value_carbohydrateSDC) %>%
      tidyr::pivot_longer(
        cols = c(beta_carbohydrateRDC, beta_carbohydrateSDC),
        names_to = "treatment",
        values_to = "beta",
        names_prefix = "beta_"
      ) %>%
      dplyr::mutate(
        se = dplyr::case_when(
          treatment == "carbohydrateRDC" ~ se_carbohydrateRDC,
          treatment == "carbohydrateSDC" ~ se_carbohydrateSDC,
          TRUE ~ NA_real_
        ),
        q_value = dplyr::case_when(
          treatment == "carbohydrateRDC" ~ q_value_carbohydrateRDC,
          treatment == "carbohydrateSDC" ~ q_value_carbohydrateSDC,
          TRUE ~ NA_real_
        ),
        ci_lower = beta - 1.96 * se,
        ci_upper = beta + 1.96 * se,
        treatment_label = dplyr::case_when(
          treatment == "carbohydrateRDC" ~ "RDC vs No Added Carb",
          treatment == "carbohydrateSDC" ~ "SDC vs No Added Carb",
          TRUE ~ treatment
        ),
        significance = dplyr::case_when(
          q_value < 0.001 ~ "q < 0.001",
          q_value < 0.01 ~ "q < 0.01",
          q_value < 0.05 ~ "q < 0.05",
          q_value < 0.1 ~ "q < 0.1",
          TRUE ~ "q ≥ 0.1"
        )
      )
    
    ancombc2_comparison_plot <- ggplot(
      comparison_plot_data,
      aes(x = beta, y = reorder(taxon_clean, beta))
    ) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
      geom_point(aes(color = significance), size = 2, position = position_dodge(width = 0.5)) +
      geom_errorbarh(
        aes(xmin = ci_lower, xmax = ci_upper, color = significance),
        height = 0.2,
        alpha = 0.8,
        position = position_dodge(width = 0.5)
      ) +
      facet_wrap(~ treatment_label, ncol = 2) +
      scale_color_manual(
        values = c("q < 0.001" = "#d73027", 
                   "q < 0.01" = "#f46d43",
                   "q < 0.05" = "#fdae61", 
                   "q < 0.1" = "#fee08b",
                   "q ≥ 0.1" = "gray70"),
        name = "FDR-adjusted\np-value"
      ) +
      labs(
        title = "ANCOM-BC2: Side-by-Side Comparison of Carbohydrate Effects",
        subtitle = "RDC vs No Added Carb and SDC vs No Added Carb",
        x = "Log-fold change (β)",
        y = "Genus"
      ) +
      theme_scfa() +
      theme(
        axis.text.y = element_text(size = 8),
        axis.text.x = element_text(size = 9),
        strip.text = element_text(size = 10, face = "bold"),
        legend.position = "bottom"
      )
    
    print(ancombc2_comparison_plot)
  } else {
    cat("No significant comparisons found for side-by-side plot\n")
  }
} else {
  cat("No ANCOM-BC2 results available for comparison\n")
}
```

### 6.5.4 ANCOM-BC2 48H Timepoint Analysis

```{r ancombc2-48h-analysis, echo=FALSE, cache=TRUE, cache.path="cache/integrated-"}
# Subset to 48H timepoint only
ps_ancombc2_48h <- phyloseq::subset_samples(ps_ancombc2_clean, timepoint == "48H")

# Check if we have enough samples
if (nsamples(ps_ancombc2_48h) < 10) {
  cat("WARNING: Insufficient samples for ANCOM-BC2 48H analysis\n")
  ancombc2_48h_res <- NULL
} else {
  # Run ANCOM-BC2 for 48H timepoint
  ancombc2_48h_res <- tryCatch({
    ANCOMBC::ancombc2(
      data = ps_ancombc2_48h,
      tax_level = "Genus",
      fix_formula = "group + carbohydrate",
      rand_formula = "(1|subject_id)",
      p_adj_method = "BH",
      prv_cut = 0.10,
      lib_cut = 0,
      s0_perc = 0.05,
      group = "carbohydrate",
      struc_zero = TRUE,
      neg_lb = TRUE,
      alpha = 0.05,
      global = TRUE
    )
  }, error = function(e) {
    cat("ANCOM-BC2 48H failed:", e$message, "\n")
    NULL
  })
}

# Extract results if available
if (!is.null(ancombc2_48h_res) && !is.null(ancombc2_48h_res$res)) {
  res_48h_data <- ancombc2_48h_res$res
  
  # Get the effect names for 48H
  effect_names_48h <- c("groupObese", "carbohydrateRDC", "carbohydrateSDC")
  
  # Create a data frame from the results
  ancombc2_48h_tbl <- purrr::map_dfr(effect_names_48h, function(effect) {
    lfc_col <- paste0("lfc_", effect)
    se_col <- paste0("se_", effect)
    W_col <- paste0("W_", effect)
    p_col <- paste0("p_", effect)
    q_col <- paste0("q_", effect)
    diff_col <- paste0("diff_", effect)
    
    if (all(c(lfc_col, se_col, W_col, p_col, q_col, diff_col) %in% colnames(res_48h_data))) {
      tibble(
        taxon = res_48h_data$taxon,
        effect = effect,
        beta = res_48h_data[[lfc_col]],
        se = res_48h_data[[se_col]],
        W = res_48h_data[[W_col]],
        p_value = res_48h_data[[p_col]],
        q_value = res_48h_data[[q_col]],
        diff_abn = res_48h_data[[diff_col]]
      )
    } else {
      tibble()
    }
  }) %>% dplyr::arrange(q_value)
} else {
  ancombc2_48h_tbl <- tibble(
    taxon = character(0),
    effect = character(0),
    beta = numeric(0),
    se = numeric(0),
    W = numeric(0),
    p_value = numeric(0),
    q_value = numeric(0),
    diff_abn = logical(0)
  )
}

if (nrow(ancombc2_48h_tbl) > 0 && "q_value" %in% names(ancombc2_48h_tbl)) {
  ancombc2_48h_sig <- ancombc2_48h_tbl %>% dplyr::filter(q_value < 0.1)
  cat("ANCOM-BC2 48H significant results (q < 0.1):", nrow(ancombc2_48h_sig), "taxon-effect combinations\n")
}
```

### 6.5.5 ANCOM-BC2 48H Forest Plots

```{r ancombc2-48h-forest-plot, echo=FALSE, fig.width=14, fig.height=10}
# Prepare data for 48H forest plot
if (nrow(ancombc2_48h_tbl) > 0 && "q_value" %in% names(ancombc2_48h_tbl)) {
  forest_48h_data <- ancombc2_48h_tbl %>%
    dplyr::filter(q_value < 0.1) %>%
    dplyr::mutate(
      ci_lower = beta - 1.96 * se,
      ci_upper = beta + 1.96 * se,
      effect_label = dplyr::case_when(
        effect == "groupObese" ~ "Group: Obese vs Healthy",
        effect == "carbohydrateRDC" ~ "RDC vs No Added Carb",
        effect == "carbohydrateSDC" ~ "SDC vs No Added Carb",
        TRUE ~ effect
      ),
      taxon_label = stringr::str_replace(taxon, "^g__", ""),
      significance = dplyr::case_when(
        q_value < 0.001 ~ "q < 0.001",
        q_value < 0.01 ~ "q < 0.01",
        q_value < 0.05 ~ "q < 0.05",
        q_value < 0.1 ~ "q < 0.1",
        TRUE ~ "q ≥ 0.1"
      )
    ) %>%
    dplyr::arrange(desc(abs(beta)))
  
  if (nrow(forest_48h_data) > 0) {
    ancombc2_48h_forest_plot <- ggplot(
      forest_48h_data,
      aes(x = beta, y = reorder(taxon_label, beta))
    ) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
      geom_point(aes(color = significance), size = 2) +
      geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper, color = significance), 
                     height = 0.2, alpha = 0.8) +
      facet_wrap(~ effect_label, scales = "free_x", ncol = 2) +
      scale_color_manual(
        values = c("q < 0.001" = "#d73027", 
                   "q < 0.01" = "#f46d43",
                   "q < 0.05" = "#fdae61", 
                   "q < 0.1" = "#fee08b"),
        name = "FDR-adjusted\np-value"
      ) +
      labs(
        title = "ANCOM-BC2 Results at 48H: Forest Plot of Significant Taxa",
        subtitle = "Effect sizes (log-fold changes) with 95% confidence intervals",
        x = "Log-fold change (β)",
        y = "Genus"
      ) +
      theme_scfa() +
      theme(
        axis.text.y = element_text(size = 8),
        axis.text.x = element_text(size = 9),
        strip.text = element_text(size = 10, face = "bold"),
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 11),
        legend.position = "bottom"
      )
    
    print(ancombc2_48h_forest_plot)
  } else {
    cat("No significant ANCOM-BC2 48H results found (q < 0.1) for forest plot\n")
  }
} else {
  cat("No ANCOM-BC2 48H results available for forest plot\n")
}
```

### 6.5.6 ANCOM-BC2 RDC vs SDC at 48H

To directly compare the effects of RDC versus SDC at the 48H timepoint, we performed a focused ANCOM-BC2 analysis including only RDC and SDC samples at 48H, excluding the No Added Carb condition.

```{r ancombc2-48h-rdc-sdc-analysis, echo=FALSE, cache=TRUE, cache.path="cache/integrated-"}
# Subset to 48H timepoint and only RDC vs SDC (exclude No Added Carb)
ps_ancombc2_48h_rdc_sdc <- phyloseq::subset_samples(
  ps_ancombc2_clean,
  timepoint == "48H" & carbohydrate %in% c("RDC", "SDC")
)

# Check sample counts
cat("48H RDC vs SDC samples for ANCOM-BC2:", nsamples(ps_ancombc2_48h_rdc_sdc), "\n")
cat("Sample distribution by carbohydrate:\n")
print(table(sample_data(ps_ancombc2_48h_rdc_sdc)$carbohydrate))

# Check if we have enough samples
if (nsamples(ps_ancombc2_48h_rdc_sdc) < 10) {
  cat("WARNING: Insufficient samples for ANCOM-BC2 48H RDC vs SDC analysis\n")
  ancombc2_48h_rdc_sdc_res <- NULL
} else {
  # Set RDC as reference level
  sample_data(ps_ancombc2_48h_rdc_sdc)$carbohydrate <- relevel(
    as.factor(sample_data(ps_ancombc2_48h_rdc_sdc)$carbohydrate),
    ref = "RDC"
  )
  
  # Run ANCOM-BC2 for 48H RDC vs SDC
  ancombc2_48h_rdc_sdc_res <- tryCatch({
    ANCOMBC::ancombc2(
      data = ps_ancombc2_48h_rdc_sdc,
      tax_level = "Genus",
      fix_formula = "group + carbohydrate",
      rand_formula = "(1|subject_id)",
      p_adj_method = "BH",
      prv_cut = 0.10,
      lib_cut = 0,
      s0_perc = 0.05,
      group = "carbohydrate",
      struc_zero = TRUE,
      neg_lb = TRUE,
      alpha = 0.05,
      global = TRUE
    )
  }, error = function(e) {
    cat("ANCOM-BC2 48H RDC vs SDC failed:", e$message, "\n")
    NULL
  })
}

# Extract results if available
if (!is.null(ancombc2_48h_rdc_sdc_res) && !is.null(ancombc2_48h_rdc_sdc_res$res)) {
  res_48h_rdc_sdc_data <- ancombc2_48h_rdc_sdc_res$res
  
  # Get the effect names for 48H RDC vs SDC
  # With RDC as reference, we'll get carbohydrateSDC effect
  effect_names_48h_rdc_sdc <- c("groupObese", "carbohydrateSDC")
  
  # Create a data frame from the results
  ancombc2_48h_rdc_sdc_tbl <- purrr::map_dfr(effect_names_48h_rdc_sdc, function(effect) {
    lfc_col <- paste0("lfc_", effect)
    se_col <- paste0("se_", effect)
    W_col <- paste0("W_", effect)
    p_col <- paste0("p_", effect)
    q_col <- paste0("q_", effect)
    diff_col <- paste0("diff_", effect)
    
    if (all(c(lfc_col, se_col, W_col, p_col, q_col, diff_col) %in% colnames(res_48h_rdc_sdc_data))) {
      tibble(
        taxon = res_48h_rdc_sdc_data$taxon,
        effect = effect,
        beta = res_48h_rdc_sdc_data[[lfc_col]],
        se = res_48h_rdc_sdc_data[[se_col]],
        W = res_48h_rdc_sdc_data[[W_col]],
        p_value = res_48h_rdc_sdc_data[[p_col]],
        q_value = res_48h_rdc_sdc_data[[q_col]],
        diff_abn = res_48h_rdc_sdc_data[[diff_col]]
      )
    } else {
      tibble()
    }
  }) %>% dplyr::arrange(q_value)
  
  # Summary of significant results
  if (nrow(ancombc2_48h_rdc_sdc_tbl) > 0 && "q_value" %in% names(ancombc2_48h_rdc_sdc_tbl)) {
    ancombc2_48h_rdc_sdc_sig <- ancombc2_48h_rdc_sdc_tbl %>%
      dplyr::filter(effect == "carbohydrateSDC", q_value < 0.1)
    cat("ANCOM-BC2 48H RDC vs SDC significant results (q < 0.1):", 
        nrow(ancombc2_48h_rdc_sdc_sig), "taxa\n")
  }
} else {
  ancombc2_48h_rdc_sdc_tbl <- tibble(
    taxon = character(0),
    effect = character(0),
    beta = numeric(0),
    se = numeric(0),
    W = numeric(0),
    p_value = numeric(0),
    q_value = numeric(0),
    diff_abn = logical(0)
  )
}
```

### 6.5.7 Forest Plot: RDC vs SDC at 48H

```{r ancombc2-48h-rdc-sdc-forest-plot, echo=FALSE, fig.width=12, fig.height=10}
# Prepare data for 48H RDC vs SDC forest plot
if (nrow(ancombc2_48h_rdc_sdc_tbl) > 0 && "q_value" %in% names(ancombc2_48h_rdc_sdc_tbl)) {
  # Focus on carbohydrateSDC effect (SDC vs RDC, with RDC as reference)
  forest_48h_rdc_sdc_data <- ancombc2_48h_rdc_sdc_tbl %>%
    dplyr::filter(effect == "carbohydrateSDC") %>%
    dplyr::mutate(
      ci_lower = beta - 1.96 * se,
      ci_upper = beta + 1.96 * se,
      taxon_label = stringr::str_replace(taxon, "^g__", ""),
      significance = dplyr::case_when(
        q_value < 0.001 ~ "q < 0.001",
        q_value < 0.01 ~ "q < 0.01",
        q_value < 0.05 ~ "q < 0.05",
        q_value < 0.1 ~ "q < 0.1",
        TRUE ~ "q ≥ 0.1"
      ),
      # Positive beta means higher in SDC, negative means higher in RDC
      direction = dplyr::if_else(beta > 0, "Higher in SDC", "Higher in RDC")
    ) %>%
    # Order by absolute effect size, but show significant ones first
    dplyr::arrange(q_value, desc(abs(beta)))
  
  # Select top taxa for display (either significant or top by effect size)
  if (sum(forest_48h_rdc_sdc_data$q_value < 0.1, na.rm = TRUE) > 0) {
    forest_48h_rdc_sdc_plot_data <- forest_48h_rdc_sdc_data %>%
      dplyr::filter(q_value < 0.1) %>%
      dplyr::arrange(desc(abs(beta))) %>%
      dplyr::slice_head(n = 25)
  } else {
    # If no significant, show top 25 by effect size
    forest_48h_rdc_sdc_plot_data <- forest_48h_rdc_sdc_data %>%
      dplyr::arrange(desc(abs(beta))) %>%
      dplyr::slice_head(n = 25)
  }
  
  if (nrow(forest_48h_rdc_sdc_plot_data) > 0) {
    # Reorder for plotting (by beta value)
    forest_48h_rdc_sdc_plot_data <- forest_48h_rdc_sdc_plot_data %>%
      dplyr::mutate(
        taxon_label = factor(taxon_label, levels = rev(unique(taxon_label[order(beta)])))
      )
    
    ancombc2_48h_rdc_sdc_forest_plot <- ggplot(
      forest_48h_rdc_sdc_plot_data,
      aes(x = beta, y = taxon_label, color = significance)
    ) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
      geom_point(size = 2.5, aes(shape = direction)) +
      geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper), 
                     height = 0.2, alpha = 0.8, linewidth = 1) +
      scale_color_manual(
        values = c("q < 0.001" = "#d73027", 
                   "q < 0.01" = "#f46d43",
                   "q < 0.05" = "#fdae61", 
                   "q < 0.1" = "#fee08b",
                   "q ≥ 0.1" = "gray70"),
        name = "FDR-adjusted\np-value",
        guide = guide_legend(order = 1)
      ) +
      scale_shape_manual(
        values = c("Higher in SDC" = 17, "Higher in RDC" = 19),
        name = "Direction",
        guide = guide_legend(order = 2)
      ) +
      labs(
        title = "ANCOM-BC2: RDC vs SDC at 48H",
        subtitle = "Effect sizes (log-fold change: SDC vs RDC) with 95% confidence intervals",
        x = "Log-fold change (SDC vs RDC)",
        y = "Genus",
        caption = "Positive values indicate higher abundance in SDC; negative values indicate higher abundance in RDC"
      ) +
      theme_scfa() +
      theme(
        axis.text.y = element_text(size = 9),
        axis.text.x = element_text(size = 10),
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 11),
        legend.position = "bottom",
        legend.box = "horizontal",
        panel.grid.minor = element_blank()
      )
    
    print(ancombc2_48h_rdc_sdc_forest_plot)
    
    # Summary statistics
    cat("\nSummary of ANCOM-BC2 RDC vs SDC at 48H:\n")
    cat("Total taxa tested:", nrow(forest_48h_rdc_sdc_data), "\n")
    cat("Significant taxa (q < 0.1):", sum(forest_48h_rdc_sdc_data$q_value < 0.1, na.rm = TRUE), "\n")
    cat("Higher in SDC (q < 0.1):", sum(forest_48h_rdc_sdc_data$beta > 0 & forest_48h_rdc_sdc_data$q_value < 0.1, na.rm = TRUE), "\n")
    cat("Higher in RDC (q < 0.1):", sum(forest_48h_rdc_sdc_data$beta < 0 & forest_48h_rdc_sdc_data$q_value < 0.1, na.rm = TRUE), "\n")
  } else {
    cat("No data available for 48H RDC vs SDC forest plot\n")
  }
} else {
  cat("No ANCOM-BC2 48H RDC vs SDC results available for forest plot\n")
}
```



### 6.5.8 MaAsLin2 Analysis

MaAsLin2 (Microbiome Multivariable Associations with Linear Models) uses generalized linear models to identify associations between microbial taxa and metadata variables while accounting for confounding factors. We applied MaAsLin2 to identify taxa associated with group, carbohydrate type, and timepoint.

```{r maaslin2-analysis, echo=FALSE, cache=TRUE, cache.path="cache/integrated-"}
# Prepare data for MaAsLin2 (needs count data)
ps_maaslin2 <- ps_genus

# Prepare count matrix and metadata
# Use proper conversion methods for phyloseq objects
maaslin2_counts <- as.data.frame(otu_table(ps_maaslin2))
maaslin2_meta <- as(sample_data(ps_maaslin2), "data.frame")
maaslin2_meta$SampleID <- rownames(maaslin2_meta)

# Ensure numeric columns are numeric
maaslin2_meta <- maaslin2_meta %>%
  dplyr::mutate(
    subject_id = as.character(subject_id),
    group = as.factor(group),
    carbohydrate = as.factor(carbohydrate),
    timepoint = as.factor(timepoint)
  )

# Run MaAsLin2
maaslin2_fit <- tryCatch({
  Maaslin2::Maaslin2(
    input_data = maaslin2_counts,
    input_metadata = maaslin2_meta,
    output = tempfile(pattern = "maaslin2_", tmpdir = getwd()),  # Create temporary output directory
    min_prevalence = 0.05,
    normalization = "TSS",
    transform = "LOG",
    analysis_method = "LM",
    fixed_effects = c("group", "carbohydrate", "timepoint"),
    random_effects = c("subject_id"),
    plot_heatmap = FALSE,
    plot_scatter = FALSE
  )
}, error = function(e) {
  cat("MaAsLin2 failed:", e$message, "\n")
  NULL
})

# Extract results
if (!is.null(maaslin2_fit) && !is.null(maaslin2_fit$results)) {
  maaslin2_tbl <- maaslin2_fit$results %>%
    dplyr::as_tibble() %>%
    dplyr::mutate(
      q = p.adjust(pval, method = "BH")  # Add BH-adjusted q-value
    ) %>%
    dplyr::arrange(q)
  
  # Add taxonomy information
  tax_table_df <- as.data.frame(tax_table(ps_maaslin2)) %>%
    tibble::rownames_to_column("feature") %>%
    dplyr::mutate(
      Genus_clean = stringr::str_replace(Genus, "^g__", ""),
      Species_clean = stringr::str_replace(Species, "^s__", ""),
      Family_clean = stringr::str_replace(Family, "^f__", ""),
      # Only use Species if Genus is valid; if Genus is NA, use Family
      genus_species = dplyr::case_when(
        !is.na(Genus_clean) & Genus_clean != "NA" & Genus_clean != "" &
          !is.na(Species_clean) & Species_clean != "NA" & Species_clean != "" ~ 
          paste0(Genus_clean, " ", Species_clean),
        !is.na(Genus_clean) & Genus_clean != "NA" & Genus_clean != "" ~ Genus_clean,
        !is.na(Family_clean) & Family_clean != "NA" & Family_clean != "" ~ Family_clean,
        TRUE ~ feature
      )
    )
  
  maaslin2_tbl <- maaslin2_tbl %>%
    dplyr::left_join(
      tax_table_df %>% dplyr::select(feature, Genus, genus_species),
      by = "feature"
    )
  
  # Summary of significant results
  maaslin2_sig <- maaslin2_tbl %>% dplyr::filter(q < 0.1)
  cat("MaAsLin2 significant results (q < 0.1):", nrow(maaslin2_sig), "feature-metadata combinations\n")
} else {
  maaslin2_tbl <- tibble(
    feature = character(0),
    metadata = character(0),
    value = character(0),
    coef = numeric(0),
    stderr = numeric(0),
    pval = numeric(0),
    qval = numeric(0),
    q = numeric(0),
    Genus = character(0),
    genus_species = character(0)
  )
  cat("No MaAsLin2 results available\n")
}
```

### 6.5.9 MaAsLin2 Summary Table

```{r maaslin2-summary-table, echo=FALSE}
# Create summary table with taxon, genus_species, method, p-value, q-value
# Fill NA values from taxonomy table and ensure taxon is always seq##
if (nrow(maaslin2_tbl) > 0) {
  # Get full taxonomy table for lookups - use original ps object before genus aggregation
  # to get ASV-level taxonomy
  tax_table_full <- as.data.frame(tax_table(ps)) %>%
    tibble::rownames_to_column("feature") %>%
    dplyr::mutate(
      # Clean taxonomy levels - remove prefixes and check for valid values
      Genus_clean = stringr::str_replace(Genus, "^g__", ""),
      Species_clean = stringr::str_replace(Species, "^s__", ""),
      Family_clean = stringr::str_replace(Family, "^f__", ""),
      Order_clean = stringr::str_replace(Order, "^o__", ""),
      Class_clean = stringr::str_replace(Class, "^c__", ""),
      Phylum_clean = stringr::str_replace(Phylum, "^p__", ""),
      Kingdom_clean = stringr::str_replace(Kingdom, "^k__", ""),
      # Get first non-NA taxonomy level (skip "NA" strings)
      # IMPORTANT: Only use Species if Genus is valid; if Genus is NA, use Family
      first_non_na_tax = dplyr::case_when(
        # Use Genus + Species only if both are valid
        !is.na(Genus_clean) & Genus_clean != "NA" & Genus_clean != "" &
          !is.na(Species_clean) & Species_clean != "NA" & Species_clean != "" ~ 
          paste0(Genus_clean, " ", Species_clean),
        # Use Genus alone if valid
        !is.na(Genus_clean) & Genus_clean != "NA" & Genus_clean != "" ~ Genus_clean,
        # If Genus is NA, use Family
        !is.na(Family_clean) & Family_clean != "NA" & Family_clean != "" ~ Family_clean,
        # Continue down the hierarchy
        !is.na(Order_clean) & Order_clean != "NA" & Order_clean != "" ~ Order_clean,
        !is.na(Class_clean) & Class_clean != "NA" & Class_clean != "" ~ Class_clean,
        !is.na(Phylum_clean) & Phylum_clean != "NA" & Phylum_clean != "" ~ Phylum_clean,
        !is.na(Kingdom_clean) & Kingdom_clean != "NA" & Kingdom_clean != "" ~ Kingdom_clean,
        TRUE ~ feature
      )
    )
  
  maaslin2_summary_tbl <- maaslin2_tbl %>%
    dplyr::filter(q < 0.1) %>%
    dplyr::left_join(
      tax_table_full %>% dplyr::select(feature, first_non_na_tax),
      by = "feature"
    ) %>%
    dplyr::mutate(
      # Fill NA genus_species with first non-NA taxonomy level (always use taxonomy, never feature)
      `Genus Species` = dplyr::case_when(
        !is.na(genus_species) & genus_species != "NA" & genus_species != "" & !stringr::str_starts(genus_species, "seq") ~ 
          stringr::str_replace_all(genus_species, "^(g__|f__|o__|c__|p__|k__)", ""),
        !is.na(first_non_na_tax) & first_non_na_tax != "NA" & first_non_na_tax != "" & !stringr::str_starts(first_non_na_tax, "seq") ~
          stringr::str_replace_all(first_non_na_tax, "^(g__|f__|o__|c__|p__|k__)", ""),
        TRUE ~ stringr::str_replace_all(first_non_na_tax, "^(g__|f__|o__|c__|p__|k__)", "")
      ),
      `DA Method` = "MaAsLin2",
      `p value` = signif(pval, 4),
      `q value` = signif(q, 4),
      `Effect Size (coef)` = signif(coef, 4),
      Variable = paste0(metadata, "=", value)
    ) %>%
    dplyr::select(taxon = feature, `Genus Species`, 
                  `DA Method`, Variable, `Effect Size (coef)`, `p value`, `q value`)
  
  # Combine with ANCOM-BC2 results if available
  if (nrow(ancombc2_tbl) > 0 && "q_value" %in% names(ancombc2_tbl)) {
    # Map ANCOM-BC2 genus-level taxa back to ASV IDs (seq##)
    # Get ASV to genus mapping from original ps object
    asv_to_genus_map <- as.data.frame(tax_table(ps)) %>%
      tibble::rownames_to_column("asv_id") %>%
      dplyr::mutate(
        genus_match = stringr::str_replace(Genus, "^g__", "")
      ) %>%
      dplyr::select(asv_id, Genus, genus_match)
    
    # Create a mapping: for each unique genus in ANCOM-BC2, find first ASV with that genus
    genus_to_asv_mapping <- asv_to_genus_map %>%
      dplyr::group_by(genus_match) %>%
      dplyr::slice_head(n = 1) %>%
      dplyr::ungroup() %>%
      dplyr::select(genus_match, asv_id)
    
    # Get full taxonomy for all ASVs (for lookup later)
    # Use the same tax_table_full that already has first_non_na_tax calculated correctly
    ancombc2_tax_lookup_full <- tax_table_full %>%
      dplyr::mutate(
        genus_match = stringr::str_replace(Genus, "^g__", "")
      )
    
    ancombc2_summary_tbl <- ancombc2_tbl %>%
      dplyr::filter(q_value < 0.1) %>%
      dplyr::mutate(
        genus_clean = stringr::str_replace(taxon, "^g__", "")
      ) %>%
      dplyr::left_join(
        genus_to_asv_mapping,
        by = c("genus_clean" = "genus_match")
      ) %>%
      dplyr::left_join(
        ancombc2_tax_lookup_full %>% dplyr::select(feature, first_non_na_tax),
        by = c("asv_id" = "feature")
      ) %>%
      dplyr::mutate(
        # Use mapped ASV ID as taxon (seq##), use first non-NA taxonomy for Genus Species
        taxon = dplyr::case_when(
          !is.na(asv_id) ~ asv_id,
          TRUE ~ taxon  # Fallback to original taxon if no match
        ),
        # Ensure Genus Species uses taxonomy, not seq## or NA
        `Genus Species` = dplyr::case_when(
          !is.na(first_non_na_tax) & first_non_na_tax != "NA" & first_non_na_tax != "" & 
            !stringr::str_starts(first_non_na_tax, "seq") ~
            stringr::str_replace_all(first_non_na_tax, "^(g__|f__|o__|c__|p__|k__)", ""),
          !is.na(genus_clean) & genus_clean != "NA" & genus_clean != "" ~ 
            stringr::str_replace_all(genus_clean, "^(g__|f__|o__|c__|p__|k__)", ""),
          TRUE ~ taxon
        ),
        `DA Method` = "ANCOM-BC2",
        `p value` = signif(p_value, 4),
        `q value` = signif(q_value, 4),
        `Effect Size (coef)` = signif(beta, 4),
        Variable = effect
      ) %>%
      dplyr::select(taxon, `Genus Species`, `DA Method`, Variable, 
                    `Effect Size (coef)`, `p value`, `q value`)
    
    # Combine both methods
    da_summary_combined <- dplyr::bind_rows(maaslin2_summary_tbl, ancombc2_summary_tbl)
  } else {
    da_summary_combined <- maaslin2_summary_tbl
  }
  
  # Final cleanup: ensure no NA or seq## in Genus Species, remove all prefixes
  # Create lookup table for any remaining missing values
  tax_lookup_final <- tax_table_full %>%
    dplyr::select(feature, first_non_na_tax_final = first_non_na_tax)
  
  da_summary_combined <- da_summary_combined %>%
    dplyr::left_join(
      tax_lookup_final,
      by = c("taxon" = "feature")
    ) %>%
    dplyr::mutate(
      # Remove all prefixes from Genus Species
      `Genus Species` = stringr::str_replace_all(`Genus Species`, "^(g__|f__|o__|c__|p__|k__)", ""),
      # If Genus Species is still NA, seq##, or empty, use lookup
      `Genus Species` = dplyr::case_when(
        is.na(`Genus Species`) | `Genus Species` == "NA" | `Genus Species` == "" | stringr::str_starts(`Genus Species`, "seq") ~
          stringr::str_replace_all(first_non_na_tax_final, "^(g__|f__|o__|c__|p__|k__)", ""),
        TRUE ~ `Genus Species`
      ),
      # Final check: ensure no seq## or NA remains
      `Genus Species` = dplyr::case_when(
        stringr::str_starts(`Genus Species`, "seq") | `Genus Species` == "NA" | is.na(`Genus Species`) ~
          stringr::str_replace_all(first_non_na_tax_final, "^(g__|f__|o__|c__|p__|k__)", ""),
        TRUE ~ `Genus Species`
      ),
      # Remove any remaining prefixes one more time
      `Genus Species` = stringr::str_replace_all(`Genus Species`, "^(g__|f__|o__|c__|p__|k__)", "")
    ) %>%
    dplyr::select(-first_non_na_tax_final) %>%
    # Sort alphabetically by Genus Species
    dplyr::arrange(`Genus Species`, `q value`)
  
  # Display as interactive table in HTML or kable otherwise
  if (knitr::is_html_output()) {
    DT::datatable(
      da_summary_combined,
      caption = "Summary of Differential Abundance Results (q < 0.1)",
      filter = "top",
      options = list(pageLength = 25, scrollX = TRUE, order = list(list(1, 'asc')))
    )
  } else {
    knitr::kable(da_summary_combined, caption = "Summary of Differential Abundance Results (q < 0.1)")
  }
} else {
  cat("No differential abundance results available for summary table\n")
}
```

### 6.5.10 MaAsLin2 Visualizations

```{r maaslin2-visualizations, echo=FALSE, fig.width=14, fig.height=16}
if (nrow(maaslin2_tbl) > 0 && nrow(maaslin2_tbl %>% dplyr::filter(q < 0.1)) > 0) {
  # Get taxonomy table for Family lookups (for NA taxa)
  tax_table_for_vis <- as.data.frame(tax_table(ps)) %>%
    tibble::rownames_to_column("feature") %>%
    dplyr::mutate(
      Family_clean = stringr::str_replace(Family, "^f__", ""),
      Order_clean = stringr::str_replace(Order, "^o__", ""),
      Class_clean = stringr::str_replace(Class, "^c__", ""),
      Phylum_clean = stringr::str_replace(Phylum, "^p__", ""),
      # Get first non-NA taxonomy level for NA taxa
      first_non_na_tax = dplyr::case_when(
        !is.na(Family_clean) & Family_clean != "NA" & Family_clean != "" ~ Family_clean,
        !is.na(Order_clean) & Order_clean != "NA" & Order_clean != "" ~ Order_clean,
        !is.na(Class_clean) & Class_clean != "NA" & Class_clean != "" ~ Class_clean,
        !is.na(Phylum_clean) & Phylum_clean != "NA" & Phylum_clean != "" ~ Phylum_clean,
        TRUE ~ feature
      )
    ) %>%
    dplyr::select(feature, first_non_na_tax)
  
  maaslin2_sig <- maaslin2_tbl %>%
    dplyr::filter(q < 0.1) %>%
    dplyr::left_join(tax_table_for_vis, by = "feature") %>%
    dplyr::mutate(
      # Replace NA genus_species with Family level (or other taxonomy)
      genus_species = dplyr::case_when(
        is.na(genus_species) | genus_species == "NA" | genus_species == "" | 
          stringr::str_starts(genus_species, "seq") ~
          stringr::str_replace_all(first_non_na_tax, "^(g__|f__|o__|c__|p__|k__)", ""),
        TRUE ~ stringr::str_replace_all(genus_species, "^(g__|f__|o__|c__|p__|k__)", "")
      ),
      # Create comparison labels for subtitle
      comparison_label = dplyr::case_when(
        metadata == "carbohydrate" & value == "RDC" ~ "RDC vs No Added Carb (positive = more in RDC)",
        metadata == "carbohydrate" & value == "SDC" ~ "SDC vs No Added Carb (positive = more in SDC)",
        metadata == "group" & value == "Obese" ~ "Obese vs Healthy (positive = more in Obese)",
        metadata == "timepoint" & value == "48H" ~ "48H vs 0H (positive = more at 48H)",
        metadata == "timepoint" & value == "72H" ~ "72H vs 0H (positive = more at 72H)",
        TRUE ~ paste0(value, " vs Reference")
      ),
      significance = dplyr::case_when(
        q < 0.001 ~ "***",
        q < 0.01 ~ "**",
        q < 0.05 ~ "*",
        TRUE ~ ""
      ),
      display_label = paste0(genus_species, " (", metadata, "=", value, ")", significance),
      full_label = display_label,
      metadata_value = paste0(metadata, "=", value)
    ) %>%
    dplyr::select(-first_non_na_tax)
  
  # Plot 1: Coefficient effect sizes by metadata variable
  maaslin2_effects_plot <- ggplot(
    maaslin2_sig,
    aes(x = coef, y = reorder(full_label, coef))
  ) +
    geom_col(aes(fill = metadata_value), alpha = 0.8) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    scale_fill_scfa() +
    labs(
      title = "MaAsLin2 Results: Significant Associations",
      subtitle = "Coefficient Effect Sizes by Variable",
      x = "Coefficient (Effect Size)",
      y = "Features",
      caption = "*** q<0.001, ** q<0.01, * q<0.05"
    ) +
    theme_scfa() +
    theme(
      axis.text.y = element_text(size = 8),
      axis.text.x = element_text(size = 10),
      legend.position = "bottom",
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank()
    )
  
  print(maaslin2_effects_plot)
  
  # Plot 2: Volcano plot with ggrepel labels
  sig_points <- maaslin2_tbl %>%
    dplyr::filter(q < 0.1) %>%
    dplyr::left_join(tax_table_for_vis, by = "feature") %>%
    dplyr::mutate(
      # Replace NA genus_species with Family level (or other taxonomy)
      genus_species_clean = dplyr::case_when(
        is.na(genus_species) | genus_species == "NA" | genus_species == "" | 
          stringr::str_starts(genus_species, "seq") ~
          stringr::str_replace_all(first_non_na_tax, "^(g__|f__|o__|c__|p__|k__)", ""),
        TRUE ~ stringr::str_replace_all(genus_species, "^(g__|f__|o__|c__|p__|k__)", "")
      ),
      genus_label = dplyr::case_when(
        !is.na(genus_species_clean) & stringr::str_detect(genus_species_clean, " ") ~ 
          stringr::str_extract(genus_species_clean, "^[^ ]+"),
        !is.na(genus_species_clean) ~ genus_species_clean,
        TRUE ~ stringr::str_replace_all(first_non_na_tax, "^(g__|f__|o__|c__|p__|k__)", "")
      )
    ) %>%
    dplyr::select(-first_non_na_tax)
  
  maaslin2_volcano_plot <- ggplot(
    maaslin2_tbl,
    aes(x = coef, y = -log10(q))
  ) +
    geom_point(aes(color = q < 0.1), alpha = 0.6, size = 2) +
    geom_hline(yintercept = -log10(0.1), linetype = "dashed", color = "red", alpha = 0.7) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    ggrepel::geom_text_repel(
      data = sig_points,
      aes(label = genus_label),
      size = 3,
      color = "#E31A1C",
      fontface = "bold",
      max.overlaps = 20,
      min.segment.length = 0.1,
      box.padding = 0.3,
      point.padding = 0.2,
      segment.color = "gray50",
      segment.alpha = 0.6
    ) +
    scale_color_manual(
      values = c("FALSE" = "gray70", "TRUE" = "#E31A1C"),
      name = "Significant",
      labels = c("FALSE" = "q ≥ 0.1", "TRUE" = "q < 0.1")
    ) +
    labs(
      title = "MaAsLin2 Volcano Plot",
      subtitle = "Effect Size vs Statistical Significance",
      x = "Coefficient (Effect Size)",
      y = "-log10(q-value)",
      caption = "Red dashed line: q = 0.1 threshold; Labels show significant taxa"
    ) +
    theme_scfa() +
    theme(
      legend.position = "bottom"
    )
  
  print(maaslin2_volcano_plot)
  
  # Plot 3: Top effects by metadata variable
  top_effects <- maaslin2_sig %>%
    dplyr::group_by(metadata) %>%
    dplyr::slice_max(abs(coef), n = 10) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      # Create dynamic subtitle per metadata variable
      subtitle_text = dplyr::case_when(
        metadata == "carbohydrate" ~ "RDC vs No Added Carb (pos=more in RDC); SDC vs No Added Carb (pos=more in SDC)",
        metadata == "group" ~ "Obese vs Healthy (positive = more in Obese)",
        metadata == "timepoint" ~ "48H vs 0H (pos=more at 48H); 72H vs 0H (pos=more at 72H)",
        TRUE ~ paste0("Comparison: ", metadata)
      )
    )
  
  # Create subtitle that combines all comparisons
  all_comparisons <- top_effects %>%
    dplyr::distinct(metadata, comparison_label) %>%
    dplyr::group_by(metadata) %>%
    dplyr::summarize(comparisons = paste(comparison_label, collapse = "; ")) %>%
    dplyr::summarize(all_text = paste(metadata, ":", comparisons, collapse = " | "))
  
  subtitle_text <- if (nrow(all_comparisons) > 0) {
    paste0("Top 10 largest effects per metadata variable. Comparisons: ", all_comparisons$all_text[1])
  } else {
    "Top 10 largest effects per metadata variable"
  }
  
  maaslin2_top_effects_plot <- ggplot(
    top_effects,
    aes(x = coef, y = reorder(genus_species, coef))
  ) +
    geom_col(aes(fill = metadata), alpha = 0.8) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    scale_fill_scfa() +
    labs(
      title = "MaAsLin2: Top Effects by Variable",
      subtitle = subtitle_text,
      x = "Coefficient (Effect Size)",
      y = "Genus Species"
    ) +
    theme_scfa() +
    theme(
      axis.text.y = element_text(size = 9),
      axis.text.x = element_text(size = 10),
      legend.position = "bottom",
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank()
    ) +
    facet_wrap(~metadata, scales = "free_y", ncol = 1)
  
  print(maaslin2_top_effects_plot)
} else {
  cat("No significant MaAsLin2 results found (q < 0.1) for visualizations\n")
}
```

### 6.5.11 MaAsLin2 48H RDC vs SDC Analysis

```{r maaslin2-48h-rdc-sdc, echo=FALSE, cache=TRUE, cache.path="cache/integrated-"}
# Subset to 48H timepoint and RDC vs SDC only
ps_maaslin2_48h <- phyloseq::subset_samples(
  ps_maaslin2,
  timepoint == "48H" & carbohydrate %in% c("RDC", "SDC")
)

# Prepare data
maaslin2_48h_counts <- as.data.frame(otu_table(ps_maaslin2_48h))
maaslin2_48h_meta <- as(sample_data(ps_maaslin2_48h), "data.frame")
maaslin2_48h_meta$SampleID <- rownames(maaslin2_48h_meta)

maaslin2_48h_meta <- maaslin2_48h_meta %>%
  dplyr::mutate(
    subject_id = as.character(subject_id),
    carbohydrate = as.factor(carbohydrate)
  )

# Run MaAsLin2 for 48H RDC vs SDC
maaslin2_48h_fit <- tryCatch({
    Maaslin2::Maaslin2(
      input_data = maaslin2_48h_counts,
      input_metadata = maaslin2_48h_meta,
      output = tempfile(pattern = "maaslin2_48h_", tmpdir = getwd()),  # Create temporary output directory
      min_prevalence = 0.05,
      normalization = "TSS",
      transform = "LOG",
      analysis_method = "LM",
      fixed_effects = c("carbohydrate"),
      random_effects = c("subject_id"),
      plot_heatmap = FALSE,
      plot_scatter = FALSE
    )
}, error = function(e) {
  cat("MaAsLin2 48H RDC vs SDC failed:", e$message, "\n")
  NULL
})

# Extract results
if (!is.null(maaslin2_48h_fit) && !is.null(maaslin2_48h_fit$results)) {
  maaslin2_48h_tbl <- maaslin2_48h_fit$results %>%
    dplyr::as_tibble() %>%
    dplyr::mutate(q = p.adjust(pval, method = "BH")) %>%
    dplyr::arrange(q) %>%
    dplyr::left_join(
      tax_table_df %>% dplyr::select(feature, Genus, genus_species),
      by = "feature"
    )
  
  cat("MaAsLin2 48H RDC vs SDC significant results (q < 0.1):", 
      nrow(maaslin2_48h_tbl %>% dplyr::filter(q < 0.1)), "\n")
} else {
  maaslin2_48h_tbl <- tibble(
    feature = character(0),
    metadata = character(0),
    value = character(0),
    coef = numeric(0),
    pval = numeric(0),
    q = numeric(0),
    genus_species = character(0)
  )
}
```

### 6.5.12 MaAsLin2 48H RDC vs SDC Volcano Plot

```{r maaslin2-48h-volcano, echo=FALSE, fig.width=12, fig.height=10}
if (nrow(maaslin2_48h_tbl) > 0) {
  sig_points_48h <- maaslin2_48h_tbl %>%
    dplyr::filter(q < 0.1) %>%
    dplyr::mutate(
      genus_label = dplyr::case_when(
        !is.na(genus_species) & stringr::str_detect(genus_species, " ") ~ 
          stringr::str_extract(genus_species, "^[^ ]+"),
        !is.na(genus_species) ~ genus_species,
        TRUE ~ feature
      )
    )
  
  maaslin2_48h_volcano_plot <- ggplot(
    maaslin2_48h_tbl,
    aes(x = coef, y = -log10(q))
  ) +
    geom_point(aes(color = q < 0.1), alpha = 0.6, size = 2) +
    geom_hline(yintercept = -log10(0.1), linetype = "dashed", color = "red", alpha = 0.7) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    ggrepel::geom_text_repel(
      data = sig_points_48h,
      aes(label = genus_label),
      size = 3,
      color = "#E31A1C",
      fontface = "bold",
      max.overlaps = 15,
      box.padding = 0.5,
      point.padding = 0.3,
      segment.color = "gray50",
      segment.size = 0.3,
      force = 2
    ) +
    scale_color_manual(
      values = c("FALSE" = "gray70", "TRUE" = "#E31A1C"),
      name = "Significant",
      labels = c("FALSE" = "q ≥ 0.1", "TRUE" = "q < 0.1")
    ) +
    labs(
      title = "MaAsLin2 Volcano Plot: 48H RDC vs SDC",
      subtitle = "Carbohydrate Effect Size vs Statistical Significance",
      x = "Coefficient (Effect Size)",
      y = "-log10(q-value)",
      caption = "Red dashed line: q = 0.1 threshold"
    ) +
    theme_scfa() +
    theme(legend.position = "bottom")
  
  print(maaslin2_48h_volcano_plot)
} else {
  cat("No MaAsLin2 48H RDC vs SDC results available for volcano plot\n")
}
```

### 6.4.4 Genus Composition at 48H: No Carbohydrate vs RDC vs SDC

This alluvial plot shows the comparison of genus-level composition at 48H across the three carbohydrate conditions (No Added Carb, RDC, and SDC), highlighting the top 25 most abundant genera and including key differentially abundant taxa.

```{r genus-alluvial-48h-carb, echo=FALSE, fig.width=16, fig.height=12}
# Prepare data for alluvial diagram showing 48H comparison across carbohydrate conditions
ps_genus_rel_48h <- ps_genus_rel

# Filter to 48H only
ps_genus_rel_48h <- phyloseq::subset_samples(ps_genus_rel_48h, timepoint == "48H")

# Extract data
otu_mat_48h <- as.matrix(otu_table(ps_genus_rel_48h))
if (!taxa_are_rows(ps_genus_rel_48h)) {
  otu_mat_48h <- t(otu_mat_48h)
}

tax_df_48h <- as.data.frame(tax_table(ps_genus_rel_48h))
meta_df_48h <- data.frame(sample_data(ps_genus_rel_48h)) %>%
  tibble::rownames_to_column("SampleID")

# Create long format data
alluvial_data_48h_raw <- otu_mat_48h %>%
  as.data.frame() %>%
  tibble::rownames_to_column("ASV") %>%
  tidyr::pivot_longer(-ASV, names_to = "SampleID", values_to = "Abundance") %>%
  dplyr::left_join(
    tax_df_48h %>% tibble::rownames_to_column("ASV") %>% dplyr::select(ASV, Genus),
    by = "ASV"
  ) %>%
  dplyr::left_join(
    meta_df_48h %>% dplyr::select(SampleID, carbohydrate),
    by = "SampleID"
  )

# Remove NA values and filter to only valid combinations
alluvial_data_48h_clean <- alluvial_data_48h_raw %>%
  dplyr::filter(
    !is.na(Abundance),
    !is.na(Genus),
    !is.na(carbohydrate),
    carbohydrate != "NA"
  )

# Group by Genus and carbohydrate to get mean abundance at 48H
alluvial_data_48h_grouped <- alluvial_data_48h_clean %>%
  dplyr::group_by(Genus, carbohydrate) %>%
  dplyr::summarise(mean_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
  dplyr::mutate(
    mean_abundance = ifelse(is.nan(mean_abundance) | is.na(mean_abundance), 0, mean_abundance)
  )

# Get top 25 genera for visualization (use max abundance across carbohydrate conditions)
top_genera_48h_by_abundance <- alluvial_data_48h_grouped %>%
  dplyr::group_by(Genus) %>%
  dplyr::summarise(max_abund = max(mean_abundance, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(desc(max_abund)) %>%
  head(25) %>%
  dplyr::pull(Genus)

# Define genera that are important from DA analysis and should be included
da_important_genera_48h <- c(
  "g_Fusicatenibacter",
  "g__Fusicatenibacter",
  "Fusicatenibacter",
  "g_Fusicatenibacters",
  "g__Fusicatenibacters",
  "Fusicatenibacters",
  "g_Hungatella",
  "g__Hungatella",
  "Hungatella",
  "g_Eggerthella",
  "g__Eggerthella",
  "Eggerthella",
  "g_Peptoniphilus",
  "g__Peptoniphilus",
  "Peptoniphilus"
)

# Check which of these genera exist in the 48H data
available_genera_48h_from_tax <- unique(tax_df_48h$Genus)

# Check exact matches first
da_genera_48h_in_data <- da_important_genera_48h[da_important_genera_48h %in% available_genera_48h_from_tax]

# Also check for variations
da_genera_48h_variations <- available_genera_48h_from_tax[
  stringr::str_detect(available_genera_48h_from_tax, regex("Fusicatenibacter|Hungatella|Eggerthella|Peptoniphilus", ignore_case = TRUE))
]

# Combine exact matches and variations
da_genera_48h_found <- unique(c(da_genera_48h_in_data, da_genera_48h_variations))

# Combine top 25 by abundance with DA-important genera
# Remove duplicates and keep up to 29 genera total (25 + up to 4 DA genera)
top_genera_48h_alluvial <- unique(c(top_genera_48h_by_abundance, da_genera_48h_found))
if (length(top_genera_48h_alluvial) > 29) {
  # If we have more than 29, prioritize top 25 by abundance, then add DA genera
  top_genera_48h_alluvial <- c(
    top_genera_48h_by_abundance,
    setdiff(da_genera_48h_found, top_genera_48h_by_abundance)
  )[1:min(29, length(top_genera_48h_alluvial))]
}

# Filter to top genera and add "Other" category
alluvial_48h_plot_data_pre_complete <- alluvial_data_48h_grouped %>%
  dplyr::mutate(
    Genus_grouped = ifelse(Genus %in% top_genera_48h_alluvial, as.character(Genus), "Other")
  ) %>%
  dplyr::group_by(Genus_grouped, carbohydrate) %>%
  dplyr::summarise(Abundance = sum(mean_abundance, na.rm = TRUE), .groups = "drop") %>%
  dplyr::mutate(
    Abundance = ifelse(is.nan(Abundance) | is.na(Abundance), 0, Abundance)
  )

# Ensure all combinations exist (Genus × carbohydrate)
alluvial_48h_plot_data <- alluvial_48h_plot_data_pre_complete %>%
  tidyr::complete(
    Genus_grouped = c(unique(alluvial_48h_plot_data_pre_complete$Genus_grouped), "Other"),
    carbohydrate = c("No Added Carb", "RDC", "SDC"),
    fill = list(Abundance = 0)
  ) %>%
  dplyr::filter(!is.na(carbohydrate))

# Clean genus names for display (remove g__ prefix)
# First create Genus_display, then filter out NA, "NA", "Other", and empty values
alluvial_48h_plot_data <- alluvial_48h_plot_data %>%
  dplyr::mutate(
    Genus_display = stringr::str_replace(Genus_grouped, "^g__", ""),
    # Order carbohydrates for plotting
    carbohydrate = factor(carbohydrate, levels = c("No Added Carb", "RDC", "SDC"))
  ) %>%
  # Filter out NA, "NA", "Other", and empty values after creating Genus_display
  dplyr::filter(
    !is.na(Genus_display),
    Genus_display != "NA",
    Genus_display != "Other",
    Genus_display != "",
    !is.na(Genus_grouped),
    Genus_grouped != "NA",
    Genus_grouped != "Other",
    Genus_grouped != ""
  )

# Create alluvial plot showing flow between carbohydrate conditions
# Add borders for better visibility
# Label ALL genera (including thin flows) - no threshold
alluvial_48h_plot_data_labeled <- alluvial_48h_plot_data %>%
  dplyr::mutate(
    # Label all genera consistently across all columns
    label_genus = as.character(Genus_display)
  )

alluvial_48h_plot <- ggplot(
  alluvial_48h_plot_data_labeled,
  aes(
    x = carbohydrate,
    y = Abundance,
    stratum = Genus_display,
    alluvium = Genus_display,
    fill = Genus_display,
    label = label_genus
  )
) +
  geom_flow(alpha = 0.7, width = 0.3, color = "white", linewidth = 0.3) +
  geom_stratum(alpha = 0.9, width = 0.3, color = "white", linewidth = 0.5) +
  geom_text(
    stat = "stratum",
    size = 3,
    angle = 0,
    hjust = 0.5,
    vjust = 0.5,
    color = "black",
    fontface = "bold",
    check_overlap = TRUE
  ) +
  scale_fill_manual(
    values = scico::scico(length(unique(alluvial_48h_plot_data$Genus_display)), palette = "batlow"),
    name = "Genus"
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), expand = c(0, 0)) +
  labs(
    title = "Genus Composition at 48H: Comparison Across Carbohydrate Conditions",
    subtitle = paste0("Alluvial diagram showing relative abundance for top ", length(top_genera_48h_alluvial), " genera (including DA-important taxa)"),
    x = "Carbohydrate Condition",
    y = "Relative Abundance",
    fill = "Genus"
  ) +
  theme_scfa() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 11),
    legend.position = "none",  # Hide legend due to many taxa
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11)
  )

print(alluvial_48h_plot)

cat("\n48H Carbohydrate Comparison Summary:\n")
cat("- Total genera included:", length(top_genera_48h_alluvial), "\n")
cat("- DA-important genera found:", length(da_genera_48h_found), "\n")
if (length(da_genera_48h_found) > 0) {
  cat("- DA-important genera:", paste(da_genera_48h_found, collapse = ", "), "\n")
}
```

## 6.6 Absolute Abundances

Relative abundance data (compositional) can mask true changes in microbial load. To address this, we integrated qPCR-derived total genome copy estimates per sample to convert relative abundances to absolute abundances (genome copies per sample). This allows us to distinguish between compositional shifts versus true changes in microbial load.

### 6.6.1 Integration of qPCR Data

```{r absolute-abundance-integration, echo=FALSE, cache=TRUE, cache.path="cache/integrated-"}
# Load qPCR data
qpcr_file <- "../microbiome_analysis/zr24558.16S_250813.zymo/00...AllSamples.Bac16Sv34/Sample_Information/absolute.abundance.csv"
qpcr_data_raw <- readr::read_csv(qpcr_file, show_col_types = FALSE)
qpcr_data <- qpcr_data_raw %>%
  janitor::clean_names()

# Find the genome copies column (original was "genome_copies_per_ul*" which becomes "genome_copies_per_ul" after clean_names)
# Check for exact match first, then look for columns containing the pattern
genome_cols <- colnames(qpcr_data)[stringr::str_detect(colnames(qpcr_data), "genome_copies_per_ul")]
if (length(genome_cols) == 0) {
  stop("Could not find genome_copies_per_ul column in qPCR data")
}
# Use the first matching column (should be the one from "genome_copies_per_ul*")
genome_col <- genome_cols[1]

# Select only customer_label and the genome copies column, then rename to standardize
qpcr_data <- qpcr_data %>%
  dplyr::select(customer_label, !!rlang::sym(genome_col)) %>%
  dplyr::rename(genome_copies_per_ul = !!rlang::sym(genome_col))

# Clean customer_label to match sample metadata format
qpcr_data <- qpcr_data %>%
  dplyr::mutate(
    customer_label = stringr::str_trim(customer_label),
    # Remove trailing periods and spaces
    customer_label = stringr::str_replace_all(customer_label, "\\.+$", ""),
    customer_label = stringr::str_replace_all(customer_label, "\\.\\.", ".")
  )

# Get sample metadata from phyloseq
sample_meta_abs <- data.frame(sample_data(ps)) %>%
  tibble::rownames_to_column("SampleID") %>%
  dplyr::mutate(
    # Create matching customer_label from SampleID if needed
    customer_label_match = dplyr::case_when(
      # Try direct match first
      SampleID %in% qpcr_data$customer_label ~ SampleID,
      # Try with different patterns
      stringr::str_detect(SampleID, "^S") ~ stringr::str_replace(SampleID, "^S", ""),
      TRUE ~ SampleID
    )
  )

# Match qPCR data to samples
# Try direct match first, then pattern matching
sample_meta_abs <- sample_meta_abs %>%
  dplyr::left_join(
    qpcr_data,
    by = c("SampleID" = "customer_label")
  ) %>%
  dplyr::mutate(
    # If no direct match, try customer_label_match
    genome_copies_per_ul = dplyr::if_else(
      is.na(genome_copies_per_ul),
      dplyr::left_join(
        data.frame(customer_label_match = .$customer_label_match),
        qpcr_data %>% dplyr::select(customer_label, genome_copies_per_ul),
        by = c("customer_label_match" = "customer_label")
      )$genome_copies_per_ul,
      genome_copies_per_ul
    )
  )

# Calculate samples matched
matched_samples <- sum(!is.na(sample_meta_abs$genome_copies_per_ul))
cat("qPCR data integration:\n")
cat("- Total samples in phyloseq:", nrow(sample_meta_abs), "\n")
cat("- Samples matched to qPCR data:", matched_samples, "\n")
cat("- Match rate:", round(matched_samples / nrow(sample_meta_abs) * 100, 1), "%\n")
```

### 6.6.2 Conversion to Absolute Abundances

```{r absolute-abundance-conversion, echo=FALSE, cache=TRUE, cache.path="cache/integrated-"}
# Convert relative abundances to absolute abundances
# For each sample: absolute_abundance = relative_abundance × total_genome_copies

# Get relative abundance phyloseq object (genus-level)
ps_genus_abs <- ps_genus_rel

# Match genome copies to samples
sample_data(ps_genus_abs)$genome_copies_per_ul <- sample_meta_abs$genome_copies_per_ul[
  match(sample_names(ps_genus_abs), sample_meta_abs$SampleID)
]

# Create absolute abundance otu table
otu_rel <- as.matrix(otu_table(ps_genus_rel))
genome_copies <- sample_data(ps_genus_abs)$genome_copies_per_ul

# Convert to absolute abundances
otu_abs <- otu_rel
for (i in 1:ncol(otu_rel)) {
  if (!is.na(genome_copies[i])) {
    # Absolute abundance = relative abundance × total genome copies
    otu_abs[, i] <- otu_rel[, i] * genome_copies[i]
  } else {
    otu_abs[, i] <- NA
  }
}

# Create phyloseq object with absolute abundances
ps_genus_abs_obj <- ps_genus_abs
otu_table(ps_genus_abs_obj) <- otu_table(otu_abs, taxa_are_rows = TRUE)

cat("Absolute abundance conversion complete:\n")
cat("- Samples with qPCR data:", sum(!is.na(genome_copies)), "\n")
cat("- Samples without qPCR data:", sum(is.na(genome_copies)), "\n")
```

### 6.6.3 Absolute Abundance Plots by Group, Carbohydrate, and Timepoint

```{r qpcr-summary-boxplots, echo=FALSE, fig.width=14, fig.height=8}
# Create summary boxplots of qPCR data (genomes/μL) by group, carbohydrate, and timepoint
# Use sample_meta_abs which has genome_copies_per_ul matched to samples

qpcr_plot_data <- sample_meta_abs %>%
  dplyr::filter(!is.na(genome_copies_per_ul)) %>%
  dplyr::filter(!is.na(group), !is.na(carbohydrate), !is.na(timepoint)) %>%
  dplyr::mutate(
    group = as.character(group),
    carbohydrate = as.character(carbohydrate),
    timepoint = as.character(timepoint),
    # Convert to log10 for better visualization if needed
    genome_copies_log10 = log10(genome_copies_per_ul + 1)
  )

# Plot 1: Boxplots by group and carbohydrate, faceted by timepoint
qpcr_plot1 <- ggplot(
  qpcr_plot_data,
  aes(x = carbohydrate, y = genome_copies_per_ul, fill = group)
) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA, position = position_dodge(width = 0.8)) +
  geom_jitter(
    aes(color = group),
    alpha = 0.5,
    size = 1.2,
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8)
  ) +
  facet_wrap(~ timepoint, ncol = 2, scales = "free_y") +
  scale_fill_scfa() +
  scale_color_scfa() +
  scale_y_continuous(labels = scales::comma, trans = "log10") +
  labs(
    title = "Total Bacterial Genome Copies per μL",
    subtitle = "qPCR-derived total bacterial load by group, carbohydrate type, and timepoint",
    x = "Carbohydrate Type",
    y = "Genome Copies per μL (log10 scale)",
    fill = "Group",
    color = "Group"
  ) +
  theme_scfa() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

print(qpcr_plot1)

# Plot 2: Over time by carbohydrate and group
qpcr_summary <- qpcr_plot_data %>%
  dplyr::group_by(group, carbohydrate, timepoint) %>%
  dplyr::summarise(
    mean_genomes = mean(genome_copies_per_ul, na.rm = TRUE),
    se_genomes = sd(genome_copies_per_ul, na.rm = TRUE) / sqrt(dplyr::n()),
    median_genomes = median(genome_copies_per_ul, na.rm = TRUE),
    n = dplyr::n(),
    .groups = "drop"
  )

qpcr_plot2 <- ggplot(
  qpcr_summary,
  aes(x = timepoint, y = mean_genomes, color = group, group = group)
) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = mean_genomes - se_genomes, ymax = mean_genomes + se_genomes),
    width = 0.1,
    size = 0.8
  ) +
  facet_wrap(~ carbohydrate, ncol = 3, scales = "free_y") +
  scale_color_scfa() +
  scale_y_continuous(labels = scales::comma, trans = "log10") +
  labs(
    title = "Total Bacterial Genome Copies per μL Over Time",
    subtitle = "Mean ± SE by carbohydrate type and group",
    x = "Timepoint",
    y = "Mean Genome Copies per μL (log10 scale)",
    color = "Group"
  ) +
  theme_scfa() +
  theme(legend.position = "bottom")

print(qpcr_plot2)

# Summary statistics
cat("\nqPCR Data Summary:\n")
cat("- Total samples with qPCR data:", nrow(qpcr_plot_data), "\n")
cat("- Samples by timepoint:\n")
print(table(qpcr_plot_data$timepoint, useNA = "ifany"))
cat("- Samples by carbohydrate:\n")
print(table(qpcr_plot_data$carbohydrate, useNA = "ifany"))
cat("- Samples by group:\n")
print(table(qpcr_plot_data$group, useNA = "ifany"))
```

### 6.6.4 Fusicatenibacter Absolute Abundance Plots

```{r fusicatenibacter-absolute-abundance, echo=FALSE, fig.width=14, fig.height=10}
# Extract absolute abundance data for g__Fusicatenibacter
# Find all ASVs assigned to this genus
fusicatenibacter_taxa <- taxa_names(ps_genus_abs_obj)[
  stringr::str_detect(tax_table(ps_genus_abs_obj)[, "Genus"], regex("Fusicatenibacter", ignore_case = TRUE))
]

if (length(fusicatenibacter_taxa) > 0) {
  # Get absolute abundance for Fusicatenibacter genus
  # Sum across all ASVs if multiple
  if (length(fusicatenibacter_taxa) > 1) {
    fusicatenibacter_abs <- colSums(otu_table(ps_genus_abs_obj)[fusicatenibacter_taxa, , drop = FALSE], na.rm = TRUE)
  } else {
    fusicatenibacter_abs <- as.numeric(otu_table(ps_genus_abs_obj)[fusicatenibacter_taxa, ])
  }
  
  # Get metadata
  meta_abs <- data.frame(sample_data(ps_genus_abs_obj)) %>%
    tibble::rownames_to_column("SampleID") %>%
    dplyr::mutate(
      fusicatenibacter_abs_abund = fusicatenibacter_abs,
      # Filter out samples without qPCR data
      has_qpcr = !is.na(genome_copies_per_ul)
    ) %>%
    dplyr::filter(has_qpcr) %>%
    dplyr::filter(!is.na(group), !is.na(carbohydrate), !is.na(timepoint))
  
  # Calculate mean and SE by group, carbohydrate, and timepoint
  fusicatenibacter_summary <- meta_abs %>%
    dplyr::group_by(group, carbohydrate, timepoint) %>%
    dplyr::summarise(
      mean_abs_abund = mean(fusicatenibacter_abs_abund, na.rm = TRUE),
      se_abs_abund = sd(fusicatenibacter_abs_abund, na.rm = TRUE) / sqrt(dplyr::n()),
      n = dplyr::n(),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      mean_abs_abund = ifelse(is.na(mean_abs_abund), 0, mean_abs_abund),
      se_abs_abund = ifelse(is.na(se_abs_abund), 0, se_abs_abund)
    )
  
  # Create plots
  # Plot 1: By group and carbohydrate, faceted by timepoint
  fusicatenibacter_plot1 <- ggplot(
    fusicatenibacter_summary,
    aes(x = carbohydrate, y = mean_abs_abund, fill = group)
  ) +
    geom_col(position = position_dodge(width = 0.8), alpha = 0.8) +
    geom_errorbar(
      aes(ymin = mean_abs_abund - se_abs_abund, ymax = mean_abs_abund + se_abs_abund),
      position = position_dodge(width = 0.8),
      width = 0.2
    ) +
    facet_wrap(~ timepoint, ncol = 2) +
    scale_fill_scfa() +
    scale_y_continuous(labels = scales::comma) +
    labs(
      title = "g__Fusicatenibacter Absolute Abundance",
      subtitle = "Mean ± SE genome copies per μL by group, carbohydrate, and timepoint",
      x = "Carbohydrate",
      y = "Absolute Abundance (genome copies/μL)",
      fill = "Group"
    ) +
    theme_scfa() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  print(fusicatenibacter_plot1)
  
  # Plot 2: Over time by carbohydrate and group
  fusicatenibacter_plot2 <- ggplot(
    fusicatenibacter_summary,
    aes(x = timepoint, y = mean_abs_abund, color = group, group = group)
  ) +
    geom_line(size = 1.2) +
    geom_point(size = 3) +
    geom_errorbar(
      aes(ymin = mean_abs_abund - se_abs_abund, ymax = mean_abs_abund + se_abs_abund),
      width = 0.1
    ) +
    facet_wrap(~ carbohydrate, ncol = 3) +
    scale_color_scfa() +
    scale_y_continuous(labels = scales::comma) +
    labs(
      title = "g__Fusicatenibacter Absolute Abundance Over Time",
      subtitle = "Mean ± SE genome copies per μL by carbohydrate type and group",
      x = "Timepoint",
      y = "Absolute Abundance (genome copies/μL)",
      color = "Group"
    ) +
    theme_scfa()
  
  print(fusicatenibacter_plot2)
  
} else {
  cat("No Fusicatenibacter taxa found in the data\n")
}
```

# 7. Integrated Analysis: Taxon–SCFA Relationships

This section integrates SCFA metabolomics with microbiome composition to identify taxa associated with SCFA concentrations. We model associations between microbial taxa and butyrate/propionate concentrations while accounting for carbohydrate type effects, using both relative and absolute abundance data.

## 7.1 Integration of SCFA Data with Microbiome Metadata

```{r scfa-microbiome-integration, echo=FALSE, cache=TRUE, cache.path="cache/integrated-"}
# Merge SCFA data with microbiome metadata
# SCFA uses: subject, group, carbohydrate_type, timepoint_hr
# Microbiome uses: subject_id, group, carbohydrate, timepoint

# Get SCFA concentrations at subject × carbohydrate × timepoint level
scfa_wide <- scfa_sample_data_long_factored %>%
  dplyr::filter(analyte %in% c("butyrate", "propionate")) %>%
  dplyr::filter(timepoint_hr %in% c("0", "48")) %>%
  tidyr::pivot_wider(
    id_cols = c(subject, group, carbohydrate_type, timepoint_hr),
    names_from = analyte,
    values_from = concentration,
    names_prefix = "scfa_"
  ) %>%
  dplyr::mutate(
    carbohydrate = dplyr::case_when(
      carbohydrate_type == "no_carbohydrate" ~ "No Added Carb",
      carbohydrate_type == "rapid_digestible" ~ "RDC",
      carbohydrate_type == "slow_digestible" ~ "SDC",
      TRUE ~ NA_character_
    ),
    timepoint = dplyr::case_when(
      timepoint_hr == "0" ~ "0H",
      timepoint_hr == "48" ~ "48H",
      TRUE ~ NA_character_
    ),
    group_micro = dplyr::case_when(
      group == "control" ~ "Healthy",
      group == "case" ~ "Obese",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::filter(!is.na(carbohydrate), !is.na(timepoint), !is.na(group_micro))

# Diagnostic: Check subject IDs in SCFA data
cat("\nSCFA Data Summary:\n")
cat("- Unique subjects:", length(unique(scfa_wide$subject)), "\n")
cat("- Unique subject × carbohydrate × timepoint combinations:", nrow(scfa_wide), "\n")
cat("- Sample subjects:", paste(head(unique(scfa_wide$subject), 10), collapse = ", "), "\n")

# Get microbiome sample metadata
micro_meta_merged <- data.frame(sample_data(ps)) %>%
  tibble::rownames_to_column("SampleID") %>%
  dplyr::mutate(
    subject_id = as.character(subject_id),
    # Extract numeric part of subject_id for matching (e.g., "84B" -> "84")
    subject_id_numeric = stringr::str_extract(subject_id, "^[0-9]+"),
    group = as.character(group),
    carbohydrate = as.character(carbohydrate),
    timepoint = as.character(timepoint)
  )

# Diagnostic: Check subject IDs in microbiome data
cat("\nMicrobiome Data Summary:\n")
cat("- Unique subject_ids:", length(unique(micro_meta_merged$subject_id)), "\n")
cat("- Unique numeric subject_ids:", length(unique(micro_meta_merged$subject_id_numeric)), "\n")
cat("- Sample subject_ids:", paste(head(unique(micro_meta_merged$subject_id), 10), collapse = ", "), "\n")
cat("- Sample numeric subject_ids:", paste(head(unique(micro_meta_merged$subject_id_numeric), 10), collapse = ", "), "\n")

# Merge SCFA data with microbiome metadata
# Match on numeric subject ID (SCFA has numeric, microbiome has "84B" format)
micro_meta_scfa <- micro_meta_merged %>%
  dplyr::left_join(
    scfa_wide,
    by = c(
      "subject_id_numeric" = "subject",
      "group" = "group_micro",
      "carbohydrate" = "carbohydrate",
      "timepoint" = "timepoint"
    )
  )

cat("SCFA-Microbiome Integration Summary:
")
cat("- Total microbiome samples:", nrow(micro_meta_scfa), "
")
cat("- Samples with butyrate data:", sum(!is.na(micro_meta_scfa$scfa_butyrate)), "
")
cat("- Samples with propionate data:", sum(!is.na(micro_meta_scfa$scfa_propionate)), "
")
cat("- Samples with both SCFAs:", sum(!is.na(micro_meta_scfa$scfa_butyrate) & 
                                        !is.na(micro_meta_scfa$scfa_propionate)), "
")
```

## 7.2 Taxon–Butyrate Associations

```{r taxon-butyrate-relative, echo=FALSE, cache=TRUE, cache.path="cache/integrated-"}
# Prepare data for analysis: relative abundance with butyrate
ps_butyrate_rel <- ps

# Update sample_data with SCFA concentrations
sample_data(ps_butyrate_rel)$butyrate <- micro_meta_scfa$scfa_butyrate[
  match(sample_names(ps_butyrate_rel), micro_meta_scfa$SampleID)
]

# Filter to samples with butyrate data
ps_butyrate_rel <- phyloseq::subset_samples(
  ps_butyrate_rel,
  !is.na(butyrate) & !is.na(carbohydrate) & !is.na(timepoint)
)

cat("Butyrate analysis (relative):", nsamples(ps_butyrate_rel), "samples
")

# Run ANCOM-BC2 for taxon ~ butyrate + carbohydrate
# Note: When using continuous variables, set global = FALSE and group = NULL
# or specify a group variable for structural zero detection
if (nsamples(ps_butyrate_rel) > 10 && ntaxa(ps_butyrate_rel) > 1) {
  ancombc2_butyrate_rel <- tryCatch({
    ANCOMBC::ancombc2(
      data = ps_butyrate_rel,
      fix_formula = "butyrate + carbohydrate",
      rand_formula = "(1|subject_id)",
      p_adj_method = "BH",
      prv_cut = 0.10,
      lib_cut = 1000,
      alpha = 0.05,
      group = "carbohydrate",  # Use carbohydrate as group for structural zeros
      struc_zero = TRUE,
      neg_lb = TRUE,
      global = TRUE,
      verbose = FALSE
    )
  }, error = function(e) {
    cat("ANCOM-BC2 butyrate (relative) failed:", e$message, "
")
    NULL
  })
  
  if (!is.null(ancombc2_butyrate_rel) && !is.null(ancombc2_butyrate_rel$res)) {
    res_data <- ancombc2_butyrate_rel$res
    ancombc2_butyrate_tbl_rel <- tibble::tibble()
    
    if ("q_butyrate" %in% colnames(res_data)) {
      ancombc2_butyrate_tbl_rel <- res_data %>%
        dplyr::filter(q_butyrate < 0.1) %>%
        dplyr::mutate(
          effect = "butyrate",
          beta = lfc_butyrate,
          q_value = q_butyrate
        ) %>%
        dplyr::select(taxon, effect, beta, q_value)
    }
    
    cat("ANCOM-BC2 butyrate (relative):", nrow(ancombc2_butyrate_tbl_rel), 
        "significant taxa (q < 0.1)
")
  } else {
    ancombc2_butyrate_tbl_rel <- tibble::tibble()
  }
} else {
  ancombc2_butyrate_tbl_rel <- tibble::tibble()
}

# Run MaAsLin2 for taxon ~ butyrate + carbohydrate
# Ensure count matrix columns match metadata rownames
maaslin2_butyrate_counts_rel <- as.data.frame(otu_table(ps_butyrate_rel))
if (!taxa_are_rows(ps_butyrate_rel)) {
  maaslin2_butyrate_counts_rel <- t(maaslin2_butyrate_counts_rel)
}

# Prepare metadata with rownames matching count matrix column names
maaslin2_butyrate_meta_rel <- data.frame(sample_data(ps_butyrate_rel)) %>%
  dplyr::mutate(
    subject_id = as.character(subject_id),
    carbohydrate = as.factor(carbohydrate),
    butyrate = as.numeric(butyrate)
  )
# Ensure rownames match count matrix column names
rownames(maaslin2_butyrate_meta_rel) <- sample_names(ps_butyrate_rel)

maaslin2_butyrate_rel <- tryCatch({
  Maaslin2::Maaslin2(
    input_data = maaslin2_butyrate_counts_rel,
    input_metadata = maaslin2_butyrate_meta_rel,
    output = tempfile(pattern = "maaslin2_butyrate_rel_", tmpdir = getwd()),
    min_prevalence = 0.05,
    normalization = "TSS",
    transform = "LOG",
    analysis_method = "LM",
    fixed_effects = c("butyrate", "carbohydrate"),
    random_effects = c("subject_id"),
    plot_heatmap = FALSE,
    plot_scatter = FALSE
  )
}, error = function(e) {
  cat("MaAsLin2 butyrate (relative) failed:", e$message, "
")
  NULL
})

if (!is.null(maaslin2_butyrate_rel) && !is.null(maaslin2_butyrate_rel$results)) {
  maaslin2_butyrate_tbl_rel <- maaslin2_butyrate_rel$results %>%
    dplyr::as_tibble() %>%
    dplyr::mutate(q = p.adjust(pval, method = "BH")) %>%
    dplyr::filter(q < 0.1 & metadata == "butyrate") %>%
    dplyr::arrange(q)
  
  cat("MaAsLin2 butyrate (relative):", nrow(maaslin2_butyrate_tbl_rel), 
      "significant associations (q < 0.1)
")
} else {
  maaslin2_butyrate_tbl_rel <- tibble::tibble()
}
```

## 7.3 Taxon–Propionate Associations

```{r taxon-propionate-relative, echo=FALSE, cache=TRUE, cache.path="cache/integrated-"}
# Prepare data for analysis: relative abundance with propionate
ps_propionate_rel <- ps

# Update sample_data with SCFA concentrations
sample_data(ps_propionate_rel)$propionate <- micro_meta_scfa$scfa_propionate[
  match(sample_names(ps_propionate_rel), micro_meta_scfa$SampleID)
]

# Filter to samples with propionate data
ps_propionate_rel <- phyloseq::subset_samples(
  ps_propionate_rel,
  !is.na(propionate) & !is.na(carbohydrate) & !is.na(timepoint)
)

cat("Propionate analysis (relative):", nsamples(ps_propionate_rel), "samples
")

# Run ANCOM-BC2 for taxon ~ propionate + carbohydrate
if (nsamples(ps_propionate_rel) > 10 && ntaxa(ps_propionate_rel) > 1) {
  ancombc2_propionate_rel <- tryCatch({
    ANCOMBC::ancombc2(
      data = ps_propionate_rel,
      fix_formula = "propionate + carbohydrate",
      rand_formula = "(1|subject_id)",
      p_adj_method = "BH",
      prv_cut = 0.10,
      lib_cut = 1000,
      alpha = 0.05,
      group = "carbohydrate",  # Use carbohydrate as group for structural zeros
      struc_zero = TRUE,
      neg_lb = TRUE,
      global = TRUE,
      verbose = FALSE
    )
  }, error = function(e) {
    cat("ANCOM-BC2 propionate (relative) failed:", e$message, "
")
    NULL
  })
  
  if (!is.null(ancombc2_propionate_rel) && !is.null(ancombc2_propionate_rel$res)) {
    res_data <- ancombc2_propionate_rel$res
    ancombc2_propionate_tbl_rel <- tibble::tibble()
    
    if ("q_propionate" %in% colnames(res_data)) {
      ancombc2_propionate_tbl_rel <- res_data %>%
        dplyr::filter(q_propionate < 0.1) %>%
        dplyr::mutate(
          effect = "propionate",
          beta = lfc_propionate,
          q_value = q_propionate
        ) %>%
        dplyr::select(taxon, effect, beta, q_value)
    }
    
    cat("ANCOM-BC2 propionate (relative):", nrow(ancombc2_propionate_tbl_rel), 
        "significant taxa (q < 0.1)
")
  } else {
    ancombc2_propionate_tbl_rel <- tibble::tibble()
  }
} else {
  ancombc2_propionate_tbl_rel <- tibble::tibble()
}

# Run MaAsLin2 for taxon ~ propionate + carbohydrate
# Ensure count matrix columns match metadata rownames
maaslin2_propionate_counts_rel <- as.data.frame(otu_table(ps_propionate_rel))
if (!taxa_are_rows(ps_propionate_rel)) {
  maaslin2_propionate_counts_rel <- t(maaslin2_propionate_counts_rel)
}

# Prepare metadata with rownames matching count matrix column names
maaslin2_propionate_meta_rel <- data.frame(sample_data(ps_propionate_rel)) %>%
  dplyr::mutate(
    subject_id = as.character(subject_id),
    carbohydrate = as.factor(carbohydrate),
    propionate = as.numeric(propionate)
  )
# Ensure rownames match count matrix column names
rownames(maaslin2_propionate_meta_rel) <- sample_names(ps_propionate_rel)

maaslin2_propionate_rel <- tryCatch({
  Maaslin2::Maaslin2(
    input_data = maaslin2_propionate_counts_rel,
    input_metadata = maaslin2_propionate_meta_rel,
    output = tempfile(pattern = "maaslin2_propionate_rel_", tmpdir = getwd()),
    min_prevalence = 0.05,
    normalization = "TSS",
    transform = "LOG",
    analysis_method = "LM",
    fixed_effects = c("propionate", "carbohydrate"),
    random_effects = c("subject_id"),
    plot_heatmap = FALSE,
    plot_scatter = FALSE
  )
}, error = function(e) {
  cat("MaAsLin2 propionate (relative) failed:", e$message, "
")
  NULL
})

if (!is.null(maaslin2_propionate_rel) && !is.null(maaslin2_propionate_rel$results)) {
  maaslin2_propionate_tbl_rel <- maaslin2_propionate_rel$results %>%
    dplyr::as_tibble() %>%
    dplyr::mutate(q = p.adjust(pval, method = "BH")) %>%
    dplyr::filter(q < 0.1 & metadata == "propionate") %>%
    dplyr::arrange(q)
  
  cat("MaAsLin2 propionate (relative):", nrow(maaslin2_propionate_tbl_rel), 
      "significant associations (q < 0.1)
")
} else {
  maaslin2_propionate_tbl_rel <- tibble::tibble()
}
```

## 7.4 Fusicatenibacter–SCFA Correlations

```{r fusicatenibacter-scfa-correlations, echo=FALSE, fig.width=14, fig.height=10}
# Extract relative abundance of g__Fusicatenibacter
fusicatenibacter_taxa_rel <- taxa_names(ps)[
  stringr::str_detect(tax_table(ps)[, "Genus"], regex("Fusicatenibacter", ignore_case = TRUE))
]

# Prepare correlation data
corr_data <- micro_meta_scfa %>%
  dplyr::select(SampleID, subject_id, group, carbohydrate, timepoint, 
                scfa_butyrate, scfa_propionate) %>%
  dplyr::filter(!is.na(scfa_butyrate) | !is.na(scfa_propionate))

# Add relative abundance of Fusicatenibacter
if (length(fusicatenibacter_taxa_rel) > 0) {
  if (length(fusicatenibacter_taxa_rel) > 1) {
    fusicatenibacter_rel_abund <- colSums(
      as.data.frame(otu_table(ps_rel))[fusicatenibacter_taxa_rel, , drop = FALSE],
      na.rm = TRUE
    )
  } else {
    fusicatenibacter_rel_abund <- as.numeric(
      otu_table(ps_rel)[fusicatenibacter_taxa_rel, ]
    )
  }
  
  names(fusicatenibacter_rel_abund) <- sample_names(ps_rel)
  
  corr_data <- corr_data %>%
    dplyr::mutate(
      fusicatenibacter_rel = fusicatenibacter_rel_abund[SampleID]
    )
  
  # Filter to samples with Fusicatenibacter data
  corr_data <- corr_data %>%
    dplyr::filter(!is.na(fusicatenibacter_rel))
  
  # Create correlation plots
  # Plot 1: Butyrate correlations by timepoint
  corr_butyrate_plot <- ggplot(
    corr_data %>% dplyr::filter(!is.na(scfa_butyrate)),
    aes(x = scfa_butyrate, y = fusicatenibacter_rel)
  ) +
    geom_point(aes(color = carbohydrate), alpha = 0.6, size = 2) +
    geom_smooth(method = "lm", se = TRUE, color = "black") +
    facet_wrap(~ timepoint, scales = "free") +
    scale_color_scfa() +
    labs(
      title = "Fusicatenibacter vs Butyrate (Relative Abundance)",
      subtitle = "Correlation by timepoint and carbohydrate type",
      x = "Butyrate Concentration",
      y = "Fusicatenibacter Relative Abundance",
      color = "Carbohydrate"
    ) +
    theme_scfa() +
    theme(legend.position = "bottom")
  
  print(corr_butyrate_plot)
  
  # Plot 2: Propionate correlations by timepoint
  corr_propionate_plot <- ggplot(
    corr_data %>% dplyr::filter(!is.na(scfa_propionate)),
    aes(x = scfa_propionate, y = fusicatenibacter_rel)
  ) +
    geom_point(aes(color = carbohydrate), alpha = 0.6, size = 2) +
    geom_smooth(method = "lm", se = TRUE, color = "black") +
    facet_wrap(~ timepoint, scales = "free") +
    scale_color_scfa() +
    labs(
      title = "Fusicatenibacter vs Propionate (Relative Abundance)",
      subtitle = "Correlation by timepoint and carbohydrate type",
      x = "Propionate Concentration",
      y = "Fusicatenibacter Relative Abundance",
      color = "Carbohydrate"
    ) +
    theme_scfa() +
    theme(legend.position = "bottom")
  
  print(corr_propionate_plot)
  
  # Calculate correlation statistics
  cat("
Fusicatenibacter-SCFA Correlations (Relative Abundance):
")
  for (tp in c("0H", "48H")) {
    corr_data_tp <- corr_data %>% dplyr::filter(timepoint == tp)
    
    # Butyrate correlation
    if (sum(!is.na(corr_data_tp$scfa_butyrate) & 
            !is.na(corr_data_tp$fusicatenibacter_rel)) > 3) {
      corr_but <- cor.test(corr_data_tp$scfa_butyrate, 
                          corr_data_tp$fusicatenibacter_rel,
                          method = "spearman")
      cat(sprintf("  %s - Butyrate: rho = %.3f, p = %.4f
", 
                  tp, corr_but$estimate, corr_but$p.value))
    }
    
    # Propionate correlation
    if (sum(!is.na(corr_data_tp$scfa_propionate) & 
            !is.na(corr_data_tp$fusicatenibacter_rel)) > 3) {
      corr_prop <- cor.test(corr_data_tp$scfa_propionate, 
                           corr_data_tp$fusicatenibacter_rel,
                           method = "spearman")
      cat(sprintf("  %s - Propionate: rho = %.3f, p = %.4f
", 
                  tp, corr_prop$estimate, corr_prop$p.value))
    }
  }
} else {
  cat("No Fusicatenibacter taxa found for correlation analysis
")
}
```

## 7.5 Responder vs Non-Responder Analysis

```{r responder-nonresponder, echo=FALSE, cache=TRUE, cache.path="cache/integrated-"}
# Define responder groups based on delta SCFA (48H - 0H)
# Calculate delta for butyrate and propionate
scfa_delta <- scfa_sample_data_long_factored %>%
  dplyr::filter(analyte %in% c("butyrate", "propionate")) %>%
  dplyr::filter(timepoint_hr %in% c("0", "48")) %>%
  dplyr::group_by(subject, group, carbohydrate_type, analyte) %>%
  dplyr::summarise(
    conc_0h = concentration[timepoint_hr == "0"][1],
    conc_48h = concentration[timepoint_hr == "48"][1],
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    delta_48h_0h = conc_48h - conc_0h,
    # Map to microbiome labels
    carbohydrate = dplyr::case_when(
      carbohydrate_type == "no_carbohydrate" ~ "No Added Carb",
      carbohydrate_type == "rapid_digestible" ~ "RDC",
      carbohydrate_type == "slow_digestible" ~ "SDC",
      TRUE ~ NA_character_
    ),
    group_micro = dplyr::case_when(
      group == "control" ~ "Healthy",
      group == "case" ~ "Obese",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::filter(!is.na(delta_48h_0h), !is.na(carbohydrate), !is.na(group_micro))

# Define responder/non-responder using median split for each analyte
# Separate by carbohydrate type for more relevant comparisons
responder_status <- scfa_delta %>%
  dplyr::group_by(analyte, carbohydrate) %>%
  dplyr::mutate(
    median_delta = median(delta_48h_0h, na.rm = TRUE),
    responder_status = dplyr::if_else(
      delta_48h_0h >= median_delta,
      "Responder",
      "Non-Responder"
    )
  ) %>%
  dplyr::ungroup() %>%
  dplyr::select(subject, group_micro, carbohydrate, analyte, responder_status)

# Pivot to wide format for overlap analysis
responder_wide <- responder_status %>%
  tidyr::pivot_wider(
    id_cols = c(subject, group_micro, carbohydrate),
    names_from = analyte,
    values_from = responder_status,
    names_prefix = "responder_"
  )

# Examine overlap in responder status
cat("Responder Status Overlap:
")
cat("- Total subjects:", nrow(responder_wide), "
")
if ("responder_butyrate" %in% colnames(responder_wide) && 
    "responder_propionate" %in% colnames(responder_wide)) {
  overlap_table <- table(
    Butyrate = responder_wide$responder_butyrate,
    Propionate = responder_wide$responder_propionate,
    useNA = "ifany"
  )
  print(overlap_table)
  
  # Calculate overlap statistics
  both_responder <- sum(responder_wide$responder_butyrate == "Responder" & 
                        responder_wide$responder_propionate == "Responder", 
                        na.rm = TRUE)
  both_non_responder <- sum(responder_wide$responder_butyrate == "Non-Responder" & 
                           responder_wide$responder_propionate == "Non-Responder", 
                           na.rm = TRUE)
  cat(sprintf("- Both Responder: %d (%.1f%%)
", both_responder, 
              100 * both_responder / nrow(responder_wide)))
  cat(sprintf("- Both Non-Responder: %d (%.1f%%)
", both_non_responder, 
              100 * both_non_responder / nrow(responder_wide)))
}

# Merge responder status with microbiome metadata at 48H
ps_48h_resp <- phyloseq::subset_samples(ps, timepoint == "48H")

# Add responder status to sample_data
# Extract numeric subject ID for matching
sample_meta_resp <- data.frame(sample_data(ps_48h_resp)) %>%
  tibble::rownames_to_column("SampleID") %>%
  dplyr::mutate(
    subject_id_numeric = stringr::str_extract(as.character(subject_id), "^[0-9]+")
  ) %>%
  dplyr::left_join(
    responder_wide,
    by = c("subject_id_numeric" = "subject", 
           "group" = "group_micro", 
           "carbohydrate" = "carbohydrate")
  )

# Update phyloseq sample_data
sample_data(ps_48h_resp)$responder_butyrate <- sample_meta_resp$responder_butyrate
sample_data(ps_48h_resp)$responder_propionate <- sample_meta_resp$responder_propionate

cat("
48H Microbiome Composition Comparison:
")
cat("- Samples with responder status:", 
    sum(!is.na(sample_data(ps_48h_resp)$responder_butyrate)), "
")
```

## 7.6 Microbial Composition Differences: Butyrate Responders vs Non-Responders

```{r butyrate-responder-composition, echo=FALSE, fig.width=14, fig.height=8}
# Compare microbial composition between butyrate responder groups at 48H
ps_48h_butyrate_resp <- phyloseq::subset_samples(
  ps_48h_resp,
  !is.na(responder_butyrate) & carbohydrate != "No Added Carb"
)

if (nsamples(ps_48h_butyrate_resp) > 5) {
  # Alpha diversity comparison
  alpha_butyrate_resp <- estimate_richness(
    ps_48h_butyrate_resp,
    measures = c("Observed", "Shannon", "Simpson")
  ) %>%
    tibble::rownames_to_column("SampleID") %>%
    dplyr::left_join(
      data.frame(sample_data(ps_48h_butyrate_resp)) %>%
        tibble::rownames_to_column("SampleID") %>%
        dplyr::select(SampleID, responder_butyrate, carbohydrate),
      by = "SampleID"
    )
  
  # Plot alpha diversity
  alpha_butyrate_resp_long <- alpha_butyrate_resp %>%
    tidyr::pivot_longer(
      cols = c(Observed, Shannon, Simpson),
      names_to = "metric",
      values_to = "value"
    )
  
  alpha_resp_plot <- ggplot(
    alpha_butyrate_resp_long,
    aes(x = responder_butyrate, y = value, fill = responder_butyrate)
  ) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.5) +
    facet_wrap(~ metric, scales = "free_y") +
    scale_fill_scfa() +
    labs(
      title = "Alpha Diversity: Butyrate Responders vs Non-Responders (48H)",
      subtitle = "Comparison across RDC and SDC conditions",
      x = "Butyrate Response Status",
      y = "Alpha Diversity",
      fill = "Response Status"
    ) +
    theme_scfa() +
    theme(legend.position = "none")
  
  print(alpha_resp_plot)
  
  # Beta diversity comparison
  ps_48h_butyrate_resp_rel <- microbiome::transform(
    ps_48h_butyrate_resp, 
    "compositional"
  )
  
  dist_bray_resp <- phyloseq::distance(
    ps_48h_butyrate_resp_rel, 
    method = "bray"
  )
  
  meta_resp <- data.frame(sample_data(ps_48h_butyrate_resp_rel))
  
  # PCoA
  pcoa_resp <- ape::pcoa(as.matrix(dist_bray_resp))
  pcoa_df_resp <- data.frame(
    PC1 = pcoa_resp$vectors[, 1],
    PC2 = pcoa_resp$vectors[, 2],
    responder_butyrate = meta_resp$responder_butyrate,
    carbohydrate = meta_resp$carbohydrate
  ) %>%
    dplyr::filter(!is.na(responder_butyrate))
  
  pcoa_plot_resp <- ggplot(
    pcoa_df_resp,
    aes(x = PC1, y = PC2, color = responder_butyrate)
  ) +
    geom_point(size = 2, alpha = 0.7) +
    stat_ellipse(aes(group = responder_butyrate), level = 0.68) +
    facet_wrap(~ carbohydrate) +
    scale_color_scfa() +
    labs(
      title = "Beta Diversity: Butyrate Responders vs Non-Responders (48H)",
      subtitle = "Bray-Curtis PCoA",
      x = paste0("PC1 (", round(100 * pcoa_resp$values$Relative_eig[1], 1), "%)"),
      y = paste0("PC2 (", round(100 * pcoa_resp$values$Relative_eig[2], 1), "%)"),
      color = "Response Status"
    ) +
    theme_scfa()
  
  print(pcoa_plot_resp)
  
  # PERMANOVA
  perm_resp <- vegan::adonis2(
    dist_bray_resp ~ responder_butyrate * carbohydrate,
    data = meta_resp,
    strata = meta_resp$subject_id,
    permutations = 999
  )
  
  cat("
PERMANOVA: Butyrate Responder Status (48H):
")
  print(perm_resp)
  
} else {
  cat("Insufficient data for butyrate responder composition comparison
")
}
```

## 7.7 Microbial Composition Differences: Propionate Responders vs Non-Responders

```{r propionate-responder-composition, echo=FALSE, fig.width=14, fig.height=8}
# Compare microbial composition between propionate responder groups at 48H
ps_48h_propionate_resp <- phyloseq::subset_samples(
  ps_48h_resp,
  !is.na(responder_propionate) & carbohydrate != "No Added Carb"
)

if (nsamples(ps_48h_propionate_resp) > 5) {
  # Alpha diversity comparison
  alpha_propionate_resp <- estimate_richness(
    ps_48h_propionate_resp,
    measures = c("Observed", "Shannon", "Simpson")
  ) %>%
    tibble::rownames_to_column("SampleID") %>%
    dplyr::left_join(
      data.frame(sample_data(ps_48h_propionate_resp)) %>%
        tibble::rownames_to_column("SampleID") %>%
        dplyr::select(SampleID, responder_propionate, carbohydrate),
      by = "SampleID"
    )
  
  alpha_propionate_resp_long <- alpha_propionate_resp %>%
    tidyr::pivot_longer(
      cols = c(Observed, Shannon, Simpson),
      names_to = "metric",
      values_to = "value"
    )
  
  alpha_prop_resp_plot <- ggplot(
    alpha_propionate_resp_long,
    aes(x = responder_propionate, y = value, fill = responder_propionate)
  ) +
    geom_boxplot(alpha = 0.7) +
    geom_jitter(width = 0.2, alpha = 0.5) +
    facet_wrap(~ metric, scales = "free_y") +
    scale_fill_scfa() +
    labs(
      title = "Alpha Diversity: Propionate Responders vs Non-Responders (48H)",
      subtitle = "Comparison across RDC and SDC conditions",
      x = "Propionate Response Status",
      y = "Alpha Diversity",
      fill = "Response Status"
    ) +
    theme_scfa() +
    theme(legend.position = "none")
  
  print(alpha_prop_resp_plot)
  
  # Beta diversity comparison
  ps_48h_propionate_resp_rel <- microbiome::transform(
    ps_48h_propionate_resp, 
    "compositional"
  )
  
  dist_bray_prop_resp <- phyloseq::distance(
    ps_48h_propionate_resp_rel, 
    method = "bray"
  )
  
  meta_prop_resp <- data.frame(sample_data(ps_48h_propionate_resp_rel))
  
  # PCoA
  pcoa_prop_resp <- ape::pcoa(as.matrix(dist_bray_prop_resp))
  pcoa_df_prop_resp <- data.frame(
    PC1 = pcoa_prop_resp$vectors[, 1],
    PC2 = pcoa_prop_resp$vectors[, 2],
    responder_propionate = meta_prop_resp$responder_propionate,
    carbohydrate = meta_prop_resp$carbohydrate
  ) %>%
    dplyr::filter(!is.na(responder_propionate))
  
  pcoa_plot_prop_resp <- ggplot(
    pcoa_df_prop_resp,
    aes(x = PC1, y = PC2, color = responder_propionate)
  ) +
    geom_point(size = 2, alpha = 0.7) +
    stat_ellipse(aes(group = responder_propionate), level = 0.68) +
    facet_wrap(~ carbohydrate) +
    scale_color_scfa() +
    labs(
      title = "Beta Diversity: Propionate Responders vs Non-Responders (48H)",
      subtitle = "Bray-Curtis PCoA",
      x = paste0("PC1 (", round(100 * pcoa_prop_resp$values$Relative_eig[1], 1), "%)"),
      y = paste0("PC2 (", round(100 * pcoa_prop_resp$values$Relative_eig[2], 1), "%)"),
      color = "Response Status"
    ) +
    theme_scfa()
  
  print(pcoa_plot_prop_resp)
  
  # PERMANOVA
  perm_prop_resp <- vegan::adonis2(
    dist_bray_prop_resp ~ responder_propionate * carbohydrate,
    data = meta_prop_resp,
    strata = meta_prop_resp$subject_id,
    permutations = 999
  )
  
  cat("
PERMANOVA: Propionate Responder Status (48H):
")
  print(perm_prop_resp)
  
} else {
  cat("Insufficient data for propionate responder composition comparison
")
}
```

# 8. Summary

```{r final-summary, echo=FALSE, results='asis'}
cat("## Key Integrated Findings

")

cat("### SCFA Metabolomics
")
cat("- Carbohydrate type effects were observed for multiple SCFAs, with significant differences in delta concentrations (48H - 0H) between carbohydrate conditions.
")
cat("- Mixed-effects models revealed significant interactions between group, carbohydrate type, and timepoint for key SCFAs.

")

cat("### Microbiome Composition
")
cat("- Alpha diversity showed differences between Healthy and Obese groups, with variations by carbohydrate type and timepoint.
")
cat("- Beta diversity analyses demonstrated significant effects of group, carbohydrate type, and timepoint on community composition.
")
cat("- Differential abundance analyses (ANCOM-BC2, MaAsLin2) identified multiple taxa associated with carbohydrate type, group, and timepoint effects.

")

cat("### Taxon–SCFA Relationships
")
if (exists("ancombc2_butyrate_tbl_rel") && nrow(ancombc2_butyrate_tbl_rel) > 0) {
  cat("- Multiple taxa were significantly associated with butyrate concentrations after adjusting for carbohydrate effects.
")
}
if (exists("ancombc2_propionate_tbl_rel") && nrow(ancombc2_propionate_tbl_rel) > 0) {
  cat("- Multiple taxa were significantly associated with propionate concentrations after adjusting for carbohydrate effects.
")
}
cat("- Fusicatenibacter showed correlations with both butyrate and propionate concentrations, with variation by timepoint.

")

cat("### Responder vs Non-Responder Analysis
")
if (exists("responder_wide") && nrow(responder_wide) > 0) {
  cat("- Responder groups were defined separately for butyrate and propionate based on median delta SCFA changes from 0H to 48H.
")
  if (exists("both_responder")) {
    cat(sprintf("- Overlap analysis revealed %d subjects (%.1f%%) were responders for both SCFAs.
", 
                both_responder, 100 * both_responder / nrow(responder_wide)))
  }
  cat("- Microbial composition differences were observed between responder and non-responder groups at 48H, indicating potential microbiome–SCFA interactions.
")
}

cat("
### Conclusions
")
cat("The integrated analysis demonstrates clear relationships between carbohydrate type, SCFA production, and microbiome composition. The identification of responder vs non-responder groups suggests individual variation in SCFA response that correlates with microbiome structure, highlighting the importance of personalized approaches to dietary interventions.
")
```


