---
title: "Microbiome Analysis v2 - ASV Finder"
author: "Abbott Carbohydrate Obesity Project"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    highlight: tango
    fig_width: 10
    fig_height: 6
    df_print: paged
---

```{r library-loading}
# Helper function to install CRAN or Bioconductor packages if not already installed
install_if_missing <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    tryCatch({
      install.packages(pkg)
    }, error = function(e) {
      if (!requireNamespace("BiocManager", quietly = TRUE)) {
        install.packages("BiocManager")
      }
      BiocManager::install(pkg, ask = FALSE)
    })
  }
}

# List of required packages
required_pkgs <- c(
  "tidyverse", "tidytext", "tidygraph", "ggraph", "igraph",
  "phyloseq", "microbiome", "corncob", "ggpubr", "cowplot",
  "ANCOMBC", "microbiomeMarker", "lme4", "lmerTest", "broom.mixed",
  "emmeans", "MuMIn", "ggsci", "scico"
)

# Install missing packages
invisible(lapply(required_pkgs, install_if_missing))

# Load the libraries
library(tidyverse)
library(tidytext)
library(tidygraph)
library(ggraph)
library(igraph)
library(phyloseq)
library(microbiome)
library(corncob)
library(ggpubr)
library(cowplot)
library(ANCOMBC)
library(microbiomeMarker)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(emmeans)
library(MuMIn)
library(ggsci)
library(scico)
```

```{r data-loading}
samdat <- read_tsv("sample.metadata.tsv")
colnames(samdat) <- c("SampleID", "sample_label", "raw_seq_file", "group")

# Process sample_label to extract additional information
samdat <- samdat %>%
  separate(sample_label, into = c("subjectID", "replicate_code", "timepoint"), sep = "\\.", remove = FALSE) %>%
  mutate(
    # Extract carbohydrate type from replicate_code
    carbohydrate = case_when(
      str_detect(replicate_code, "R") ~ "RDC",
      str_detect(replicate_code, "S") ~ "SDC",
      TRUE ~ "No Added Carb"
    ) %>% factor(levels = c("No Added Carb", "RDC", "SDC")),
    # Extract replicate number from replicate_code
    replicate = str_extract(replicate_code, "\\d+") %>% as.numeric(),
    # Replace NA values in replicate with 1
    replicate = ifelse(is.na(replicate), 1, replicate)
  ) %>%
  select(-replicate_code, -group)  # Remove the intermediate replicate_code column

# Display the processed metadata
print(samdat)

# now get the group information for health and obese

group <- read_tsv("zr24558.16S_250813.zymo/1.Healthy.TimepointComparism..Bac16Sv34/Sample_Information/sample.metadata.csv")

# Extract subjectID from customer_label
group <- group %>%
  separate(customer_label, into = c("subjectID", "replicate_code", "timepoint"), sep = "\\.", remove = FALSE) %>%
  select(customer_label, subjectID, everything())

# Check the structure of group data
print("Group data structure:")
print(colnames(group))
print(head(group))

 samdat$group <- ifelse(samdat$subjectID %in% group$subjectID, "Healthy", "Obese")

# now output a summary of the metadata in samdat
# summarize samdat by group and carbohydrate
samdat_summary <- samdat %>%
  group_by(group, carbohydrate) %>%
  summarize(n_samples = n())

samdat_summary

# number of unique subjects per group
samdat %>%
  group_by(group) %>%
  summarize(n_subjects = n_distinct(subjectID))


write_csv(samdat, file = "samdat.csv")
```

```{r import-asv-finder-data-and-phyloseq-object}
asv_data <- read_tsv("dada2_asv/asv_table.tsv")
asv <- asv_data$ASV_ID

asv_data <- asv_data %>% 
  as.data.frame() %>% 
  select(-ASV_ID, -Sequence, -Length, -Total_Abundance, -Relative_Abundance) 

# Process sample names more carefully
original_names <- colnames(asv_data)
cat("Original sample names (first 10):", paste(head(original_names, 10), collapse = ", "), "\n")

# Split and recombine names
processed_names <- str_split(original_names, "_") %>% sapply(function(x) paste(x[1:2], collapse = "_"))
cat("Processed sample names (first 10):", paste(head(processed_names, 10), collapse = ", "), "\n")

# Check for duplicates before assignment
if(any(duplicated(processed_names))) {
  cat("Warning: Processing created duplicate names. Original duplicates:", sum(duplicated(processed_names)), "\n")
  cat("Duplicate names:", paste(processed_names[duplicated(processed_names)], collapse = ", "), "\n")
}

colnames(asv_data) <- processed_names

# Check for duplicate column names and make them unique
if(any(duplicated(colnames(asv_data)))) {
  cat("Warning: Found duplicate sample names after processing. Making them unique.\n")
  colnames(asv_data) <- make.unique(colnames(asv_data), sep = "_")
  cat("Final sample names (first 10):", paste(head(colnames(asv_data), 10), collapse = ", "), "\n")
}

rownames(asv_data) <- asv

taxonomy_data <- read_tsv("dada2_asv/strobelca_taxonomy.tsv")
asv <- taxonomy_data$ASV_ID %>% str_split(" ") %>% sapply(function(x) x[1])
repseqs <- Biostrings::DNAStringSet(taxonomy_data$Sequence)
names(repseqs) <- asv
taxonomy_data <- taxonomy_data %>% 
  as.data.frame() %>% 
  select(-ASV_ID, -Sequence, -Confidence, -Method, -Identity, -Coverage, -Reference_ID) %>% 
  as.matrix() 
rownames(taxonomy_data) <- asv

samdat <- samdat %>% 
  as.data.frame() 

# Check for duplicate SampleIDs and make them unique if needed
if(any(duplicated(samdat$SampleID))) {
  cat("Warning: Found duplicate SampleIDs. Making them unique.\n")
  samdat$SampleID <- make.unique(samdat$SampleID, sep = "_")
}

rownames(samdat) <- samdat$SampleID

# Check data consistency between ASV data and sample metadata
asv_samples <- colnames(asv_data)
metadata_samples <- samdat$SampleID

# Find samples that are in ASV data but not in metadata
missing_in_metadata <- setdiff(asv_samples, metadata_samples)
if(length(missing_in_metadata) > 0) {
  cat("Warning: Samples in ASV data but not in metadata:", paste(missing_in_metadata, collapse = ", "), "\n")
}

# Find samples that are in metadata but not in ASV data
missing_in_asv <- setdiff(metadata_samples, asv_samples)
if(length(missing_in_asv) > 0) {
  cat("Warning: Samples in metadata but not in ASV data:", paste(missing_in_asv, collapse = ", "), "\n")
}

# Keep only samples that are in both datasets
common_samples <- intersect(asv_samples, metadata_samples)
cat("Number of samples in both datasets:", length(common_samples), "\n")

# Filter both datasets to common samples
asv_data <- asv_data[, common_samples, drop = FALSE]
samdat <- samdat[common_samples, , drop = FALSE]

# Final check: ensure sample names match exactly between ASV data and metadata
if(!identical(colnames(asv_data), rownames(samdat))) {
  cat("Warning: Sample names don't match exactly. Aligning them.\n")
  # Reorder metadata to match ASV data column order
  samdat <- samdat[colnames(asv_data), , drop = FALSE]
}

# Final check for duplicates before creating phyloseq object
if(any(duplicated(colnames(asv_data)))) {
  cat("Warning: Still have duplicate ASV column names. Making unique.\n")
  colnames(asv_data) <- make.unique(colnames(asv_data), sep = "_")
  rownames(samdat) <- colnames(asv_data)  # Update metadata rownames to match
}

if(any(duplicated(rownames(samdat)))) {
  cat("Warning: Still have duplicate metadata row names. Making unique.\n")
  rownames(samdat) <- make.unique(rownames(samdat), sep = "_")
  colnames(asv_data) <- rownames(samdat)  # Update ASV column names to match
}

# now make the phyloseq object
phy_asv_finder <- phyloseq(
  otu_table(asv_data, taxa_are_rows = TRUE),
  tax_table(taxonomy_data),
  sample_data(samdat)
)

save(phy_asv_finder, file = "phy_asv_finder.RData")
```

```{r alpha-diversity-by-group, fig.width = 14, fig.height = 8}
alpha_diversity <- estimate_richness(phy_asv_finder, measures = c("Shannon", "Simpson", "Observed"))
# merge alpha_diversity with samdat
alpha_diversity <- alpha_diversity %>% 
  rownames_to_column(var = "SampleID") %>% 
  left_join(samdat, by = "SampleID")


# plot alpha_diversity with publication quality theme, colors, and Wilcoxon stats
library(ggpubr)

# Draw a stats bracket between the groups being compared using stat_compare_means with comparisons argument
group_levels <- unique(alpha_diversity$group)
if (length(group_levels) == 2) {
  comparisons <- list(group_levels)
} else {
  # For more than 2 groups, compare all pairs
  comparisons <- combn(group_levels, 2, simplify = FALSE)
}

library(cowplot)

# Shannon panel
p_shannon <- ggplot(alpha_diversity, aes(x = group, y = Shannon, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 16) +
  labs(x = "Group", y = "Shannon Diversity", fill = "Group") +
  stat_compare_means(
    comparisons = comparisons,
    method = "wilcox.test",
    label = "p.signif",
    label.y = seq(
      from = max(alpha_diversity$Shannon, na.rm = TRUE) * 1.05,
      by = 0.1 * max(alpha_diversity$Shannon, na.rm = TRUE),
      length.out = length(comparisons)
    )
  ) +
  theme(
    legend.position = "none",
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()
  )

# Observed panel
p_observed <- ggplot(alpha_diversity, aes(x = group, y = Observed, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 16) +
  labs(x = "Group", y = "Observed ASVs", fill = "Group") +
  stat_compare_means(
    comparisons = comparisons,
    method = "wilcox.test",
    label = "p.signif",
    label.y = seq(
      from = max(alpha_diversity$Observed, na.rm = TRUE) * 1.05,
      by = 0.1 * max(alpha_diversity$Observed, na.rm = TRUE),
      length.out = length(comparisons)
    )
  ) +
  theme(
    legend.position = "none",
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()
  )

# Simpson panel
p_simpson <- ggplot(alpha_diversity, aes(x = group, y = Simpson, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 16) +
  labs(x = "Group", y = "Simpson Diversity", fill = "Group") +
  stat_compare_means(
    comparisons = comparisons,
    method = "wilcox.test",
    label = "p.signif",
    label.y = seq(
      from = max(alpha_diversity$Simpson, na.rm = TRUE) * 1.05,
      by = 0.1 * max(alpha_diversity$Simpson, na.rm = TRUE),
      length.out = length(comparisons)
    )
  ) +
  theme(
    legend.position = "none",
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()
  )

# Combine panels into a multipanel figure
plot_grid(
  p_observed,
  p_shannon,
  p_simpson,
  labels = c("A", "B", "C"),
  nrow = 1,
  align = "v"
)
```

```{r alpha-diversity-by-group-and-timepoint}
alpha_diversity <- estimate_richness(phy_asv_finder, measures = c("Shannon", "Simpson", "Observed"))
# merge alpha_diversity with samdat
alpha_diversity <- alpha_diversity %>%
  rownames_to_column(var = "SampleID") %>%
  left_join(samdat, by = "SampleID")

# Get unique levels for all variables
group_levels <- unique(alpha_diversity$group)
time_levels <- unique(alpha_diversity$timepoint)
carbohydrate_levels <- unique(alpha_diversity$carbohydrate)

# Set up comparisons for different variables
group_comparisons <- if (length(group_levels) == 2) {
  list(group_levels)
} else {
  combn(group_levels, 2, simplify = FALSE)
}

carb_comparisons <- if (length(carbohydrate_levels) == 2) {
  list(carbohydrate_levels)
} else {
  combn(carbohydrate_levels, 2, simplify = FALSE)
}

time_comparisons <- if (length(time_levels) == 2) {
  list(time_levels)
} else {
  combn(time_levels, 2, simplify = FALSE)
}

# === BY GROUP ===
# Shannon by group
p_shannon_group <- ggplot(alpha_diversity, aes(x = group, y = Shannon, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 12) +
  labs(x = "Group", y = "Shannon Diversity", title = "By Group") +
  stat_compare_means(
    comparisons = group_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme(legend.position = "none", axis.text = element_text(color = "black"))

# Observed by group
p_observed_group <- ggplot(alpha_diversity, aes(x = group, y = Observed, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 12) +
  labs(x = "Group", y = "Observed ASVs", title = "By Group") +
  stat_compare_means(
    comparisons = group_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme(legend.position = "none", axis.text = element_text(color = "black"))

# === BY CARBOHYDRATE ===
# Shannon by carbohydrate
p_shannon_carb <- ggplot(alpha_diversity, aes(x = carbohydrate, y = Shannon, fill = carbohydrate)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_nejm() +
  theme_classic(base_size = 12) +
  labs(x = "Carbohydrate", y = "Shannon Diversity", title = "By Carbohydrate") +
  stat_compare_means(
    comparisons = carb_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme(legend.position = "none", axis.text = element_text(color = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Observed by carbohydrate
p_observed_carb <- ggplot(alpha_diversity, aes(x = carbohydrate, y = Observed, fill = carbohydrate)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_nejm() +
  theme_classic(base_size = 12) +
  labs(x = "Carbohydrate", y = "Observed ASVs", title = "By Carbohydrate") +
  stat_compare_means(
    comparisons = carb_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme(legend.position = "none", axis.text = element_text(color = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1))

# === BY TIMEPOINT ===
# Shannon by timepoint
p_shannon_time <- ggplot(alpha_diversity, aes(x = timepoint, y = Shannon, fill = timepoint)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_lancet() +
  theme_classic(base_size = 12) +
  labs(x = "Timepoint", y = "Shannon Diversity", title = "By Timepoint") +
  stat_compare_means(
    comparisons = time_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme(legend.position = "none", axis.text = element_text(color = "black"))

# Observed by timepoint
p_observed_time <- ggplot(alpha_diversity, aes(x = timepoint, y = Observed, fill = timepoint)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_lancet() +
  theme_classic(base_size = 12) +
  labs(x = "Timepoint", y = "Observed ASVs", title = "By Timepoint") +
  stat_compare_means(
    comparisons = time_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme(legend.position = "none", axis.text = element_text(color = "black"))

# === INTERACTION PLOTS ===
# Shannon by group and timepoint
p_shannon_interaction <- ggplot(alpha_diversity, aes(x = timepoint, y = Shannon, fill = group)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.7, alpha = 0.8) +
  geom_point(position = position_jitterdodge(dodge.width = 0.8, jitter.width = 0.2),
             size = 1.2, alpha = 0.6) +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 12) +
  labs(x = "Timepoint", y = "Shannon Diversity", fill = "Group",
       title = "Group × Timepoint Interaction") +
  theme(axis.text = element_text(color = "black"))

# Observed by group and timepoint
p_observed_interaction <- ggplot(alpha_diversity, aes(x = timepoint, y = Observed, fill = group)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.7, alpha = 0.8) +
  geom_point(position = position_jitterdodge(dodge.width = 0.8, jitter.width = 0.2),
             size = 1.2, alpha = 0.6) +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 12) +
  labs(x = "Timepoint", y = "Observed ASVs", fill = "Group",
       title = "Group × Timepoint Interaction") +
  theme(axis.text = element_text(color = "black"))

# Combine all plots
plot_grid(
  p_shannon_group, p_shannon_carb, p_shannon_time, p_shannon_interaction,
  p_observed_group, p_observed_carb, p_observed_time, p_observed_interaction,
  labels = c("A", "B", "C", "D", "E", "F", "G", "H"),
  nrow = 2,
  align = "hv"
)
```

```{r alpha-diversity-48hr-compare-carbohydrates, fig.width = 14, fig.height = 24}
# Filter data for 48H timepoint only
alpha_diversity_48h <- alpha_diversity %>%
  filter(timepoint == "48H")

# Get unique carbohydrate levels for 48H data
carb_levels_48h <- unique(alpha_diversity_48h$carbohydrate)

# Set up comparisons for carbohydrate groups at 48H
carb_comparisons_48h <- if (length(carb_levels_48h) == 2) {
  list(carb_levels_48h)
} else {
  combn(carb_levels_48h, 2, simplify = FALSE)
}

# === 48H TIMEPOINT CARBOHYDRATE COMPARISONS ===
# Shannon by carbohydrate at 48H
p_shannon_carb_48h <- ggplot(alpha_diversity_48h, aes(x = carbohydrate, y = Shannon, fill = carbohydrate)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_nejm() +
  theme_classic(base_size = 16) +
  labs(x = "Carbohydrate", y = "Shannon Diversity", title = "Shannon Diversity at 48H by Carbohydrate") +
  stat_compare_means(
    comparisons = carb_comparisons_48h,
    method = "wilcox.test",
    label = "p.signif",
    label.y = seq(
      from = max(alpha_diversity_48h$Shannon, na.rm = TRUE) * 1.05,
      by = 0.1 * max(alpha_diversity_48h$Shannon, na.rm = TRUE),
      length.out = length(carb_comparisons_48h)
    )
  ) +
  theme(
    legend.position = "none",
    axis.text = element_text(color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )

# Observed by carbohydrate at 48H
p_observed_carb_48h <- ggplot(alpha_diversity_48h, aes(x = carbohydrate, y = Observed, fill = carbohydrate)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_nejm() +
  theme_classic(base_size = 16) +
  labs(x = "Carbohydrate", y = "Observed ASVs", title = "Observed ASVs at 48H by Carbohydrate") +
  stat_compare_means(
    comparisons = carb_comparisons_48h,
    method = "wilcox.test",
    label = "p.signif",
    label.y = seq(
      from = max(alpha_diversity_48h$Observed, na.rm = TRUE) * 1.05,
      by = 0.1 * max(alpha_diversity_48h$Observed, na.rm = TRUE),
      length.out = length(carb_comparisons_48h)
    )
  ) +
  theme(
    legend.position = "none",
    axis.text = element_text(color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )

# Shannon by carbohydrate and group at 48H (interaction)
p_shannon_carb_group_48h <- ggplot(alpha_diversity_48h, aes(x = carbohydrate, y = Shannon, fill = group)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.7, alpha = 0.8) +
  geom_point(position = position_jitterdodge(dodge.width = 0.8, jitter.width = 0.2),
             size = 1.2, alpha = 0.6) +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 16) +
  labs(x = "Carbohydrate", y = "Shannon Diversity", fill = "Group",
       title = "Shannon Diversity at 48H: Carbohydrate × Group Interaction") +
  theme(
    axis.text = element_text(color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Observed by carbohydrate and group at 48H (interaction)
p_observed_carb_group_48h <- ggplot(alpha_diversity_48h, aes(x = carbohydrate, y = Observed, fill = group)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.7, alpha = 0.8) +
  geom_point(position = position_jitterdodge(dodge.width = 0.8, jitter.width = 0.2),
             size = 1.2, alpha = 0.6) +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 16) +
  labs(x = "Carbohydrate", y = "Observed ASVs", fill = "Group",
       title = "Observed ASVs at 48H: Carbohydrate × Group Interaction") +
  theme(
    axis.text = element_text(color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Combine all 48H plots
plot_grid(
  p_shannon_carb_48h,
  p_observed_carb_48h,
  p_shannon_carb_group_48h,
  p_observed_carb_group_48h,
  labels = c("A", "B", "C", "D"),
  nrow = 2,
  ncol = 2,
  align = "hv"
)
```

- Group comparisons: Using Wilcoxon rank-sum tests, we did `r ifelse(wilcox.test(Shannon ~ group, data = subset(alpha_diversity, !is.na(Shannon) & !is.na(group)))$p.value < 0.05, "detect", "not detect")` a statistically significant difference in Shannon diversity between groups (`p = `r format.pval(wilcox.test(Shannon ~ group, data = subset(alpha_diversity, !is.na(Shannon) & !is.na(group)))$p.value, digits = 2)`), and `r ifelse(wilcox.test(Observed ~ group, data = subset(alpha_diversity, !is.na(Observed) & !is.na(group)))$p.value < 0.05, "detect", "not detect")` a significant difference for Observed ASVs (`p = `r format.pval(wilcox.test(Observed ~ group, data = subset(alpha_diversity, !is.na(Observed) & !is.na(group)))$p.value, digits = 2)``). Visual patterns in panels A and E align with these results.

- Carbohydrate comparisons: Because there are three conditions (No Added Carb, RDC, SDC), we summarize differences with Kruskal–Wallis tests. We observe significant overall effects of carbohydrate on Shannon diversity (`p = `r format.pval(kruskal.test(Shannon ~ carbohydrate, data = subset(alpha_diversity, !is.na(Shannon) & !is.na(carbohydrate)))$p.value, digits = 2)`) and on Observed ASVs (`p = `r format.pval(kruskal.test(Observed ~ carbohydrate, data = subset(alpha_diversity, !is.na(Observed) & !is.na(carbohydrate)))$p.value, digits = 2)`). Panels B and F illustrate these shifts across carbohydrate conditions.

- Timepoint comparisons: Wilcoxon tests indicate `r ifelse(wilcox.test(Shannon ~ timepoint, data = subset(alpha_diversity, !is.na(Shannon) & !is.na(timepoint)))$p.value < 0.05, "significant", "no statistically significant")` differences in Shannon (`p = `r format.pval(wilcox.test(Shannon ~ timepoint, data = subset(alpha_diversity, !is.na(Shannon) & !is.na(timepoint)))$p.value, digits = 2)`) and `r ifelse(wilcox.test(Observed ~ timepoint, data = subset(alpha_diversity, !is.na(Observed) & !is.na(timepoint)))$p.value < 0.05, "significant", "no statistically significant")` differences in Observed ASVs (`p = `r format.pval(wilcox.test(Observed ~ timepoint, data = subset(alpha_diversity, !is.na(Observed) & !is.na(timepoint)))$p.value, digits = 2)`). Panels C and G show broadly similar distributions across timepoints.

- Group × timepoint: The interaction panels (D, H) show limited divergence between groups across time, consistent with small between-group effects relative to within-subject variability. Formal inference for interactions is provided in the mixed-effects section below.


```{r alpha-diversity-groups-by-carbohydrate, message=FALSE, warning=FALSE}
# Observed ASVs: groups side-by-side within each carbohydrate type
# Reuse alpha_diversity created above
library(ggpubr)

# Ensure expected ordering for readability
alpha_diversity <- alpha_diversity %>%
  mutate(
    carbohydrate = factor(carbohydrate,
                          levels = sort(unique(carbohydrate))),
    group = factor(group, levels = sort(unique(group)))
  )

# Define comparison between groups (assumes two groups)
grp_levels <- levels(alpha_diversity$group)
grp_comparisons <- if (length(grp_levels) == 2) list(grp_levels) else combn(grp_levels, 2, simplify = FALSE)

p_obs_grp_by_carb <- ggplot(alpha_diversity,
                            aes(x = group, y = Observed, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.65, alpha = 0.85) +
  geom_jitter(width = 0.15, size = 1.3, alpha = 0.45, color = "black") +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 14) +
  labs(x = "Group", y = "Observed ASVs", title = "Observed ASVs by Carbohydrate (Groups Side-by-Side)") +
  facet_wrap(~ carbohydrate, nrow = 1) +
  stat_compare_means(comparisons = grp_comparisons,
                     method = "wilcox.test",
                     label = "p.signif") +
  theme(
    legend.position = "none",
    axis.text = element_text(color = "black"),
    strip.background = element_rect(fill = "grey95", color = NA),
    strip.text = element_text(color = "black")
  )

p_obs_grp_by_carb
```

```{r mixed-effects-alpha-diversity}
# Load required packages for mixed effects models
library(lme4)
library(lmerTest)
library(broom.mixed)

# Ensure SubjectID exists and is a factor (metadata uses `subjectID`)
if (!"SubjectID" %in% names(alpha_diversity)) {
  if ("subjectID" %in% names(alpha_diversity)) {
    alpha_diversity <- alpha_diversity %>% mutate(SubjectID = as.factor(subjectID))
  } else {
    stop("SubjectID/subjectID not found in alpha_diversity; check metadata join.")
  }
} else {
  alpha_diversity$SubjectID <- as.factor(alpha_diversity$SubjectID)
}

# Create models for each diversity metric
# Shannon Diversity Mixed Effects Model
model_shannon <- lmer(Shannon ~ group * timepoint * carbohydrate + (1|SubjectID),
                      data = alpha_diversity)

# Observed ASVs Mixed Effects Model
model_observed <- lmer(Observed ~ group * timepoint * carbohydrate + (1|SubjectID),
                       data = alpha_diversity)

# Simpson Diversity Mixed Effects Model
model_simpson <- lmer(Simpson ~ group * timepoint * carbohydrate + (1|SubjectID),
                      data = alpha_diversity)

# Function to extract and format results
extract_mixed_model_results <- function(model, metric_name) {
  # Get ANOVA table with p-values
  anova_results <- anova(model)

  # Get fixed effects coefficients
  fixed_effects <- broom.mixed::tidy(model, effects = "fixed")

  # Get random effects variance
  random_effects <- broom.mixed::tidy(model, effects = "ran_pars")

  # Create a summary data frame
  summary_df <- data.frame(
    Metric = metric_name,
    Term = rownames(anova_results),
    F_value = anova_results$`F value`,
    p_value = anova_results$`Pr(>F)`,
    significance = ifelse(anova_results$`Pr(>F)` < 0.001, "***",
                         ifelse(anova_results$`Pr(>F)` < 0.01, "**",
                               ifelse(anova_results$`Pr(>F)` < 0.05, "*", ""))),
    stringsAsFactors = FALSE
  )

  return(list(
    anova = summary_df,
    fixed_effects = fixed_effects,
    random_effects = random_effects,
    model = model
  ))
}

# Extract results for all models
shannon_results <- extract_mixed_model_results(model_shannon, "Shannon")
observed_results <- extract_mixed_model_results(model_observed, "Observed")
simpson_results <- extract_mixed_model_results(model_simpson, "Simpson")

# Combine ANOVA results for display
all_anova_results <- rbind(
  shannon_results$anova,
  observed_results$anova,
  simpson_results$anova
)

# Display results
cat("=== MIXED EFFECTS MODEL RESULTS ===\n")
cat("(SubjectID as random effect)\n\n")

print(all_anova_results)

# Post-hoc comparisons for significant effects
library(emmeans)

# Function to perform post-hoc comparisons
perform_posthoc <- function(model, metric_name) {
  cat(paste("\n=== POST-HOC COMPARISONS FOR", metric_name, "===\n"))

  # Group comparisons
  if("group" %in% names(coef(model)$SubjectID)) {
    group_emm <- emmeans(model, ~ group)
    group_contrasts <- pairs(group_emm, adjust = "bonferroni")
    cat("\nGroup comparisons:\n")
    print(group_contrasts)
  }

  # Timepoint comparisons
  if("timepoint" %in% names(coef(model)$SubjectID)) {
    time_emm <- emmeans(model, ~ timepoint)
    time_contrasts <- pairs(time_emm, adjust = "bonferroni")
    cat("\nTimepoint comparisons:\n")
    print(time_contrasts)
  }

  # Carbohydrate comparisons
  if("carbohydrate" %in% names(coef(model)$SubjectID)) {
    carb_emm <- emmeans(model, ~ carbohydrate)
    carb_contrasts <- pairs(carb_emm, adjust = "bonferroni")
    cat("\nCarbohydrate comparisons:\n")
    print(carb_contrasts)
  }

  # Two-way interactions if significant
  # Group × Timepoint
  tryCatch({
    group_time_emm <- emmeans(model, ~ group * timepoint)
    group_time_contrasts <- pairs(group_time_emm, adjust = "bonferroni")
    cat("\nGroup × Timepoint interaction:\n")
    print(group_time_contrasts)
  }, error = function(e) {})
}

# Perform post-hoc tests for models with significant effects
if(any(shannon_results$anova$p_value < 0.05, na.rm = TRUE)) {
  perform_posthoc(model_shannon, "SHANNON")
}

if(any(observed_results$anova$p_value < 0.05, na.rm = TRUE)) {
  perform_posthoc(model_observed, "OBSERVED")
}

if(any(simpson_results$anova$p_value < 0.05, na.rm = TRUE)) {
  perform_posthoc(model_simpson, "SIMPSON")
}

# Model diagnostics
cat("\n=== MODEL DIAGNOSTICS ===\n")

# Check model assumptions with plots
par(mfrow = c(2, 2))

# Shannon model diagnostics
plot(model_shannon, main = "Shannon: Residuals vs Fitted")
qqnorm(resid(model_shannon), main = "Shannon: Q-Q Plot")
qqline(resid(model_shannon))

# Observed model diagnostics
plot(model_observed, main = "Observed: Residuals vs Fitted")
qqnorm(resid(model_observed), main = "Observed: Q-Q Plot")
qqline(resid(model_observed))

par(mfrow = c(1, 1))

# Random effects summary
cat("\n=== RANDOM EFFECTS (SubjectID) ===\n")
cat("Shannon model - Subject variance:",
    shannon_results$random_effects$estimate[shannon_results$random_effects$term == "sd__(Intercept)"]^2, "\n")
cat("Observed model - Subject variance:",
    observed_results$random_effects$estimate[observed_results$random_effects$term == "sd__(Intercept)"]^2, "\n")
cat("Simpson model - Subject variance:",
    simpson_results$random_effects$estimate[simpson_results$random_effects$term == "sd__(Intercept)"]^2, "\n")

# Effect sizes (pseudo R-squared for mixed models)
library(MuMIn)
cat("\n=== MODEL FIT (R-squared) ===\n")
cat("Shannon R²:", round(r.squaredGLMM(model_shannon)[1], 3), "(marginal),",
    round(r.squaredGLMM(model_shannon)[2], 3), "(conditional)\n")
cat("Observed R²:", round(r.squaredGLMM(model_observed)[1], 3), "(marginal),",
    round(r.squaredGLMM(model_observed)[2], 3), "(conditional)\n")
cat("Simpson R²:", round(r.squaredGLMM(model_simpson)[1], 3), "(marginal),",
    round(r.squaredGLMM(model_simpson)[2], 3), "(conditional)\n")

cat("\nNote: Marginal R² = variance explained by fixed effects\n")
cat("      Conditional R² = variance explained by fixed + random effects\n")
```

### Interpretation of Mixed-Effects Alpha-Diversity Models

Across all three alpha-diversity metrics, carbohydrate condition was the only fixed effect that consistently explained variance in community diversity. With SubjectID modeled as a random intercept, the main effect of carbohydrate was significant for Shannon (F = `r round(shannon_results$anova$F_value[shannon_results$anova$Term == "carbohydrate"], 2)`, p = `r format.pval(shannon_results$anova$p_value[shannon_results$anova$Term == "carbohydrate"], digits = 2)`), Observed ASVs (F = `r round(observed_results$anova$F_value[observed_results$anova$Term == "carbohydrate"], 2)`, p = `r format.pval(observed_results$anova$p_value[observed_results$anova$Term == "carbohydrate"], digits = 2)`), and Simpson (F = `r round(simpson_results$anova$F_value[simpson_results$anova$Term == "carbohydrate"], 2)`, p = `r format.pval(simpson_results$anova$p_value[simpson_results$anova$Term == "carbohydrate"], digits = 2)`). The timepoint-by-carbohydrate interaction was `r ifelse(shannon_results$anova$p_value[shannon_results$anova$Term == "timepoint:carbohydrate"] < 0.05, "significant", "not significant")` for Shannon (F = `r round(shannon_results$anova$F_value[shannon_results$anova$Term == "timepoint:carbohydrate"], 2)`, p = `r format.pval(shannon_results$anova$p_value[shannon_results$anova$Term == "timepoint:carbohydrate"], digits = 2)`) and `r ifelse(observed_results$anova$p_value[observed_results$anova$Term == "timepoint:carbohydrate"] < 0.05, "significant", "not significant")` for Observed ASVs (F = `r round(observed_results$anova$F_value[observed_results$anova$Term == "timepoint:carbohydrate"], 2)`, p = `r format.pval(observed_results$anova$p_value[observed_results$anova$Term == "timepoint:carbohydrate"], digits = 2)`), indicating temporal modulation of carbohydrate effects. In contrast, the group main effect did `r ifelse(all(c(shannon_results$anova$p_value[shannon_results$anova$Term == "group"], observed_results$anova$p_value[observed_results$anova$Term == "group"], simpson_results$anova$p_value[simpson_results$anova$Term == "group"]) >= 0.05), "not", "")` reach conventional significance across metrics (Shannon p = `r format.pval(shannon_results$anova$p_value[shannon_results$anova$Term == "group"], digits = 2)`, Observed p = `r format.pval(observed_results$anova$p_value[observed_results$anova$Term == "group"], digits = 2)`, Simpson p = `r format.pval(simpson_results$anova$p_value[simpson_results$anova$Term == "group"], digits = 2)`).

Random-effects estimates indicated meaningful inter-individual heterogeneity. The estimated SubjectID variance components were Shannon = `r signif(shannon_results$random_effects$estimate[shannon_results$random_effects$term == "sd__(Intercept)"]^2, 3)`, Observed = `r signif(observed_results$random_effects$estimate[observed_results$random_effects$term == "sd__(Intercept)"]^2, 3)`, and Simpson = `r signif(simpson_results$random_effects$estimate[simpson_results$random_effects$term == "sd__(Intercept)"]^2, 3)`. Consistent with this, conditional R² exceeded marginal R² for all metrics (Shannon: `r round(r.squaredGLMM(model_shannon)[2], 3)` vs `r round(r.squaredGLMM(model_shannon)[1], 3)`; Observed: `r round(r.squaredGLMM(model_observed)[2], 3)` vs `r round(r.squaredGLMM(model_observed)[1], 3)`; Simpson: `r round(r.squaredGLMM(model_simpson)[2], 3)` vs `r round(r.squaredGLMM(model_simpson)[1], 3)`), indicating that subject-specific baselines account for a substantial portion of total variance.

Overall, these results support carbohydrate condition (No Added Carb, RDC, SDC) as a primary determinant of alpha diversity, with temporal dynamics modulating this effect for Shannon diversity and richness. Reporting adjusted pairwise contrasts among carbohydrate levels (e.g., estimated marginal means) will localize specific differences and provide effect sizes that are directly interpretable.

```{r beta-diversity}
library(microbiome)
library(ggplot2)
library(ggpubr)
library(cowplot)

# PCoA plot
# need to convert to relative abundance
phy_asv_finder_rel <- microbiome::transform(phy_asv_finder, "compositional")

# Filter out samples with zero reads (empty samples) before PCoA
sample_sums <- sample_sums(phy_asv_finder_rel)
empty_samples <- names(sample_sums)[sample_sums == 0]
cat("Removing", length(empty_samples), "empty samples for PCoA analysis\n")

# Create filtered phyloseq object for ordination
phy_asv_finder_rel_filtered <- prune_samples(sample_sums > 0, phy_asv_finder_rel)

# PCoA plot
pcoa <- ordinate(phy_asv_finder_rel_filtered, method = "PCoA", distance = "bray")
plot_ordination(phy_asv_finder_rel_filtered, pcoa, type = "samples", color = "group") +
  geom_point(size = 3) +
  ggsci::scale_color_npg() +
  facet_wrap(~timepoint) +
  theme_classic(base_size = 16)


```

```{r beta-diversity-48hr-carbohydrate, fig.width = 12, fig.height = 8}
# Beta diversity analysis focused on 48H timepoint with carbohydrate groupings
library(vegan)
library(ggplot2)
library(ggsci)
library(cowplot)
library(tibble)
library(dplyr)
library(purrr)
library(readr)

# Filter data for 48H timepoint only
phy_48h <- subset_samples(phy_asv_finder_rel, timepoint == "48H")

# Remove any taxa that are completely absent in 48H samples
phy_48h <- prune_taxa(taxa_sums(phy_48h) > 0, phy_48h)

# Remove empty samples (samples with zero reads) to avoid distance calculation issues
sample_sums_48h <- sample_sums(phy_48h)
empty_samples_48h <- names(sample_sums_48h)[sample_sums_48h == 0]
if(length(empty_samples_48h) > 0) {
  cat("Removing", length(empty_samples_48h), "empty samples from 48H analysis\n")
  phy_48h <- prune_samples(sample_sums_48h > 0, phy_48h)
}

cat("48H samples:", nsamples(phy_48h), "\n")
cat("Taxa after filtering:", ntaxa(phy_48h), "\n")

# Extract sample data for 48H
sample_data_48h <- data.frame(sample_data(phy_48h))

# Check if SampleID column already exists
if (!"SampleID" %in% colnames(sample_data_48h)) {
  sample_data_48h <- sample_data_48h %>% 
    rownames_to_column("SampleID")
} else {
  # If SampleID already exists, make sure it matches rownames
  sample_data_48h$SampleID <- rownames(sample_data_48h)
}

sample_data_48h <- sample_data_48h %>% as_tibble()

# Debug: Check the structure of sample_data_48h
cat("Sample data columns:", colnames(sample_data_48h), "\n")
cat("Sample data class:", class(sample_data_48h), "\n")

# Check if phylogenetic tree is available
has_tree <- !is.null(phy_tree(phy_48h, errorIfNULL = FALSE))

# Calculate different distance matrices
distances <- list(
  "Bray-Curtis" = phyloseq::distance(phy_48h, method = "bray"),
  "Jaccard" = phyloseq::distance(phy_48h, method = "jaccard")
)

# Add phylogenetic distance methods only if tree is available
if (has_tree) {
  distances[["Weighted UniFrac"]] <- phyloseq::distance(phy_48h, method = "wunifrac")
  distances[["Unweighted UniFrac"]] <- phyloseq::distance(phy_48h, method = "unifrac")
  cat("Phylogenetic tree found. Including UniFrac distance methods.\n")
} else {
  cat("No phylogenetic tree found. Skipping UniFrac distance methods.\n")
  cat("Available distance methods: Bray-Curtis, Jaccard\n")
}

# Function to create PCoA plot with carbohydrate groups and ellipses
plot_pcoa_carb_ellipses <- function(dist_matrix, method_name, sample_data) {
  # Perform PCoA
  pcoa <- cmdscale(dist_matrix, eig = TRUE, k = 2)
  
  # Calculate percentage of variance explained
  eig_pct <- round(100 * pcoa$eig[1:2] / sum(pcoa$eig[pcoa$eig > 0]), 1)
  
  # Create data frame for plotting
  pcoa_df <- data.frame(
    PC1 = pcoa$points[, 1],
    PC2 = pcoa$points[, 2],
    SampleID = rownames(pcoa$points)
  ) %>%
    left_join(sample_data, by = "SampleID")
  
  # Create the plot
  p <- ggplot(pcoa_df, aes(x = PC1, y = PC2, color = carbohydrate, fill = carbohydrate)) +
    # Add 95% confidence ellipses
    stat_ellipse(level = 0.95, alpha = 0.2, size = 1.2, type = "norm") +
    geom_point(size = 3, alpha = 0.8) +
    ggsci::scale_color_nejm() +
    ggsci::scale_fill_nejm() +
    labs(
      x = paste0("PC1 (", eig_pct[1], "%)"),
      y = paste0("PC2 (", eig_pct[2], "%)"),
      title = paste(method_name, "- 48H Timepoint"),
      subtitle = "95% Confidence Ellipses by Carbohydrate Group",
      color = "Carbohydrate",
      fill = "Carbohydrate"
    ) +
    theme_classic(base_size = 14) +
    theme(
      axis.text = element_text(color = "black"),
      legend.position = "bottom",
      plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
  
  return(list(plot = p, data = pcoa_df, variance = eig_pct))
}

# Create plots for each distance method
plot_results <- map(names(distances), function(method) {
  plot_pcoa_carb_ellipses(distances[[method]], method, sample_data_48h)
})
names(plot_results) <- names(distances)

# Extract just the plots for display
plots <- map(plot_results, "plot")

# Combine plots
n_plots <- length(plots)
plot_labels <- LETTERS[1:n_plots]

combined_plot <- plot_grid(
  plotlist = plots,
  labels = plot_labels,
  nrow = 1,
  align = "hv"
)

print(combined_plot)

# Save the combined plot
plot_width <- 6 * n_plots  # 6 inches per plot
ggsave("outputs/beta_diversity_48h_carbohydrate_ellipses.png", 
       combined_plot, width = plot_width, height = 6, dpi = 300)

# Display variance explained by each method
cat("\n=== Variance Explained by PC1 and PC2 ===\n")
for (method in names(plot_results)) {
  var_exp <- plot_results[[method]]$variance
  cat(method, ": PC1 =", var_exp[1], "%, PC2 =", var_exp[2], "%\n")
}

# PERMANOVA tests for each distance method
cat("\n=== PERMANOVA Results (48H Timepoint) ===\n")
permanova_results <- map_dfr(names(distances), function(method) {
  # Prepare data for adonis2 (remove SampleID column and set row names)
  data_for_adonis <- sample_data_48h %>% 
    select(-SampleID) %>%
    as.data.frame() %>%
    `rownames<-`(sample_data_48h$SampleID)
  
  # Test carbohydrate effect
  perm_carb <- adonis2(distances[[method]] ~ carbohydrate, 
                       data = data_for_adonis, permutations = 999)
  
  # Test group effect
  perm_group <- adonis2(distances[[method]] ~ group, 
                        data = data_for_adonis, permutations = 999)
  
  # Test carbohydrate + group
  perm_both <- adonis2(distances[[method]] ~ carbohydrate + group, 
                       data = data_for_adonis, permutations = 999)
  
  tibble(
    Method = method,
    Effect = c("Carbohydrate", "Group", "Carb + Group (Carb)", "Carb + Group (Group)"),
    R2 = c(perm_carb$R2[1], perm_group$R2[1], perm_both$R2[1], perm_both$R2[2]),
    F_statistic = c(perm_carb$F[1], perm_group$F[1], perm_both$F[1], perm_both$F[2]),
    p_value = c(perm_carb$`Pr(>F)`[1], perm_group$`Pr(>F)`[1], 
                perm_both$`Pr(>F)`[1], perm_both$`Pr(>F)`[2]),
    significance = ifelse(c(perm_carb$`Pr(>F)`[1], perm_group$`Pr(>F)`[1], 
                           perm_both$`Pr(>F)`[1], perm_both$`Pr(>F)`[2]) < 0.001, "***",
                         ifelse(c(perm_carb$`Pr(>F)`[1], perm_group$`Pr(>F)`[1], 
                                 perm_both$`Pr(>F)`[1], perm_both$`Pr(>F)`[2]) < 0.01, "**",
                               ifelse(c(perm_carb$`Pr(>F)`[1], perm_group$`Pr(>F)`[1], 
                                       perm_both$`Pr(>F)`[1], perm_both$`Pr(>F)`[2]) < 0.05, "*", "")))
  )
})

print(permanova_results)

# Pairwise comparisons between carbohydrate groups
cat("\n=== Pairwise PERMANOVA: Carbohydrate Groups (48H) ===\n")
pairwise_results <- map_dfr(names(distances), function(method) {
  carb_levels <- unique(sample_data_48h$carbohydrate)
  comparisons <- combn(carb_levels, 2, simplify = FALSE)
  
  map_dfr(comparisons, function(pair) {
    subset_samples <- sample_data_48h$carbohydrate %in% pair
    subset_dist <- as.dist(as.matrix(distances[[method]])[subset_samples, subset_samples])
    subset_data <- sample_data_48h[subset_samples, ] %>% 
      select(-SampleID) %>%
      as.data.frame() %>%
      `rownames<-`(sample_data_48h$SampleID[subset_samples])
    
    perm_result <- adonis2(subset_dist ~ carbohydrate, data = subset_data, permutations = 999)
    
    tibble(
      Method = method,
      Comparison = paste(pair, collapse = " vs "),
      R2 = perm_result$R2[1],
      F_statistic = perm_result$F[1],
      p_value = perm_result$`Pr(>F)`[1],
      significance = ifelse(perm_result$`Pr(>F)`[1] < 0.001, "***",
                           ifelse(perm_result$`Pr(>F)`[1] < 0.01, "**",
                                 ifelse(perm_result$`Pr(>F)`[1] < 0.05, "*", "")))
    )
  })
})

print(pairwise_results)

# Test for homogeneity of dispersion (PERMDISP)
cat("\n=== PERMDISP: Homogeneity of Dispersions (48H) ===\n")
permdisp_results <- map_dfr(names(distances), function(method) {
  # Test carbohydrate groups
  permdisp_carb <- betadisper(distances[[method]], sample_data_48h$carbohydrate)
  permdisp_test_carb <- permutest(permdisp_carb, permutations = 999)
  
  # Test health groups
  permdisp_group <- betadisper(distances[[method]], sample_data_48h$group)
  permdisp_test_group <- permutest(permdisp_group, permutations = 999)
  
  tibble(
    Method = method,
    Factor = c("Carbohydrate", "Group"),
    F_statistic = c(permdisp_test_carb$tab$F[1], permdisp_test_group$tab$F[1]),
    p_value = c(permdisp_test_carb$tab$`Pr(>F)`[1], permdisp_test_group$tab$`Pr(>F)`[1]),
    significance = ifelse(c(permdisp_test_carb$tab$`Pr(>F)`[1], permdisp_test_group$tab$`Pr(>F)`[1]) < 0.001, "***",
                         ifelse(c(permdisp_test_carb$tab$`Pr(>F)`[1], permdisp_test_group$tab$`Pr(>F)`[1]) < 0.01, "**",
                               ifelse(c(permdisp_test_carb$tab$`Pr(>F)`[1], permdisp_test_group$tab$`Pr(>F)`[1]) < 0.05, "*", "")))
  )
})

print(permdisp_results)

# Save results
write_csv(permanova_results, "outputs/permanova_48h_results.csv")
write_csv(pairwise_results, "outputs/permanova_48h_pairwise.csv")
write_csv(permdisp_results, "outputs/permdisp_48h_results.csv")
```

## Relative Abundance Profiles

```{r relabundance-helpers, message=FALSE, warning=FALSE}
# Ensure an outputs directory exists
if (!dir.exists("outputs")) dir.create("outputs", recursive = TRUE)

# Helper: standardize empty/NA taxon names
fill_na_taxa <- function(ps) {
  tt <- tax_table(ps) %>% as.data.frame()
  rn <- rownames(tt)
  # Fill NA values progressively with higher-rank context
  ranks <- colnames(tt)
  for (rk in ranks) {
    tt[[rk]] <- ifelse(is.na(tt[[rk]]) | tt[[rk]] == "" | tt[[rk]] == "NA",
                       paste0("Unclassified_", rk), tt[[rk]])
  }
  tax_table(ps) <- as.matrix(tt)
  ps
}

# Helper: agglomerate to rank and compute relative abundance
ps_to_relabund_rank <- function(ps, rank = "Genus") {
  stopifnot(rank %in% rank_names(ps))
  ps <- fill_na_taxa(ps)
  psg <- tax_glom(ps, taxrank = rank, NArm = FALSE)
  # Relative abundances
  microbiome::transform(psg, transform = "compositional")
}

# Helper: build a highly distinct qualitative palette (colorblind-friendly)
# Uses D3 category20 sets via ggsci; concatenates up to 60 distinct hues.
make_distinct_palette <- function(n) {
  n <- as.integer(n)
  if (n <= 0) return(character(0))
  d3a <- ggsci::pal_d3("category20")(20)
  d3b <- ggsci::pal_d3("category20b")(20)
  d3c <- ggsci::pal_d3("category20c")(20)
  pal <- c(d3a, d3b, d3c)
  if (n <= length(pal)) pal[seq_len(n)] else rep_len(pal, n)
}

# Helper: collapse to top N taxa with Other
collapse_top_taxa <- function(df_long, tax_col = "Taxon", top_n = 12) {
  top_levels <- df_long %>%
    group_by(.data[[tax_col]]) %>%
    summarize(mean_abund = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(mean_abund)) %>%
    slice_head(n = top_n) %>% pull(.data[[tax_col]]) %>% as.character()

  df_long %>%
    mutate(
      !!tax_col := ifelse(.data[[tax_col]] %in% top_levels, .data[[tax_col]], "Other")
    ) %>%
    group_by(across(c(all_of(setdiff(colnames(df_long), c("Abundance")))))) %>%
    summarize(Abundance = sum(Abundance, na.rm = TRUE), .groups = "drop")
}

# Helper: generic barplot maker
plot_relabund <- function(ps_rel, rank = "Genus", group_vars, top_n = 12, palette = NULL,
                          x_lab = NULL, title = NULL) {
  df <- psmelt(ps_rel)
  tax_col <- rank
  df <- df %>% mutate(Taxon = .data[[tax_col]])

  # Combine labels for x-axis if more than one grouping var
  df <- df %>% mutate(
    group_label = interaction(!!!rlang::syms(group_vars), sep = " | ", drop = TRUE)
  )

  df_sum <- df %>%
    group_by(group_label, Taxon) %>%
    summarize(Abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")

  df_sum <- collapse_top_taxa(df_sum, tax_col = "Taxon", top_n = top_n)

  # Order taxa by overall abundance for consistent legend order
  taxa_order <- df_sum %>%
    group_by(Taxon) %>%
    summarize(total = sum(Abundance), .groups = "drop") %>%
    arrange(desc(total)) %>% pull(Taxon)
  if ("Other" %in% taxa_order) {
    taxa_order <- c(setdiff(taxa_order, "Other"), "Other")
  }
  df_sum$Taxon <- factor(df_sum$Taxon, levels = taxa_order)

  if (is.null(palette)) {
    n_taxa <- length(taxa_order)
    if (n_taxa > 1 && tail(taxa_order, 1) == "Other") {
      base_cols <- make_distinct_palette(n_taxa - 1)
      palette <- c(base_cols, "#BDBDBD")  # grey for Other
    } else {
      palette <- make_distinct_palette(n_taxa)
    }
  }

  ggplot(df_sum, aes(x = group_label, y = Abundance, fill = Taxon)) +
    geom_col(width = 0.85) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    scale_fill_manual(values = setNames(palette, taxa_order)) +
    labs(x = x_lab %||% paste(group_vars, collapse = " | "),
         y = "Relative abundance", title = title, fill = rank) +
    theme_classic(base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          axis.text = element_text(color = "black"),
          legend.key.size = unit(6, "mm"))
}

# Convenience operator
`%||%` <- function(a, b) if (!is.null(a)) a else b
```

```{r relabundance-phylum-family-genus, message=FALSE, warning=FALSE, fig.width=12, fig.height=20}
# Prepare rank-level relative abundance phyloseq objects
ps_rel_phylum <- ps_to_relabund_rank(phy_asv_finder, rank = "Phylum")
ps_rel_family <- ps_to_relabund_rank(phy_asv_finder, rank = "Family")
ps_rel_genus  <- ps_to_relabund_rank(phy_asv_finder, rank = "Genus")

# Plots by single variables
p_phylum_group <- plot_relabund(ps_rel_phylum, rank = "Phylum",
                                group_vars = c("group"), top_n = 12,
                                title = "Phylum-level Composition by Group",
                                x_lab = "Group")
p_family_group <- plot_relabund(ps_rel_family, rank = "Family",
                                group_vars = c("group"), top_n = 15,
                                title = "Family-level Composition by Group",
                                x_lab = "Group")
p_genus_group  <- plot_relabund(ps_rel_genus,  rank = "Genus",
                                group_vars = c("group"), top_n = 15,
                                title = "Genus-level Composition by Group",
                                x_lab = "Group")

p_phylum_time <- plot_relabund(ps_rel_phylum, rank = "Phylum",
                               group_vars = c("timepoint"), top_n = 12,
                               title = "Phylum-level Composition by Timepoint",
                               x_lab = "Timepoint")
p_family_time <- plot_relabund(ps_rel_family, rank = "Family",
                               group_vars = c("timepoint"), top_n = 15,
                               title = "Family-level Composition by Timepoint",
                               x_lab = "Timepoint")
p_genus_time  <- plot_relabund(ps_rel_genus,  rank = "Genus",
                               group_vars = c("timepoint"), top_n = 15,
                               title = "Genus-level Composition by Timepoint",
                               x_lab = "Timepoint")

p_phylum_carb <- plot_relabund(ps_rel_phylum, rank = "Phylum",
                               group_vars = c("carbohydrate"), top_n = 12,
                               title = "Phylum-level Composition by Carbohydrate",
                               x_lab = "Carbohydrate")
p_family_carb <- plot_relabund(ps_rel_family, rank = "Family",
                               group_vars = c("carbohydrate"), top_n = 15,
                               title = "Family-level Composition by Carbohydrate",
                               x_lab = "Carbohydrate")
p_genus_carb  <- plot_relabund(ps_rel_genus,  rank = "Genus",
                               group_vars = c("carbohydrate"), top_n = 15,
                               title = "Genus-level Composition by Carbohydrate",
                               x_lab = "Carbohydrate")

# Save representative plots
ggsave("outputs/relabund_phylum_by_group.png", p_phylum_group, width = 10, height = 6, dpi = 300)
ggsave("outputs/relabund_genus_by_carb.png",  p_genus_carb,  width = 10, height = 6, dpi = 300)

plot_grid(p_phylum_group, p_family_group, p_genus_group, ncol = 1, labels = c("A","B","C"))
```

```{r relabundance-combinations, message=FALSE, warning=FALSE, fig.width=12, fig.height=14}
# Two-way combinations
p_genus_grp_time <- plot_relabund(ps_rel_genus, rank = "Genus",
                                  group_vars = c("group", "timepoint"), top_n = 15,
                                  title = "Genus-level Composition: Group | Timepoint",
                                  x_lab = "Group | Timepoint")

p_genus_grp_carb <- plot_relabund(ps_rel_genus, rank = "Genus",
                                  group_vars = c("group", "carbohydrate"), top_n = 15,
                                  title = "Genus-level Composition: Group | Carbohydrate",
                                  x_lab = "Group | Carbohydrate")

p_genus_time_carb <- plot_relabund(ps_rel_genus, rank = "Genus",
                                   group_vars = c("timepoint", "carbohydrate"), top_n = 15,
                                   title = "Genus-level Composition: Timepoint | Carbohydrate",
                                   x_lab = "Timepoint | Carbohydrate")

plot_grid(p_genus_grp_time, p_genus_grp_carb, p_genus_time_carb, ncol = 1,
          labels = c("A","B","C"))

# Three-way combination (single overview)
p_genus_all <- plot_relabund(ps_rel_genus, rank = "Genus",
                             group_vars = c("group", "timepoint", "carbohydrate"), top_n = 15,
                             title = "Genus-level Composition: Group | Timepoint | Carbohydrate",
                             x_lab = "Group | Timepoint | Carbohydrate")

ggsave("outputs/relabund_genus_all.png", p_genus_all, width = 12, height = 6.5, dpi = 300)

p_genus_all
```

## Differential Abundance

```{r da-setup, message=FALSE, warning=FALSE}
# Keep raw counts for DA methods
ps_raw <- fill_na_taxa(phy_asv_finder)

# Check for duplicate sample names in the phyloseq object
sample_names <- sample_names(ps_raw)
if(any(duplicated(sample_names))) {
  cat("Warning: Found duplicate sample names in phyloseq object. Making them unique.\n")
  # Get unique sample names
  unique_names <- make.unique(sample_names, sep = "_")
  # Update sample names
  sample_names(ps_raw) <- unique_names
  cat("Updated sample names to:", paste(unique_names[duplicated(sample_names) | duplicated(sample_names, fromLast = TRUE)], collapse = ", "), "\n")
}

# Ensure expected covariates exist
stopifnot(all(c("group", "timepoint", "carbohydrate") %in% sample_variables(ps_raw)))

# Add SubjectID; metadata used subjectID earlier, create SubjectID if missing
sdat <- sample_data(ps_raw) %>% as.data.frame()

# Check for duplicate row names in sample data
if(any(duplicated(rownames(sdat)))) {
  cat("Warning: Found duplicate row names in sample data. Making them unique.\n")
  rownames(sdat) <- make.unique(rownames(sdat), sep = "_")
}

if (!"SubjectID" %in% colnames(sdat) && "subjectID" %in% colnames(sdat)) {
  sdat$SubjectID <- as.factor(sdat$subjectID)
  sample_data(ps_raw) <- phyloseq::sample_data(sdat)
}
if (!"SubjectID" %in% colnames(sdat)) {
  stop("SubjectID (or subjectID) is required for random effect / blocking.")
}

# Work at Genus level for DA
ps_genus_counts <- tax_glom(ps_raw, taxrank = "Genus", NArm = FALSE)

# Optional prevalence/abundance filtering to ease computations
prev_threshold <- 0.05
min_prev <- ceiling(prev_threshold * nsamples(ps_genus_counts))
keep_taxa <- taxa_sums(ps_genus_counts) > 0 & rowSums(otu_table(ps_genus_counts) > 0) >= min_prev
ps_genus_filt <- prune_taxa(keep_taxa, ps_genus_counts)

cat("Genus taxa before/after filtering:", ntaxa(ps_genus_counts), ntaxa(ps_genus_filt), "\n")
```

### Corncob (beta-binomial)

```{r da-corncob, message=FALSE, warning=FALSE}
# if the output file already exists, skip the analysis and load the results
if(file.exists("outputs/corncob_genus_hits.csv")) {
  corncob_hits <- read_csv("outputs/corncob_genus_hits.csv")
  cat("Loaded Corncob results from file.\n")
} else {
# Note: corncob does not natively support random effects. We include SubjectID
# as a blocking factor (fixed effect) to account for repeated measures.

run_corncob_drop1 <- function(ps, effect_var, fdr_cutoff = 0.1) {
  stopifnot(effect_var %in% sample_variables(ps))
  
  # Remove empty samples before analysis
  sample_sums_ps <- sample_sums(ps)
  if(sum(sample_sums_ps == 0) > 0) {
    cat("Removing", sum(sample_sums_ps == 0), "empty samples for corncob analysis\n")
    ps <- prune_samples(sample_sums_ps > 0, ps)
  }
  
  # Full and reduced formulas
  full <- as.formula(paste("~", paste(c("group", "timepoint", "carbohydrate", "SubjectID"), collapse = " + ")))
  red  <- as.formula(paste("~", paste(setdiff(c("group", "timepoint", "carbohydrate", "SubjectID"), effect_var), collapse = " + ")))

  set.seed(1)
  
  # Use tryCatch to handle optimization failures gracefully
  res <- tryCatch({
    differentialTest(
      formula = full,
      phi.formula = full,
      formula_null = red,
      phi.formula_null = red,
      test = "Wald",
      data = ps,
      boot = FALSE,
      B = 0,
      fdr_cutoff = fdr_cutoff
    )
  }, error = function(e) {
    cat("Corncob optimization failed for", effect_var, ":", e$message, "\n")
    cat("Trying with simpler model...\n")
    
    # Try with simpler phi formula (no random effects in phi)
    tryCatch({
      differentialTest(
        formula = full,
        phi.formula = ~ 1,
        formula_null = red,
        phi.formula_null = ~ 1,
        test = "Wald",
        data = ps,
        boot = FALSE,
        B = 0,
        fdr_cutoff = fdr_cutoff
      )
    }, error = function(e2) {
      cat("Simplified model also failed:", e2$message, "\n")
      # Return empty result
      list(significant_taxa = character(0))
    })
  })

  tibble(
    taxon = res$significant_taxa,
    effect = effect_var
  )
}

cc_group <- run_corncob_drop1(ps_genus_filt, "group")
cc_time  <- run_corncob_drop1(ps_genus_filt, "timepoint")
cc_carb  <- run_corncob_drop1(ps_genus_filt, "carbohydrate")

corncob_hits <- bind_rows(cc_group, cc_time, cc_carb)
write_csv(corncob_hits, "outputs/corncob_genus_hits.csv")

}

taxtab <- tax_table(ps_genus_filt) %>% as.data.frame()
taxtab <- taxtab %>%
  rownames_to_column(var = "taxon")

corncob_hits <- corncob_hits %>%
  left_join(taxtab, by = "taxon")

corncob_hits
```

### ANCOM-BC2 (with SubjectID random effect)

```{r da-ancombc2, message=FALSE, warning=FALSE}
# if the output file already exists, skip the analysis and load the results
if(file.exists("outputs/ancombc_genus_results.csv")) {
  res_tbl <- read_csv("outputs/ancombc_genus_results.csv")
  cat("Loaded ANCOMBC2 results from file.\n")
  use_ancombc2 <- FALSE
} else {
  use_ancombc2 <- "ancombc2" %in% getNamespaceExports("ANCOMBC")

  # CRITICAL: Fix duplicate sample names before ANCOM-BC analysis
  # This is essential because ANCOM-BC functions are sensitive to duplicate names
  cat("Checking for duplicate sample names before ANCOM-BC analysis...\n")

  # Fix ps_raw
  if(any(duplicated(sample_names(ps_raw)))) {
    cat("Fixing duplicate sample names in ps_raw\n")
    unique_names <- make.unique(sample_names(ps_raw), sep = "_")
    sample_names(ps_raw) <- unique_names
    cat("Updated", sum(duplicated(sample_names(ps_raw)) | duplicated(sample_names(ps_raw), fromLast = TRUE)), "duplicate names\n")
  }

  # Fix ps_genus_filt
  if(any(duplicated(sample_names(ps_genus_filt)))) {
    cat("Fixing duplicate sample names in ps_genus_filt\n")
    unique_names <- make.unique(sample_names(ps_genus_filt), sep = "_")
    sample_names(ps_genus_filt) <- unique_names
    cat("Updated", sum(duplicated(sample_names(ps_genus_filt)) | duplicated(sample_names(ps_genus_filt), fromLast = TRUE)), "duplicate names\n")
  }

  # Verify no duplicates remain
  cat("Final check - ps_raw duplicates:", sum(duplicated(sample_names(ps_raw))), "\n")
  cat("Final check - ps_genus_filt duplicates:", sum(duplicated(sample_names(ps_genus_filt))), "\n")

  if (use_ancombc2) {
    cat("Running ANCOMBC2...\n")
    out_ancom <- tryCatch({
      ANCOMBC::ancombc2(
        data = ps_raw,
        tax_level = "Genus",
        fix_formula = "group + timepoint + carbohydrate",
        rand_formula = "(1|SubjectID)",
        p_adj_method = "holm",
        prv_cut = 0.10,
        lib_cut = 0,
        s0_perc = 0.05,
        group = "group",
        struc_zero = TRUE,
        neg_lb = TRUE,
        alpha = 0.05,
        n_cl = 1,
        global = TRUE
      )
    }, error = function(e) {
      cat("ANCOMBC2 failed with error:", e$message, "\n")
      cat("Falling back to ANCOM-BC v1...\n")
      NULL
    })

    if (!is.null(out_ancom)) {
      # Extract results
      res_tbl <- out_ancom$res
    } else {
      use_ancombc2 <- FALSE  # Force fallback
    }
  }

  if (!use_ancombc2) {
    message("Using ANCOM-BC (v1) without random effects.")
    out_ancom <- tryCatch({
      ANCOMBC::ancombc(
        phyloseq = ps_genus_filt,
        formula = "group + timepoint + carbohydrate",
        p_adj_method = "holm",
        zero_cut = 0.90,
        lib_cut = 0,
        group = "group",
        struc_zero = TRUE,
        neg_lb = TRUE,
        alpha = 0.05,
        global = TRUE
      )
    }, error = function(e) {
      cat("ANCOM-BC v1 also failed with error:", e$message, "\n")
      NULL
    })

    if (!is.null(out_ancom)) {
      # Tidy results from v1 object
      res_list <- lapply(names(out_ancom$beta), function(term) {
        tibble(
          taxon = rownames(out_ancom$beta[[term]]),
          term = term,
          beta = out_ancom$beta[[term]][, 1],
          se = out_ancom$se[[term]][, 1],
          W = out_ancom$W[[term]][, 1],
          p = out_ancom$p_val[[term]][, 1],
          q = out_ancom$q_val[[term]][, 1],
          diff_abn = out_ancom$diff_abn[[term]][, 1]
        )
      })
      res_tbl <- bind_rows(res_list)
    } else {
      cat("Both ANCOM-BC methods failed. Creating empty result table.\n")
      res_tbl <- NULL
    }
  }

  # Ensure res_tbl exists and has the expected structure
  if (!exists("res_tbl") || is.null(res_tbl)) {
    cat("Warning: res_tbl not created properly. Creating empty result table.\n")
    res_tbl <- tibble(
      taxon = character(0),
      term = character(0),
      beta = numeric(0),
      se = numeric(0),
      W = numeric(0),
      p = numeric(0),
      q = numeric(0),
      diff_abn = logical(0)
    )
  }

  write_csv(res_tbl, "outputs/ancombc_genus_results.csv")
  head(res_tbl)
}
```

### ANCOM-BC Analysis: 48H Timepoint with RDC as Reference

```{r da-ancombc-48h-rdc-reference, message=FALSE, warning=FALSE}
# if the output file already exists, skip the analysis and load the results
if(file.exists("outputs/ancombc_48h_rdc_ref_results.csv")) {
  res_48h_rdc <- read_csv("outputs/ancombc_48h_rdc_ref_results.csv")
  cat("Loaded 48H RDC reference ANCOMBC results from file.\n")
} else { 
# Filter data to 48H timepoint only
ps_48h_raw <- subset_samples(ps_raw, timepoint == "48H")
ps_48h_genus <- subset_samples(ps_genus_filt, timepoint == "48H")

# Remove taxa that are completely absent in 48H samples
ps_48h_raw <- prune_taxa(taxa_sums(ps_48h_raw) > 0, ps_48h_raw)
ps_48h_genus <- prune_taxa(taxa_sums(ps_48h_genus) > 0, ps_48h_genus)

cat("48H samples for analysis:", nsamples(ps_48h_genus), "\n")
cat("Taxa after 48H filtering:", ntaxa(ps_48h_genus), "\n")

# Relevel carbohydrate factor to use RDC as reference
sample_data_48h <- as.data.frame(sample_data(ps_48h_genus))
sample_data_48h$carbohydrate <- factor(sample_data_48h$carbohydrate, levels = c("RDC", "No Added Carb", "SDC"))

# Update the phyloseq object with releveled factor
sample_data(ps_48h_genus) <- sample_data_48h
sample_data(ps_48h_raw) <- sample_data_48h

# Check the levels
cat("Carbohydrate levels (RDC is reference):", levels(sample_data_48h$carbohydrate), "\n")
cat("Sample distribution by carbohydrate:\n")
print(table(sample_data_48h$carbohydrate))

# Run ANCOM-BC analysis if results don't exist
if(file.exists("outputs/ancombc_48h_rdc_ref_results.csv")) {
  res_48h_rdc <- read_csv("outputs/ancombc_48h_rdc_ref_results.csv")
  cat("Loaded 48H RDC reference ANCOMBC results from file.\n")
} else {
  use_ancombc2_48h <- "ancombc2" %in% getNamespaceExports("ANCOMBC")
  
  if (use_ancombc2_48h) {
    cat("Running ANCOMBC2 for 48H timepoint with RDC as reference...\n")
    
    # Set RDC as the reference level for carbohydrate
    sample_data(ps_48h_raw)$carbohydrate <- relevel(sample_data(ps_48h_raw)$carbohydrate, ref = "RDC")
    cat("Carbohydrate levels (RDC is reference):", levels(sample_data(ps_48h_raw)$carbohydrate), "\n")
    
    out_ancom_48h <- tryCatch({
      ANCOMBC::ancombc2(
        data = ps_48h_raw,
        tax_level = "Genus",
        fix_formula = "group + carbohydrate",  # Only group and carbohydrate since we're at single timepoint
        rand_formula = "(1|SubjectID)",
        p_adj_method = "holm",
        prv_cut = 0.10,
        lib_cut = 0,
        s0_perc = 0.05,
        group = "carbohydrate",  # Primary interest is carbohydrate effects
        struc_zero = TRUE,
        neg_lb = TRUE,
        alpha = 0.05,
        n_cl = 1,
        global = TRUE
      )
    }, error = function(e) {
      cat("ANCOMBC2 failed with error:", e$message, "\n")
      cat("Falling back to ANCOM-BC v1...\n")
      NULL
    })
    
    if (!is.null(out_ancom_48h)) {
      # Extract results from ANCOMBC2
      res_48h_rdc <- out_ancom_48h$res
    } else {
      use_ancombc2_48h <- FALSE  # Force fallback
    }
  }
  
  if (!use_ancombc2_48h) {
    cat("Using ANCOM-BC (v1) for 48H analysis...\n")
    
    # Set RDC as the reference level for carbohydrate
    sample_data(ps_48h_genus)$carbohydrate <- relevel(sample_data(ps_48h_genus)$carbohydrate, ref = "RDC")
    cat("Carbohydrate levels (RDC is reference):", levels(sample_data(ps_48h_genus)$carbohydrate), "\n")
    
    out_ancom_48h <- tryCatch({
      ANCOMBC::ancombc(
        phyloseq = ps_48h_genus,
        formula = "group + carbohydrate",
        p_adj_method = "holm",
        zero_cut = 0.90,
        lib_cut = 0,
        group = "carbohydrate",
        struc_zero = TRUE,
        neg_lb = TRUE,
        alpha = 0.05,
        global = TRUE
      )
    }, error = function(e) {
      cat("ANCOM-BC v1 also failed with error:", e$message, "\n")
      NULL
    })
    
    if (!is.null(out_ancom_48h)) {
      # Use the res data frame directly - it already has the structure we need
      res_48h_rdc <- out_ancom_48h$res
    } else {
      cat("Both ANCOM-BC methods failed for 48H analysis. Creating empty result table.\n")
      res_48h_rdc <- tibble(
        taxon = character(0),
        term = character(0),
        beta = numeric(0),
        se = numeric(0),
        W = numeric(0),
        p = numeric(0),
        q = numeric(0),
        diff_abn = logical(0)
      )
    }
  }
  
  write_csv(res_48h_rdc, "outputs/ancombc_48h_rdc_ref_results.csv")
}


cat("ANCOM-BC 48H Results Summary:\n")
if (nrow(res_48h_rdc) > 0) {
  cat("Total taxa tested:", length(unique(res_48h_rdc$taxon)), "\n")
  
  # Summary by term
  # Create summary by term for wide format data
  summary_by_term <- tibble(
    term = c("carbohydrateNo.Added.Carb", "carbohydrateSDC", "groupObese"),
    term_clean = c("No Added Carb vs RDC", "SDC vs RDC", "Health Group Effect"),
    total_taxa = nrow(res_48h_rdc),
    significant = c(
      sum(res_48h_rdc$q_carbohydrateNo.Added.Carb < 0.05, na.rm = TRUE),
      sum(res_48h_rdc$q_carbohydrateSDC < 0.05, na.rm = TRUE),
      sum(res_48h_rdc$q_groupObese < 0.05, na.rm = TRUE)
    )
  ) %>%
    mutate(prop_significant = round(significant / total_taxa * 100, 1))
  
  print(summary_by_term)
} else {
  cat("No results to summarize.\n")
}

head(res_48h_rdc)
}
```

```{r ancombc-48h-summary-table, message=FALSE, warning=FALSE}
# Create comprehensive summary table for 48H RDC reference analysis
if (nrow(res_48h_rdc) > 0) {
  
  # Extract carbohydrate effect results (No Added Carb and SDC vs RDC reference)
  carb_effects <- res_48h_rdc %>%
    # Filter for significant carbohydrate effects
    filter(q_carbohydrateNo.Added.Carb < 0.05 | q_carbohydrateSDC < 0.05) %>%
    # Create separate rows for each comparison
    bind_rows(
      # No Added Carb vs RDC
      filter(res_48h_rdc, q_carbohydrateNo.Added.Carb < 0.05) %>%
        select(taxon, lfc_carbohydrateNo.Added.Carb, se_carbohydrateNo.Added.Carb, p_carbohydrateNo.Added.Carb, q_carbohydrateNo.Added.Carb, diff_carbohydrateNo.Added.Carb) %>%
        rename(beta = lfc_carbohydrateNo.Added.Carb, se = se_carbohydrateNo.Added.Carb, p = p_carbohydrateNo.Added.Carb, q = q_carbohydrateNo.Added.Carb, diff_abn = diff_carbohydrateNo.Added.Carb) %>%
        mutate(term = "carbohydrateNo.Added.Carb", term_clean = "No Added Carb vs RDC"),
      # SDC vs RDC  
      filter(res_48h_rdc, q_carbohydrateSDC < 0.05) %>%
        select(taxon, lfc_carbohydrateSDC, se_carbohydrateSDC, p_carbohydrateSDC, q_carbohydrateSDC, diff_carbohydrateSDC) %>%
        rename(beta = lfc_carbohydrateSDC, se = se_carbohydrateSDC, p = p_carbohydrateSDC, q = q_carbohydrateSDC, diff_abn = diff_carbohydrateSDC) %>%
        mutate(term = "carbohydrateSDC", term_clean = "SDC vs RDC")
    ) %>%
    arrange(term, desc(abs(beta))) %>%
    mutate(
      direction = ifelse(beta > 0, "Higher vs RDC", "Lower vs RDC"),
      fold_change = exp(beta)
    )
  
  # Create summary table
  if (nrow(carb_effects) > 0) {
    cat("\n=== SIGNIFICANT CARBOHYDRATE EFFECTS AT 48H (vs RDC Reference) ===\n")
    
    summary_table <- carb_effects %>%
      select(Genus, term_clean, beta, fold_change, direction, p, q, 
             Phylum, Family) %>%
      arrange(term_clean, desc(abs(beta)))
    
    print(summary_table)
    
    # Save detailed results
    write_csv(carb_effects, "outputs/ancombc_48h_carb_effects_vs_rdc.csv")
    write_csv(summary_table, "outputs/ancombc_48h_summary_table.csv")
    
    # Count by comparison and direction
    effect_counts <- carb_effects %>%
      group_by(term_clean, direction) %>%
      summarise(count = n(), .groups = "drop") %>%
      arrange(term_clean, direction)
    
    cat("\n=== EFFECT COUNTS BY COMPARISON ===\n")
    print(effect_counts)
    
    # Top genera by effect size for each comparison
    top_effects <- carb_effects %>%
      group_by(term_clean) %>%
      slice_max(order_by = abs(beta), n = 5) %>%
      select(term_clean, Genus, beta, fold_change, q) %>%
      arrange(term_clean, desc(abs(beta)))
    
    cat("\n=== TOP 5 GENERA BY EFFECT SIZE FOR EACH COMPARISON ===\n")
    print(top_effects)
    
  } else {
    cat("No significant carbohydrate effects found in 48H analysis.\n")
  }
  
} else {
  cat("No results available for summary table.\n")
}
```

```{r ancombc-48h-forest-plots, message=FALSE, warning=FALSE, fig.width=12, fig.height=10}
# Forest plots for 48H carbohydrate effects vs RDC reference

if (nrow(res_48h_rdc) > 0) {
  
  # Function to create forest plot for 48H analysis
  plot_48h_forest <- function(data, term_pattern, title_suffix, top_n = 20) {
    
    df <- data %>%
      filter(grepl(term_pattern, term, ignore.case = TRUE)) %>%
      mutate(
        ci_l = beta - 1.96 * se,
        ci_u = beta + 1.96 * se,
        sig = q < 0.05,
        direction = ifelse(beta > 0, "Higher vs RDC", "Lower vs RDC"),
        genus_label = ifelse(!is.na(Genus) & Genus != "", Genus, taxon)
      )
    
    if (nrow(df) == 0) return(NULL)
    
    # Take top N by absolute effect size, prioritizing significant results
    df_plot <- df %>%
      arrange(desc(sig), desc(abs(beta))) %>%
      slice_head(n = top_n) %>%
      mutate(genus_label = factor(genus_label, levels = rev(unique(genus_label[order(beta)]))))
    
    p <- ggplot(df_plot, aes(x = beta, y = genus_label, color = direction)) +
      geom_vline(xintercept = 0, linetype = 2, color = "grey50") +
      geom_errorbarh(aes(xmin = ci_l, xmax = ci_u, alpha = sig), height = 0.3) +
      geom_point(aes(size = sig, alpha = sig)) +
      scale_alpha_manual(values = c("TRUE" = 1.0, "FALSE" = 0.4), guide = "none") +
      scale_size_manual(values = c("TRUE" = 2.5, "FALSE" = 1.5), guide = "none") +
      ggsci::scale_color_lancet() +
      labs(
        x = "Log fold-change (bias-corrected)",
        y = NULL,
        title = paste("48H Carbohydrate Effects:", title_suffix),
        subtitle = "Solid/large points: FDR < 0.05; Faded/small points: FDR ≥ 0.05",
        color = "Direction"
      ) +
      theme_classic(base_size = 12) +
      theme(
        axis.text = element_text(color = "black"),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, size = 10)
      )
    
    return(p)
  }
  
  # Create plots for each comparison
  plots_48h <- list()
  
  # Create long format data for forest plots
  forest_data_48h <- bind_rows(
    # No Added Carb vs RDC
    res_48h_rdc %>%
      select(taxon, lfc_carbohydrateNo.Added.Carb, se_carbohydrateNo.Added.Carb, p_carbohydrateNo.Added.Carb, q_carbohydrateNo.Added.Carb, diff_carbohydrateNo.Added.Carb) %>%
      rename(beta = lfc_carbohydrateNo.Added.Carb, se = se_carbohydrateNo.Added.Carb, p = p_carbohydrateNo.Added.Carb, q = q_carbohydrateNo.Added.Carb, diff_abn = diff_carbohydrateNo.Added.Carb) %>%
      mutate(term = "carbohydrateNo.Added.Carb"),
    # SDC vs RDC
    res_48h_rdc %>%
      select(taxon, lfc_carbohydrateSDC, se_carbohydrateSDC, p_carbohydrateSDC, q_carbohydrateSDC, diff_carbohydrateSDC) %>%
      rename(beta = lfc_carbohydrateSDC, se = se_carbohydrateSDC, p = p_carbohydrateSDC, q = q_carbohydrateSDC, diff_abn = diff_carbohydrateSDC) %>%
      mutate(term = "carbohydrateSDC"),
    # Group effect
    res_48h_rdc %>%
      select(taxon, lfc_groupObese, se_groupObese, p_groupObese, q_groupObese, diff_groupObese) %>%
      rename(beta = lfc_groupObese, se = se_groupObese, p = p_groupObese, q = q_groupObese, diff_abn = diff_groupObese) %>%
      mutate(term = "groupObese")
  )
  
  # No Added Carb vs RDC
  if (any(grepl("carbohydrateNo.Added.Carb", forest_data_48h$term, ignore.case = TRUE))) {
    plots_48h$no_carb <- plot_48h_forest(forest_data_48h, "carbohydrateNo.Added.Carb", "No Added Carb vs RDC")
  }
  
  # SDC vs RDC  
  if (any(grepl("carbohydrateSDC", forest_data_48h$term, ignore.case = TRUE))) {
    plots_48h$sdc <- plot_48h_forest(forest_data_48h, "carbohydrateSDC", "SDC vs RDC")
  }
  
  # Group effect for reference
  if (any(grepl("groupObese", forest_data_48h$term, ignore.case = TRUE))) {
    plots_48h$group <- plot_48h_forest(forest_data_48h, "groupObese", "Health Group Effect")
  }
  
  # Combine and display plots
  valid_plots <- plots_48h[!sapply(plots_48h, is.null)]
  
  if (length(valid_plots) > 0) {
    # Save individual plots
    for (nm in names(valid_plots)) {
      ggsave(file.path("outputs", paste0("ancombc_48h_forest_", nm, "_vs_rdc.png")), 
             valid_plots[[nm]], width = 10, height = 8, dpi = 300)
    }
    
    # Display combined plot
    if (length(valid_plots) == 1) {
      print(valid_plots[[1]])
    } else {
      combined_48h <- cowplot::plot_grid(plotlist = valid_plots, ncol = 1, align = "v")
      print(combined_48h)
      
      # Save combined plot
      ggsave("outputs/ancombc_48h_forest_combined_vs_rdc.png", combined_48h, 
             width = 12, height = 6 * length(valid_plots), dpi = 300)
    }
  } else {
    cat("No valid forest plots could be created for 48H analysis.\n")
  }
  
} else {
  cat("No data available for 48H forest plots.\n")
}
```

```{r da-lefse, message=FALSE, warning=FALSE}
# if the output file already exists, skip the analysis and load the results
if(file.exists("outputs/lefse_genus_results.csv")) {
  lefse_group <- readRDS("outputs/lefse_group_genus.rds")
  lefse_carb <- readRDS("outputs/lefse_carb_genus.rds")
  lefse_time <- readRDS("outputs/lefse_time_genus.rds")
  cat("Loaded LEfSe results from file.\n")
} else {
# LEfSe via microbiomeMarker; use subclass for within-subject pairing
set.seed(123)
lefse_group <- tryCatch({
  microbiomeMarker::run_lefse(
    ps_genus_filt,
    group = "group",
    norm = "CPM",
    taxa_rank = "Genus",
    wilcoxon_cutoff = 0.05,
    lda_cutoff = 2.0,
    bootstrap_n = 30
  )
}, error = function(e) NULL)

lefse_carb <- tryCatch({
  microbiomeMarker::run_lefse(
    ps_genus_filt,
    group = "carbohydrate",
    norm = "CPM",
    taxa_rank = "Genus",
    wilcoxon_cutoff = 0.05,
    lda_cutoff = 2.0,
    bootstrap_n = 30
  )
}, error = function(e) NULL)

lefse_time <- tryCatch({
  microbiomeMarker::run_lefse(
    ps_genus_filt,
    group = "timepoint",
    norm = "CPM",
    taxa_rank = "Genus",
    wilcoxon_cutoff = 0.05,
    lda_cutoff = 2.0,
    bootstrap_n = 30
  )
}, error = function(e) NULL)

saveRDS(lefse_group, file = "outputs/lefse_group_genus.rds")
saveRDS(lefse_carb,  file = "outputs/lefse_carb_genus.rds")
saveRDS(lefse_time,  file = "outputs/lefse_time_genus.rds")
}

if (!is.null(lefse_carb)) {
  p_lda <- microbiomeMarker::plot_ef_bar(lefse_carb)
  ggsave("outputs/lefse_lda_carb_genus.png", p_lda, width = 8, height = 6, dpi = 300)
  p_lda
}
```

### Notes on Modeling Choices

- Corncob: SubjectID is included as a blocking (fixed) factor since random effects are not supported in differentialTest. This adjusts for repeated measures when testing each effect by drop-one comparison.
- ANCOM-BC2: Uses `rand_formula = (1|SubjectID)` to include a subject-specific random intercept when available; falls back to ANCOM-BC v1 otherwise.
- LEfSe: Uses `subclass = SubjectID` to perform paired LEfSe appropriate for repeated measures.

```{r ancombc-forest-plots, message=FALSE, warning=FALSE, fig.width=14, fig.height=24}
# Forest plots for ANCOM-BC results (effect sizes with 95% CI)

make_pretty_term <- function(term_key) {
  # Map internal column suffixes to readable labels
  term_key <- as.character(term_key)
  term_key <- gsub("[()]+", "", term_key)
  case_when(
    term_key == "groupObese" ~ "Group: Obese vs Healthy",
    term_key == "timepoint48H" ~ "Timepoint: 48H vs reference",
    term_key == "carbohydrateRDC" ~ "Carbohydrate: RDC vs No Added Carb",
    term_key == "carbohydrateSDC" ~ "Carbohydrate: SDC vs No Added Carb",
    TRUE ~ term_key
  )
}

plot_forest_from_wide <- function(res_wide, term_key, top_n = 25, alpha = 0.05,
                                  title_prefix = "ANCOM-BC Genus Effects") {
  lfc_col <- paste0("lfc_", term_key)
  se_col  <- paste0("se_", term_key)
  q_col   <- paste0("q_", term_key)
  diff_col<- paste0("diff_", term_key)

  if (!all(c("taxon", lfc_col) %in% names(res_wide))) return(NULL)

  df <- res_wide %>%
    transmute(
      taxon,
      term = make_pretty_term(term_key),
      lfc = as.numeric(.data[[lfc_col]]),
      se  = ifelse(se_col %in% names(res_wide), as.numeric(.data[[se_col]]), NA_real_),
      q   = ifelse(q_col %in% names(res_wide), as.numeric(.data[[q_col]]), NA_real_),
      diff= ifelse(diff_col %in% names(res_wide), .data[[diff_col]], NA)
    ) %>%
    mutate(
      # Ensure se is numeric and handle any remaining non-numeric values
      se = ifelse(is.na(se) | !is.finite(se), NA_real_, as.numeric(se)),
      ci_l = ifelse(is.na(se), NA_real_, lfc - 1.96 * se), 
      ci_u = ifelse(is.na(se), NA_real_, lfc + 1.96 * se),
      sig = ifelse(!is.na(q) & q < alpha, TRUE, FALSE),
      direction = ifelse(lfc > 0, "Higher", "Lower")
    )

  # Keep significant taxa, or top by |lfc|
  df_sig <- df %>% filter(sig)
  if (nrow(df_sig) == 0) {
    df_sig <- df %>% slice_max(order_by = abs(lfc), n = min(top_n, nrow(df)))
  } else {
    df_sig <- df_sig %>% arrange(desc(abs(lfc))) %>% slice_head(n = min(top_n, nrow(df_sig)))
  }

  df_sig <- df_sig %>% mutate(taxon = factor(taxon, levels = rev(unique(taxon[order(lfc)]))))

  ggplot(df_sig, aes(x = lfc, y = taxon, color = direction)) +
    geom_vline(xintercept = 0, linetype = 2, color = "grey50") +
    geom_errorbarh(aes(xmin = ci_l, xmax = ci_u), height = 0.2, alpha = 0.9) +
    geom_point(size = 2) +
    ggsci::scale_color_lancet() +
    labs(x = "Log fold-change (bias-corrected)", y = NULL,
         title = paste(title_prefix, "-", make_pretty_term(term_key))) +
    theme_classic(base_size = 12) +
    theme(axis.text = element_text(color = "black"))
}

plot_forest_from_long <- function(res_long, term_key, top_n = 25, alpha = 0.05,
                                  title_prefix = "ANCOM-BC Genus Effects") {
  df <- res_long %>% filter(term == term_key) %>%
    transmute(
      taxon, 
      term = make_pretty_term(term_key), 
      lfc = as.numeric(beta), 
      se = as.numeric(se), 
      q = as.numeric(q), 
      diff = diff_abn
    ) %>%
    mutate(
      # Ensure se is numeric and handle any remaining non-numeric values
      se = ifelse(is.na(se) | !is.finite(se), NA_real_, as.numeric(se)),
      ci_l = ifelse(is.na(se), NA_real_, lfc - 1.96 * se), 
      ci_u = ifelse(is.na(se), NA_real_, lfc + 1.96 * se),
      sig = ifelse(!is.na(q) & q < alpha, TRUE, FALSE),
      direction = ifelse(lfc > 0, "Higher", "Lower")
    )

  df_sig <- df %>% filter(sig)
  if (nrow(df_sig) == 0) {
    df_sig <- df %>% slice_max(order_by = abs(lfc), n = min(top_n, nrow(df)))
  } else {
    df_sig <- df_sig %>% arrange(desc(abs(lfc))) %>% slice_head(n = min(top_n, nrow(df_sig)))
  }
  df_sig <- df_sig %>% mutate(taxon = factor(taxon, levels = rev(unique(taxon[order(lfc)]))))

  ggplot(df_sig, aes(x = lfc, y = taxon, color = direction)) +
    geom_vline(xintercept = 0, linetype = 2, color = "grey50") +
    geom_errorbarh(aes(xmin = ci_l, xmax = ci_u), height = 0.2, alpha = 0.9) +
    geom_point(size = 2) +
    ggsci::scale_color_lancet() +
    labs(x = "Log fold-change (bias-corrected)", y = NULL,
         title = paste(title_prefix, "-", make_pretty_term(term_key))) +
    theme_classic(base_size = 12) +
    theme(axis.text = element_text(color = "black"))
}

# Decide which plotting path to use based on res_tbl structure
is_wide <- any(grepl("^lfc_", names(res_tbl)))

terms_to_plot <- c("groupObese", "timepoint48H", "carbohydrateRDC", "carbohydrateSDC")
plots <- list()
for (tm in terms_to_plot) {
  p <- if (is_wide) plot_forest_from_wide(res_tbl, tm) else plot_forest_from_long(res_tbl, tm)
  if (!is.null(p)) plots[[tm]] <- p
}

# Save individual plots and show a combined grid of available ones
plot_list <- plots[!vapply(plots, is.null, logical(1))]
if (length(plot_list) > 0) {
  for (nm in names(plot_list)) {
    ggsave(file.path("outputs", paste0("ancombc_forest_", nm, ".png")), plot_list[[nm]],
           width = 8.5, height = 6.5, dpi = 300)
  }
  cowplot::plot_grid(plotlist = plot_list, ncol = 1)
}
```


