

```{r library-loading}
# Helper function to install CRAN or Bioconductor packages if not already installed
install_if_missing <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    tryCatch({
      install.packages(pkg)
    }, error = function(e) {
      if (!requireNamespace("BiocManager", quietly = TRUE)) {
        install.packages("BiocManager")
      }
      BiocManager::install(pkg, ask = FALSE)
    })
  }
}

# List of required packages
required_pkgs <- c(
  "tidyverse", "tidytext", "tidygraph", "ggraph", "igraph",
  "phyloseq", "microbiome", "corncob", "ggpubr", "cowplot",
  "ANCOMBC", "microbiomeMarker", "lme4", "lmerTest", "broom.mixed",
  "emmeans", "MuMIn"
)

# Install missing packages
invisible(lapply(required_pkgs, install_if_missing))

# Load the libraries
library(tidyverse)
library(tidytext)
library(tidygraph)
library(ggraph)
library(igraph)
library(phyloseq)
library(microbiome)
library(corncob)
library(ggpubr)
library(cowplot)
library(ANCOMBC)
library(microbiomeMarker)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(emmeans)
library(MuMIn)
```

```{r data-loading}
samdat <- read_tsv("sample.metadata.tsv")
colnames(samdat) <- c("SampleID", "sample_label", "raw_seq_file", "group")

# Process sample_label to extract additional information
samdat <- samdat %>%
  separate(sample_label, into = c("subjectID", "replicate_code", "timepoint"), sep = "\\.", remove = FALSE) %>%
  mutate(
    # Extract carbohydrate type from replicate_code
    carbohydrate = case_when(
      str_detect(replicate_code, "R") ~ "RDC",
      str_detect(replicate_code, "S") ~ "SDC",
      TRUE ~ "No Added Carb"
    ) %>% factor(levels = c("No Added Carb", "RDC", "SDC")),
    # Extract replicate number from replicate_code
    replicate = str_extract(replicate_code, "\\d+") %>% as.numeric(),
    # Replace NA values in replicate with 1
    replicate = ifelse(is.na(replicate), 1, replicate)
  ) %>%
  select(-replicate_code, -group)  # Remove the intermediate replicate_code column

# Display the processed metadata
print(samdat)

# now get the group information for health and obese

group <- read_tsv("zr24558.16S_250813.zymo/1.Healthy.TimepointComparism..Bac16Sv34/Sample_Information/sample.metadata.csv")

# Extract subjectID from customer_label
group <- group %>%
  separate(customer_label, into = c("subjectID", "replicate_code", "timepoint"), sep = "\\.", remove = FALSE) %>%
  select(customer_label, subjectID, everything())

# Check the structure of group data
print("Group data structure:")
print(colnames(group))
print(head(group))

 samdat$group <- ifelse(samdat$subjectID %in% group$subjectID, "Healthy", "Obese")

# now output a summary of the metadata in samdat
# summarize samdat by group and carbohydrate
samdat_summary <- samdat %>%
  group_by(group, carbohydrate) %>%
  summarize(n_samples = n())

samdat_summary

# number of unique subjects per group
samdat %>%
  group_by(group) %>%
  summarize(n_subjects = n_distinct(subjectID))


write_csv(samdat, file = "samdat.csv")
```

```{r import-asv-finder-data-and-phyloseq-object}
asv_data <- read_tsv("dada2_asv/asv_table.tsv")
asv <- asv_data$ASV_ID

asv_data <- asv_data %>% 
  as.data.frame() %>% 
  select(-ASV_ID, -Sequence, -Length, -Total_Abundance, -Relative_Abundance) 
colnames(asv_data) <- str_split(colnames(asv_data), "_") %>% sapply(function(x) paste(x[1:2], collapse = "_"))
rownames(asv_data) <- asv

taxonomy_data <- read_tsv("dada2_asv/strobelca_taxonomy.tsv")
asv <- taxonomy_data$ASV_ID %>% str_split(" ") %>% sapply(function(x) x[1])
repseqs <- Biostrings::DNAStringSet(taxonomy_data$Sequence)
names(repseqs) <- asv
taxonomy_data <- taxonomy_data %>% 
  as.data.frame() %>% 
  select(-ASV_ID, -Sequence, -Confidence, -Method, -Identity, -Coverage, -Reference_ID) %>% 
  as.matrix() 
rownames(taxonomy_data) <- asv

samdat <- samdat %>% 
  as.data.frame() 
rownames(samdat) <- samdat$SampleID

# now make the phyloseq object
phy_asv_finder <- phyloseq(
  otu_table(asv_data, taxa_are_rows = TRUE),
  tax_table(taxonomy_data),
  sample_data(samdat)
)

save(phy_asv_finder, file = "phy_asv_finder.RData")
```

```{r alpha-diversity-by-group}
alpha_diversity <- estimate_richness(phy_asv_finder, measures = c("Shannon", "Simpson", "Observed"))
# merge alpha_diversity with samdat
alpha_diversity <- alpha_diversity %>% 
  rownames_to_column(var = "SampleID") %>% 
  left_join(samdat, by = "SampleID")


# plot alpha_diversity with publication quality theme, colors, and Wilcoxon stats
library(ggpubr)

# Draw a stats bracket between the groups being compared using stat_compare_means with comparisons argument
group_levels <- unique(alpha_diversity$group)
if (length(group_levels) == 2) {
  comparisons <- list(group_levels)
} else {
  # For more than 2 groups, compare all pairs
  comparisons <- combn(group_levels, 2, simplify = FALSE)
}

library(cowplot)

# Shannon panel
p_shannon <- ggplot(alpha_diversity, aes(x = group, y = Shannon, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  scale_fill_brewer(palette = "Set2") +
  theme_classic(base_size = 16) +
  labs(x = "Group", y = "Shannon Diversity", fill = "Group") +
  stat_compare_means(
    comparisons = comparisons,
    method = "wilcox.test",
    label = "p.signif",
    label.y = seq(
      from = max(alpha_diversity$Shannon, na.rm = TRUE) * 1.05,
      by = 0.1 * max(alpha_diversity$Shannon, na.rm = TRUE),
      length.out = length(comparisons)
    )
  ) +
  theme(
    legend.position = "none",
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()
  )

# Observed panel
p_observed <- ggplot(alpha_diversity, aes(x = group, y = Observed, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  scale_fill_brewer(palette = "Set2") +
  theme_classic(base_size = 16) +
  labs(x = "Group", y = "Observed ASVs", fill = "Group") +
  stat_compare_means(
    comparisons = comparisons,
    method = "wilcox.test",
    label = "p.signif",
    label.y = seq(
      from = max(alpha_diversity$Observed, na.rm = TRUE) * 1.05,
      by = 0.1 * max(alpha_diversity$Observed, na.rm = TRUE),
      length.out = length(comparisons)
    )
  ) +
  theme(
    legend.position = "none",
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()
  )

# Simpson panel
p_simpson <- ggplot(alpha_diversity, aes(x = group, y = Simpson, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  scale_fill_brewer(palette = "Set2") +
  theme_classic(base_size = 16) +
  labs(x = "Group", y = "Simpson Diversity", fill = "Group") +
  stat_compare_means(
    comparisons = comparisons,
    method = "wilcox.test",
    label = "p.signif",
    label.y = seq(
      from = max(alpha_diversity$Simpson, na.rm = TRUE) * 1.05,
      by = 0.1 * max(alpha_diversity$Simpson, na.rm = TRUE),
      length.out = length(comparisons)
    )
  ) +
  theme(
    legend.position = "none",
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()
  )

# Combine panels into a multipanel figure
plot_grid(
  p_observed,
  p_shannon,
  p_simpson,
  labels = c("A", "B", "C"),
  nrow = 1,
  align = "v"
)
```

```{r alpha-diversity-by-group-and-timepoint}
alpha_diversity <- estimate_richness(phy_asv_finder, measures = c("Shannon", "Simpson", "Observed"))
# merge alpha_diversity with samdat
alpha_diversity <- alpha_diversity %>%
  rownames_to_column(var = "SampleID") %>%
  left_join(samdat, by = "SampleID")

# Get unique levels for all variables
group_levels <- unique(alpha_diversity$group)
time_levels <- unique(alpha_diversity$timepoint)
carbohydrate_levels <- unique(alpha_diversity$carbohydrate)

# Set up comparisons for different variables
group_comparisons <- if (length(group_levels) == 2) {
  list(group_levels)
} else {
  combn(group_levels, 2, simplify = FALSE)
}

carb_comparisons <- if (length(carbohydrate_levels) == 2) {
  list(carbohydrate_levels)
} else {
  combn(carbohydrate_levels, 2, simplify = FALSE)
}

time_comparisons <- if (length(time_levels) == 2) {
  list(time_levels)
} else {
  combn(time_levels, 2, simplify = FALSE)
}

# === BY GROUP ===
# Shannon by group
p_shannon_group <- ggplot(alpha_diversity, aes(x = group, y = Shannon, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  scale_fill_brewer(palette = "Set2") +
  theme_classic(base_size = 12) +
  labs(x = "Group", y = "Shannon Diversity", title = "By Group") +
  stat_compare_means(
    comparisons = group_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme(legend.position = "none", axis.text = element_text(color = "black"))

# Observed by group
p_observed_group <- ggplot(alpha_diversity, aes(x = group, y = Observed, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  scale_fill_brewer(palette = "Set2") +
  theme_classic(base_size = 12) +
  labs(x = "Group", y = "Observed ASVs", title = "By Group") +
  stat_compare_means(
    comparisons = group_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme(legend.position = "none", axis.text = element_text(color = "black"))

# === BY CARBOHYDRATE ===
# Shannon by carbohydrate
p_shannon_carb <- ggplot(alpha_diversity, aes(x = carbohydrate, y = Shannon, fill = carbohydrate)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  scale_fill_brewer(palette = "Set3") +
  theme_classic(base_size = 12) +
  labs(x = "Carbohydrate", y = "Shannon Diversity", title = "By Carbohydrate") +
  stat_compare_means(
    comparisons = carb_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme(legend.position = "none", axis.text = element_text(color = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Observed by carbohydrate
p_observed_carb <- ggplot(alpha_diversity, aes(x = carbohydrate, y = Observed, fill = carbohydrate)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  scale_fill_brewer(palette = "Set3") +
  theme_classic(base_size = 12) +
  labs(x = "Carbohydrate", y = "Observed ASVs", title = "By Carbohydrate") +
  stat_compare_means(
    comparisons = carb_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme(legend.position = "none", axis.text = element_text(color = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1))

# === BY TIMEPOINT ===
# Shannon by timepoint
p_shannon_time <- ggplot(alpha_diversity, aes(x = timepoint, y = Shannon, fill = timepoint)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  scale_fill_brewer(palette = "Paired") +
  theme_classic(base_size = 12) +
  labs(x = "Timepoint", y = "Shannon Diversity", title = "By Timepoint") +
  stat_compare_means(
    comparisons = time_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme(legend.position = "none", axis.text = element_text(color = "black"))

# Observed by timepoint
p_observed_time <- ggplot(alpha_diversity, aes(x = timepoint, y = Observed, fill = timepoint)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  scale_fill_brewer(palette = "Paired") +
  theme_classic(base_size = 12) +
  labs(x = "Timepoint", y = "Observed ASVs", title = "By Timepoint") +
  stat_compare_means(
    comparisons = time_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme(legend.position = "none", axis.text = element_text(color = "black"))

# === INTERACTION PLOTS ===
# Shannon by group and timepoint
p_shannon_interaction <- ggplot(alpha_diversity, aes(x = timepoint, y = Shannon, fill = group)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.7, alpha = 0.8) +
  geom_point(position = position_jitterdodge(dodge.width = 0.8, jitter.width = 0.2),
             size = 1.2, alpha = 0.6) +
  scale_fill_brewer(palette = "Set2") +
  theme_classic(base_size = 12) +
  labs(x = "Timepoint", y = "Shannon Diversity", fill = "Group",
       title = "Group × Timepoint Interaction") +
  theme(axis.text = element_text(color = "black"))

# Observed by group and timepoint
p_observed_interaction <- ggplot(alpha_diversity, aes(x = timepoint, y = Observed, fill = group)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.7, alpha = 0.8) +
  geom_point(position = position_jitterdodge(dodge.width = 0.8, jitter.width = 0.2),
             size = 1.2, alpha = 0.6) +
  scale_fill_brewer(palette = "Set2") +
  theme_classic(base_size = 12) +
  labs(x = "Timepoint", y = "Observed ASVs", fill = "Group",
       title = "Group × Timepoint Interaction") +
  theme(axis.text = element_text(color = "black"))

# Combine all plots
plot_grid(
  p_shannon_group, p_shannon_carb, p_shannon_time, p_shannon_interaction,
  p_observed_group, p_observed_carb, p_observed_time, p_observed_interaction,
  labels = c("A", "B", "C", "D", "E", "F", "G", "H"),
  nrow = 2,
  align = "hv"
)
```

### Interpretation of Alpha Diversity by Group, Carbohydrate, and Timepoint

- Group comparisons: Using Wilcoxon rank-sum tests, we did `r ifelse(wilcox.test(Shannon ~ group, data = subset(alpha_diversity, !is.na(Shannon) & !is.na(group)))$p.value < 0.05, "detect", "not detect")` a statistically significant difference in Shannon diversity between groups (`p = `r format.pval(wilcox.test(Shannon ~ group, data = subset(alpha_diversity, !is.na(Shannon) & !is.na(group)))$p.value, digits = 2)`), and `r ifelse(wilcox.test(Observed ~ group, data = subset(alpha_diversity, !is.na(Observed) & !is.na(group)))$p.value < 0.05, "detect", "not detect")` a significant difference for Observed ASVs (`p = `r format.pval(wilcox.test(Observed ~ group, data = subset(alpha_diversity, !is.na(Observed) & !is.na(group)))$p.value, digits = 2)``). Visual patterns in panels A and E align with these results.

- Carbohydrate comparisons: Because there are three conditions (No Added Carb, RDC, SDC), we summarize differences with Kruskal–Wallis tests. We observe significant overall effects of carbohydrate on Shannon diversity (`p = `r format.pval(kruskal.test(Shannon ~ carbohydrate, data = subset(alpha_diversity, !is.na(Shannon) & !is.na(carbohydrate)))$p.value, digits = 2)`) and on Observed ASVs (`p = `r format.pval(kruskal.test(Observed ~ carbohydrate, data = subset(alpha_diversity, !is.na(Observed) & !is.na(carbohydrate)))$p.value, digits = 2)`). Panels B and F illustrate these shifts across carbohydrate conditions.

- Timepoint comparisons: Wilcoxon tests indicate `r ifelse(wilcox.test(Shannon ~ timepoint, data = subset(alpha_diversity, !is.na(Shannon) & !is.na(timepoint)))$p.value < 0.05, "significant", "no statistically significant")` differences in Shannon (`p = `r format.pval(wilcox.test(Shannon ~ timepoint, data = subset(alpha_diversity, !is.na(Shannon) & !is.na(timepoint)))$p.value, digits = 2)`) and `r ifelse(wilcox.test(Observed ~ timepoint, data = subset(alpha_diversity, !is.na(Observed) & !is.na(timepoint)))$p.value < 0.05, "significant", "no statistically significant")` differences in Observed ASVs (`p = `r format.pval(wilcox.test(Observed ~ timepoint, data = subset(alpha_diversity, !is.na(Observed) & !is.na(timepoint)))$p.value, digits = 2)`). Panels C and G show broadly similar distributions across timepoints.

- Group × timepoint: The interaction panels (D, H) show limited divergence between groups across time, consistent with small between-group effects relative to within-subject variability. Formal inference for interactions is provided in the mixed-effects section below.


```{r alpha-diversity-groups-by-carbohydrate, message=FALSE, warning=FALSE}
# Observed ASVs: groups side-by-side within each carbohydrate type
# Reuse alpha_diversity created above
library(ggpubr)

# Ensure expected ordering for readability
alpha_diversity <- alpha_diversity %>%
  mutate(
    carbohydrate = factor(carbohydrate,
                          levels = sort(unique(carbohydrate))),
    group = factor(group, levels = sort(unique(group)))
  )

# Define comparison between groups (assumes two groups)
grp_levels <- levels(alpha_diversity$group)
grp_comparisons <- if (length(grp_levels) == 2) list(grp_levels) else combn(grp_levels, 2, simplify = FALSE)

p_obs_grp_by_carb <- ggplot(alpha_diversity,
                            aes(x = group, y = Observed, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.65, alpha = 0.85) +
  geom_jitter(width = 0.15, size = 1.3, alpha = 0.45, color = "black") +
  scale_fill_brewer(palette = "Set2") +
  theme_classic(base_size = 14) +
  labs(x = "Group", y = "Observed ASVs", title = "Observed ASVs by Carbohydrate (Groups Side-by-Side)") +
  facet_wrap(~ carbohydrate, nrow = 1) +
  stat_compare_means(comparisons = grp_comparisons,
                     method = "wilcox.test",
                     label = "p.signif") +
  theme(
    legend.position = "none",
    axis.text = element_text(color = "black"),
    strip.background = element_rect(fill = "grey95", color = NA),
    strip.text = element_text(color = "black")
  )

p_obs_grp_by_carb
```

```{r mixed-effects-alpha-diversity}
# Load required packages for mixed effects models
library(lme4)
library(lmerTest)
library(broom.mixed)

# Ensure SubjectID exists and is a factor (metadata uses `subjectID`)
if (!"SubjectID" %in% names(alpha_diversity)) {
  if ("subjectID" %in% names(alpha_diversity)) {
    alpha_diversity <- alpha_diversity %>% mutate(SubjectID = as.factor(subjectID))
  } else {
    stop("SubjectID/subjectID not found in alpha_diversity; check metadata join.")
  }
} else {
  alpha_diversity$SubjectID <- as.factor(alpha_diversity$SubjectID)
}

# Create models for each diversity metric
# Shannon Diversity Mixed Effects Model
model_shannon <- lmer(Shannon ~ group * timepoint * carbohydrate + (1|SubjectID),
                      data = alpha_diversity)

# Observed ASVs Mixed Effects Model
model_observed <- lmer(Observed ~ group * timepoint * carbohydrate + (1|SubjectID),
                       data = alpha_diversity)

# Simpson Diversity Mixed Effects Model
model_simpson <- lmer(Simpson ~ group * timepoint * carbohydrate + (1|SubjectID),
                      data = alpha_diversity)

# Function to extract and format results
extract_mixed_model_results <- function(model, metric_name) {
  # Get ANOVA table with p-values
  anova_results <- anova(model)

  # Get fixed effects coefficients
  fixed_effects <- broom.mixed::tidy(model, effects = "fixed")

  # Get random effects variance
  random_effects <- broom.mixed::tidy(model, effects = "ran_pars")

  # Create a summary data frame
  summary_df <- data.frame(
    Metric = metric_name,
    Term = rownames(anova_results),
    F_value = anova_results$`F value`,
    p_value = anova_results$`Pr(>F)`,
    significance = ifelse(anova_results$`Pr(>F)` < 0.001, "***",
                         ifelse(anova_results$`Pr(>F)` < 0.01, "**",
                               ifelse(anova_results$`Pr(>F)` < 0.05, "*", ""))),
    stringsAsFactors = FALSE
  )

  return(list(
    anova = summary_df,
    fixed_effects = fixed_effects,
    random_effects = random_effects,
    model = model
  ))
}

# Extract results for all models
shannon_results <- extract_mixed_model_results(model_shannon, "Shannon")
observed_results <- extract_mixed_model_results(model_observed, "Observed")
simpson_results <- extract_mixed_model_results(model_simpson, "Simpson")

# Combine ANOVA results for display
all_anova_results <- rbind(
  shannon_results$anova,
  observed_results$anova,
  simpson_results$anova
)

# Display results
cat("=== MIXED EFFECTS MODEL RESULTS ===\n")
cat("(SubjectID as random effect)\n\n")

print(all_anova_results)

# Post-hoc comparisons for significant effects
library(emmeans)

# Function to perform post-hoc comparisons
perform_posthoc <- function(model, metric_name) {
  cat(paste("\n=== POST-HOC COMPARISONS FOR", metric_name, "===\n"))

  # Group comparisons
  if("group" %in% names(coef(model)$SubjectID)) {
    group_emm <- emmeans(model, ~ group)
    group_contrasts <- pairs(group_emm, adjust = "bonferroni")
    cat("\nGroup comparisons:\n")
    print(group_contrasts)
  }

  # Timepoint comparisons
  if("timepoint" %in% names(coef(model)$SubjectID)) {
    time_emm <- emmeans(model, ~ timepoint)
    time_contrasts <- pairs(time_emm, adjust = "bonferroni")
    cat("\nTimepoint comparisons:\n")
    print(time_contrasts)
  }

  # Carbohydrate comparisons
  if("carbohydrate" %in% names(coef(model)$SubjectID)) {
    carb_emm <- emmeans(model, ~ carbohydrate)
    carb_contrasts <- pairs(carb_emm, adjust = "bonferroni")
    cat("\nCarbohydrate comparisons:\n")
    print(carb_contrasts)
  }

  # Two-way interactions if significant
  # Group × Timepoint
  tryCatch({
    group_time_emm <- emmeans(model, ~ group * timepoint)
    group_time_contrasts <- pairs(group_time_emm, adjust = "bonferroni")
    cat("\nGroup × Timepoint interaction:\n")
    print(group_time_contrasts)
  }, error = function(e) {})
}

# Perform post-hoc tests for models with significant effects
if(any(shannon_results$anova$p_value < 0.05, na.rm = TRUE)) {
  perform_posthoc(model_shannon, "SHANNON")
}

if(any(observed_results$anova$p_value < 0.05, na.rm = TRUE)) {
  perform_posthoc(model_observed, "OBSERVED")
}

if(any(simpson_results$anova$p_value < 0.05, na.rm = TRUE)) {
  perform_posthoc(model_simpson, "SIMPSON")
}

# Model diagnostics
cat("\n=== MODEL DIAGNOSTICS ===\n")

# Check model assumptions with plots
par(mfrow = c(2, 2))

# Shannon model diagnostics
plot(model_shannon, main = "Shannon: Residuals vs Fitted")
qqnorm(resid(model_shannon), main = "Shannon: Q-Q Plot")
qqline(resid(model_shannon))

# Observed model diagnostics
plot(model_observed, main = "Observed: Residuals vs Fitted")
qqnorm(resid(model_observed), main = "Observed: Q-Q Plot")
qqline(resid(model_observed))

par(mfrow = c(1, 1))

# Random effects summary
cat("\n=== RANDOM EFFECTS (SubjectID) ===\n")
cat("Shannon model - Subject variance:",
    shannon_results$random_effects$estimate[shannon_results$random_effects$term == "sd__(Intercept)"]^2, "\n")
cat("Observed model - Subject variance:",
    observed_results$random_effects$estimate[observed_results$random_effects$term == "sd__(Intercept)"]^2, "\n")
cat("Simpson model - Subject variance:",
    simpson_results$random_effects$estimate[simpson_results$random_effects$term == "sd__(Intercept)"]^2, "\n")

# Effect sizes (pseudo R-squared for mixed models)
library(MuMIn)
cat("\n=== MODEL FIT (R-squared) ===\n")
cat("Shannon R²:", round(r.squaredGLMM(model_shannon)[1], 3), "(marginal),",
    round(r.squaredGLMM(model_shannon)[2], 3), "(conditional)\n")
cat("Observed R²:", round(r.squaredGLMM(model_observed)[1], 3), "(marginal),",
    round(r.squaredGLMM(model_observed)[2], 3), "(conditional)\n")
cat("Simpson R²:", round(r.squaredGLMM(model_simpson)[1], 3), "(marginal),",
    round(r.squaredGLMM(model_simpson)[2], 3), "(conditional)\n")

cat("\nNote: Marginal R² = variance explained by fixed effects\n")
cat("      Conditional R² = variance explained by fixed + random effects\n")
```

### Interpretation of Mixed-Effects Alpha-Diversity Models

Across all three alpha-diversity metrics, carbohydrate condition was the only fixed effect that consistently explained variance in community diversity. With SubjectID modeled as a random intercept, the main effect of carbohydrate was significant for Shannon (F = `r round(shannon_results$anova$F_value[shannon_results$anova$Term == "carbohydrate"], 2)`, p = `r format.pval(shannon_results$anova$p_value[shannon_results$anova$Term == "carbohydrate"], digits = 2)`), Observed ASVs (F = `r round(observed_results$anova$F_value[observed_results$anova$Term == "carbohydrate"], 2)`, p = `r format.pval(observed_results$anova$p_value[observed_results$anova$Term == "carbohydrate"], digits = 2)`), and Simpson (F = `r round(simpson_results$anova$F_value[simpson_results$anova$Term == "carbohydrate"], 2)`, p = `r format.pval(simpson_results$anova$p_value[simpson_results$anova$Term == "carbohydrate"], digits = 2)`). The timepoint-by-carbohydrate interaction was `r ifelse(shannon_results$anova$p_value[shannon_results$anova$Term == "timepoint:carbohydrate"] < 0.05, "significant", "not significant")` for Shannon (F = `r round(shannon_results$anova$F_value[shannon_results$anova$Term == "timepoint:carbohydrate"], 2)`, p = `r format.pval(shannon_results$anova$p_value[shannon_results$anova$Term == "timepoint:carbohydrate"], digits = 2)`) and `r ifelse(observed_results$anova$p_value[observed_results$anova$Term == "timepoint:carbohydrate"] < 0.05, "significant", "not significant")` for Observed ASVs (F = `r round(observed_results$anova$F_value[observed_results$anova$Term == "timepoint:carbohydrate"], 2)`, p = `r format.pval(observed_results$anova$p_value[observed_results$anova$Term == "timepoint:carbohydrate"], digits = 2)`), indicating temporal modulation of carbohydrate effects. In contrast, the group main effect did `r ifelse(all(c(shannon_results$anova$p_value[shannon_results$anova$Term == "group"], observed_results$anova$p_value[observed_results$anova$Term == "group"], simpson_results$anova$p_value[simpson_results$anova$Term == "group"]) >= 0.05), "not", "")` reach conventional significance across metrics (Shannon p = `r format.pval(shannon_results$anova$p_value[shannon_results$anova$Term == "group"], digits = 2)`, Observed p = `r format.pval(observed_results$anova$p_value[observed_results$anova$Term == "group"], digits = 2)`, Simpson p = `r format.pval(simpson_results$anova$p_value[simpson_results$anova$Term == "group"], digits = 2)`).

Random-effects estimates indicated meaningful inter-individual heterogeneity. The estimated SubjectID variance components were Shannon = `r signif(shannon_results$random_effects$estimate[shannon_results$random_effects$term == "sd__(Intercept)"]^2, 3)`, Observed = `r signif(observed_results$random_effects$estimate[observed_results$random_effects$term == "sd__(Intercept)"]^2, 3)`, and Simpson = `r signif(simpson_results$random_effects$estimate[simpson_results$random_effects$term == "sd__(Intercept)"]^2, 3)`. Consistent with this, conditional R² exceeded marginal R² for all metrics (Shannon: `r round(r.squaredGLMM(model_shannon)[2], 3)` vs `r round(r.squaredGLMM(model_shannon)[1], 3)`; Observed: `r round(r.squaredGLMM(model_observed)[2], 3)` vs `r round(r.squaredGLMM(model_observed)[1], 3)`; Simpson: `r round(r.squaredGLMM(model_simpson)[2], 3)` vs `r round(r.squaredGLMM(model_simpson)[1], 3)`), indicating that subject-specific baselines account for a substantial portion of total variance.

Overall, these results support carbohydrate condition (No Added Carb, RDC, SDC) as a primary determinant of alpha diversity, with temporal dynamics modulating this effect for Shannon diversity and richness. Reporting adjusted pairwise contrasts among carbohydrate levels (e.g., estimated marginal means) will localize specific differences and provide effect sizes that are directly interpretable.

```{r beta-diversity}
library(microbiome)
library(ggplot2)
library(ggpubr)
library(cowplot)

# PCoA plot
# need to convert to relative abundance
phy_asv_finder_rel <- microbiome::transform(phy_asv_finder, "compositional")
# PCoA plot
pcoa <- ordinate(phy_asv_finder_rel, method = "PCoA", distance = "bray")
plot_ordination(phy_asv_finder_rel, pcoa, type = "samples", color = "group") +
  geom_point(size = 3) +
  facet_wrap(~timepoint) +
  theme_classic(base_size = 16)


```

## Relative Abundance Profiles

```{r relabundance-helpers, message=FALSE, warning=FALSE}
# Ensure an outputs directory exists
if (!dir.exists("outputs")) dir.create("outputs", recursive = TRUE)

# Helper: standardize empty/NA taxon names
fill_na_taxa <- function(ps) {
  tt <- tax_table(ps) %>% as.data.frame()
  rn <- rownames(tt)
  # Fill NA values progressively with higher-rank context
  ranks <- colnames(tt)
  for (rk in ranks) {
    tt[[rk]] <- ifelse(is.na(tt[[rk]]) | tt[[rk]] == "" | tt[[rk]] == "NA",
                       paste0("Unclassified_", rk), tt[[rk]])
  }
  tax_table(ps) <- as.matrix(tt)
  ps
}

# Helper: agglomerate to rank and compute relative abundance
ps_to_relabund_rank <- function(ps, rank = "Genus") {
  stopifnot(rank %in% rank_names(ps))
  ps <- fill_na_taxa(ps)
  psg <- tax_glom(ps, taxrank = rank, NArm = FALSE)
  # Relative abundances
  microbiome::transform(psg, transform = "compositional")
}

# Helper: collapse to top N taxa with Other
collapse_top_taxa <- function(df_long, tax_col = "Taxon", top_n = 12) {
  top_levels <- df_long %>%
    group_by(.data[[tax_col]]) %>%
    summarize(mean_abund = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(mean_abund)) %>%
    slice_head(n = top_n) %>% pull(.data[[tax_col]]) %>% as.character()

  df_long %>%
    mutate(
      !!tax_col := ifelse(.data[[tax_col]] %in% top_levels, .data[[tax_col]], "Other")
    ) %>%
    group_by(across(c(all_of(setdiff(colnames(df_long), c("Abundance")))))) %>%
    summarize(Abundance = sum(Abundance, na.rm = TRUE), .groups = "drop")
}

# Helper: generic barplot maker
plot_relabund <- function(ps_rel, rank = "Genus", group_vars, top_n = 12, palette = NULL,
                          x_lab = NULL, title = NULL) {
  df <- psmelt(ps_rel)
  tax_col <- rank
  df <- df %>% mutate(Taxon = .data[[tax_col]])

  # Combine labels for x-axis if more than one grouping var
  df <- df %>% mutate(
    group_label = interaction(!!!rlang::syms(group_vars), sep = " | ", drop = TRUE)
  )

  df_sum <- df %>%
    group_by(group_label, Taxon) %>%
    summarize(Abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")

  df_sum <- collapse_top_taxa(df_sum, tax_col = "Taxon", top_n = top_n)

  # Order taxa by overall abundance for consistent legend order
  taxa_order <- df_sum %>%
    group_by(Taxon) %>%
    summarize(total = sum(Abundance), .groups = "drop") %>%
    arrange(desc(total)) %>% pull(Taxon)
  df_sum$Taxon <- factor(df_sum$Taxon, levels = taxa_order)

  if (is.null(palette)) {
    palette <- scales::hue_pal()(length(taxa_order))
  }

  ggplot(df_sum, aes(x = group_label, y = Abundance, fill = Taxon)) +
    geom_col(width = 0.85) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    scale_fill_manual(values = setNames(palette, taxa_order)) +
    labs(x = x_lab %||% paste(group_vars, collapse = " | "),
         y = "Relative abundance", title = title, fill = rank) +
    theme_classic(base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          axis.text = element_text(color = "black"),
          legend.key.size = unit(6, "mm"))
}

# Convenience operator
`%||%` <- function(a, b) if (!is.null(a)) a else b
```

```{r relabundance-phylum-family-genus, message=FALSE, warning=FALSE, fig.width=12, fig.height=7}
# Prepare rank-level relative abundance phyloseq objects
ps_rel_phylum <- ps_to_relabund_rank(phy_asv_finder, rank = "Phylum")
ps_rel_family <- ps_to_relabund_rank(phy_asv_finder, rank = "Family")
ps_rel_genus  <- ps_to_relabund_rank(phy_asv_finder, rank = "Genus")

# Plots by single variables
p_phylum_group <- plot_relabund(ps_rel_phylum, rank = "Phylum",
                                group_vars = c("group"), top_n = 12,
                                title = "Phylum-level Composition by Group",
                                x_lab = "Group")
p_family_group <- plot_relabund(ps_rel_family, rank = "Family",
                                group_vars = c("group"), top_n = 15,
                                title = "Family-level Composition by Group",
                                x_lab = "Group")
p_genus_group  <- plot_relabund(ps_rel_genus,  rank = "Genus",
                                group_vars = c("group"), top_n = 15,
                                title = "Genus-level Composition by Group",
                                x_lab = "Group")

p_phylum_time <- plot_relabund(ps_rel_phylum, rank = "Phylum",
                               group_vars = c("timepoint"), top_n = 12,
                               title = "Phylum-level Composition by Timepoint",
                               x_lab = "Timepoint")
p_family_time <- plot_relabund(ps_rel_family, rank = "Family",
                               group_vars = c("timepoint"), top_n = 15,
                               title = "Family-level Composition by Timepoint",
                               x_lab = "Timepoint")
p_genus_time  <- plot_relabund(ps_rel_genus,  rank = "Genus",
                               group_vars = c("timepoint"), top_n = 15,
                               title = "Genus-level Composition by Timepoint",
                               x_lab = "Timepoint")

p_phylum_carb <- plot_relabund(ps_rel_phylum, rank = "Phylum",
                               group_vars = c("carbohydrate"), top_n = 12,
                               title = "Phylum-level Composition by Carbohydrate",
                               x_lab = "Carbohydrate")
p_family_carb <- plot_relabund(ps_rel_family, rank = "Family",
                               group_vars = c("carbohydrate"), top_n = 15,
                               title = "Family-level Composition by Carbohydrate",
                               x_lab = "Carbohydrate")
p_genus_carb  <- plot_relabund(ps_rel_genus,  rank = "Genus",
                               group_vars = c("carbohydrate"), top_n = 15,
                               title = "Genus-level Composition by Carbohydrate",
                               x_lab = "Carbohydrate")

# Save representative plots
ggsave("outputs/relabund_phylum_by_group.png", p_phylum_group, width = 10, height = 6, dpi = 300)
ggsave("outputs/relabund_genus_by_carb.png",  p_genus_carb,  width = 10, height = 6, dpi = 300)

plot_grid(p_phylum_group, p_family_group, p_genus_group, ncol = 1, labels = c("A","B","C"))
```

```{r relabundance-combinations, message=FALSE, warning=FALSE, fig.width=12, fig.height=7}
# Two-way combinations
p_genus_grp_time <- plot_relabund(ps_rel_genus, rank = "Genus",
                                  group_vars = c("group", "timepoint"), top_n = 15,
                                  title = "Genus-level Composition: Group | Timepoint",
                                  x_lab = "Group | Timepoint")

p_genus_grp_carb <- plot_relabund(ps_rel_genus, rank = "Genus",
                                  group_vars = c("group", "carbohydrate"), top_n = 15,
                                  title = "Genus-level Composition: Group | Carbohydrate",
                                  x_lab = "Group | Carbohydrate")

p_genus_time_carb <- plot_relabund(ps_rel_genus, rank = "Genus",
                                   group_vars = c("timepoint", "carbohydrate"), top_n = 15,
                                   title = "Genus-level Composition: Timepoint | Carbohydrate",
                                   x_lab = "Timepoint | Carbohydrate")

plot_grid(p_genus_grp_time, p_genus_grp_carb, p_genus_time_carb, ncol = 1,
          labels = c("A","B","C"))

# Three-way combination (single overview)
p_genus_all <- plot_relabund(ps_rel_genus, rank = "Genus",
                             group_vars = c("group", "timepoint", "carbohydrate"), top_n = 15,
                             title = "Genus-level Composition: Group | Timepoint | Carbohydrate",
                             x_lab = "Group | Timepoint | Carbohydrate")

ggsave("outputs/relabund_genus_all.png", p_genus_all, width = 12, height = 6.5, dpi = 300)

p_genus_all
```

## Differential Abundance

```{r da-setup, message=FALSE, warning=FALSE}
# Keep raw counts for DA methods
ps_raw <- fill_na_taxa(phy_asv_finder)

# Ensure expected covariates exist
stopifnot(all(c("group", "timepoint", "carbohydrate") %in% sample_variables(ps_raw)))

# Add SubjectID; metadata used subjectID earlier, create SubjectID if missing
sdat <- sample_data(ps_raw) %>% as.data.frame()
if (!"SubjectID" %in% colnames(sdat) && "subjectID" %in% colnames(sdat)) {
  sdat$SubjectID <- as.factor(sdat$subjectID)
  sample_data(ps_raw) <- sample_data(sdat)
}
if (!"SubjectID" %in% colnames(sdat)) {
  stop("SubjectID (or subjectID) is required for random effect / blocking.")
}

# Work at Genus level for DA
ps_genus_counts <- tax_glom(ps_raw, taxrank = "Genus", NArm = FALSE)

# Optional prevalence/abundance filtering to ease computations
prev_threshold <- 0.05
min_prev <- ceiling(prev_threshold * nsamples(ps_genus_counts))
keep_taxa <- taxa_sums(ps_genus_counts) > 0 & rowSums(otu_table(ps_genus_counts) > 0) >= min_prev
ps_genus_filt <- prune_taxa(keep_taxa, ps_genus_counts)

cat("Genus taxa before/after filtering:", ntaxa(ps_genus_counts), ntaxa(ps_genus_filt), "\n")
```

### Corncob (beta-binomial)

```{r da-corncob, message=FALSE, warning=FALSE}
# Note: corncob does not natively support random effects. We include SubjectID
# as a blocking factor (fixed effect) to account for repeated measures.

run_corncob_drop1 <- function(ps, effect_var, fdr_cutoff = 0.1) {
  stopifnot(effect_var %in% sample_variables(ps))
  # Full and reduced formulas
  full <- as.formula(paste("~", paste(c("group", "timepoint", "carbohydrate", "SubjectID"), collapse = " + ")))
  red  <- as.formula(paste("~", paste(setdiff(c("group", "timepoint", "carbohydrate", "SubjectID"), effect_var), collapse = " + ")))

  set.seed(1)
  res <- differentialTest(
    formula = full,
    phi.formula = full,
    formula_null = red,
    phi.formula_null = red,
    test = "Wald",
    data = ps,
    boot = FALSE,
    B = 0,
    fdr_cutoff = fdr_cutoff
  )

  tibble(
    taxon = res$significant_taxa,
    effect = effect_var
  )
}

cc_group <- run_corncob_drop1(ps_genus_filt, "group")
cc_time  <- run_corncob_drop1(ps_genus_filt, "timepoint")
cc_carb  <- run_corncob_drop1(ps_genus_filt, "carbohydrate")

corncob_hits <- bind_rows(cc_group, cc_time, cc_carb)
write_csv(corncob_hits, "outputs/corncob_genus_hits.csv")
corncob_hits
```

### ANCOM-BC2 (with SubjectID random effect)

```{r da-ancombc2, message=FALSE, warning=FALSE}
# Try ANCOMBC2 if available; otherwise fall back to ANCOM-BC v1
use_ancombc2 <- "ancombc2" %in% getNamespaceExports("ANCOMBC")

if (use_ancombc2) {
  out_ancom <- ANCOMBC::ancombc2(
    data = ps_raw,
    tax_level = "Genus",
    fix_formula = "group + timepoint + carbohydrate",
    rand_formula = "(1|SubjectID)",
    p_adj_method = "holm",
    prv_cut = 0.10,
    lib_cut = 0,
    s0_perc = 0.05,
    group = "group",
    struc_zero = TRUE,
    neg_lb = TRUE,
    alpha = 0.05,
    n_cl = 1,
    global = TRUE
  )

  # Extract results
  res_tbl <- out_ancom$res
} else {
  message("ANCOMBC2 not found; using ANCOM-BC (v1) without random effects.")
  out_ancom <- ANCOMBC::ancombc(
    phyloseq = ps_genus_filt,
    formula = "group + timepoint + carbohydrate",
    p_adj_method = "holm",
    zero_cut = 0.90,
    lib_cut = 0,
    group = "group",
    struc_zero = TRUE,
    neg_lb = TRUE,
    alpha = 0.05,
    global = TRUE
  )

  # Tidy results from v1 object
  res_list <- lapply(names(out_ancom$beta), function(term) {
    tibble(
      taxon = rownames(out_ancom$beta[[term]]),
      term = term,
      beta = out_ancom$beta[[term]][, 1],
      se = out_ancom$se[[term]][, 1],
      W = out_ancom$W[[term]][, 1],
      p = out_ancom$p_val[[term]][, 1],
      q = out_ancom$q_val[[term]][, 1],
      diff_abn = out_ancom$diff_abn[[term]][, 1]
    )
  })
  res_tbl <- bind_rows(res_list)
}

write_csv(res_tbl, "outputs/ancombc_genus_results.csv")
head(res_tbl)
```

### LEfSe (paired by SubjectID)

```{r da-lefse, message=FALSE, warning=FALSE}
# LEfSe via microbiomeMarker; use subclass for within-subject pairing
set.seed(123)
lefse_group <- tryCatch({
  microbiomeMarker::run_lefse(
    ps_genus_filt,
    class = "group",
    subclass = "SubjectID",
    norm = "CPM",
    taxa_rank = "Genus",
    wilcoxon_cutoff = 0.05,
    lda_cutoff = 2.0,
    bootstrap_n = 0
  )
}, error = function(e) NULL)

lefse_carb <- tryCatch({
  microbiomeMarker::run_lefse(
    ps_genus_filt,
    class = "carbohydrate",
    subclass = "SubjectID",
    norm = "CPM",
    taxa_rank = "Genus",
    wilcoxon_cutoff = 0.05,
    lda_cutoff = 2.0,
    bootstrap_n = 0
  )
}, error = function(e) NULL)

lefse_time <- tryCatch({
  microbiomeMarker::run_lefse(
    ps_genus_filt,
    class = "timepoint",
    subclass = "SubjectID",
    norm = "CPM",
    taxa_rank = "Genus",
    wilcoxon_cutoff = 0.05,
    lda_cutoff = 2.0,
    bootstrap_n = 0
  )
}, error = function(e) NULL)

saveRDS(lefse_group, file = "outputs/lefse_group_genus.rds")
saveRDS(lefse_carb,  file = "outputs/lefse_carb_genus.rds")
saveRDS(lefse_time,  file = "outputs/lefse_time_genus.rds")

if (!is.null(lefse_carb)) {
  p_lda <- plot(lefse_carb, taxa_level = "Genus", ldascore = 2.0)
  ggsave("outputs/lefse_lda_carb_genus.png", p_lda, width = 8, height = 6, dpi = 300)
  p_lda
}
```

### Notes on Modeling Choices

- Corncob: SubjectID is included as a blocking (fixed) factor since random effects are not supported in differentialTest. This adjusts for repeated measures when testing each effect by drop-one comparison.
- ANCOM-BC2: Uses `rand_formula = (1|SubjectID)` to include a subject-specific random intercept when available; falls back to ANCOM-BC v1 otherwise.
- LEfSe: Uses `subclass = SubjectID` to perform paired LEfSe appropriate for repeated measures.
