---
title: "Ex Vivo Carbohydrate–Microbiome Adaptation: 16S Analysis"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
set.seed(20250101)
options(stringsAsFactors = FALSE)
# Use sum-to-zero contrasts for Type III tests and balanced estimability
options(contrasts = c("contr.sum", "contr.poly"))

# Knitr default chunk options for inline figures and cleaner logs
knitr::opts_chunk$set(
  fig.width = 7, fig.height = 4, dpi = 150,
  message = FALSE, warning = FALSE,
  fig.align = "center"
)

dirs <- c("outputs", "outputs/figures", "outputs/tables", "outputs/da")
invisible(lapply(dirs, \(d) if(!dir.exists(d)) dir.create(d, recursive = TRUE)))

# Simple config (avoids YAML params for interactive chunk testing)
cfg <- list(
  min_reads = 1000L,
  feature_min_total = 10L,
  prevalence_min = 0.05
)
```

```{r libraries}
# Helper to install CRAN/Bioc packages as needed
install_if_missing <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    tryCatch({
      install.packages(pkg)
    }, error = function(e) {
      if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
      BiocManager::install(pkg, ask = FALSE)
    })
  }
}

pkgs <- c(
  "tidyverse", "stringr", "tibble", "readr", "glue",
  "phyloseq", "vegan", "microbiome",
  "lme4", "lmerTest", "car", "emmeans",
  "broom", "broom.mixed",
  "ggplot2", "ggpubr", "cowplot"
)
invisible(lapply(pkgs, install_if_missing))

library(tidyverse)
library(phyloseq)
library(vegan)
library(microbiome)
library(lme4)
library(lmerTest)
library(car)
library(emmeans)
library(broom)
library(broom.mixed)
library(glue)
library(ggpubr)
library(cowplot)
```

# Introduction

This report follows the plan in `ANALYSIS_PLAN_PROJECT2.md` to evaluate how adolescent gut microbiota from Healthy vs Obese donors adapt ex vivo to rapid- vs slow-digestible carbohydrates over time using 16S rDNA.

Objectives:

- Describe sample balance and sequencing depth; flag replicate inconsistencies.
- Test alpha diversity changes across group, carbohydrate, and timepoint with mixed-effects where supported.
- Test beta diversity differences via PERMANOVA with permutations constrained within subjects; visualize with PCoA.
- Explore differential abundance with ANCOM-BC2 using subject-level random effects.

```{r load-metadata}
# Load and parse metadata based on available files in repo
md_raw <- readr::read_tsv("sample.metadata.tsv", show_col_types = FALSE)
md <- md_raw %>%
  transmute(
    SampleID    = .data$sample_id,
    sample_label= .data$customer_label,
    raw_seq_file= .data$raw_seq_files
  ) %>%
  separate(sample_label, into = c("subjectID", "replicate_code", "timepoint"), sep = "\\.", remove = FALSE) %>%
  mutate(
    carbohydrate = case_when(
      str_detect(replicate_code, "R") ~ "RDC",
      str_detect(replicate_code, "S") ~ "SDC",
      TRUE ~ "No Added Carb"
    ),
    replicate = str_extract(replicate_code, "\\d+") %>% as.integer() %>% replace_na(1L)
  ) %>%
  select(-replicate_code)

# Derive Healthy vs Obese from Zymo report (present in repo)
zymo_md <- readr::read_tsv(
  "zr24558.16S_250813.zymo/1.Healthy.TimepointComparism..Bac16Sv34/Sample_Information/sample.metadata.csv",
  show_col_types = FALSE
)
zymo_subjects <- zymo_md %>%
  separate(customer_label, into = c("subjectID", "rc", "tp"), sep = "\\.", remove = FALSE) %>%
  distinct(subjectID) %>% pull(subjectID)

md <- md %>% mutate(group = if_else(subjectID %in% zymo_subjects, "Healthy", "Obese"))

# Factor coding (explicit reference levels)
md <- md %>% mutate(
  group        = factor(group,        levels = c("Healthy", "Obese")),
  carbohydrate = factor(carbohydrate, levels = c("No Added Carb", "RDC", "SDC")),
  timepoint    = factor(timepoint,    levels = c("0H", "48H", "72H")),
  subjectID    = factor(subjectID),
  replicate    = factor(replicate)
)

# Keys for nested random effects
md <- md %>% mutate(
  cell = interaction(subjectID, carbohydrate, timepoint, drop = TRUE),
  replicate_id = interaction(subjectID, carbohydrate, timepoint, replicate, drop = TRUE)
)

readr::write_csv(md, "samdat.csv")
md %>% count(group, carbohydrate, timepoint, name = "n") %>% arrange(group, carbohydrate, timepoint) %>%
  readr::write_csv("outputs/tables/cell_counts.csv")

# Inline: sample balance across factors
md %>% count(group, carbohydrate, timepoint, name = "n") %>% arrange(group, carbohydrate, timepoint) %>%
  knitr::kable(caption = "Sample counts by group × carbohydrate × timepoint")
```

```{r load-asv-tax, message=FALSE}
# Use processed DADA2 outputs already present in repo
asv_tab <- readr::read_tsv("dada2_asv/asv_table.tsv", show_col_types = FALSE) %>% as.data.frame()
rownames(asv_tab) <- asv_tab$ASV_ID
counts <- asv_tab %>% select(-ASV_ID, -Sequence, -Length, -Total_Abundance, -Relative_Abundance)

# Harmonize sample IDs to metadata: remove trailing "_filtered"
colnames(counts) <- gsub("_filtered$", "", colnames(counts))

# Keep only intersection and order by metadata
common_samples <- intersect(colnames(counts), md$SampleID)
if (length(common_samples) == 0) stop("No overlapping samples between counts and metadata.")
counts <- counts[, common_samples, drop = FALSE]
md_use <- md %>% filter(SampleID %in% common_samples) %>% arrange(match(SampleID, common_samples))
stopifnot(all(colnames(counts) == md_use$SampleID))

# Load taxonomy from strobelca table
tax_tab <- readr::read_tsv("dada2_asv/strobelca_taxonomy.tsv", show_col_types = FALSE) %>% as.data.frame()
tax_tab$ASV_ID <- sub(" .*$", "", tax_tab$ASV_ID)  # keep ASV id up to first space
rownames(tax_tab) <- tax_tab$ASV_ID
TAX <- tax_tab %>% select(Kingdom, Phylum, Class, Order, Family, Genus, Species) %>% as.matrix()

# Align taxa rows between counts and taxonomy
common_asv <- intersect(rownames(counts), rownames(TAX))
counts <- counts[common_asv, , drop = FALSE]
TAX    <- TAX[common_asv, , drop = FALSE]

# Assemble phyloseq
OTU  <- otu_table(as.matrix(counts), taxa_are_rows = TRUE)
META <- sample_data(as.data.frame(md_use) %>% tibble::column_to_rownames("SampleID"))
TAXO <- tax_table(TAX)
ps   <- phyloseq(OTU, TAXO, META)

saveRDS(ps, "outputs/ps_raw.rds")
```

## Data Sources

```{r data-sources}
data_sources <- tibble::tibble(
  source = c("Metadata", "ASV table", "Taxonomy"),
  path = c("sample.metadata.tsv", "dada2_asv/asv_table.tsv", "dada2_asv/strobelca_taxonomy.tsv"),
  total_records = c(nrow(md_raw), nrow(asv_tab), nrow(tax_tab)),
  matched_samples = c(length(common_samples), NA_integer_, NA_integer_),
  matched_features = c(NA_integer_, length(common_asv), length(common_asv))
)
knitr::kable(data_sources, caption = "Data sources and matched counts")
```

# Methods (Summary)

- Assembly: Built a phyloseq object from ASV counts, taxonomy, and curated metadata with explicit factor levels (Healthy vs Obese; No Added Carb, RDC, SDC; 0H, 48H, 72H).
- Filtering: Removed low-depth samples below the max of 1,000 reads or the 10th percentile of library sizes; no rarefaction for inference.
- Replicates: Kept biological replicates and used mixed-effects where identifiable; subject is the clustering unit.
- Alpha diversity: Fitted per-metric models using the most complex estimable fixed-effects structure among 3-way, all two-way, or main effects; random intercept for subject where non-singular. Type III tests with sum contrasts; EMMs for carbohydrate within time×group.
- Beta diversity: Bray–Curtis on relative abundance; PERMANOVA with permutations stratified by subject; PCoA visualization by carbohydrate/timepoint and faceted by group.
- Differential abundance: ANCOM-BC2 with random effects `(1|subjectID) + (1|subjectID:cell)` when available.

## Quality Control

```{r qc}
# QC: sequencing depth and basic filtering thresholds
lib_sizes <- sample_sums(ps)
min_depth <- max(cfg$min_reads, as.numeric(stats::quantile(lib_sizes, 0.10)))
keep <- lib_sizes >= min_depth
ps_filt <- prune_samples(keep, ps)

readr::write_csv(
  tibble(SampleID = names(lib_sizes), lib_size = as.numeric(lib_sizes), keep = keep),
  "outputs/tables/qc_library_sizes.csv"
)

# Replicate concordance at Genus level
ps_genus <- tax_glom(ps_filt, "Genus")
ps_rel   <- transform_sample_counts(ps_genus, function(x) x/sum(x))
md_rel   <- as.data.frame(sample_data(ps_rel)) %>% tibble::rownames_to_column("SampleID")

dist_bray <- phyloseq::distance(ps_rel, "bray")
get_within <- function(cell_id){
  sids <- md_rel$SampleID[md_rel$cell == cell_id]
  if(length(sids) < 2) return(NA_real_)
  as.matrix(dist_bray)[sids, sids] %>% .[upper.tri(., diag=FALSE)] %>% mean()
}
wd <- sapply(unique(md_rel$cell), get_within)
thresh <- stats::quantile(wd, 0.95, na.rm=TRUE)
qc_cells_flag <- names(wd)[wd > thresh]

# Create the tibble with proper handling of NA values
qc_tibble <- tibble(
  cell = names(wd), 
  mean_within_bray = wd, 
  flag = if(length(qc_cells_flag) > 0) names(wd) %in% qc_cells_flag else rep(FALSE, length(wd))
)
readr::write_csv(qc_tibble, "outputs/tables/qc_within_cell_bray.csv")

# Inline QC summary table
tibble(
  n_samples_raw = nsamples(ps),
  n_samples_kept = nsamples(ps_filt),
  n_subjects = dplyr::n_distinct(sample_data(ps)$subjectID),
  depth_min = min(lib_sizes),
  depth_median = stats::median(lib_sizes),
  depth_max = max(lib_sizes)
) %>% knitr::kable(caption = "QC summary: library sizes and sample counts before/after filtering")
```

## Alpha Diversity

```{r alpha}
alpha_metrics <- c("Observed","Shannon","Simpson","InvSimpson")
alpha_df <- estimate_richness(ps_filt, measures = alpha_metrics) %>%
  tibble::rownames_to_column("SampleID") %>%
  left_join(as.data.frame(sample_data(ps_filt)) %>% tibble::rownames_to_column("SampleID"), by = "SampleID")

# Check data structure for random effects
cat("Number of observations:", nrow(alpha_df), "\n")
cat("Number of subjects:", length(unique(alpha_df$subjectID)), "\n")
cat("Number of subject:cell combinations:", length(unique(alpha_df$cell)), "\n")

# Choose the most complex non-singular fixed-effects specification available
infer_fixed_terms <- function(df) {
  # Try 3-way interaction
  mm3 <- model.matrix(~ group*carbohydrate*timepoint, data = df)
  if (qr(mm3)$rank == ncol(mm3)) return("group*carbohydrate*timepoint")
  # Else try all 2-way interactions
  mm2 <- model.matrix(~ (group + carbohydrate + timepoint)^2, data = df)
  if (qr(mm2)$rank == ncol(mm2)) return("(group + carbohydrate + timepoint)^2")
  # Fallback to main effects only
  return("group + carbohydrate + timepoint")
}

fixed_str <- infer_fixed_terms(alpha_df)
cat("Fixed-effects used for alpha models:", fixed_str, "\n")

# Mixed models per metric with robust optimizer; fall back to lm if random effect is singular
fits <- purrr::map(alpha_metrics, function(m){
  fm_mixed <- as.formula(paste0(m, " ~ ", fixed_str, " + (1|subjectID)"))
  fit <- try(
    lme4::lmer(
      fm_mixed, data = alpha_df, REML = TRUE,
      control = lme4::lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
    ),
    silent = TRUE
  )
  if (inherits(fit, "try-error") || (inherits(fit, "merMod") && lme4::isSingular(fit, tol = 1e-4))) {
    # Fallback to marginal model without random effect
    fm_lm <- as.formula(paste0(m, " ~ ", fixed_str))
    fit <- stats::lm(fm_lm, data = alpha_df)
  }
  fit
}) %>% rlang::set_names(alpha_metrics)

anova_tbl <- purrr::imap_dfr(fits, function(fit, m){
  suppressWarnings(car::Anova(fit, type = "III")) %>% broom::tidy() %>% mutate(metric = m)
})
readr::write_csv(anova_tbl, "outputs/tables/alpha_typeIII_anova.csv")

# Show Type III ANOVA results inline (top rows per metric)
anova_tbl %>%
  filter(term != "(Intercept)") %>%
  arrange(metric, p.value) %>%
  group_by(metric) %>%
  slice_head(n = 10) %>%
  ungroup() %>%
  mutate(p.value = signif(p.value, 3)) %>%
  knitr::kable(caption = "Alpha diversity: Type III ANOVA (top terms by metric)")

# Example EMMs: carbohydrate contrasts at each timepoint within group
emm_list <- list()
for (m in alpha_metrics) {
  fit <- fits[[m]]
  emm <- emmeans::emmeans(fit, ~ carbohydrate | timepoint * group)
  emm_list[[m]] <- summary(pairs(emm), adjust = "fdr") %>% as.data.frame() %>% mutate(metric = m)
}
emm_tbl <- dplyr::bind_rows(emm_list)
readr::write_csv(emm_tbl, "outputs/tables/alpha_emmeans_contrasts.csv")

# Brief narrative: list significant fixed effects per metric
alpha_sig <- anova_tbl %>% filter(term != "(Intercept)", p.value < 0.05) %>% arrange(metric, p.value)
if (nrow(alpha_sig) > 0) {
  cat("Significant fixed effects detected in alpha models by metric:\n")
  alpha_sig %>% group_by(metric) %>% summarize(terms = paste(unique(term), collapse = ", ")) %>%
    mutate(text = paste0("- ", metric, ": ", terms)) %>% pull(text) %>% cat(sep = "\n")
} else {
  cat("No fixed effects reached p < 0.05 across alpha metrics; see EMMs for trends.\n")
}

# Inline: Shannon carbohydrate contrasts by timepoint × group (FDR-adjusted)
emm_tbl %>%
  filter(metric == "Shannon") %>%
  {
    cols <- intersect(c("timepoint","group","contrast","estimate","SE","df","t.ratio","p.value"), colnames(.))
    dplyr::select(., dplyr::all_of(cols))
  } %>%
  mutate(p.value = signif(p.value, 3)) %>%
  arrange(timepoint, group, contrast) %>%
  knitr::kable(caption = "Shannon: carbohydrate contrasts by timepoint × group (FDR-adjusted p)")
```

```{r alpha-plots, message=FALSE, warning=FALSE}
p_alpha <- alpha_df %>%
  ggplot(aes(x = carbohydrate, y = Shannon, fill = carbohydrate)) +
  geom_violin(trim = TRUE, alpha = 0.7) +
  geom_boxplot(width = 0.1, outlier.shape = NA) +
  geom_jitter(aes(color = timepoint), width = 0.15, height = 0, size = 1, alpha = 0.5) +
  facet_wrap(~ group) +
  scale_fill_brewer(palette = "Set3") +
  theme_bw() + labs(title = "Alpha Diversity (Shannon)", x = "Carbohydrate", y = "Shannon")
ggsave("outputs/figures/alpha_shannon.png", p_alpha, width = 8, height = 5, dpi = 300)
p_alpha
```

## Beta Diversity

```{r beta}
ps_rel_all <- transform_sample_counts(ps_filt, function(x) x/sum(x))
bray <- phyloseq::distance(ps_rel_all, "bray")
md_b <- data.frame(sample_data(ps_rel_all), stringsAsFactors = FALSE)

set.seed(42)
permanova <- vegan::adonis2(bray ~ group*carbohydrate*timepoint, data = md_b, strata = md_b$subjectID, permutations = 999)
capture.output(permanova, file = "outputs/tables/permanova_bray.txt")

# Show PERMANOVA table inline
as.data.frame(permanova) %>% tibble::rownames_to_column("term") %>%
  knitr::kable(caption = "PERMANOVA on Bray–Curtis distances (stratified by subject)")

ord <- ordinate(ps_rel_all, method = "PCoA", distance = "bray")
p_ord <- plot_ordination(ps_rel_all, ord, color = "carbohydrate", shape = "timepoint") +
  geom_point(size = 2, alpha = 0.85) + facet_wrap(~group) + theme_bw()
ggsave("outputs/figures/pcoa_bray.png", p_ord, width = 8, height = 5, dpi = 300)
p_ord
```

## Sensitivity: Collapsed Replicates

```{r sensitivity-collapsed}
# Collapse biological replicates to subject×carbohydrate×timepoint (cell)
ps_coll <- phyloseq::merge_samples(ps_filt, "cell")

# Rebuild minimal sample data for collapsed object
md_coll <- as.data.frame(sample_data(ps_filt)) %>%
  tibble::rownames_to_column("SampleID") %>%
  dplyr::group_by(cell) %>%
  dplyr::summarise(
    subjectID = dplyr::first(subjectID),
    group = dplyr::first(group),
    carbohydrate = dplyr::first(carbohydrate),
    timepoint = dplyr::first(timepoint),
    .groups = "drop"
  ) %>%
  as.data.frame()

# Set row names and create proper sample_data object
rownames(md_coll) <- md_coll$cell
md_coll$cell <- NULL  # Remove the cell column since it's now the row name
sample_data(ps_coll) <- sample_data(md_coll)
ps_coll_rel <- transform_sample_counts(ps_coll, function(x) x/sum(x))

bray_coll <- phyloseq::distance(ps_coll_rel, "bray")
md_bc <- sample_data(ps_coll_rel) %>%
  as("data.frame") %>%
  tibble::rownames_to_column("SampleID") %>%
  mutate(
    group = factor(group, levels = c("Healthy", "Obese")),
    carbohydrate = factor(carbohydrate, levels = c("No Added Carb", "RDC", "SDC")),
    timepoint = factor(timepoint, levels = c("0H", "48H", "72H")),
    subjectID = factor(subjectID)
  )

set.seed(42)
permanova_coll <- vegan::adonis2(bray_coll ~ group*carbohydrate*timepoint, data = md_bc, strata = md_bc$subjectID, permutations = 999)

as.data.frame(permanova_coll) %>% tibble::rownames_to_column("term") %>%
  knitr::kable(caption = "PERMANOVA (collapsed replicates) on Bray–Curtis distances")

# Compare three-way interaction between primary and collapsed
tri <- "group:carbohydrate:timepoint"
if (tri %in% rownames(permanova) && tri %in% rownames(permanova_coll)) {
  p1 <- permanova$`Pr(>F)`[tri]; r1 <- permanova$R2[tri]
  p2 <- permanova_coll$`Pr(>F)`[tri]; r2 <- permanova_coll$R2[tri]
  cat(glue::glue("Three-way interaction: primary R²={round(r1,3)}, p={signif(p1,3)}; collapsed R²={round(r2,3)}, p={signif(p2,3)}"))
}
```

## Differential Abundance

```{r da-ancombc2, message=FALSE, warning=FALSE}
# Primary DA: ANCOM-BC2 with random effects (optional)
if (requireNamespace("ANCOMBC", quietly = TRUE)) {
  set.seed(42)
  res_ab <- ANCOMBC::ancombc2(
    data         = ps_filt,
    fix_formula  = "group * carbohydrate * timepoint",
    rand_formula = "(1|subjectID) + (1|subjectID:cell)",
    group        = "group",
    global       = TRUE,
    p_adj_method = "fdr",
    prv_cut = 0.10, lib_cut = cfg$min_reads, alpha = 0.05, verbose = TRUE
  )
  saveRDS(res_ab, "outputs/da/ancombc2_results.rds")

  # Flatten and export
  if (!is.null(res_ab$res)) {
    res_df <- tryCatch({
      tibble::as_tibble(res_ab$res)
    }, error = function(e) {
      # Fall back to combining elements if list-like
      dplyr::bind_rows(lapply(res_ab, function(x) if (is.data.frame(x)) x else NULL), .id = "component")
    })
    readr::write_csv(res_df, "outputs/tables/ancombc2_all_results.csv")

    # Show a snapshot inline if likely columns exist
    poss <- intersect(c("taxon", "feature", "q_val", "q.value", "p_adj", "lfc", "effect"), colnames(res_df))
    if (length(poss) > 0) {
      cols <- unique(c("taxon", "feature", "lfc", "effect", "q_val", "q.value", "p_adj"))
      cols <- cols[cols %in% colnames(res_df)]
      res_df %>% select(dplyr::all_of(cols)) %>%
        mutate(across(matches("q_?val|q.value|p_adj"), ~signif(., 3))) %>%
        head(20) %>% knitr::kable(caption = "ANCOM-BC2 results (top 20; selected columns)")
    }
  }
} else {
  message("ANCOMBC not installed; skipping DA. See installation notes in the report.")
  readr::write_lines(
    "ANCOMBC not installed; DA step skipped. Install system libs: libgmp-dev, libmpfr-dev, then install.packages('CVXR') or BiocManager::install('ANCOMBC').",
    "outputs/tables/da_skipped.txt"
  )
}
```

```{r dynamic-interpretation}
interp_permanova <- function(perma){
  rn <- rownames(perma)
  tri <- "group:carbohydrate:timepoint"
  if (!tri %in% rn) return("PERMANOVA summary unavailable.")
  r2 <- perma$R2[tri]; p <- perma$`Pr(>F)`[tri]
  if(!is.na(p) && p < 0.05) glue::glue("Community composition differed by the group×carbohydrate×timepoint interaction (R² = {round(r2,3)}, p = {signif(p,3)}).")
  else "No significant three-way interaction; see main effects."
}

cat(interp_permanova(permanova))

# Also list any main or two-way effects with p < 0.05
perm_df <- as.data.frame(permanova) %>% tibble::rownames_to_column("term")
perm_sig <- perm_df %>% filter(!term %in% c("Df", "Residual"), `Pr(>F)` < 0.05)
if (nrow(perm_sig) > 0) {
  cat("\nAdditional PERMANOVA effects with p < 0.05:\n")
  perm_sig %>% transmute(text = paste0("- ", term, ": R²=", round(R2, 3), ", p=", signif(`Pr(>F)`, 3))) %>%
    pull(text) %>% cat(sep = "\n")
}
```

## Session Info

```{r session}
sessionInfo()
```
