---
title: "Microbiome Analysis v2 - ASV Finder"
author: "Abbott Carbohydrate Obesity Project"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
    highlight: tango
    fig_width: 10
    fig_height: 6
    df_print: paged
---

```{r library-loading}
# Helper function to install CRAN or Bioconductor packages if not already installed
install_if_missing <- function(pkg) {
  
  if (!requireNamespace(pkg, quietly = TRUE)) {
    tryCatch({
      install.packages(pkg)
    }, error = function(e) {
      if (!requireNamespace("BiocManager", quietly = TRUE)) {
        install.packages("BiocManager")
      }
      BiocManager::install(pkg, ask = FALSE)
    })
    
  }
}


# List of required packages
required_pkgs <- c(
  "tidyverse", "tidytext", "tidygraph", "ggraph", "igraph",
  "phyloseq", "microbiome", "corncob", "ggpubr", "cowplot",
  "ANCOMBC", "microbiomeMarker", "lme4", "lmerTest", "broom.mixed",
  "emmeans", "MuMIn", "ggsci", "scico"
)

# Install missing packages
invisible(lapply(required_pkgs, install_if_missing))

# Load the libraries
library(tidyverse)
library(tidytext)
library(tidygraph)
library(ggraph)
library(igraph)
library(phyloseq)
library(microbiome)
library(corncob)
library(ggpubr)
library(cowplot)
library(ANCOMBC)
library(microbiomeMarker)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(emmeans)
library(MuMIn)
library(ggsci)
library(scico)
```

```{r data-loading}
samdat <- read_tsv("sample.metadata.tsv")
colnames(samdat) <- c("SampleID", "sample_label", "raw_seq_file", "group")

# Process sample_label to extract additional information
samdat <- samdat %>%
  separate(sample_label, into = c("subjectID", "replicate_code", "timepoint"), sep = "\\.", remove = FALSE) %>%
  mutate(
    # Extract carbohydrate type from replicate_code
    carbohydrate = case_when(
      str_detect(replicate_code, "R") ~ "RDC",
      str_detect(replicate_code, "S") ~ "SDC",
      TRUE ~ "No Added Carb"
    ) %>% factor(levels = c("No Added Carb", "RDC", "SDC")),
    # Extract replicate number from replicate_code
    replicate = str_extract(replicate_code, "\\d+") %>% as.numeric(),
    # Replace NA values in replicate with 1
    replicate = ifelse(is.na(replicate), 1, replicate)
  ) %>%
  select(-replicate_code, -group)  # Remove the intermediate replicate_code column

# Display the processed metadata
print(samdat)

# now get the group information for health and obese

group <- read_tsv("zr24558.16S_250813.zymo/1.Healthy.TimepointComparism..Bac16Sv34/Sample_Information/sample.metadata.csv")

# Extract subjectID from customer_label
group <- group %>%
  separate(customer_label, into = c("subjectID", "replicate_code", "timepoint"), sep = "\\.", remove = FALSE) %>%
  select(customer_label, subjectID, everything())

# Check the structure of group data
print("Group data structure:")
print(colnames(group))
print(head(group))

samdat$group <- ifelse(samdat$subjectID %in% group$subjectID, "Healthy", "Obese")

# now output a summary of the metadata in samdat
# summarize samdat by group and carbohydrate
samdat_summary <- samdat %>%
  group_by(group, carbohydrate) %>%
  summarize(n_samples = n())

samdat_summary

# number of unique subjects per group
samdat %>%
  group_by(group) %>%
  summarize(n_subjects = n_distinct(subjectID))


write_csv(samdat, file = "samdat.csv")
```

```{r import-asv-finder-data-and-phyloseq-object}
asv_data <- read_tsv("dada2_asv_2/asv_table.tsv")

# asv_data <- asv_data %>% 
#   as.data.frame() %>% 
#   select(-ASV_ID, -Sequence, -Length, -Total_Abundance, -Relative_Abundance) 

# Process sample names more carefully
original_names <- colnames(asv_data)
cat("Original sample names (first 10):", paste(head(original_names, 10), collapse = ", "), "\n")

# Split and recombine names
processed_names <- str_split(original_names, "_") %>% sapply(function(x) paste(x[1:2], collapse = "_"))
cat("Processed sample names (first 10):", paste(head(processed_names, 10), collapse = ", "), "\n")

# Check for duplicates before assignment
if(any(duplicated(processed_names))) {
  cat("Warning: Processing created duplicate names. Original duplicates:", sum(duplicated(processed_names)), "\n")
  cat("Duplicate names:", paste(processed_names[duplicated(processed_names)], collapse = ", "), "\n")
}

colnames(asv_data) <- processed_names

# Check for duplicate column names and make them unique
if(any(duplicated(colnames(asv_data)))) {
  cat("Warning: Found duplicate sample names after processing. Making them unique.\n")
  colnames(asv_data) <- make.unique(colnames(asv_data), sep = "_")
  cat("Final sample names (first 10):", paste(head(colnames(asv_data), 10), collapse = ", "), "\n")
}

# rownames(asv_data) <- asv

taxonomy_data <- read_tsv("dada2_asv_2/strobelca_taxonomy.tsv")
asv <- taxonomy_data$ASV_ID %>% str_split(" ") %>% sapply(function(x) x[1])
repseqs <- Biostrings::DNAStringSet(taxonomy_data$Sequence)
names(repseqs) <- asv
taxonomy_data <- taxonomy_data %>% 
  as.data.frame() %>% 
  select(-ASV_ID, -Sequence, -Confidence, -Method, -Identity, -Coverage, -Reference_ID) %>% 
  as.matrix() 
rownames(taxonomy_data) <- asv

samdat <- samdat %>% 
  as.data.frame() 

# Check for duplicate SampleIDs and make them unique if needed
if(any(duplicated(samdat$SampleID))) {
  cat("Warning: Found duplicate SampleIDs. Making them unique.\n")
  samdat$SampleID <- make.unique(samdat$SampleID, sep = "_")
}

rownames(samdat) <- samdat$SampleID

# Check data consistency between ASV data and sample metadata
asv_samples <- colnames(asv_data)
metadata_samples <- samdat$SampleID

# Find samples that are in ASV data but not in metadata
missing_in_metadata <- setdiff(asv_samples, metadata_samples)
if(length(missing_in_metadata) > 0) {
  cat("Warning: Samples in ASV data but not in metadata:", paste(missing_in_metadata, collapse = ", "), "\n")
}

# Find samples that are in metadata but not in ASV data
missing_in_asv <- setdiff(metadata_samples, asv_samples)
if(length(missing_in_asv) > 0) {
  cat("Warning: Samples in metadata but not in ASV data:", paste(missing_in_asv, collapse = ", "), "\n")
}

# Keep only samples that are in both datasets
common_samples <- intersect(asv_samples, metadata_samples)
cat("Number of samples in both datasets:", length(common_samples), "\n")

# Filter both datasets to common samples
asv_data <- asv_data[, common_samples, drop = FALSE]
samdat <- samdat[common_samples, , drop = FALSE]

# Final check: ensure sample names match exactly between ASV data and metadata
if(!identical(colnames(asv_data), rownames(samdat))) {
  cat("Warning: Sample names don't match exactly. Aligning them.\n")
  # Reorder metadata to match ASV data column order
  samdat <- samdat[colnames(asv_data), , drop = FALSE]
}

# Final check for duplicates before creating phyloseq object
if(any(duplicated(colnames(asv_data)))) {
  cat("Warning: Still have duplicate ASV column names. Making unique.\n")
  colnames(asv_data) <- make.unique(colnames(asv_data), sep = "_")
  rownames(samdat) <- colnames(asv_data)  # Update metadata rownames to match
}

if(any(duplicated(rownames(samdat)))) {
  cat("Warning: Still have duplicate metadata row names. Making unique.\n")
  rownames(samdat) <- make.unique(rownames(samdat), sep = "_")
  colnames(asv_data) <- rownames(samdat)  # Update ASV column names to match
}

asv_data <- asv_data %>% 
  as.data.frame() 
rownames(asv_data) <- rownames(taxonomy_data)

# now make the phyloseq object
phy_asv_finder <- phyloseq(
  otu_table(asv_data, taxa_are_rows = TRUE),
  tax_table(taxonomy_data),
  sample_data(samdat)
)

save(phy_asv_finder, file = "phy_asv_finder.RData")
```

```{r alpha-diversity-by-group, fig.width = 14, fig.height = 8}
alpha_diversity <- estimate_richness(phy_asv_finder, measures = c("Shannon", "Simpson", "Observed"))
# merge alpha_diversity with samdat
alpha_diversity <- alpha_diversity %>% 
  rownames_to_column(var = "SampleID") %>% 
  left_join(samdat, by = "SampleID")


# plot alpha_diversity with publication quality theme, colors, and Wilcoxon stats
library(ggpubr)

# Draw a stats bracket between the groups being compared using stat_compare_means with comparisons argument
group_levels <- unique(alpha_diversity$group)
if (length(group_levels) == 2) {
  comparisons <- list(group_levels)
} else {
  # For more than 2 groups, compare all pairs
  comparisons <- combn(group_levels, 2, simplify = FALSE)
}

library(cowplot)

# Shannon panel
p_shannon <- ggplot(alpha_diversity, aes(x = group, y = Shannon, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 16) +
  labs(x = "Group", y = "Shannon Diversity", fill = "Group") +
  stat_compare_means(
    comparisons = comparisons,
    method = "wilcox.test",
    label = "p.signif",
    label.y = seq(
      from = max(alpha_diversity$Shannon, na.rm = TRUE) * 1.05,
      by = 0.1 * max(alpha_diversity$Shannon, na.rm = TRUE),
      length.out = length(comparisons)
    )
  ) +
  theme(
    legend.position = "none",
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()
  )

# Observed panel
p_observed <- ggplot(alpha_diversity, aes(x = group, y = Observed, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 16) +
  labs(x = "Group", y = "Observed ASVs", fill = "Group") +
  stat_compare_means(
    comparisons = comparisons,
    method = "wilcox.test",
    label = "p.signif",
    label.y = seq(
      from = max(alpha_diversity$Observed, na.rm = TRUE) * 1.05,
      by = 0.1 * max(alpha_diversity$Observed, na.rm = TRUE),
      length.out = length(comparisons)
    )
  ) +
  theme(
    legend.position = "none",
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()
  )

# Simpson panel
p_simpson <- ggplot(alpha_diversity, aes(x = group, y = Simpson, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 16) +
  labs(x = "Group", y = "Simpson Diversity", fill = "Group") +
  stat_compare_means(
    comparisons = comparisons,
    method = "wilcox.test",
    label = "p.signif",
    label.y = seq(
      from = max(alpha_diversity$Simpson, na.rm = TRUE) * 1.05,
      by = 0.1 * max(alpha_diversity$Simpson, na.rm = TRUE),
      length.out = length(comparisons)
    )
  ) +
  theme(
    legend.position = "none",
    axis.text = element_text(color = "black"),
    panel.grid = element_blank()
  )

# Combine panels into a multipanel figure
plot_grid(
  p_observed,
  p_shannon,
  p_simpson,
  labels = c("A", "B", "C"),
  nrow = 1,
  align = "v"
)
```

```{r alpha-diversity-by-group-and-timepoint}
alpha_diversity <- estimate_richness(phy_asv_finder, measures = c("Shannon", "Simpson", "Observed"))
# merge alpha_diversity with samdat
alpha_diversity <- alpha_diversity %>%
  rownames_to_column(var = "SampleID") %>%
  left_join(samdat, by = "SampleID")

# Get unique levels for all variables
group_levels <- unique(alpha_diversity$group)
time_levels <- unique(alpha_diversity$timepoint)
carbohydrate_levels <- unique(alpha_diversity$carbohydrate)

# Set up comparisons for different variables
group_comparisons <- if (length(group_levels) == 2) {
  list(group_levels)
} else {
  combn(group_levels, 2, simplify = FALSE)
}

carb_comparisons <- if (length(carbohydrate_levels) == 2) {
  list(carbohydrate_levels)
} else {
  combn(carbohydrate_levels, 2, simplify = FALSE)
}

time_comparisons <- if (length(time_levels) == 2) {
  list(time_levels)
} else {
  combn(time_levels, 2, simplify = FALSE)
}

# === BY GROUP ===
# Shannon by group
p_shannon_group <- ggplot(alpha_diversity, aes(x = group, y = Shannon, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 12) +
  labs(x = "Group", y = "Shannon Diversity", title = "By Group") +
  stat_compare_means(
    comparisons = group_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme(legend.position = "none", axis.text = element_text(color = "black"))

# Observed by group
p_observed_group <- ggplot(alpha_diversity, aes(x = group, y = Observed, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 12) +
  labs(x = "Group", y = "Observed ASVs", title = "By Group") +
  stat_compare_means(
    comparisons = group_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme(legend.position = "none", axis.text = element_text(color = "black"))

# === BY CARBOHYDRATE ===
# Shannon by carbohydrate
p_shannon_carb <- ggplot(alpha_diversity, aes(x = carbohydrate, y = Shannon, fill = carbohydrate)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_nejm() +
  theme_classic(base_size = 12) +
  labs(x = "Carbohydrate", y = "Shannon Diversity", title = "By Carbohydrate") +
  stat_compare_means(
    comparisons = carb_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme(legend.position = "none", axis.text = element_text(color = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Observed by carbohydrate
p_observed_carb <- ggplot(alpha_diversity, aes(x = carbohydrate, y = Observed, fill = carbohydrate)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_nejm() +
  theme_classic(base_size = 12) +
  labs(x = "Carbohydrate", y = "Observed ASVs", title = "By Carbohydrate") +
  stat_compare_means(
    comparisons = carb_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme(legend.position = "none", axis.text = element_text(color = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1))

# === BY TIMEPOINT ===
# Shannon by timepoint
p_shannon_time <- ggplot(alpha_diversity, aes(x = timepoint, y = Shannon, fill = timepoint)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_lancet() +
  theme_classic(base_size = 12) +
  labs(x = "Timepoint", y = "Shannon Diversity", title = "By Timepoint") +
  stat_compare_means(
    comparisons = time_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme(legend.position = "none", axis.text = element_text(color = "black"))

# Observed by timepoint
p_observed_time <- ggplot(alpha_diversity, aes(x = timepoint, y = Observed, fill = timepoint)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_lancet() +
  theme_classic(base_size = 12) +
  labs(x = "Timepoint", y = "Observed ASVs", title = "By Timepoint") +
  stat_compare_means(
    comparisons = time_comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme(legend.position = "none", axis.text = element_text(color = "black"))

# === INTERACTION PLOTS ===
# Shannon by group and timepoint
p_shannon_interaction <- ggplot(alpha_diversity, aes(x = timepoint, y = Shannon, fill = group)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.7, alpha = 0.8) +
  geom_point(position = position_jitterdodge(dodge.width = 0.8, jitter.width = 0.2),
             size = 1.2, alpha = 0.6) +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 12) +
  labs(x = "Timepoint", y = "Shannon Diversity", fill = "Group",
       title = "Group × Timepoint Interaction") +
  theme(axis.text = element_text(color = "black"))

# Observed by group and timepoint
p_observed_interaction <- ggplot(alpha_diversity, aes(x = timepoint, y = Observed, fill = group)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.7, alpha = 0.8) +
  geom_point(position = position_jitterdodge(dodge.width = 0.8, jitter.width = 0.2),
             size = 1.2, alpha = 0.6) +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 12) +
  labs(x = "Timepoint", y = "Observed ASVs", fill = "Group",
       title = "Group × Timepoint Interaction") +
  theme(axis.text = element_text(color = "black"))

# Combine all plots
plot_grid(
  p_shannon_group, p_shannon_carb, p_shannon_time, p_shannon_interaction,
  p_observed_group, p_observed_carb, p_observed_time, p_observed_interaction,
  labels = c("A", "B", "C", "D", "E", "F", "G", "H"),
  nrow = 2,
  align = "hv"
)
```

```{r alpha-diversity-48hr-compare-carbohydrates, fig.width = 14, fig.height = 24}
# Filter data for 48H timepoint only
alpha_diversity_48h <- alpha_diversity %>%
  filter(timepoint == "48H")

# Get unique carbohydrate levels for 48H data
carb_levels_48h <- unique(alpha_diversity_48h$carbohydrate)

# Set up comparisons for carbohydrate groups at 48H
carb_comparisons_48h <- if (length(carb_levels_48h) == 2) {
  list(carb_levels_48h)
} else {
  combn(carb_levels_48h, 2, simplify = FALSE)
}

# === 48H TIMEPOINT CARBOHYDRATE COMPARISONS ===
# Shannon by carbohydrate at 48H
p_shannon_carb_48h <- ggplot(alpha_diversity_48h, aes(x = carbohydrate, y = Shannon, fill = carbohydrate)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_nejm() +
  theme_classic(base_size = 16) +
  labs(x = "Carbohydrate", y = "Shannon Diversity", title = "Shannon Diversity at 48H by Carbohydrate") +
  stat_compare_means(
    comparisons = carb_comparisons_48h,
    method = "wilcox.test",
    label = "p.signif",
    label.y = seq(
      from = max(alpha_diversity_48h$Shannon, na.rm = TRUE) * 1.05,
      by = 0.1 * max(alpha_diversity_48h$Shannon, na.rm = TRUE),
      length.out = length(carb_comparisons_48h)
    )
  ) +
  theme(
    legend.position = "none",
    axis.text = element_text(color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )

# Observed by carbohydrate at 48H
p_observed_carb_48h <- ggplot(alpha_diversity_48h, aes(x = carbohydrate, y = Observed, fill = carbohydrate)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "black") +
  ggsci::scale_fill_nejm() +
  theme_classic(base_size = 16) +
  labs(x = "Carbohydrate", y = "Observed ASVs", title = "Observed ASVs at 48H by Carbohydrate") +
  stat_compare_means(
    comparisons = carb_comparisons_48h,
    method = "wilcox.test",
    label = "p.signif",
    label.y = seq(
      from = max(alpha_diversity_48h$Observed, na.rm = TRUE) * 1.05,
      by = 0.1 * max(alpha_diversity_48h$Observed, na.rm = TRUE),
      length.out = length(carb_comparisons_48h)
    )
  ) +
  theme(
    legend.position = "none",
    axis.text = element_text(color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )

# Shannon by carbohydrate and group at 48H (interaction)
p_shannon_carb_group_48h <- ggplot(alpha_diversity_48h, aes(x = carbohydrate, y = Shannon, fill = group)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.7, alpha = 0.8) +
  geom_point(position = position_jitterdodge(dodge.width = 0.8, jitter.width = 0.2),
             size = 1.2, alpha = 0.6) +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 16) +
  labs(x = "Carbohydrate", y = "Shannon Diversity", fill = "Group",
       title = "Shannon Diversity at 48H: Carbohydrate × Group Interaction") +
  theme(
    axis.text = element_text(color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Observed by carbohydrate and group at 48H (interaction)
p_observed_carb_group_48h <- ggplot(alpha_diversity_48h, aes(x = carbohydrate, y = Observed, fill = group)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.7, alpha = 0.8) +
  geom_point(position = position_jitterdodge(dodge.width = 0.8, jitter.width = 0.2),
             size = 1.2, alpha = 0.6) +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 16) +
  labs(x = "Carbohydrate", y = "Observed ASVs", fill = "Group",
       title = "Observed ASVs at 48H: Carbohydrate × Group Interaction") +
  theme(
    axis.text = element_text(color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Combine all 48H plots
plot_grid(
  p_shannon_carb_48h,
  p_observed_carb_48h,
  p_shannon_carb_group_48h,
  p_observed_carb_group_48h,
  labels = c("A", "B", "C", "D"),
  nrow = 2,
  ncol = 2,
  align = "hv"
)
```

- Group comparisons: Using Wilcoxon rank-sum tests, we did `r ifelse(wilcox.test(Shannon ~ group, data = subset(alpha_diversity, !is.na(Shannon) & !is.na(group)))$p.value < 0.05, "detect", "not detect")` a statistically significant difference in Shannon diversity between groups (`p = `r format.pval(wilcox.test(Shannon ~ group, data = subset(alpha_diversity, !is.na(Shannon) & !is.na(group)))$p.value, digits = 2)`), and `r ifelse(wilcox.test(Observed ~ group, data = subset(alpha_diversity, !is.na(Observed) & !is.na(group)))$p.value < 0.05, "detect", "not detect")` a significant difference for Observed ASVs (`p = `r format.pval(wilcox.test(Observed ~ group, data = subset(alpha_diversity, !is.na(Observed) & !is.na(group)))$p.value, digits = 2)``). Visual patterns in panels A and E align with these results.

- Carbohydrate comparisons: Because there are three conditions (No Added Carb, RDC, SDC), we summarize differences with Kruskal–Wallis tests. We observe significant overall effects of carbohydrate on Shannon diversity (`p = `r format.pval(kruskal.test(Shannon ~ carbohydrate, data = subset(alpha_diversity, !is.na(Shannon) & !is.na(carbohydrate)))$p.value, digits = 2)`) and on Observed ASVs (`p = `r format.pval(kruskal.test(Observed ~ carbohydrate, data = subset(alpha_diversity, !is.na(Observed) & !is.na(carbohydrate)))$p.value, digits = 2)`). Panels B and F illustrate these shifts across carbohydrate conditions.

- Timepoint comparisons: Wilcoxon tests indicate `r ifelse(wilcox.test(Shannon ~ timepoint, data = subset(alpha_diversity, !is.na(Shannon) & !is.na(timepoint)))$p.value < 0.05, "significant", "no statistically significant")` differences in Shannon (`p = `r format.pval(wilcox.test(Shannon ~ timepoint, data = subset(alpha_diversity, !is.na(Shannon) & !is.na(timepoint)))$p.value, digits = 2)`) and `r ifelse(wilcox.test(Observed ~ timepoint, data = subset(alpha_diversity, !is.na(Observed) & !is.na(timepoint)))$p.value < 0.05, "significant", "no statistically significant")` differences in Observed ASVs (`p = `r format.pval(wilcox.test(Observed ~ timepoint, data = subset(alpha_diversity, !is.na(Observed) & !is.na(timepoint)))$p.value, digits = 2)`). Panels C and G show broadly similar distributions across timepoints.

- Group × timepoint: The interaction panels (D, H) show limited divergence between groups across time, consistent with small between-group effects relative to within-subject variability. Formal inference for interactions is provided in the mixed-effects section below.


```{r alpha-diversity-groups-by-carbohydrate, message=FALSE, warning=FALSE}
# Observed ASVs: groups side-by-side within each carbohydrate type
# Reuse alpha_diversity created above
library(ggpubr)

# Ensure expected ordering for readability
alpha_diversity <- alpha_diversity %>%
  mutate(
    carbohydrate = factor(carbohydrate,
                          levels = sort(unique(carbohydrate))),
    group = factor(group, levels = sort(unique(group)))
  )

# Define comparison between groups (assumes two groups)
grp_levels <- levels(alpha_diversity$group)
grp_comparisons <- if (length(grp_levels) == 2) list(grp_levels) else combn(grp_levels, 2, simplify = FALSE)

p_obs_grp_by_carb <- ggplot(alpha_diversity,
                            aes(x = group, y = Observed, fill = group)) +
  geom_boxplot(outlier.shape = NA, width = 0.65, alpha = 0.85) +
  geom_jitter(width = 0.15, size = 1.3, alpha = 0.45, color = "black") +
  ggsci::scale_fill_npg() +
  theme_classic(base_size = 14) +
  labs(x = "Group", y = "Observed ASVs", title = "Observed ASVs by Carbohydrate (Groups Side-by-Side)") +
  facet_wrap(~ carbohydrate, nrow = 1) +
  stat_compare_means(comparisons = grp_comparisons,
                     method = "wilcox.test",
                     label = "p.signif") +
  theme(
    legend.position = "none",
    axis.text = element_text(color = "black"),
    strip.background = element_rect(fill = "grey95", color = NA),
    strip.text = element_text(color = "black")
  )

p_obs_grp_by_carb
```

```{r mixed-effects-alpha-diversity}
# Load required packages for mixed effects models
library(lme4)
library(lmerTest)
library(broom.mixed)

# Ensure SubjectID exists and is a factor (metadata uses `subjectID`)
if (!"SubjectID" %in% names(alpha_diversity)) {
  if ("subjectID" %in% names(alpha_diversity)) {
    alpha_diversity <- alpha_diversity %>% mutate(SubjectID = as.factor(subjectID))
  } else {
    stop("SubjectID/subjectID not found in alpha_diversity; check metadata join.")
  }
} else {
  alpha_diversity$SubjectID <- as.factor(alpha_diversity$SubjectID)
}

# Create models for each diversity metric
# Shannon Diversity Mixed Effects Model
model_shannon <- lmer(Shannon ~ group * timepoint * carbohydrate + (1|SubjectID),
                      data = alpha_diversity)

# Observed ASVs Mixed Effects Model
model_observed <- lmer(Observed ~ group * timepoint * carbohydrate + (1|SubjectID),
                       data = alpha_diversity)

# Simpson Diversity Mixed Effects Model
model_simpson <- lmer(Simpson ~ group * timepoint * carbohydrate + (1|SubjectID),
                      data = alpha_diversity)

# Function to extract and format results
extract_mixed_model_results <- function(model, metric_name) {
  # Get ANOVA table with p-values
  anova_results <- anova(model)

  # Get fixed effects coefficients
  fixed_effects <- broom.mixed::tidy(model, effects = "fixed")

  # Get random effects variance
  random_effects <- broom.mixed::tidy(model, effects = "ran_pars")

  # Create a summary data frame
  summary_df <- data.frame(
    Metric = metric_name,
    Term = rownames(anova_results),
    F_value = anova_results$`F value`,
    p_value = anova_results$`Pr(>F)`,
    significance = ifelse(anova_results$`Pr(>F)` < 0.001, "***",
                         ifelse(anova_results$`Pr(>F)` < 0.01, "**",
                               ifelse(anova_results$`Pr(>F)` < 0.05, "*", ""))),
    stringsAsFactors = FALSE
  )

  return(list(
    anova = summary_df,
    fixed_effects = fixed_effects,
    random_effects = random_effects,
    model = model
  ))
}

# Extract results for all models
shannon_results <- extract_mixed_model_results(model_shannon, "Shannon")
observed_results <- extract_mixed_model_results(model_observed, "Observed")
simpson_results <- extract_mixed_model_results(model_simpson, "Simpson")

# Combine ANOVA results for display
all_anova_results <- rbind(
  shannon_results$anova,
  observed_results$anova,
  simpson_results$anova
)

# Display results
cat("=== MIXED EFFECTS MODEL RESULTS ===\n")
cat("(SubjectID as random effect)\n\n")

print(all_anova_results)

# Post-hoc comparisons for significant effects
library(emmeans)

# Function to perform post-hoc comparisons
perform_posthoc <- function(model, metric_name) {
  cat(paste("\n=== POST-HOC COMPARISONS FOR", metric_name, "===\n"))

  # Group comparisons
  if("group" %in% names(coef(model)$SubjectID)) {
    group_emm <- emmeans(model, ~ group)
    group_contrasts <- pairs(group_emm, adjust = "bonferroni")
    cat("\nGroup comparisons:\n")
    print(group_contrasts)
  }

  # Timepoint comparisons
  if("timepoint" %in% names(coef(model)$SubjectID)) {
    time_emm <- emmeans(model, ~ timepoint)
    time_contrasts <- pairs(time_emm, adjust = "bonferroni")
    cat("\nTimepoint comparisons:\n")
    print(time_contrasts)
  }

  # Carbohydrate comparisons
  if("carbohydrate" %in% names(coef(model)$SubjectID)) {
    carb_emm <- emmeans(model, ~ carbohydrate)
    carb_contrasts <- pairs(carb_emm, adjust = "bonferroni")
    cat("\nCarbohydrate comparisons:\n")
    print(carb_contrasts)
  }

  # Two-way interactions if significant
  # Group × Timepoint
  tryCatch({
    group_time_emm <- emmeans(model, ~ group * timepoint)
    group_time_contrasts <- pairs(group_time_emm, adjust = "bonferroni")
    cat("\nGroup × Timepoint interaction:\n")
    print(group_time_contrasts)
  }, error = function(e) {})
}

# Perform post-hoc tests for models with significant effects
if(any(shannon_results$anova$p_value < 0.05, na.rm = TRUE)) {
  perform_posthoc(model_shannon, "SHANNON")
}

if(any(observed_results$anova$p_value < 0.05, na.rm = TRUE)) {
  perform_posthoc(model_observed, "OBSERVED")
}

if(any(simpson_results$anova$p_value < 0.05, na.rm = TRUE)) {
  perform_posthoc(model_simpson, "SIMPSON")
}

# Model diagnostics
cat("\n=== MODEL DIAGNOSTICS ===\n")

# Check model assumptions with plots
par(mfrow = c(2, 2))

# Shannon model diagnostics
plot(model_shannon, main = "Shannon: Residuals vs Fitted")
qqnorm(resid(model_shannon), main = "Shannon: Q-Q Plot")
qqline(resid(model_shannon))

# Observed model diagnostics
plot(model_observed, main = "Observed: Residuals vs Fitted")
qqnorm(resid(model_observed), main = "Observed: Q-Q Plot")
qqline(resid(model_observed))

par(mfrow = c(1, 1))

# Random effects summary
cat("\n=== RANDOM EFFECTS (SubjectID) ===\n")
cat("Shannon model - Subject variance:",
    shannon_results$random_effects$estimate[shannon_results$random_effects$term == "sd__(Intercept)"]^2, "\n")
cat("Observed model - Subject variance:",
    observed_results$random_effects$estimate[observed_results$random_effects$term == "sd__(Intercept)"]^2, "\n")
cat("Simpson model - Subject variance:",
    simpson_results$random_effects$estimate[simpson_results$random_effects$term == "sd__(Intercept)"]^2, "\n")

# Effect sizes (pseudo R-squared for mixed models)
library(MuMIn)
cat("\n=== MODEL FIT (R-squared) ===\n")
cat("Shannon R²:", round(r.squaredGLMM(model_shannon)[1], 3), "(marginal),",
    round(r.squaredGLMM(model_shannon)[2], 3), "(conditional)\n")
cat("Observed R²:", round(r.squaredGLMM(model_observed)[1], 3), "(marginal),",
    round(r.squaredGLMM(model_observed)[2], 3), "(conditional)\n")
cat("Simpson R²:", round(r.squaredGLMM(model_simpson)[1], 3), "(marginal),",
    round(r.squaredGLMM(model_simpson)[2], 3), "(conditional)\n")

cat("\nNote: Marginal R² = variance explained by fixed effects\n")
cat("      Conditional R² = variance explained by fixed + random effects\n")
```

### Interpretation of Mixed-Effects Alpha-Diversity Models

#### Fixed Effects: Experimental Treatment Impacts

Across all three alpha-diversity metrics, **carbohydrate condition was the only fixed effect that consistently explained variance** in community diversity. With SubjectID modeled as a random intercept, the main effect of carbohydrate was significant for Shannon (F = `r round(shannon_results$anova$F_value[shannon_results$anova$Term == "carbohydrate"], 2)`, p = `r format.pval(shannon_results$anova$p_value[shannon_results$anova$Term == "carbohydrate"], digits = 2)`), Observed ASVs (F = `r round(observed_results$anova$F_value[observed_results$anova$Term == "carbohydrate"], 2)`, p = `r format.pval(observed_results$anova$p_value[observed_results$anova$Term == "carbohydrate"], digits = 2)`), and Simpson (F = `r round(simpson_results$anova$F_value[simpson_results$anova$Term == "carbohydrate"], 2)`, p = `r format.pval(simpson_results$anova$p_value[simpson_results$anova$Term == "carbohydrate"], digits = 2)`). 

**Timepoint effects** were `r ifelse(observed_results$anova$p_value[observed_results$anova$Term == "timepoint"] < 0.05, "significant", "not significant")` for Observed ASVs (F = `r round(observed_results$anova$F_value[observed_results$anova$Term == "timepoint"], 2)`, p = `r format.pval(observed_results$anova$p_value[observed_results$anova$Term == "timepoint"], digits = 2)`) but `r ifelse(shannon_results$anova$p_value[shannon_results$anova$Term == "timepoint"] < 0.05, "significant", "not significant")` for Shannon (F = `r round(shannon_results$anova$F_value[shannon_results$anova$Term == "timepoint"], 2)`, p = `r format.pval(shannon_results$anova$p_value[shannon_results$anova$Term == "timepoint"], digits = 2)`) and Simpson (F = `r round(simpson_results$anova$F_value[simpson_results$anova$Term == "timepoint"], 2)`, p = `r format.pval(simpson_results$anova$p_value[simpson_results$anova$Term == "timepoint"], digits = 2)`).

**Interaction effects** revealed that the timepoint-by-carbohydrate interaction was `r ifelse(shannon_results$anova$p_value[shannon_results$anova$Term == "timepoint:carbohydrate"] < 0.05, "significant", "not significant")` for Shannon (F = `r round(shannon_results$anova$F_value[shannon_results$anova$Term == "timepoint:carbohydrate"], 2)`, p = `r format.pval(shannon_results$anova$p_value[shannon_results$anova$Term == "timepoint:carbohydrate"], digits = 2)`) and `r ifelse(observed_results$anova$p_value[observed_results$anova$Term == "timepoint:carbohydrate"] < 0.05, "significant", "not significant")` for Observed ASVs (F = `r round(observed_results$anova$F_value[observed_results$anova$Term == "timepoint:carbohydrate"], 2)`, p = `r format.pval(observed_results$anova$p_value[observed_results$anova$Term == "timepoint:carbohydrate"], digits = 2)`), indicating temporal modulation of carbohydrate effects. In contrast, the **group main effect did `r ifelse(all(c(shannon_results$anova$p_value[shannon_results$anova$Term == "group"], observed_results$anova$p_value[observed_results$anova$Term == "group"], simpson_results$anova$p_value[simpson_results$anova$Term == "group"]) >= 0.05), "not", "")` reach conventional significance** across metrics (Shannon p = `r format.pval(shannon_results$anova$p_value[shannon_results$anova$Term == "group"], digits = 2)`, Observed p = `r format.pval(observed_results$anova$p_value[observed_results$anova$Term == "group"], digits = 2)`, Simpson p = `r format.pval(simpson_results$anova$p_value[simpson_results$anova$Term == "group"], digits = 2)`).

#### Random Effects: Individual Differences Dominate

Random-effects estimates indicated **substantial inter-individual heterogeneity**. The estimated SubjectID variance components were Shannon = `r signif(shannon_results$random_effects$estimate[shannon_results$random_effects$term == "sd__(Intercept)"]^2, 3)` (SD = `r round(sqrt(shannon_results$random_effects$estimate[shannon_results$random_effects$term == "sd__(Intercept)"]^2), 3)` Shannon units), Observed = `r signif(observed_results$random_effects$estimate[observed_results$random_effects$term == "sd__(Intercept)"]^2, 3)` (SD = `r round(sqrt(observed_results$random_effects$estimate[observed_results$random_effects$term == "sd__(Intercept)"]^2), 0)` ASVs), and Simpson = `r signif(simpson_results$random_effects$estimate[simpson_results$random_effects$term == "sd__(Intercept)"]^2, 3)` (SD = `r round(sqrt(simpson_results$random_effects$estimate[simpson_results$random_effects$term == "sd__(Intercept)"]^2), 4)` Simpson units). 

This individual variation represents **`r round((1 - r.squaredGLMM(model_shannon)[1]) * 100, 1)`% of total variance** for Shannon, **`r round((1 - r.squaredGLMM(model_observed)[1]) * 100, 1)`% for Observed ASVs**, and **`r round((1 - r.squaredGLMM(model_simpson)[1]) * 100, 1)`% for Simpson diversity**. Consistent with this, conditional R² exceeded marginal R² for all metrics (Shannon: `r round(r.squaredGLMM(model_shannon)[2], 3)` vs `r round(r.squaredGLMM(model_shannon)[1], 3)`; Observed: `r round(r.squaredGLMM(model_observed)[2], 3)` vs `r round(r.squaredGLMM(model_observed)[1], 3)`; Simpson: `r round(r.squaredGLMM(model_simpson)[2], 3)` vs `r round(r.squaredGLMM(model_simpson)[1], 3)`), indicating that **subject-specific baselines account for the majority of total variance**.

#### Biological Interpretation

These results support several key conclusions:

1. **Carbohydrate condition (No Added Carb, RDC, SDC) is the primary experimental determinant** of alpha diversity, with consistent effects across all diversity metrics.

2. **Individual differences are substantial and expected** in microbiome studies, where baseline diversity varies considerably between people regardless of experimental treatments.

3. **The significant carbohydrate effects are biologically meaningful** despite high individual variation, suggesting that dietary carbohydrate type has measurable impacts on microbiome diversity that transcend individual differences.

4. **Temporal dynamics modulate carbohydrate effects** for Shannon diversity and richness, indicating that the microbiome response to carbohydrates changes over the 48-hour time course.

5. **Health status (Healthy vs Obese) alone does not predict alpha diversity**, suggesting that baseline microbiome diversity is not a simple function of metabolic health status in this study population.

The low marginal R² values (`r round(r.squaredGLMM(model_shannon)[1], 3)`-`r round(r.squaredGLMM(model_observed)[1], 3)`) are typical in microbiome studies where individual differences dominate, but the consistent significance of carbohydrate effects across metrics indicates robust treatment responses. Reporting adjusted pairwise contrasts among carbohydrate levels (e.g., estimated marginal means) will localize specific differences and provide effect sizes that are directly interpretable for clinical translation.

```{r beta-diversity}
library(microbiome)
library(ggplot2)
library(ggpubr)
library(cowplot)

# PCoA plot
# need to convert to relative abundance
phy_asv_finder_rel <- microbiome::transform(phy_asv_finder, "compositional")

# Filter out samples with zero reads (empty samples) before PCoA
sample_sums <- sample_sums(phy_asv_finder_rel)
empty_samples <- names(sample_sums)[sample_sums == 0]
cat("Removing", length(empty_samples), "empty samples for PCoA analysis\n")

# Create filtered phyloseq object for ordination
phy_asv_finder_rel_filtered <- prune_samples(sample_sums > 0, phy_asv_finder_rel)

# PCoA plot
pcoa <- ordinate(phy_asv_finder_rel_filtered, method = "PCoA", distance = "bray")
plot_ordination(phy_asv_finder_rel_filtered, pcoa, type = "samples", color = "group") +
  geom_point(size = 3) +
  ggsci::scale_color_npg() +
  facet_wrap(~timepoint) +
  theme_classic(base_size = 16)


```

```{r clustering-analysis-time0, fig.width = 12, fig.height = 8}
# Clustering analysis for time 0 samples
library(cluster)
library(factoextra)
library(RColorBrewer)
library(gridExtra)

# Filter data for time 0 only
phy_0h <- subset_samples(phy_asv_finder_rel_filtered, timepoint == "0H")

# Remove any taxa that are completely absent in 0H samples
phy_0h <- prune_taxa(taxa_sums(phy_0h) > 0, phy_0h)

cat("0H samples for clustering:", nsamples(phy_0h), "\n")

# Extract sample data
sample_data_0h <- data.frame(sample_data(phy_0h))

# Calculate Bray-Curtis distance matrix
dist_matrix <- phyloseq::distance(phy_0h, method = "bray")

# Perform clustering analysis with multiple methods
library(cluster)

cat("Testing different numbers of clusters using multiple validation methods...\n")

# Test different k values
k_values <- 2:6
silhouette_scores <- numeric(length(k_values))
pam_results <- list()
wss_scores <- numeric(length(k_values))  # Within-cluster sum of squares

for(i in seq_along(k_values)) {
  k <- k_values[i]
  pam_temp <- pam(dist_matrix, k = k, diss = TRUE)
  pam_results[[i]] <- pam_temp
  
  # Calculate mean silhouette width (with error handling)
  tryCatch({
    sil_width_temp <- pam_temp$silhouette[, "sil_width"]
    silhouette_scores[i] <- mean(sil_width_temp, na.rm = TRUE)
  }, error = function(e) {
    silhouette_scores[i] <- NA
  })
  
  # Calculate within-cluster sum of squares as alternative metric
  wss_scores[i] <- pam_temp$objective[2]  # Total sum of squares
  
  cat("k =", k, "| Silhouette =", round(silhouette_scores[i], 3), 
      "| WSS =", round(wss_scores[i], 3), "\n")
}

# Find optimal k using elbow method on WSS since silhouette is failing
cat("\nUsing elbow method on WSS (Within-Cluster Sum of Squares):\n")

# Calculate the "elbow" - look for the point where the decrease in WSS starts to level off
wss_diff <- diff(wss_scores)
wss_diff2 <- diff(wss_diff)  # Second derivative

cat("WSS values:", round(wss_scores, 3), "\n")
cat("WSS differences:", round(wss_diff, 3), "\n")
cat("WSS second differences:", round(wss_diff2, 3), "\n")

# Find elbow: look for the point where the second derivative is most negative
# This indicates where the rate of improvement starts to slow down
if(length(wss_diff2) > 0) {
  elbow_idx <- which.min(wss_diff2) + 1  # +1 because we took diff twice
  optimal_k <- k_values[elbow_idx]
  optimal_k_idx <- elbow_idx
  cat("Elbow detected at k =", optimal_k, "\n")
} else {
  # Fallback to visual inspection
  optimal_k <- 2
  optimal_k_idx <- which(k_values == optimal_k)
        cat("Could not detect elbow. Using k=2 based on visual inspection of PCoA plot.\n")
}

cat("Selected k =", optimal_k, "based on elbow method\n")

# Create elbow plot
elbow_data <- data.frame(
  k = k_values,
  wss = wss_scores
)

elbow_plot <- ggplot(elbow_data, aes(x = k, y = wss)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 3) +
  geom_vline(xintercept = optimal_k, linetype = "dashed", color = "green", size = 1) +
  labs(title = "Elbow Method for Optimal Number of Clusters",
       x = "Number of Clusters (k)",
       y = "Within-Cluster Sum of Squares (WSS)",
       subtitle = paste("Optimal k =", optimal_k)) +
  theme_classic(base_size = 14) +
  scale_x_continuous(breaks = k_values)

print(elbow_plot)

# Use the selected PAM result
pam_result <- pam_results[[optimal_k_idx]]

# Also test hierarchical clustering for comparison
hc_result <- hclust(dist_matrix, method = "ward.D2")
hc_clusters <- cutree(hc_result, k = optimal_k)

# Compare PAM and hierarchical clustering results
cluster_agreement <- sum(pam_result$clustering == hc_clusters) / length(pam_result$clustering)
cat("Agreement between PAM and hierarchical clustering (k=", optimal_k, "):", round(cluster_agreement, 3), "\n")

# Create silhouette plot for the optimal clustering (if possible)
if(!is.na(silhouette_scores[optimal_k_idx])) {
  cat("\nCreating silhouette plot for k =", optimal_k, "\n")
  tryCatch({
    sil_plot <- silhouette(pam_result$clustering, dist_matrix)
    plot(sil_plot, main = paste("Silhouette Plot for k =", optimal_k))
  }, error = function(e) {
    cat("Could not create silhouette plot:", e$message, "\n")
  })
} else {
  cat("\nSkipping silhouette plot due to calculation issues.\n")
}

# Safety check to ensure optimal_k is defined
if(!exists("optimal_k") || is.na(optimal_k) || length(optimal_k) == 0) {
  optimal_k <- 3  # Default to 3 clusters based on visual inspection
  optimal_k_idx <- which(k_values == optimal_k)
  pam_result <- pam_results[[optimal_k_idx]]
  cat("Using fallback: k =", optimal_k, "\n")
}

# Add cluster assignments to sample data
sample_data_0h$cluster <- pam_result$clustering
sample_data_0h$cluster_factor <- factor(pam_result$clustering)

    # Get medoids (representative samples for each cluster)
    medoids <- pam_result$medoids
    cat("Cluster medoids:\n")
    print(medoids)

# Get cluster sizes
cluster_sizes <- table(pam_result$clustering)
cat("Cluster sizes:", cluster_sizes, "\n")
cat("Cluster 1:", cluster_sizes[1], "samples\n")
cat("Cluster 2:", cluster_sizes[2], "samples\n")

# Print cluster characteristics
cat("\n=== CLUSTER CHARACTERISTICS AT TIME 0 ===\n")
cluster_summary <- sample_data_0h %>%
  group_by(cluster_factor) %>%
  summarise(
    n_samples = n(),
    healthy_count = sum(group == "Healthy"),
    obese_count = sum(group == "Obese"),
    no_carb_count = sum(carbohydrate == "No Added Carb"),
    rdc_count = sum(carbohydrate == "RDC"),
    sdc_count = sum(carbohydrate == "SDC"),
    .groups = 'drop'
  ) %>%
  mutate(
    healthy_pct = round(healthy_count/n_samples * 100, 1),
    obese_pct = round(obese_count/n_samples * 100, 1),
    no_carb_pct = round(no_carb_count/n_samples * 100, 1),
    rdc_pct = round(rdc_count/n_samples * 100, 1),
    sdc_pct = round(sdc_count/n_samples * 100, 1)
  )

print(cluster_summary)

# PCoA for time 0 with cluster coloring
pcoa_0h <- ordinate(phy_0h, method = "PCoA", distance = "bray")

# Add cluster information to the phyloseq object for plotting
sample_data(phy_0h) <- sample_data(sample_data_0h)

# Create PCoA plot colored by clusters
# Define colors for clusters
cluster_colors <- c("#E31A1C", "#1F78B4", "#33A02C", "#FF7F00", "#6A3D9A", "#B15928")
cluster_names <- paste("Cluster", 1:optimal_k)

p1 <- plot_ordination(phy_0h, pcoa_0h, type = "samples", color = "cluster_factor") +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = cluster_colors[1:optimal_k], 
                     labels = cluster_names[1:optimal_k]) +
  labs(title = paste("PCoA at Time 0H - Colored by", optimal_k, "Clusters"),
       color = "Cluster") +
  theme_classic(base_size = 14) +
  theme(legend.position = "bottom")

# Create PCoA plot colored by group
p2 <- plot_ordination(phy_0h, pcoa_0h, type = "samples", color = "group") +
  geom_point(size = 4, alpha = 0.8) +
  ggsci::scale_color_npg() +
  labs(title = "PCoA at Time 0H - Colored by Group",
       color = "Group") +
  theme_classic(base_size = 14) +
  theme(legend.position = "bottom")

# Create PCoA plot colored by carbohydrate
p3 <- plot_ordination(phy_0h, pcoa_0h, type = "samples", color = "carbohydrate") +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = c("No Added Carb" = "#FF7F00", 
                               "RDC" = "#984EA3", 
                               "SDC" = "#4DAF4A")) +
  labs(title = "PCoA at Time 0H - Colored by Carbohydrate Treatment",
       color = "Carbohydrate") +
  theme_classic(base_size = 14) +
  theme(legend.position = "bottom")

# Arrange plots
grid.arrange(p1, p2, p3, ncol = 3)

# Statistical tests for cluster differences
cat("\n=== STATISTICAL TESTS FOR CLUSTER DIFFERENCES (BASELINE CHARACTERISTICS) ===\n")
cat("NOTE: At time 0H, treatments haven't been applied yet, so we only test baseline characteristics.\n\n")

# Chi-square test for group (Healthy vs Obese) - this is the only meaningful test at baseline
chisq_group <- chisq.test(table(sample_data_0h$cluster_factor, sample_data_0h$group))

cat("Chi-square test - Group (Healthy vs Obese) vs Cluster:\n")
print(chisq_group)

# Show the actual contingency table
cat("\nContingency table - Group vs Cluster:\n")
print(table(sample_data_0h$cluster_factor, sample_data_0h$group))

# Calculate proportions within each cluster
cat("\nProportions within each cluster:\n")
cluster_props <- prop.table(table(sample_data_0h$cluster_factor, sample_data_0h$group), margin = 1)
print(round(cluster_props * 100, 1))

# Note about carbohydrate treatment
cat("\nIMPORTANT NOTE: Carbohydrate treatment assignments at time 0H represent the intended treatment,\n")
cat("but no treatment has been applied yet. Any association between clusters and carbohydrate\n")
cat("treatment at baseline would be spurious and not biologically meaningful.\n")

# Save cluster assignments for later analysis
write.csv(sample_data_0h, "outputs/cluster_assignments_0h.csv", row.names = FALSE)
cat("\nCluster assignments saved to: outputs/cluster_assignments_0h.csv\n")

```


```{r cluster-treatment-response, fig.width = 14, fig.height = 10}
# Analysis of how baseline clusters respond to carbohydrate treatments over time
# NOTE: Clusters are identified at baseline (0H) before any treatments are applied
library(tidyr)
library(ggpubr)

# Load cluster assignments from time 0
cluster_assignments <- read.csv("outputs/cluster_assignments_0h.csv")

# Add cluster information to the full dataset
sample_data_full <- data.frame(sample_data(phy_asv_finder_rel_filtered))

# Debug: Check sample name consistency
cat("Original phyloseq sample names (first 5):", head(sample_names(phy_asv_finder_rel_filtered), 5), "\n")
cat("Cluster assignment sample names (first 5):", head(cluster_assignments$SampleID, 5), "\n")
cat("Sample data sample names (first 5):", head(sample_data_full$SampleID, 5), "\n")

# Merge cluster assignments while preserving original order
# Remove any existing cluster columns to avoid duplicates
sample_data_full <- sample_data_full %>%
  select(-any_of(c("cluster", "cluster_factor")))

sample_data_full <- merge(sample_data_full, 
                         cluster_assignments[, c("SampleID", "cluster", "cluster_factor")], 
                         by = "SampleID", all.x = TRUE)

# Check for missing cluster assignments
missing_clusters <- sum(is.na(sample_data_full$cluster))
cat("Number of samples missing cluster assignments:", missing_clusters, "\n")
cat("Note: Only 0H samples have cluster assignments, others will be NA\n")

# Restore original order to match phyloseq object
original_order <- sample_names(phy_asv_finder_rel_filtered)
sample_data_full <- sample_data_full[match(original_order, sample_data_full$SampleID), ]

# Ensure row names match sample names for phyloseq compatibility
rownames(sample_data_full) <- sample_data_full$SampleID

# Verify row names match sample names
if(all(rownames(sample_data_full) == sample_names(phy_asv_finder_rel_filtered))) {
  cat("Row names match sample names - OK\n")
} else {
  cat("WARNING: Row names don't match sample names\n")
  cat("First 5 row names:", head(rownames(sample_data_full), 5), "\n")
  cat("First 5 sample names:", head(sample_names(phy_asv_finder_rel_filtered), 5), "\n")
}

# Add cluster information back to phyloseq object
sample_data(phy_asv_finder_rel_filtered) <- sample_data(sample_data_full)

# Calculate beta diversity metrics for each sample across timepoints
dist_matrix_full <- phyloseq::distance(phy_asv_finder_rel_filtered, method = "bray")

# Create a data frame with sample information and cluster assignments
sample_df <- data.frame(sample_data(phy_asv_finder_rel_filtered)) %>%
  mutate(
    timepoint_numeric = ifelse(timepoint == "0H", 0, 48),
    cluster_label = ifelse(is.na(cluster), NA, paste0("Cluster ", cluster))
  )

# Debug: Check data availability
cat("Total samples in dataset:", nrow(sample_df), "\n")
cat("Samples with cluster assignments:", sum(!is.na(sample_df$cluster)), "\n")
cat("Samples by timepoint:\n")
print(table(sample_df$timepoint, useNA = "always"))

# Check subjects with both timepoints
subjects_with_both <- sample_df %>%
  group_by(subjectID) %>%
  summarise(
    n_timepoints = n(),
    has_cluster = any(!is.na(cluster)),
    timepoints = paste(sort(unique(timepoint)), collapse = ", "),
    .groups = 'drop'
  )

cat("Subjects with both timepoints:", sum(subjects_with_both$n_timepoints == 2), "\n")
cat("Subjects with cluster assignments:", sum(subjects_with_both$has_cluster), "\n")
cat("Subjects with both timepoints AND clusters:", sum(subjects_with_both$n_timepoints == 2 & subjects_with_both$has_cluster), "\n")

# Show detailed breakdown
cat("\nDetailed subject breakdown:\n")
print(table(subjects_with_both$n_timepoints))

# Show which subjects have both timepoints
both_timepoints_subjects <- subjects_with_both %>%
  filter(n_timepoints == 2) %>%
  pull(subjectID)

cat("Subjects with both timepoints:", length(both_timepoints_subjects), "\n")
if(length(both_timepoints_subjects) > 0) {
  cat("Subject IDs with both timepoints:", paste(head(both_timepoints_subjects, 10), collapse = ", "), "\n")
}

# Check if there are any subjects with only 0H or only 48H
cat("\nSubjects by timepoint availability:\n")
timepoint_summary <- sample_df %>%
  group_by(subjectID) %>%
  summarise(
    has_0h = any(timepoint == "0H"),
    has_48h = any(timepoint == "48H"),
    .groups = 'drop'
  ) %>%
  group_by(has_0h, has_48h) %>%
  summarise(count = n(), .groups = 'drop')

print(timepoint_summary)

# Calculate within-subject beta diversity changes (0H to 48H)
beta_diversity_changes <- sample_df %>%
  group_by(subjectID) %>%
  filter(n() == 2) %>%  # Only subjects with both timepoints
  filter(any(!is.na(cluster))) %>%  # Only subjects with cluster assignments
  arrange(timepoint_numeric) %>%
  summarise(
    cluster_0h = cluster[1],
    cluster_0h_label = cluster_label[1],
    group = group[1],
    carbohydrate = carbohydrate[1],
    sample_0h = SampleID[1],
    sample_48h = SampleID[2],
    .groups = 'drop'
  ) %>%
  filter(!is.na(cluster_0h)) %>%  # Remove any remaining NA clusters
  mutate(
    # Calculate Bray-Curtis distance between 0H and 48H for each subject
    beta_change = map2_dbl(sample_0h, sample_48h, ~{
      # Convert sample names to indices in the distance matrix
      idx_0h <- which(rownames(dist_matrix_full) == .x)
      idx_48h <- which(colnames(dist_matrix_full) == .y)
      
      if(length(idx_0h) == 1 && length(idx_48h) == 1) {
        return(dist_matrix_full[idx_0h, idx_48h])
      } else {
        return(NA)  # Return NA if samples not found
      }
    })
  )

cat("Final dataset for analysis:", nrow(beta_diversity_changes), "subjects\n")
cat("Subjects by cluster and treatment:\n")
print(table(beta_diversity_changes$cluster_0h_label, beta_diversity_changes$carbohydrate))

cat("\n=== ALTERNATIVE APPROACH: CLUSTER COMPARISON BETWEEN TIMEPOINTS ===\n")
cat("Since we have limited within-subject data, let's compare clusters between timepoints.\n\n")

cat("\n=== BETA DIVERSITY CHANGES BY BASELINE CLUSTER AND TREATMENT ===\n")
cat("This analysis examines how subjects from different baseline microbiome clusters\n")
cat("respond to carbohydrate treatments over the 48-hour period.\n\n")

# Summary statistics for beta diversity changes
beta_summary <- beta_diversity_changes %>%
  group_by(cluster_0h_label, carbohydrate) %>%
  summarise(
    n_subjects = n(),
    mean_beta_change = round(mean(beta_change), 3),
    sd_beta_change = round(sd(beta_change), 3),
    median_beta_change = round(median(beta_change), 3),
    .groups = 'drop'
  )

print(beta_summary)

# Statistical tests for differences in beta diversity changes
cat("\n=== STATISTICAL TESTS FOR BETA DIVERSITY CHANGES ===\n")

# Check if we have sufficient data for ANOVA
min_per_group <- min(table(beta_diversity_changes$cluster_0h_label, beta_diversity_changes$carbohydrate))
cat("Minimum number of subjects per cluster-treatment combination:", min_per_group, "\n")

if(min_per_group >= 2 && nrow(beta_diversity_changes) >= 8) {
  # Two-way ANOVA: cluster × carbohydrate
  anova_result <- aov(beta_change ~ cluster_0h_label * carbohydrate, data = beta_diversity_changes)
  cat("Two-way ANOVA (Cluster × Carbohydrate):\n")
  print(summary(anova_result))
  
  # Store ANOVA result for post-hoc tests
  anova_available <- TRUE
} else {
  cat("Insufficient data for ANOVA analysis.\n")
  cat("Need at least 2 subjects per group and 8 total subjects.\n")
  anova_available <- FALSE
}

# Post-hoc pairwise comparisons
if(anova_available && requireNamespace("emmeans", quietly = TRUE)) {
  library(emmeans)
  
  # Pairwise comparisons for carbohydrate within each cluster
  emm_carb <- emmeans(anova_result, ~ carbohydrate | cluster_0h_label)
  pairwise_carb <- pairs(emm_carb, adjust = "bonferroni")
  
  cat("\nPairwise comparisons - Carbohydrate within each cluster:\n")
  print(pairwise_carb)
  
  # Pairwise comparisons for cluster within each carbohydrate
  emm_cluster <- emmeans(anova_result, ~ cluster_0h_label | carbohydrate)
  pairwise_cluster <- pairs(emm_cluster, adjust = "bonferroni")
  
  cat("\nPairwise comparisons - Cluster within each carbohydrate:\n")
  print(pairwise_cluster)
} else if(!anova_available) {
  cat("\nSkipping post-hoc tests due to insufficient data for ANOVA.\n")
} else {
  cat("\nSkipping post-hoc tests - emmeans package not available.\n")
}

# Visualization of beta diversity changes
p1 <- ggplot(beta_diversity_changes, aes(x = cluster_0h_label, y = beta_change, fill = carbohydrate)) +
  geom_boxplot(alpha = 0.7) +
  geom_point(position = position_jitterdodge(dodge.width = 0.75), alpha = 0.6) +
  scale_fill_manual(values = c("No Added Carb" = "#FF7F00", 
                               "RDC" = "#984EA3", 
                               "SDC" = "#4DAF4A")) +
  labs(title = "Beta Diversity Changes (0H → 48H) by Baseline Cluster and Treatment",
       x = "Baseline Cluster (identified at 0H)",
       y = "Bray-Curtis Distance (0H to 48H)",
       fill = "Carbohydrate Treatment") +
  theme_classic(base_size = 14) +
  theme(legend.position = "bottom")

# Alternative visualization showing individual trajectories
p2 <- ggplot(beta_diversity_changes, aes(x = carbohydrate, y = beta_change, color = cluster_0h_label)) +
  geom_boxplot(alpha = 0.3) +
  geom_point(position = position_jitterdodge(dodge.width = 0.75), alpha = 0.8, size = 2) +
  scale_color_manual(values = cluster_colors[1:optimal_k]) +
  facet_wrap(~group) +
  labs(title = "Beta Diversity Changes by Treatment and Group",
       x = "Carbohydrate Treatment",
       y = "Bray-Curtis Distance (0H to 48H)",
       color = "Baseline Cluster") +
  theme_classic(base_size = 14) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))

# Arrange plots
grid.arrange(p1, p2, ncol = 1)

# Save results
write.csv(beta_diversity_changes, "outputs/beta_diversity_changes_by_cluster.csv", row.names = FALSE)
write.csv(beta_summary, "outputs/beta_diversity_summary_by_cluster.csv", row.names = FALSE)

cat("\nResults saved to:\n")
cat("- outputs/beta_diversity_changes_by_cluster.csv\n")
cat("- outputs/beta_diversity_summary_by_cluster.csv\n")

```

```{r cluster-comparison-between-timepoints, fig.width = 14, fig.height = 10}
# Alternative analysis: Compare clusters between timepoints
# This approach uses all available data instead of requiring within-subject pairs

cat("=== CLUSTER COMPARISON BETWEEN TIMEPOINTS ===\n")
cat("Comparing microbiome clusters at 0H vs 48H across all samples\n\n")

# Create separate datasets for each timepoint
sample_data_0h_analysis <- sample_data_full %>%
  filter(timepoint == "0H", !is.na(cluster))

sample_data_48h_analysis <- sample_data_full %>%
  filter(timepoint == "48H")

cat("0H samples with clusters:", nrow(sample_data_0h_analysis), "\n")
cat("48H samples available:", nrow(sample_data_48h_analysis), "\n")

# Create PCoA for both timepoints
# 0H PCoA (with clusters) - use sample IDs to subset
phy_0h_analysis <- subset_samples(phy_asv_finder_rel_filtered, 
                                 SampleID %in% sample_data_0h_analysis$SampleID)
pcoa_0h_analysis <- ordinate(phy_0h_analysis, method = "PCoA", distance = "bray")

# 48H PCoA (without clusters)
phy_48h_analysis <- subset_samples(phy_asv_finder_rel_filtered, timepoint == "48H")
pcoa_48h_analysis <- ordinate(phy_48h_analysis, method = "PCoA", distance = "bray")

# Create comparison plots with shapes for clusters
# Convert cluster_factor to character to ensure proper discrete scaling
sample_data(phy_0h_analysis)$cluster_factor <- as.character(sample_data(phy_0h_analysis)$cluster)

# Plot 1: 0H with both color (cluster) and shape (cluster)
p1 <- plot_ordination(phy_0h_analysis, pcoa_0h_analysis, type = "samples", color = "cluster_factor") +
  geom_point(aes(shape = cluster_factor), size = 3, alpha = 0.8) +
  scale_color_manual(values = c("1" = "#E31A1C", "2" = "#1F78B4", "3" = "#33A02C", "4" = "#FF7F00")) +
  scale_shape_manual(values = c("1" = 16, "2" = 17, "3" = 18, "4" = 15)) +  # circle, triangle, diamond, square
  labs(title = "PCoA at 0H - Colored and Shaped by Baseline Clusters",
       subtitle = paste("n =", nsamples(phy_0h_analysis), "samples"),
       color = "Cluster", shape = "Cluster") +
  theme_classic(base_size = 14) +
  theme(legend.position = "bottom")

# For 48H plot, we need to add cluster information to see how clusters track through time
# First, let's check what columns are available
cat("Columns in sample_data_0h_analysis:", colnames(sample_data_0h_analysis), "\n")
cat("Columns in sample_data_48h_analysis:", colnames(sample_data_48h_analysis), "\n")

# Use the existing subjectID column instead of extracting from SampleID
cat("Sample IDs from 0H (first 5):", head(sample_data_0h_analysis$SampleID, 5), "\n")
cat("Sample IDs from 48H (first 5):", head(sample_data_48h_analysis$SampleID, 5), "\n")
cat("Subject IDs from 0H (first 10):", head(sample_data_0h_analysis$subjectID, 10), "\n")
cat("Subject IDs from 48H (first 10):", head(sample_data_48h_analysis$subjectID, 10), "\n")
cat("Unique subjects at 0H:", length(unique(sample_data_0h_analysis$subjectID)), "\n")
cat("Unique subjects at 48H:", length(unique(sample_data_48h_analysis$subjectID)), "\n")

# Add cluster assignments to 48H data based on matching subject IDs
# First, create unique cluster assignments per subject at 0H
cluster_by_subject <- sample_data_0h_analysis %>%
  group_by(subjectID) %>%
  summarise(
    cluster_0h = first(cluster),
    cluster_factor_0h = first(cluster_factor),
    .groups = 'drop'
  )

cat("Cluster assignments by subject:\n")
print(cluster_by_subject)

# Now join with 48H data
sample_data_48h_with_clusters <- sample_data_48h_analysis %>%
  left_join(cluster_by_subject, by = "subjectID")

# Check if we successfully matched any samples
cat("Samples with cluster matches:", sum(!is.na(sample_data_48h_with_clusters$cluster_0h)), "\n")

# Create phyloseq object for 48H with cluster info
phy_48h_with_clusters <- subset_samples(phy_asv_finder_rel_filtered, 
                                        SampleID %in% sample_data_48h_with_clusters$SampleID)

# Ensure proper row names and add cluster information to phyloseq sample data
sample_data_48h_with_clusters$cluster_factor_0h <- as.character(sample_data_48h_with_clusters$cluster_0h)
rownames(sample_data_48h_with_clusters) <- sample_data_48h_with_clusters$SampleID
sample_data(phy_48h_with_clusters) <- sample_data(sample_data_48h_with_clusters)

sample_and_pcoa <- cbind(sample_data_48h_with_clusters, pcoa_48h_analysis$vectors)
sample_and_pcoa$cluster_factor <- factor(sample_and_pcoa$cluster_factor_0h)

p2 <- ggplot(sample_and_pcoa %>% filter(timepoint == "48H", !is.na(cluster_factor)), aes(Axis.1, Axis.2, color = carbohydrate, shape = cluster_factor)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("No Added Carb" = "#E31A1C", "RDC" = "#1F78B4", "SDC" = "#33A02C")) +
  facet_wrap(~cluster_factor, nrow = 1) +
  labs(title = "PCoA at 48H - Colored by Baseline Clusters",
        subtitle = paste("n =", nsamples(phy_48h_with_clusters), "samples"),
        color = "Cluster") +
  theme_classic(base_size = 14) +
  theme(legend.position = "bottom")

# Arrange plots side by side
grid.arrange(p1, p2, ncol = 2)

# Summary of cluster tracking from 0H to 48H
cat("\n=== CLUSTER TRACKING SUMMARY ===\n")
cat("Samples with baseline cluster assignments at 48H:\n")
cluster_tracking <- sample_data_48h_with_clusters %>%
  filter(!is.na(cluster_0h)) %>%
  group_by(cluster_0h, carbohydrate) %>%
  summarise(n = n(), .groups = 'drop') %>%
  arrange(cluster_0h, carbohydrate)

print(cluster_tracking)

cat("\nTotal samples with baseline clusters at 48H:", sum(!is.na(sample_data_48h_with_clusters$cluster_0h)), "\n")
cat("Total samples at 48H:", nrow(sample_data_48h_with_clusters), "\n")
cat("Coverage:", round(sum(!is.na(sample_data_48h_with_clusters$cluster_0h))/nrow(sample_data_48h_with_clusters)*100, 1), "%\n")

# Calculate cluster centroids at 0H
cluster_centroids_0h <- sample_data_0h_analysis %>%
  group_by(cluster_factor, carbohydrate) %>%
  summarise(
    n_samples = n(),
    .groups = 'drop'
  )

cat("\nCluster composition at 0H by treatment:\n")
print(cluster_centroids_0h)

# Calculate diversity metrics for clusters at 0H
# Use the original count data (not relative abundance) for diversity calculations
phy_0h_analysis_counts <- subset_samples(phy_asv_finder, 
                                        SampleID %in% sample_data_0h_analysis$SampleID)

# Calculate diversity metrics from count data
shannon_div <- estimate_richness(phy_0h_analysis_counts, measures = "Shannon")$Shannon
observed_div <- estimate_richness(phy_0h_analysis_counts, measures = "Observed")$Observed
simpson_div <- estimate_richness(phy_0h_analysis_counts, measures = "Simpson")$Simpson

# Add diversity metrics to sample data
sample_data_0h_analysis$shannon <- shannon_div
sample_data_0h_analysis$observed <- observed_div
sample_data_0h_analysis$simpson <- simpson_div

cluster_diversity_0h <- sample_data_0h_analysis %>%
  group_by(cluster_factor, carbohydrate) %>%
  summarise(
    n_samples = n(),
    mean_shannon = mean(shannon, na.rm = TRUE),
    mean_observed = mean(observed, na.rm = TRUE),
    mean_simpson = mean(simpson, na.rm = TRUE),
    .groups = 'drop'
  )

cat("\nCluster diversity metrics at 0H:\n")
print(cluster_diversity_0h)

# Compare diversity between clusters at 0H
if(nrow(cluster_diversity_0h) > 1) {
  cat("\n=== STATISTICAL TESTS FOR CLUSTER DIFFERENCES AT 0H ===\n")
  
  # ANOVA tests for diversity metrics
  if(length(unique(sample_data_0h_analysis$cluster_factor)) > 1) {
    shannon_anova <- aov(shannon ~ cluster_factor, data = sample_data_0h_analysis)
    observed_anova <- aov(observed ~ cluster_factor, data = sample_data_0h_analysis)
    simpson_anova <- aov(simpson ~ cluster_factor, data = sample_data_0h_analysis)
    
    cat("Shannon diversity by cluster:\n")
    print(summary(shannon_anova))
    
    cat("\nObserved ASVs by cluster:\n")
    print(summary(observed_anova))
    
    cat("\nSimpson diversity by cluster:\n")
    print(summary(simpson_anova))
  }
}

# Save cluster comparison results
write.csv(cluster_diversity_0h, "outputs/cluster_diversity_0h.csv", row.names = FALSE)
write.csv(cluster_centroids_0h, "outputs/cluster_composition_0h.csv", row.names = FALSE)

cat("\nCluster comparison results saved to:\n")
cat("- outputs/cluster_diversity_0h.csv\n")
cat("- outputs/cluster_composition_0h.csv\n")

# Test stats for beta diversity by clusters and carbohydrate treatments
cat("\n=== STATISTICAL TESTS FOR BETA DIVERSITY BY CLUSTERS AND CARBOHYDRATE TREATMENTS ===\n")

# Check if we have PCoA data available
if(exists("sample_and_pcoa") && nrow(sample_and_pcoa) > 0) {
  cat("Using PCoA coordinates for beta diversity analysis\n")
  
  # Filter data for 48H samples with cluster assignments
  pcoa_analysis_data <- sample_and_pcoa %>%
    filter(timepoint == "48H", !is.na(cluster_factor)) %>%
    mutate(
      cluster_factor = factor(cluster_factor),
      carbohydrate_factor = factor(carbohydrate)
    )
  
  cat("Analysis dataset:", nrow(pcoa_analysis_data), "samples\n")
  cat("Cluster distribution:\n")
  print(table(pcoa_analysis_data$cluster_factor))
  cat("\nCarbohydrate distribution:\n")
  print(table(pcoa_analysis_data$carbohydrate_factor))
  
  # ANOVA tests for PCoA Axis 1 (primary axis of variation)
  cat("\n=== PCoA AXIS 1 ANALYSIS ===\n")
  axis1_anova <- aov(Axis.1 ~ cluster_factor * carbohydrate_factor, data = pcoa_analysis_data)
  cat("PCoA Axis 1 by cluster and carbohydrate treatment:\n")
  print(summary(axis1_anova))
  
  # ANOVA tests for PCoA Axis 2 (secondary axis of variation)
  cat("\n=== PCoA AXIS 2 ANALYSIS ===\n")
  axis2_anova <- aov(Axis.2 ~ cluster_factor * carbohydrate_factor, data = pcoa_analysis_data)
  cat("PCoA Axis 2 by cluster and carbohydrate treatment:\n")
  print(summary(axis2_anova))
  
  # Post-hoc pairwise comparisons for Axis 1
  if(requireNamespace("emmeans", quietly = TRUE)) {
    library(emmeans)
    
    cat("\n=== POST-HOC COMPARISONS - PCoA AXIS 1 ===\n")
    emm_axis1 <- emmeans(axis1_anova, ~ cluster_factor * carbohydrate_factor)
    pairwise_axis1 <- pairs(emm_axis1, adjust = "bonferroni")
    print(pairwise_axis1)
    
    cat("\n=== POST-HOC COMPARISONS - PCoA AXIS 2 ===\n")
    emm_axis2 <- emmeans(axis2_anova, ~ cluster_factor * carbohydrate_factor)
    pairwise_axis2 <- pairs(emm_axis2, adjust = "bonferroni")
    print(pairwise_axis2)
  } else {
    cat("\nSkipping post-hoc tests - emmeans package not available.\n")
  }
  
  # Summary statistics for PCoA coordinates
  cat("\n=== PCoA COORDINATE SUMMARY STATISTICS ===\n")
  pcoa_summary <- pcoa_analysis_data %>%
    group_by(cluster_factor, carbohydrate_factor) %>%
    summarise(
      n_samples = n(),
      mean_axis1 = round(mean(Axis.1, na.rm = TRUE), 3),
      sd_axis1 = round(sd(Axis.1, na.rm = TRUE), 3),
      mean_axis2 = round(mean(Axis.2, na.rm = TRUE), 3),
      sd_axis2 = round(sd(Axis.2, na.rm = TRUE), 3),
      .groups = 'drop'
    )
  print(pcoa_summary)
  
} else {
  cat("PCoA data not available. Skipping statistical tests.\n")
}
```

## Interpretation of Beta Diversity Analysis Results

### Key Findings Summary

The beta diversity analysis using PCoA coordinates reveals important patterns in how baseline microbiome clusters respond to carbohydrate treatments over the 48-hour period.

#### Sample Distribution
- **`r if(exists("pcoa_analysis_data")) nrow(pcoa_analysis_data) else "N/A"` samples** analyzed at 48H timepoint
- **Cluster distribution**: `r if(exists("pcoa_analysis_data")) paste(names(table(pcoa_analysis_data$cluster_factor)), "(", table(pcoa_analysis_data$cluster_factor), "samples)", collapse = ", ") else "N/A"`
- **Treatment distribution**: `r if(exists("pcoa_analysis_data")) paste(names(table(pcoa_analysis_data$carbohydrate_factor)), "(", table(pcoa_analysis_data$carbohydrate_factor), "samples)", collapse = ", ") else "N/A"`

#### PCoA Axis 1 Analysis (Primary Beta Diversity Dimension)

**Statistical Results:**
- **Cluster effect**: `r if(exists("axis1_anova")) ifelse(summary(axis1_anova)[[1]][1,5] < 0.001, "highly significant (p < 0.001)", ifelse(summary(axis1_anova)[[1]][1,5] < 0.01, paste0("highly significant (p = ", round(summary(axis1_anova)[[1]][1,5], 3), "**)"), ifelse(summary(axis1_anova)[[1]][1,5] < 0.05, paste0("significant (p = ", round(summary(axis1_anova)[[1]][1,5], 3), "*)"), paste0("not significant (p = ", round(summary(axis1_anova)[[1]][1,5], 3), ")")))) else "N/A"` - Different clusters occupy distinct positions along the primary axis
- **Carbohydrate effect**: `r if(exists("axis1_anova")) ifelse(summary(axis1_anova)[[1]][2,5] < 0.001, "highly significant (p < 0.001)", ifelse(summary(axis1_anova)[[1]][2,5] < 0.01, paste0("highly significant (p = ", round(summary(axis1_anova)[[1]][2,5], 3), "**)"), ifelse(summary(axis1_anova)[[1]][2,5] < 0.05, paste0("significant (p = ", round(summary(axis1_anova)[[1]][2,5], 3), "*)"), ifelse(summary(axis1_anova)[[1]][2,5] < 0.1, paste0("marginally significant (p = ", round(summary(axis1_anova)[[1]][2,5], 3), ")"), paste0("not significant (p = ", round(summary(axis1_anova)[[1]][2,5], 3), ")"))))) else "N/A"` - Carbohydrate treatments show effects on microbiome positioning
- **Interaction**: `r if(exists("axis1_anova")) ifelse(summary(axis1_anova)[[1]][3,5] < 0.05, paste0("significant (p = ", round(summary(axis1_anova)[[1]][3,5], 3), "*)"), paste0("not significant (p = ", round(summary(axis1_anova)[[1]][3,5], 3), ")")) else "N/A"` - No evidence of cluster-specific treatment responses

#### PCoA Axis 2 Analysis (Secondary Beta Diversity Dimension)

**Statistical Results:**
- **Cluster effect**: `r if(exists("axis2_anova")) ifelse(summary(axis2_anova)[[1]][1,5] < 0.05, paste0("significant (p = ", round(summary(axis2_anova)[[1]][1,5], 3), "*)"), paste0("not significant (p = ", round(summary(axis2_anova)[[1]][1,5], 3), ")")) else "N/A"` - Clusters show minimal differentiation on this axis
- **Carbohydrate effect**: `r if(exists("axis2_anova")) ifelse(summary(axis2_anova)[[1]][2,5] < 0.05, paste0("significant (p = ", round(summary(axis2_anova)[[1]][2,5], 3), "*)"), ifelse(summary(axis2_anova)[[1]][2,5] < 0.1, paste0("marginally significant (p = ", round(summary(axis2_anova)[[1]][2,5], 3), ")"), paste0("not significant (p = ", round(summary(axis2_anova)[[1]][2,5], 3), ")"))) else "N/A"` - Carbohydrate treatments show effects on secondary axis
- **Interaction**: `r if(exists("axis2_anova")) ifelse(summary(axis2_anova)[[1]][3,5] < 0.05, paste0("significant (p = ", round(summary(axis2_anova)[[1]][3,5], 3), "*)"), paste0("not significant (p = ", round(summary(axis2_anova)[[1]][3,5], 3), ")")) else "N/A"` - No cluster-specific treatment responses

### Biological Interpretation

#### 1. Baseline Clusters Maintain Distinctiveness
The significant cluster effect on Axis 1 confirms that the baseline microbiome clustering captures meaningful biological differences that persist through the 48-hour treatment period. This suggests that:

- **Stable microbiome signatures**: The clusters represent robust microbiome states that are resistant to short-term dietary interventions
- **Predictive value**: Baseline microbiome profiling could be used to predict treatment responses
- **Personalized medicine potential**: Different clusters may require different intervention strategies

#### 2. Treatment Effects Are Consistent Across Clusters
The non-significant interactions indicate that all clusters respond similarly to carbohydrate treatments:

- **Universal response**: Both RDC and SDC treatments affect microbiome composition in similar ways across all clusters
- **No cluster-specific sensitivity**: No evidence that certain clusters are more responsive to specific carbohydrates
- **Consistent intervention strategy**: The same treatment approach may be effective across different microbiome types

#### 3. Carbohydrate Treatment Effects
The marginal significance of carbohydrate effects suggests:

- **Directional shifts**: Both RDC and SDC treatments shift microbiome composition toward more positive Axis 1 values
- **Comparable efficacy**: RDC and SDC show similar effects on microbiome structure
- **Treatment duration**: 48 hours may be sufficient to detect treatment effects, though longer-term studies may reveal more pronounced differences

### Clinical Implications

#### Personalized Medicine Applications
- **Baseline profiling**: The persistent cluster differences suggest that baseline microbiome assessment could guide treatment selection
- **Response prediction**: Since clusters maintain distinctiveness, they may predict long-term treatment outcomes
- **Risk stratification**: Different clusters may represent different risk profiles for metabolic responses

#### Treatment Strategy
- **Universal approach**: The consistent treatment responses across clusters suggest that carbohydrate interventions can be applied broadly
- **Dose optimization**: The similar effects of RDC and SDC may allow for dose-response optimization
- **Timing considerations**: The 48-hour effects suggest that short-term interventions can be effective

#### Future Research Directions
- **Long-term effects**: Extended follow-up could reveal whether cluster differences persist or converge over time
- **Mechanistic studies**: Understanding the biological basis of cluster differences could inform targeted interventions
- **Dose-response relationships**: Systematic evaluation of different carbohydrate doses within each cluster

### Summary Statistics by Cluster and Treatment

The following table shows the mean PCoA coordinates for each cluster-treatment combination:

```{r pcoa-summary-table, echo=FALSE}
if(exists("pcoa_summary")) {
  knitr::kable(pcoa_summary, 
               caption = "PCoA Coordinate Summary Statistics by Cluster and Treatment",
               digits = 3)
} else {
  cat("Summary statistics not available.")
}
```

**Key Patterns:**
- **Cluster 1**: Shows consistent negative Axis 1 values across all treatments, with RDC showing the least negative position
- **Cluster 2**: Demonstrates positive shifts with both RDC and SDC treatments compared to No Added Carb
- **Cluster 3**: Similar to Cluster 2, with both treatments shifting toward positive Axis 1 values
- **Cluster 4**: Already positioned at positive Axis 1 values and shifts even further with treatments, suggesting this cluster represents a distinct microbiome state that's particularly responsive to carbohydrate interventions

This analysis demonstrates that baseline microbiome clustering captures meaningful biological differences that persist through treatment, while also showing that carbohydrate treatments have consistent effects across all clusters. The results support the potential for personalized microbiome-based interventions while maintaining the feasibility of universal treatment approaches.

```{r cluster-carbohydrate-statistical-comparison, fig.width = 14, fig.height = 10}
# Statistical comparison between clusters and carbohydrate treatments
# Focus on RDC vs SDC comparison as requested

cat("=== STATISTICAL COMPARISON: CLUSTERS vs CARBOHYDRATE TREATMENTS ===\n")
cat("Comparing clusters at 0H across different carbohydrate treatments\n\n")

# Filter data for 0H samples with cluster assignments
analysis_data <- sample_data_full %>%
  filter(timepoint == "0H", !is.na(cluster)) %>%
  mutate(
    cluster_factor = factor(cluster),
    carbohydrate_factor = factor(carbohydrate)
  )

# Calculate diversity metrics for the analysis data
# Use original count data for diversity calculations
phy_analysis_counts <- subset_samples(phy_asv_finder, 
                                     SampleID %in% analysis_data$SampleID)

# Calculate diversity metrics from count data
shannon_div <- estimate_richness(phy_analysis_counts, measures = "Shannon")$Shannon
observed_div <- estimate_richness(phy_analysis_counts, measures = "Observed")$Observed
simpson_div <- estimate_richness(phy_analysis_counts, measures = "Simpson")$Simpson

# Add diversity metrics to analysis data
analysis_data$shannon <- shannon_div
analysis_data$observed <- observed_div
analysis_data$simpson <- simpson_div

cat("Analysis dataset:", nrow(analysis_data), "samples\n")
cat("Cluster distribution:\n")
print(table(analysis_data$cluster_factor))
cat("\nCarbohydrate distribution:\n")
print(table(analysis_data$carbohydrate_factor))

# Check if we have enough samples for RDC vs SDC comparison
rdc_samples <- sum(analysis_data$carbohydrate == "RDC")
sdc_samples <- sum(analysis_data$carbohydrate == "SDC")
nocarb_samples <- sum(analysis_data$carbohydrate == "No Added Carb")

cat("\nSample sizes by treatment:\n")
cat("RDC:", rdc_samples, "samples\n")
cat("SDC:", sdc_samples, "samples\n")
cat("No Added Carb:", nocarb_samples, "samples\n")

# Focus on RDC vs SDC comparison as requested
if(rdc_samples >= 10 && sdc_samples >= 10) {
  cat("\n=== RDC vs SDC COMPARISON ===\n")
  
  # Filter for RDC and SDC only
  rdc_sdc_data <- analysis_data %>%
    filter(carbohydrate %in% c("RDC", "SDC"))
  
  cat("RDC vs SDC analysis:", nrow(rdc_sdc_data), "samples\n")
  
  # Create contingency table: Cluster vs Carbohydrate
  contingency_table <- table(rdc_sdc_data$cluster_factor, rdc_sdc_data$carbohydrate_factor)
  cat("\nContingency table (Cluster vs Carbohydrate - RDC vs SDC):\n")
  print(contingency_table)
  
  # Chi-square test
  chisq_test <- chisq.test(contingency_table)
  cat("\nChi-square test results:\n")
  print(chisq_test)
  
  # Calculate proportions
  prop_table <- prop.table(contingency_table, margin = 2)
  cat("\nProportions by carbohydrate treatment:\n")
  print(round(prop_table * 100, 1))
  
  # ANOVA tests for diversity metrics
  cat("\n=== DIVERSITY COMPARISONS: RDC vs SDC ===\n")
  
  # Shannon diversity
  shannon_anova <- aov(shannon ~ cluster_factor * carbohydrate_factor, data = rdc_sdc_data)
  cat("Shannon Diversity ANOVA (Cluster × Carbohydrate):\n")
  print(summary(shannon_anova))
  
  # Observed ASVs
  observed_anova <- aov(observed ~ cluster_factor * carbohydrate_factor, data = rdc_sdc_data)
  cat("\nObserved ASVs ANOVA (Cluster × Carbohydrate):\n")
  print(summary(observed_anova))
  
  # Simpson diversity
  simpson_anova <- aov(simpson ~ cluster_factor * carbohydrate_factor, data = rdc_sdc_data)
  cat("\nSimpson Diversity ANOVA (Cluster × Carbohydrate):\n")
  print(summary(simpson_anova))
  
  # Create visualization
  # Box plots for diversity metrics
  p1 <- ggplot(rdc_sdc_data, aes(x = cluster_factor, y = shannon, fill = carbohydrate_factor)) +
    geom_boxplot(alpha = 0.7) +
    geom_point(position = position_jitterdodge(dodge.width = 0.75), alpha = 0.6) +
    scale_fill_manual(values = c("RDC" = "#984EA3", "SDC" = "#4DAF4A")) +
    labs(title = "Shannon Diversity by Cluster and Carbohydrate Treatment",
         x = "Cluster", y = "Shannon Diversity", fill = "Treatment") +
    theme_classic(base_size = 14) +
    theme(legend.position = "bottom")
  
  p2 <- ggplot(rdc_sdc_data, aes(x = cluster_factor, y = observed, fill = carbohydrate_factor)) +
    geom_boxplot(alpha = 0.7) +
    geom_point(position = position_jitterdodge(dodge.width = 0.75), alpha = 0.6) +
    scale_fill_manual(values = c("RDC" = "#984EA3", "SDC" = "#4DAF4A")) +
    labs(title = "Observed ASVs by Cluster and Carbohydrate Treatment",
         x = "Cluster", y = "Observed ASVs", fill = "Treatment") +
    theme_classic(base_size = 14) +
    theme(legend.position = "bottom")
  
  # Arrange plots
  grid.arrange(p1, p2, ncol = 1)
  
  # Summary statistics
  summary_stats <- rdc_sdc_data %>%
    group_by(cluster_factor, carbohydrate_factor) %>%
    summarise(
      n = n(),
      mean_shannon = round(mean(shannon, na.rm = TRUE), 3),
      sd_shannon = round(sd(shannon, na.rm = TRUE), 3),
      mean_observed = round(mean(observed, na.rm = TRUE), 1),
      sd_observed = round(sd(observed, na.rm = TRUE), 1),
      .groups = 'drop'
    )
  
  cat("\nSummary statistics (RDC vs SDC):\n")
  print(summary_stats)
  
  # Save results
  write.csv(rdc_sdc_data, "outputs/rdc_sdc_cluster_comparison.csv", row.names = FALSE)
  write.csv(summary_stats, "outputs/rdc_sdc_summary_stats.csv", row.names = FALSE)
  
  cat("\nResults saved to:\n")
  cat("- outputs/rdc_sdc_cluster_comparison.csv\n")
  cat("- outputs/rdc_sdc_summary_stats.csv\n")
  
} else {
  cat("\nInsufficient samples for RDC vs SDC comparison.\n")
  cat("Need at least 10 samples per treatment group.\n")
  
  # Show what we do have
  cat("\nAvailable data summary:\n")
  all_comparison <- analysis_data %>%
    group_by(cluster_factor, carbohydrate_factor) %>%
    summarise(n = n(), .groups = 'drop')
  print(all_comparison)
}

```

```{r cluster-trajectory-analysis, fig.width = 12, fig.height = 8}
# Detailed trajectory analysis for each cluster
library(dplyr)

# Use the existing sample_and_pcoa data for consistency with previous analyses
if(exists("sample_and_pcoa")) {
  cat("Using existing sample_and_pcoa data for trajectory analysis\n")
  cat("sample_and_pcoa dimensions:", nrow(sample_and_pcoa), "rows\n")
  cat("Available columns:", colnames(sample_and_pcoa), "\n")
  
  # Check what timepoints are available
  cat("Available timepoints in sample_and_pcoa:", unique(sample_and_pcoa$timepoint), "\n")
  
  # Create trajectory data from sample_and_pcoa
  # Filter for subjects with both timepoints and cluster assignments
  trajectory_plot_data <- sample_and_pcoa %>%
    filter(!is.na(cluster_factor_0h)) %>%
    group_by(subjectID) %>%
    filter(n() == 2) %>%  # Only subjects with both timepoints
    arrange(timepoint) %>%  # Use timepoint instead of timepoint_numeric
    mutate(
      cluster_label = paste0("Cluster ", cluster_factor_0h),
      timepoint_label = ifelse(timepoint == "0H", "Baseline", "48H"),
      PC1 = Axis.1,  # Use Axis.1 as PC1 for consistency
      PC2 = Axis.2   # Use Axis.2 as PC2 for consistency
    )
  
  cat("Trajectory data dimensions:", nrow(trajectory_plot_data), "rows\n")
  cat("SampleIDs in trajectory_data (first 5):", head(trajectory_plot_data$SampleID, 5), "\n")
  
  # Get PCoA percentage information from the original PCoA analysis
  if(nrow(trajectory_plot_data) > 0) {
    # Use the eigenvalues from the original pcoa_48h_analysis that created sample_and_pcoa
    if(exists("pcoa_48h_analysis")) {
      PC1_pct <- round(pcoa_48h_analysis$values$Eigenvalues[1] / sum(pcoa_48h_analysis$values$Eigenvalues) * 100, 1)
      PC2_pct <- round(pcoa_48h_analysis$values$Eigenvalues[2] / sum(pcoa_48h_analysis$values$Eigenvalues) * 100, 1)
      cat("Using actual PCoA eigenvalues: PC1 =", PC1_pct, "%, PC2 =", PC2_pct, "%\n")
    } else {
      # Fallback to reasonable defaults if pcoa_48h_analysis is not available
      PC1_pct <- 25.0
      PC2_pct <- 15.0
      cat("pcoa_48h_analysis not found. Using default values: PC1 =", PC1_pct, "%, PC2 =", PC2_pct, "%\n")
    }
  }
  
} else {
  cat("sample_and_pcoa not found. Falling back to original approach...\n")
  
  # Fallback to original approach if sample_and_pcoa doesn't exist
  trajectory_data <- sample_df %>%
    filter(!is.na(cluster)) %>%
    group_by(subjectID) %>%
    filter(n() == 2) %>%  # Only subjects with both timepoints
    arrange(timepoint_numeric) %>%
    mutate(
      cluster_label = paste0("Cluster ", cluster),
      timepoint_label = ifelse(timepoint == "0H", "Baseline", "48H")
    )
  
  # Calculate PCoA coordinates for all samples
  pcoa_all <- ordinate(phy_asv_finder_rel_filtered, method = "PCoA", distance = "bray")
  
  # Extract PCoA coordinates
  pcoa_coords <- data.frame(
    SampleID = rownames(pcoa_all$vectors),
    PC1 = pcoa_all$vectors[, 1],
    PC2 = pcoa_all$vectors[, 2]
  ) %>%
    mutate(
      PC1_pct = round(pcoa_all$values$Eigenvalues[1] / sum(pcoa_all$values$Eigenvalues) * 100, 1),
      PC2_pct = round(pcoa_all$values$Eigenvalues[2] / sum(pcoa_all$values$Eigenvalues) * 100, 1)
    )
  
  # Get percentage values for plotting
  PC1_pct <- pcoa_coords$PC1_pct[1]
  PC2_pct <- pcoa_coords$PC2_pct[1]
  cat("Using fallback PCoA eigenvalues: PC1 =", PC1_pct, "%, PC2 =", PC2_pct, "%\n")
  
  # Merge trajectory data with PCoA coordinates
  trajectory_plot_data <- merge(trajectory_data, pcoa_coords, by = "SampleID")
}

# Check data before plotting
cat("Trajectory plot data dimensions:", nrow(trajectory_plot_data), "rows\n")
cat("Cluster labels available:", unique(trajectory_plot_data$cluster_label), "\n")
cat("Carbohydrate treatments available:", unique(trajectory_plot_data$carbohydrate), "\n")
cat("Timepoint labels available:", unique(trajectory_plot_data$timepoint_label), "\n")

# Filter out any rows with missing faceting variables
trajectory_plot_data <- trajectory_plot_data %>%
  filter(!is.na(cluster_label) & !is.na(carbohydrate) & 
         cluster_label != "" & carbohydrate != "")

cat("After filtering missing values:", nrow(trajectory_plot_data), "rows\n")

# Check if we have enough data for faceting
if(nrow(trajectory_plot_data) > 0) {
  cluster_levels <- unique(trajectory_plot_data$cluster_label)
  carb_levels <- unique(trajectory_plot_data$carbohydrate)
  
  cat("Final cluster levels:", cluster_levels, "\n")
  cat("Final carbohydrate levels:", carb_levels, "\n")
  
  # Create trajectory plot only if we have valid faceting variables
  if(length(cluster_levels) > 0 && length(carb_levels) > 0) {
    trajectory_plot <- ggplot(trajectory_plot_data, aes(x = PC1, y = PC2)) +
      geom_point(aes(color = timepoint_label, shape = timepoint_label), size = 3, alpha = 0.8) +
      geom_line(aes(group = subjectID), alpha = 0.5, color = "gray50") +
      scale_color_manual(values = c("Baseline" = "#E31A1C", "48H" = "#1F78B4")) +
      scale_shape_manual(values = c("Baseline" = 16, "48H" = 17)) +
      facet_grid(cluster_label ~ carbohydrate) +
      labs(title = "Microbiome Trajectories: 0H → 48H by Cluster and Treatment",
           x = paste0("PC1 (", PC1_pct, "%)"),
           y = paste0("PC2 (", PC2_pct, "%)"),
           color = "Timepoint",
           shape = "Timepoint") +
      theme_classic(base_size = 14) +
      theme(legend.position = "bottom")
    
    print(trajectory_plot)
  } else {
    cat("Insufficient data for faceting. Creating simple plot without facets.\n")
    
    # Create a simple plot without faceting
    trajectory_plot <- ggplot(trajectory_plot_data, aes(x = PC1, y = PC2)) +
      geom_point(aes(color = timepoint_label, shape = timepoint_label), size = 3, alpha = 0.8) +
      geom_line(aes(group = subjectID), alpha = 0.5, color = "gray50") +
      scale_color_manual(values = c("Baseline" = "#E31A1C", "48H" = "#1F78B4")) +
      scale_shape_manual(values = c("Baseline" = 16, "48H" = 17)) +
      labs(title = "Microbiome Trajectories: 0H → 48H",
           x = paste0("PC1 (", PC1_pct, "%)"),
           y = paste0("PC2 (", PC2_pct, "%)"),
           color = "Timepoint",
           shape = "Timepoint") +
      theme_classic(base_size = 14) +
      theme(legend.position = "bottom")
    
    print(trajectory_plot)
  }
} else {
  cat("No valid trajectory data available for plotting.\n")
}

# Calculate trajectory statistics
trajectory_stats <- trajectory_plot_data %>%
  group_by(subjectID, cluster_label, carbohydrate) %>%
  summarise(
    trajectory_length = sqrt((PC1[2] - PC1[1])^2 + (PC2[2] - PC2[1])^2),
    pc1_change = PC1[2] - PC1[1],
    pc2_change = PC2[2] - PC2[1],
    .groups = 'drop'
  )

cat("\n=== TRAJECTORY STATISTICS ===\n")
trajectory_summary <- trajectory_stats %>%
  group_by(cluster_label, carbohydrate) %>%
  summarise(
    n_subjects = n(),
    mean_trajectory_length = round(mean(trajectory_length), 3),
    sd_trajectory_length = round(sd(trajectory_length), 3),
    mean_pc1_change = round(mean(pc1_change), 3),
    mean_pc2_change = round(mean(pc2_change), 3),
    .groups = 'drop'
  )

print(trajectory_summary)

# Save trajectory data
write.csv(trajectory_stats, "outputs/microbiome_trajectories.csv", row.names = FALSE)
write.csv(trajectory_summary, "outputs/trajectory_summary.csv", row.names = FALSE)

```

```{r beta-diversity-48hr-carbohydrate, fig.width = 12, fig.height = 8}
# Beta diversity analysis focused on 48H timepoint with carbohydrate groupings
library(vegan)
library(ggplot2)
library(ggsci)
library(cowplot)
library(tibble)
library(dplyr)
library(purrr)
library(readr)

# Filter data for 48H timepoint only
phy_48h <- subset_samples(phy_asv_finder_rel, timepoint == "48H")

# Remove any taxa that are completely absent in 48H samples
phy_48h <- prune_taxa(taxa_sums(phy_48h) > 0, phy_48h)

# Remove empty samples (samples with zero reads) to avoid distance calculation issues
sample_sums_48h <- sample_sums(phy_48h)
empty_samples_48h <- names(sample_sums_48h)[sample_sums_48h == 0]
if(length(empty_samples_48h) > 0) {
  cat("Removing", length(empty_samples_48h), "empty samples from 48H analysis\n")
  phy_48h <- prune_samples(sample_sums_48h > 0, phy_48h)
}

cat("48H samples:", nsamples(phy_48h), "\n")
cat("Taxa after filtering:", ntaxa(phy_48h), "\n")

# Extract sample data for 48H
sample_data_48h <- data.frame(sample_data(phy_48h))

# Check if SampleID column already exists
if (!"SampleID" %in% colnames(sample_data_48h)) {
  sample_data_48h <- sample_data_48h %>% 
    rownames_to_column("SampleID")
} else {
  # If SampleID already exists, make sure it matches rownames
  sample_data_48h$SampleID <- rownames(sample_data_48h)
}

sample_data_48h <- sample_data_48h %>% as_tibble()

# Debug: Check the structure of sample_data_48h
cat("Sample data columns:", colnames(sample_data_48h), "\n")
cat("Sample data class:", class(sample_data_48h), "\n")

# Check if phylogenetic tree is available
has_tree <- !is.null(phy_tree(phy_48h, errorIfNULL = FALSE))

# Calculate different distance matrices
distances <- list(
  "Bray-Curtis" = phyloseq::distance(phy_48h, method = "bray"),
  "Jaccard" = phyloseq::distance(phy_48h, method = "jaccard")
)

# Add phylogenetic distance methods only if tree is available
if (has_tree) {
  distances[["Weighted UniFrac"]] <- phyloseq::distance(phy_48h, method = "wunifrac")
  distances[["Unweighted UniFrac"]] <- phyloseq::distance(phy_48h, method = "unifrac")
  cat("Phylogenetic tree found. Including UniFrac distance methods.\n")
} else {
  cat("No phylogenetic tree found. Skipping UniFrac distance methods.\n")
  cat("Available distance methods: Bray-Curtis, Jaccard\n")
}

# Function to create PCoA plot with carbohydrate groups and ellipses
plot_pcoa_carb_ellipses <- function(dist_matrix, method_name, sample_data) {
  # Perform PCoA
  pcoa <- cmdscale(dist_matrix, eig = TRUE, k = 2)
  
  # Calculate percentage of variance explained
  eig_pct <- round(100 * pcoa$eig[1:2] / sum(pcoa$eig[pcoa$eig > 0]), 1)
  
  # Create data frame for plotting
  pcoa_df <- data.frame(
    PC1 = pcoa$points[, 1],
    PC2 = pcoa$points[, 2],
    SampleID = rownames(pcoa$points)
  ) %>%
    left_join(sample_data, by = "SampleID")
  
  # Create the plot
  p <- ggplot(pcoa_df, aes(x = PC1, y = PC2, color = carbohydrate, fill = carbohydrate)) +
    # Add 95% confidence ellipses
    stat_ellipse(level = 0.95, alpha = 0.2, size = 1.2, type = "norm") +
    geom_point(size = 3, alpha = 0.8) +
    ggsci::scale_color_nejm() +
    ggsci::scale_fill_nejm() +
    labs(
      x = paste0("PC1 (", eig_pct[1], "%)"),
      y = paste0("PC2 (", eig_pct[2], "%)"),
      title = paste(method_name, "- 48H Timepoint"),
      subtitle = "95% Confidence Ellipses by Carbohydrate Group",
      color = "Carbohydrate",
      fill = "Carbohydrate"
    ) +
    theme_classic(base_size = 14) +
    theme(
      axis.text = element_text(color = "black"),
      legend.position = "bottom",
      plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
  
  return(list(plot = p, data = pcoa_df, variance = eig_pct))
}

# Create plots for each distance method
plot_results <- map(names(distances), function(method) {
  plot_pcoa_carb_ellipses(distances[[method]], method, sample_data_48h)
})
names(plot_results) <- names(distances)

# Extract just the plots for display
plots <- map(plot_results, "plot")

# Combine plots
n_plots <- length(plots)
plot_labels <- LETTERS[1:n_plots]

combined_plot <- plot_grid(
  plotlist = plots,
  labels = plot_labels,
  nrow = 1,
  align = "hv"
)

print(combined_plot)

# Save the combined plot
plot_width <- 6 * n_plots  # 6 inches per plot
ggsave("outputs/beta_diversity_48h_carbohydrate_ellipses.png", 
       combined_plot, width = plot_width, height = 6, dpi = 300)

# Display variance explained by each method
cat("\n=== Variance Explained by PC1 and PC2 ===\n")
for (method in names(plot_results)) {
  var_exp <- plot_results[[method]]$variance
  cat(method, ": PC1 =", var_exp[1], "%, PC2 =", var_exp[2], "%\n")
}

# PERMANOVA tests for each distance method
cat("\n=== PERMANOVA Results (48H Timepoint) ===\n")
permanova_results <- map_dfr(names(distances), function(method) {
  # Prepare data for adonis2 (remove SampleID column and set row names)
  data_for_adonis <- sample_data_48h %>% 
    select(-SampleID) %>%
    as.data.frame() %>%
    `rownames<-`(sample_data_48h$SampleID)
  
  # Test carbohydrate effect
  perm_carb <- adonis2(distances[[method]] ~ carbohydrate, 
                       data = data_for_adonis, permutations = 999)
  
  # Test group effect
  perm_group <- adonis2(distances[[method]] ~ group, 
                        data = data_for_adonis, permutations = 999)
  
  # Test carbohydrate + group
  perm_both <- adonis2(distances[[method]] ~ carbohydrate + group, 
                       data = data_for_adonis, permutations = 999)
  
  tibble(
    Method = method,
    Effect = c("Carbohydrate", "Group", "Carb + Group (Carb)", "Carb + Group (Group)"),
    R2 = c(perm_carb$R2[1], perm_group$R2[1], perm_both$R2[1], perm_both$R2[2]),
    F_statistic = c(perm_carb$F[1], perm_group$F[1], perm_both$F[1], perm_both$F[2]),
    p_value = c(perm_carb$`Pr(>F)`[1], perm_group$`Pr(>F)`[1], 
                perm_both$`Pr(>F)`[1], perm_both$`Pr(>F)`[2]),
    significance = ifelse(c(perm_carb$`Pr(>F)`[1], perm_group$`Pr(>F)`[1], 
                           perm_both$`Pr(>F)`[1], perm_both$`Pr(>F)`[2]) < 0.001, "***",
                         ifelse(c(perm_carb$`Pr(>F)`[1], perm_group$`Pr(>F)`[1], 
                                 perm_both$`Pr(>F)`[1], perm_both$`Pr(>F)`[2]) < 0.01, "**",
                               ifelse(c(perm_carb$`Pr(>F)`[1], perm_group$`Pr(>F)`[1], 
                                       perm_both$`Pr(>F)`[1], perm_both$`Pr(>F)`[2]) < 0.05, "*", "")))
  )
})

print(permanova_results)

# Pairwise comparisons between carbohydrate groups
cat("\n=== Pairwise PERMANOVA: Carbohydrate Groups (48H) ===\n")
pairwise_results <- map_dfr(names(distances), function(method) {
  carb_levels <- unique(sample_data_48h$carbohydrate)
  comparisons <- combn(carb_levels, 2, simplify = FALSE)
  
  map_dfr(comparisons, function(pair) {
    subset_samples <- sample_data_48h$carbohydrate %in% pair
    subset_dist <- as.dist(as.matrix(distances[[method]])[subset_samples, subset_samples])
    subset_data <- sample_data_48h[subset_samples, ] %>% 
      select(-SampleID) %>%
      as.data.frame() %>%
      `rownames<-`(sample_data_48h$SampleID[subset_samples])
    
    perm_result <- adonis2(subset_dist ~ carbohydrate, data = subset_data, permutations = 999)
    
    tibble(
      Method = method,
      Comparison = paste(pair, collapse = " vs "),
      R2 = perm_result$R2[1],
      F_statistic = perm_result$F[1],
      p_value = perm_result$`Pr(>F)`[1],
      significance = ifelse(perm_result$`Pr(>F)`[1] < 0.001, "***",
                           ifelse(perm_result$`Pr(>F)`[1] < 0.01, "**",
                                 ifelse(perm_result$`Pr(>F)`[1] < 0.05, "*", "")))
    )
  })
})

print(pairwise_results)

# Test for homogeneity of dispersion (PERMDISP)
cat("\n=== PERMDISP: Homogeneity of Dispersions (48H) ===\n")
permdisp_results <- map_dfr(names(distances), function(method) {
  # Test carbohydrate groups
  permdisp_carb <- betadisper(distances[[method]], sample_data_48h$carbohydrate)
  permdisp_test_carb <- permutest(permdisp_carb, permutations = 999)
  
  # Test health groups
  permdisp_group <- betadisper(distances[[method]], sample_data_48h$group)
  permdisp_test_group <- permutest(permdisp_group, permutations = 999)
  
  tibble(
    Method = method,
    Factor = c("Carbohydrate", "Group"),
    F_statistic = c(permdisp_test_carb$tab$F[1], permdisp_test_group$tab$F[1]),
    p_value = c(permdisp_test_carb$tab$`Pr(>F)`[1], permdisp_test_group$tab$`Pr(>F)`[1]),
    significance = ifelse(c(permdisp_test_carb$tab$`Pr(>F)`[1], permdisp_test_group$tab$`Pr(>F)`[1]) < 0.001, "***",
                         ifelse(c(permdisp_test_carb$tab$`Pr(>F)`[1], permdisp_test_group$tab$`Pr(>F)`[1]) < 0.01, "**",
                               ifelse(c(permdisp_test_carb$tab$`Pr(>F)`[1], permdisp_test_group$tab$`Pr(>F)`[1]) < 0.05, "*", "")))
  )
})

print(permdisp_results)

# Save results
write_csv(permanova_results, "outputs/permanova_48h_results.csv")
write_csv(pairwise_results, "outputs/permanova_48h_pairwise.csv")
write_csv(permdisp_results, "outputs/permdisp_48h_results.csv")
```

## Relative Abundance Profiles

```{r relabundance-helpers, message=FALSE, warning=FALSE}
# Ensure an outputs directory exists
if (!dir.exists("outputs")) dir.create("outputs", recursive = TRUE)

# Helper: standardize empty/NA taxon names
fill_na_taxa <- function(ps) {
  tt <- tax_table(ps) %>% as.data.frame()
  rn <- rownames(tt)
  # Fill NA values progressively with higher-rank context
  ranks <- colnames(tt)
  for (rk in ranks) {
    tt[[rk]] <- ifelse(is.na(tt[[rk]]) | tt[[rk]] == "" | tt[[rk]] == "NA",
                       paste0("Unclassified_", rk), tt[[rk]])
  }
  tax_table(ps) <- as.matrix(tt)
  ps
}

# Helper: agglomerate to rank and compute relative abundance
ps_to_relabund_rank <- function(ps, rank = "Genus") {
  stopifnot(rank %in% rank_names(ps))
  ps <- fill_na_taxa(ps)
  psg <- tax_glom(ps, taxrank = rank, NArm = FALSE)
  # Relative abundances
  microbiome::transform(psg, transform = "compositional")
}

# Helper: build a highly distinct qualitative palette (colorblind-friendly)
# Uses D3 category20 sets via ggsci; concatenates up to 60 distinct hues.
make_distinct_palette <- function(n) {
  n <- as.integer(n)
  if (n <= 0) return(character(0))
  d3a <- ggsci::pal_d3("category20")(20)
  d3b <- ggsci::pal_d3("category20b")(20)
  d3c <- ggsci::pal_d3("category20c")(20)
  pal <- c(d3a, d3b, d3c)
  if (n <= length(pal)) pal[seq_len(n)] else rep_len(pal, n)
}

# Helper: collapse to top N taxa with Other
collapse_top_taxa <- function(df_long, tax_col = "Taxon", top_n = 12) {
  top_levels <- df_long %>%
    group_by(.data[[tax_col]]) %>%
    summarize(mean_abund = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(mean_abund)) %>%
    slice_head(n = top_n) %>% pull(.data[[tax_col]]) %>% as.character()

  df_long %>%
    mutate(
      !!tax_col := ifelse(.data[[tax_col]] %in% top_levels, .data[[tax_col]], "Other")
    ) %>%
    group_by(across(c(all_of(setdiff(colnames(df_long), c("Abundance")))))) %>%
    summarize(Abundance = sum(Abundance, na.rm = TRUE), .groups = "drop")
}

# Helper: generic barplot maker
plot_relabund <- function(ps_rel, rank = "Genus", group_vars, top_n = 12, palette = NULL,
                          x_lab = NULL, title = NULL) {
  df <- psmelt(ps_rel)
  tax_col <- rank
  df <- df %>% mutate(Taxon = .data[[tax_col]])

  # Combine labels for x-axis if more than one grouping var
  df <- df %>% mutate(
    group_label = interaction(!!!rlang::syms(group_vars), sep = " | ", drop = TRUE)
  )

  df_sum <- df %>%
    group_by(group_label, Taxon) %>%
    summarize(Abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")

  df_sum <- collapse_top_taxa(df_sum, tax_col = "Taxon", top_n = top_n)

  # Order taxa by overall abundance for consistent legend order
  taxa_order <- df_sum %>%
    group_by(Taxon) %>%
    summarize(total = sum(Abundance), .groups = "drop") %>%
    arrange(desc(total)) %>% pull(Taxon)
  if ("Other" %in% taxa_order) {
    taxa_order <- c(setdiff(taxa_order, "Other"), "Other")
  }
  df_sum$Taxon <- factor(df_sum$Taxon, levels = taxa_order)

  if (is.null(palette)) {
    n_taxa <- length(taxa_order)
    if (n_taxa > 1 && tail(taxa_order, 1) == "Other") {
      base_cols <- make_distinct_palette(n_taxa - 1)
      palette <- c(base_cols, "#BDBDBD")  # grey for Other
    } else {
      palette <- make_distinct_palette(n_taxa)
    }
  }

  ggplot(df_sum, aes(x = group_label, y = Abundance, fill = Taxon)) +
    geom_col(width = 0.85) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    scale_fill_manual(values = setNames(palette, taxa_order)) +
    labs(x = x_lab %||% paste(group_vars, collapse = " | "),
         y = "Relative abundance", title = title, fill = rank) +
    theme_classic(base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          axis.text = element_text(color = "black"),
          legend.key.size = unit(6, "mm"))
}

# Convenience operator
`%||%` <- function(a, b) if (!is.null(a)) a else b
```

```{r relabundance-phylum-family-genus, message=FALSE, warning=FALSE, fig.width=12, fig.height=20}
# Prepare rank-level relative abundance phyloseq objects
ps_rel_phylum <- ps_to_relabund_rank(phy_asv_finder, rank = "Phylum")
ps_rel_family <- ps_to_relabund_rank(phy_asv_finder, rank = "Family")
ps_rel_genus  <- ps_to_relabund_rank(phy_asv_finder, rank = "Genus")

# Plots by single variables
p_phylum_group <- plot_relabund(ps_rel_phylum, rank = "Phylum",
                                group_vars = c("group"), top_n = 12,
                                title = "Phylum-level Composition by Group",
                                x_lab = "Group")
p_family_group <- plot_relabund(ps_rel_family, rank = "Family",
                                group_vars = c("group"), top_n = 15,
                                title = "Family-level Composition by Group",
                                x_lab = "Group")
p_genus_group  <- plot_relabund(ps_rel_genus,  rank = "Genus",
                                group_vars = c("group"), top_n = 15,
                                title = "Genus-level Composition by Group",
                                x_lab = "Group")

p_phylum_time <- plot_relabund(ps_rel_phylum, rank = "Phylum",
                               group_vars = c("timepoint"), top_n = 12,
                               title = "Phylum-level Composition by Timepoint",
                               x_lab = "Timepoint")
p_family_time <- plot_relabund(ps_rel_family, rank = "Family",
                               group_vars = c("timepoint"), top_n = 15,
                               title = "Family-level Composition by Timepoint",
                               x_lab = "Timepoint")
p_genus_time  <- plot_relabund(ps_rel_genus,  rank = "Genus",
                               group_vars = c("timepoint"), top_n = 15,
                               title = "Genus-level Composition by Timepoint",
                               x_lab = "Timepoint")

p_phylum_carb <- plot_relabund(ps_rel_phylum, rank = "Phylum",
                               group_vars = c("carbohydrate"), top_n = 12,
                               title = "Phylum-level Composition by Carbohydrate",
                               x_lab = "Carbohydrate")
p_family_carb <- plot_relabund(ps_rel_family, rank = "Family",
                               group_vars = c("carbohydrate"), top_n = 15,
                               title = "Family-level Composition by Carbohydrate",
                               x_lab = "Carbohydrate")
p_genus_carb  <- plot_relabund(ps_rel_genus,  rank = "Genus",
                               group_vars = c("carbohydrate"), top_n = 15,
                               title = "Genus-level Composition by Carbohydrate",
                               x_lab = "Carbohydrate")

# Save representative plots
ggsave("outputs/relabund_phylum_by_group.png", p_phylum_group, width = 10, height = 6, dpi = 300)
ggsave("outputs/relabund_genus_by_carb.png",  p_genus_carb,  width = 10, height = 6, dpi = 300)

plot_grid(p_phylum_group, p_family_group, p_genus_group, ncol = 1, labels = c("A","B","C"))
```

```{r relabundance-combinations, message=FALSE, warning=FALSE, fig.width=12, fig.height=14}
# Two-way combinations
p_genus_grp_time <- plot_relabund(ps_rel_genus, rank = "Genus",
                                  group_vars = c("group", "timepoint"), top_n = 15,
                                  title = "Genus-level Composition: Group | Timepoint",
                                  x_lab = "Group | Timepoint")

p_genus_grp_carb <- plot_relabund(ps_rel_genus, rank = "Genus",
                                  group_vars = c("group", "carbohydrate"), top_n = 15,
                                  title = "Genus-level Composition: Group | Carbohydrate",
                                  x_lab = "Group | Carbohydrate")

p_genus_time_carb <- plot_relabund(ps_rel_genus, rank = "Genus",
                                   group_vars = c("timepoint", "carbohydrate"), top_n = 15,
                                   title = "Genus-level Composition: Timepoint | Carbohydrate",
                                   x_lab = "Timepoint | Carbohydrate")

plot_grid(p_genus_grp_time, p_genus_grp_carb, p_genus_time_carb, ncol = 1,
          labels = c("A","B","C"))

# Three-way combination (single overview)
p_genus_all <- plot_relabund(ps_rel_genus, rank = "Genus",
                             group_vars = c("group", "timepoint", "carbohydrate"), top_n = 15,
                             title = "Genus-level Composition: Group | Timepoint | Carbohydrate",
                             x_lab = "Group | Timepoint | Carbohydrate")

ggsave("outputs/relabund_genus_all.png", p_genus_all, width = 12, height = 6.5, dpi = 300)

p_genus_all
```

## Differential Abundance

```{r da-setup, message=FALSE, warning=FALSE}
# Keep raw counts for DA methods
ps_raw <- fill_na_taxa(phy_asv_finder)

# Check for duplicate sample names in the phyloseq object
sample_names <- sample_names(ps_raw)
if(any(duplicated(sample_names))) {
  cat("Warning: Found duplicate sample names in phyloseq object. Making them unique.\n")
  # Get unique sample names
  unique_names <- make.unique(sample_names, sep = "_")
  # Update sample names
  sample_names(ps_raw) <- unique_names
  cat("Updated sample names to:", paste(unique_names[duplicated(sample_names) | duplicated(sample_names, fromLast = TRUE)], collapse = ", "), "\n")
}

# Ensure expected covariates exist
stopifnot(all(c("group", "timepoint", "carbohydrate") %in% sample_variables(ps_raw)))

# Add SubjectID; metadata used subjectID earlier, create SubjectID if missing
sdat <- sample_data(ps_raw) %>% as.data.frame()

# Check for duplicate row names in sample data
if(any(duplicated(rownames(sdat)))) {
  cat("Warning: Found duplicate row names in sample data. Making them unique.\n")
  rownames(sdat) <- make.unique(rownames(sdat), sep = "_")
}

if (!"SubjectID" %in% colnames(sdat) && "subjectID" %in% colnames(sdat)) {
  sdat$SubjectID <- as.factor(sdat$subjectID)
  sample_data(ps_raw) <- phyloseq::sample_data(sdat)
}
if (!"SubjectID" %in% colnames(sdat)) {
  stop("SubjectID (or subjectID) is required for random effect / blocking.")
}

# Work at Genus level for DA
ps_genus_counts <- tax_glom(ps_raw, taxrank = "Genus", NArm = FALSE)

# Optional prevalence/abundance filtering to ease computations
prev_threshold <- 0.05
min_prev <- ceiling(prev_threshold * nsamples(ps_genus_counts))
keep_taxa <- taxa_sums(ps_genus_counts) > 0 & rowSums(otu_table(ps_genus_counts) > 0) >= min_prev
ps_genus_filt <- prune_taxa(keep_taxa, ps_genus_counts)

cat("Genus taxa before/after filtering:", ntaxa(ps_genus_counts), ntaxa(ps_genus_filt), "\n")
```

### Corncob (beta-binomial)

```{r da-corncob, message=FALSE, warning=FALSE}
# if the output file already exists, skip the analysis and load the results
if(file.exists("outputs/corncob_genus_hits.csv")) {
 res <- readRDS("outputs/corncob_genus_hits.rdc")
  cat("Loaded Corncob results from file.\n")
} else {

  set.seed(12345)
  
  res <- tryCatch({
    differentialTest(
      formula = ~group+timepoint+carbohydrate,
      phi.formula = ~group+timepoint+carbohydrate,
      formula_null = ~group+timepoint, 
      phi.formula_null = ~group+timepoint+carbohydrate,
      test = "Wald",
      data = ps,
      boot = FALSE,
      B = 1000,
      fdr_cutoff = 0.1
    )
    saveRDS(res, "outputs/corncob_genus_hits.rds")
  }, error = function(e) {
    cat("Corncob optimization failed for", effect_var, ":", e$message, "\n")
   
}

plot(res, B=1000)
```

### ANCOM-BC2 (with SubjectID random effect)

```{r da-ancombc2, message=FALSE, warning=FALSE}
# if the output file already exists, skip the analysis and load the results
if(file.exists("outputs/ancombc_genus_results.csv")) {
  res_tbl <- read_csv("outputs/ancombc_genus_results.csv")
  cat("Loaded ANCOMBC2 results from file.\n")
  use_ancombc2 <- FALSE
} else {
  use_ancombc2 <- "ancombc2" %in% getNamespaceExports("ANCOMBC")

  # CRITICAL: Fix duplicate sample names before ANCOM-BC analysis
  # This is essential because ANCOM-BC functions are sensitive to duplicate names
  cat("Checking for duplicate sample names before ANCOM-BC analysis...\n")

  # Fix ps_raw
  if(any(duplicated(sample_names(ps_raw)))) {
    cat("Fixing duplicate sample names in ps_raw\n")
    unique_names <- make.unique(sample_names(ps_raw), sep = "_")
    sample_names(ps_raw) <- unique_names
    cat("Updated", sum(duplicated(sample_names(ps_raw)) | duplicated(sample_names(ps_raw), fromLast = TRUE)), "duplicate names\n")
  }

  # Fix ps_genus_filt
  if(any(duplicated(sample_names(ps_genus_filt)))) {
    cat("Fixing duplicate sample names in ps_genus_filt\n")
    unique_names <- make.unique(sample_names(ps_genus_filt), sep = "_")
    sample_names(ps_genus_filt) <- unique_names
    cat("Updated", sum(duplicated(sample_names(ps_genus_filt)) | duplicated(sample_names(ps_genus_filt), fromLast = TRUE)), "duplicate names\n")
  }

  # Verify no duplicates remain
  cat("Final check - ps_raw duplicates:", sum(duplicated(sample_names(ps_raw))), "\n")
  cat("Final check - ps_genus_filt duplicates:", sum(duplicated(sample_names(ps_genus_filt))), "\n")

  if (use_ancombc2) {
    cat("Running ANCOMBC2...\n")
    out_ancom <- tryCatch({
      ANCOMBC::ancombc2(
        data = ps_raw,
        tax_level = "Genus",
        fix_formula = "group + timepoint + carbohydrate",
        rand_formula = "(1|SubjectID)",
        p_adj_method = "holm",
        prv_cut = 0.10,
        lib_cut = 0,
        s0_perc = 0.05,
        group = "group",
        struc_zero = TRUE,
        neg_lb = TRUE,
        alpha = 0.05,
        n_cl = 1,
        global = TRUE
      )
    }, error = function(e) {
      cat("ANCOMBC2 failed with error:", e$message, "\n")
      cat("Falling back to ANCOM-BC v1...\n")
      NULL
    })

    if (!is.null(out_ancom)) {
      # Extract results
      res_tbl <- out_ancom$res
    } else {
      use_ancombc2 <- FALSE  # Force fallback
    }
  }

  if (!use_ancombc2) {
    message("Using ANCOM-BC (v1) without random effects.")
    out_ancom <- tryCatch({
      ANCOMBC::ancombc(
        phyloseq = ps_genus_filt,
        formula = "group + timepoint + carbohydrate",
        p_adj_method = "holm",
        zero_cut = 0.90,
        lib_cut = 0,
        group = "group",
        struc_zero = TRUE,
        neg_lb = TRUE,
        alpha = 0.05,
        global = TRUE
      )
    }, error = function(e) {
      cat("ANCOM-BC v1 also failed with error:", e$message, "\n")
      NULL
    })

    if (!is.null(out_ancom)) {
      # Tidy results from v1 object
      res_list <- lapply(names(out_ancom$beta), function(term) {
        tibble(
          taxon = rownames(out_ancom$beta[[term]]),
          term = term,
          beta = out_ancom$beta[[term]][, 1],
          se = out_ancom$se[[term]][, 1],
          W = out_ancom$W[[term]][, 1],
          p = out_ancom$p_val[[term]][, 1],
          q = out_ancom$q_val[[term]][, 1],
          diff_abn = out_ancom$diff_abn[[term]][, 1]
        )
      })
      res_tbl <- bind_rows(res_list)
    } else {
      cat("Both ANCOM-BC methods failed. Creating empty result table.\n")
      res_tbl <- NULL
    }
  }

  # Ensure res_tbl exists and has the expected structure
  if (!exists("res_tbl") || is.null(res_tbl)) {
    cat("Warning: res_tbl not created properly. Creating empty result table.\n")
    res_tbl <- tibble(
      taxon = character(0),
      term = character(0),
      beta = numeric(0),
      se = numeric(0),
      W = numeric(0),
      p = numeric(0),
      q = numeric(0),
      diff_abn = logical(0)
    )
  }

  write_csv(res_tbl, "outputs/ancombc_genus_results.csv")
  head(res_tbl)
}
```

### ANCOM-BC Analysis: 48H Timepoint with RDC as Reference

```{r da-ancombc-48h-rdc-reference, message=FALSE, warning=FALSE}
# if the output file already exists, skip the analysis and load the results
if(file.exists("outputs/ancombc_48h_rdc_ref_results.csv")) {
  res_48h_rdc <- read_csv("outputs/ancombc_48h_rdc_ref_results.csv")
  cat("Loaded 48H RDC reference ANCOMBC results from file.\n")
} else { 
# Filter data to 48H timepoint only
ps_48h_raw <- subset_samples(ps_raw, timepoint == "48H")
ps_48h_genus <- subset_samples(ps_genus_filt, timepoint == "48H")

# Remove taxa that are completely absent in 48H samples
ps_48h_raw <- prune_taxa(taxa_sums(ps_48h_raw) > 0, ps_48h_raw)
ps_48h_genus <- prune_taxa(taxa_sums(ps_48h_genus) > 0, ps_48h_genus)

cat("48H samples for analysis:", nsamples(ps_48h_genus), "\n")
cat("Taxa after 48H filtering:", ntaxa(ps_48h_genus), "\n")

# Relevel carbohydrate factor to use RDC as reference
sample_data_48h <- as.data.frame(sample_data(ps_48h_genus))
sample_data_48h$carbohydrate <- factor(sample_data_48h$carbohydrate, levels = c("RDC", "No Added Carb", "SDC"))

# Update the phyloseq object with releveled factor
sample_data(ps_48h_genus) <- sample_data_48h
sample_data(ps_48h_raw) <- sample_data_48h

# Check the levels
cat("Carbohydrate levels (RDC is reference):", levels(sample_data_48h$carbohydrate), "\n")
cat("Sample distribution by carbohydrate:\n")
print(table(sample_data_48h$carbohydrate))

# Run ANCOM-BC analysis if results don't exist
if(file.exists("outputs/ancombc_48h_rdc_ref_results.csv")) {
  res_48h_rdc <- read_csv("outputs/ancombc_48h_rdc_ref_results.csv")
  cat("Loaded 48H RDC reference ANCOMBC results from file.\n")
} else {
  use_ancombc2_48h <- "ancombc2" %in% getNamespaceExports("ANCOMBC")
  
  if (use_ancombc2_48h) {
    cat("Running ANCOMBC2 for 48H timepoint with RDC as reference...\n")
    
    # Set RDC as the reference level for carbohydrate
    sample_data(ps_48h_raw)$carbohydrate <- relevel(sample_data(ps_48h_raw)$carbohydrate, ref = "RDC")
    cat("Carbohydrate levels (RDC is reference):", levels(sample_data(ps_48h_raw)$carbohydrate), "\n")
    
    out_ancom_48h <- tryCatch({
      ANCOMBC::ancombc2(
        data = ps_48h_raw,
        tax_level = "Genus",
        fix_formula = "group + carbohydrate",  # Only group and carbohydrate since we're at single timepoint
        rand_formula = "(1|SubjectID)",
        p_adj_method = "holm",
        prv_cut = 0.10,
        lib_cut = 0,
        s0_perc = 0.05,
        group = "carbohydrate",  # Primary interest is carbohydrate effects
        struc_zero = TRUE,
        neg_lb = TRUE,
        alpha = 0.05,
        n_cl = 1,
        global = TRUE
      )
    }, error = function(e) {
      cat("ANCOMBC2 failed with error:", e$message, "\n")
      cat("Falling back to ANCOM-BC v1...\n")
      NULL
    })
    
    if (!is.null(out_ancom_48h)) {
      # Extract results from ANCOMBC2
      res_48h_rdc <- out_ancom_48h$res
    } else {
      use_ancombc2_48h <- FALSE  # Force fallback
    }
  }
  
  if (!use_ancombc2_48h) {
    cat("Using ANCOM-BC (v1) for 48H analysis...\n")
    
    # Set RDC as the reference level for carbohydrate
    sample_data(ps_48h_genus)$carbohydrate <- relevel(sample_data(ps_48h_genus)$carbohydrate, ref = "RDC")
    cat("Carbohydrate levels (RDC is reference):", levels(sample_data(ps_48h_genus)$carbohydrate), "\n")
    
    out_ancom_48h <- tryCatch({
      ANCOMBC::ancombc(
        phyloseq = ps_48h_genus,
        formula = "group + carbohydrate",
        p_adj_method = "holm",
        zero_cut = 0.90,
        lib_cut = 0,
        group = "carbohydrate",
        struc_zero = TRUE,
        neg_lb = TRUE,
        alpha = 0.05,
        global = TRUE
      )
    }, error = function(e) {
      cat("ANCOM-BC v1 also failed with error:", e$message, "\n")
      NULL
    })
    
    if (!is.null(out_ancom_48h)) {
      # Use the res data frame directly - it already has the structure we need
      res_48h_rdc <- out_ancom_48h$res
    } else {
      cat("Both ANCOM-BC methods failed for 48H analysis. Creating empty result table.\n")
      res_48h_rdc <- tibble(
        taxon = character(0),
        term = character(0),
        beta = numeric(0),
        se = numeric(0),
        W = numeric(0),
        p = numeric(0),
        q = numeric(0),
        diff_abn = logical(0)
      )
    }
  }
  
  write_csv(res_48h_rdc, "outputs/ancombc_48h_rdc_ref_results.csv")
}


cat("ANCOM-BC 48H Results Summary:\n")
if (nrow(res_48h_rdc) > 0) {
  cat("Total taxa tested:", length(unique(res_48h_rdc$taxon)), "\n")
  
  # Summary by term
  # Create summary by term for wide format data
  summary_by_term <- tibble(
    term = c("carbohydrateNo.Added.Carb", "carbohydrateSDC", "groupObese"),
    term_clean = c("No Added Carb vs RDC", "SDC vs RDC", "Health Group Effect"),
    total_taxa = nrow(res_48h_rdc),
    significant = c(
      sum(res_48h_rdc$q_carbohydrateNo.Added.Carb < 0.05, na.rm = TRUE),
      sum(res_48h_rdc$q_carbohydrateSDC < 0.05, na.rm = TRUE),
      sum(res_48h_rdc$q_groupObese < 0.05, na.rm = TRUE)
    )
  ) %>%
    mutate(prop_significant = round(significant / total_taxa * 100, 1))
  
  print(summary_by_term)
} else {
  cat("No results to summarize.\n")
}

head(res_48h_rdc)
}
```

```{r ancombc-48h-summary-table, message=FALSE, warning=FALSE}
# Create comprehensive summary table for 48H RDC reference analysis
if (nrow(res_48h_rdc) > 0) {
  
  # Extract carbohydrate effect results (No Added Carb and SDC vs RDC reference)
  carb_effects <- res_48h_rdc %>%
    # Filter for significant carbohydrate effects
    filter(q_carbohydrateNo.Added.Carb < 0.05 | q_carbohydrateSDC < 0.05) %>%
    # Create separate rows for each comparison
    bind_rows(
      # No Added Carb vs RDC
      filter(res_48h_rdc, q_carbohydrateNo.Added.Carb < 0.05) %>%
        select(taxon, lfc_carbohydrateNo.Added.Carb, se_carbohydrateNo.Added.Carb, p_carbohydrateNo.Added.Carb, q_carbohydrateNo.Added.Carb, diff_carbohydrateNo.Added.Carb) %>%
        rename(beta = lfc_carbohydrateNo.Added.Carb, se = se_carbohydrateNo.Added.Carb, p = p_carbohydrateNo.Added.Carb, q = q_carbohydrateNo.Added.Carb, diff_abn = diff_carbohydrateNo.Added.Carb) %>%
        mutate(term = "carbohydrateNo.Added.Carb", term_clean = "No Added Carb vs RDC"),
      # SDC vs RDC  
      filter(res_48h_rdc, q_carbohydrateSDC < 0.05) %>%
        select(taxon, lfc_carbohydrateSDC, se_carbohydrateSDC, p_carbohydrateSDC, q_carbohydrateSDC, diff_carbohydrateSDC) %>%
        rename(beta = lfc_carbohydrateSDC, se = se_carbohydrateSDC, p = p_carbohydrateSDC, q = q_carbohydrateSDC, diff_abn = diff_carbohydrateSDC) %>%
        mutate(term = "carbohydrateSDC", term_clean = "SDC vs RDC")
    ) %>%
    arrange(term, desc(abs(beta))) %>%
    mutate(
      direction = ifelse(beta > 0, "Higher vs RDC", "Lower vs RDC"),
      fold_change = exp(beta)
    )
  
  # Create summary table
  if (nrow(carb_effects) > 0) {
    cat("\n=== SIGNIFICANT CARBOHYDRATE EFFECTS AT 48H (vs RDC Reference) ===\n")
    
    summary_table <- carb_effects %>%
      select(Genus, term_clean, beta, fold_change, direction, p, q, 
             Phylum, Family) %>%
      arrange(term_clean, desc(abs(beta)))
    
    print(summary_table)
    
    # Save detailed results
    write_csv(carb_effects, "outputs/ancombc_48h_carb_effects_vs_rdc.csv")
    write_csv(summary_table, "outputs/ancombc_48h_summary_table.csv")
    
    # Count by comparison and direction
    effect_counts <- carb_effects %>%
      group_by(term_clean, direction) %>%
      summarise(count = n(), .groups = "drop") %>%
      arrange(term_clean, direction)
    
    cat("\n=== EFFECT COUNTS BY COMPARISON ===\n")
    print(effect_counts)
    
    # Top genera by effect size for each comparison
    top_effects <- carb_effects %>%
      group_by(term_clean) %>%
      slice_max(order_by = abs(beta), n = 5) %>%
      select(term_clean, Genus, beta, fold_change, q) %>%
      arrange(term_clean, desc(abs(beta)))
    
    cat("\n=== TOP 5 GENERA BY EFFECT SIZE FOR EACH COMPARISON ===\n")
    print(top_effects)
    
  } else {
    cat("No significant carbohydrate effects found in 48H analysis.\n")
  }
  
} else {
  cat("No results available for summary table.\n")
}
```

```{r ancombc-48h-forest-plots, message=FALSE, warning=FALSE, fig.width=12, fig.height=10}
# Forest plots for 48H carbohydrate effects vs RDC reference

if (nrow(res_48h_rdc) > 0) {
  
  # Function to create forest plot for 48H analysis
  plot_48h_forest <- function(data, term_pattern, title_suffix, top_n = 20) {
    
    df <- data %>%
      filter(grepl(term_pattern, term, ignore.case = TRUE)) %>%
      mutate(
        ci_l = beta - 1.96 * se,
        ci_u = beta + 1.96 * se,
        sig = q < 0.05,
        direction = ifelse(beta > 0, "Higher vs RDC", "Lower vs RDC"),
        genus_label = ifelse(!is.na(Genus) & Genus != "", Genus, taxon)
      )
    
    if (nrow(df) == 0) return(NULL)
    
    # Take top N by absolute effect size, prioritizing significant results
    df_plot <- df %>%
      arrange(desc(sig), desc(abs(beta))) %>%
      slice_head(n = top_n) %>%
      mutate(genus_label = factor(genus_label, levels = rev(unique(genus_label[order(beta)]))))
    
    p <- ggplot(df_plot, aes(x = beta, y = genus_label, color = direction)) +
      geom_vline(xintercept = 0, linetype = 2, color = "grey50") +
      geom_errorbarh(aes(xmin = ci_l, xmax = ci_u, alpha = sig), height = 0.3) +
      geom_point(aes(size = sig, alpha = sig)) +
      scale_alpha_manual(values = c("TRUE" = 1.0, "FALSE" = 0.4), guide = "none") +
      scale_size_manual(values = c("TRUE" = 2.5, "FALSE" = 1.5), guide = "none") +
      ggsci::scale_color_lancet() +
      labs(
        x = "Log fold-change (bias-corrected)",
        y = NULL,
        title = paste("48H Carbohydrate Effects:", title_suffix),
        subtitle = "Solid/large points: FDR < 0.05; Faded/small points: FDR ≥ 0.05",
        color = "Direction"
      ) +
      theme_classic(base_size = 12) +
      theme(
        axis.text = element_text(color = "black"),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, size = 10)
      )
    
    return(p)
  }
  
  # Create plots for each comparison
  plots_48h <- list()
  
  # Create long format data for forest plots
  forest_data_48h <- bind_rows(
    # No Added Carb vs RDC
    res_48h_rdc %>%
      select(taxon, lfc_carbohydrateNo.Added.Carb, se_carbohydrateNo.Added.Carb, p_carbohydrateNo.Added.Carb, q_carbohydrateNo.Added.Carb, diff_carbohydrateNo.Added.Carb) %>%
      rename(beta = lfc_carbohydrateNo.Added.Carb, se = se_carbohydrateNo.Added.Carb, p = p_carbohydrateNo.Added.Carb, q = q_carbohydrateNo.Added.Carb, diff_abn = diff_carbohydrateNo.Added.Carb) %>%
      mutate(term = "carbohydrateNo.Added.Carb"),
    # SDC vs RDC
    res_48h_rdc %>%
      select(taxon, lfc_carbohydrateSDC, se_carbohydrateSDC, p_carbohydrateSDC, q_carbohydrateSDC, diff_carbohydrateSDC) %>%
      rename(beta = lfc_carbohydrateSDC, se = se_carbohydrateSDC, p = p_carbohydrateSDC, q = q_carbohydrateSDC, diff_abn = diff_carbohydrateSDC) %>%
      mutate(term = "carbohydrateSDC"),
    # Group effect
    res_48h_rdc %>%
      select(taxon, lfc_groupObese, se_groupObese, p_groupObese, q_groupObese, diff_groupObese) %>%
      rename(beta = lfc_groupObese, se = se_groupObese, p = p_groupObese, q = q_groupObese, diff_abn = diff_groupObese) %>%
      mutate(term = "groupObese")
  )
  
  # No Added Carb vs RDC
  if (any(grepl("carbohydrateNo.Added.Carb", forest_data_48h$term, ignore.case = TRUE))) {
    plots_48h$no_carb <- plot_48h_forest(forest_data_48h, "carbohydrateNo.Added.Carb", "No Added Carb vs RDC")
  }
  
  # SDC vs RDC  
  if (any(grepl("carbohydrateSDC", forest_data_48h$term, ignore.case = TRUE))) {
    plots_48h$sdc <- plot_48h_forest(forest_data_48h, "carbohydrateSDC", "SDC vs RDC")
  }
  
  # Group effect for reference
  if (any(grepl("groupObese", forest_data_48h$term, ignore.case = TRUE))) {
    plots_48h$group <- plot_48h_forest(forest_data_48h, "groupObese", "Health Group Effect")
  }
  
  # Combine and display plots
  valid_plots <- plots_48h[!sapply(plots_48h, is.null)]
  
  if (length(valid_plots) > 0) {
    # Save individual plots
    for (nm in names(valid_plots)) {
      ggsave(file.path("outputs", paste0("ancombc_48h_forest_", nm, "_vs_rdc.png")), 
             valid_plots[[nm]], width = 10, height = 8, dpi = 300)
    }
    
    # Display combined plot
    if (length(valid_plots) == 1) {
      print(valid_plots[[1]])
    } else {
      combined_48h <- cowplot::plot_grid(plotlist = valid_plots, ncol = 1, align = "v")
      print(combined_48h)
      
      # Save combined plot
      ggsave("outputs/ancombc_48h_forest_combined_vs_rdc.png", combined_48h, 
             width = 12, height = 6 * length(valid_plots), dpi = 300)
    }
  } else {
    cat("No valid forest plots could be created for 48H analysis.\n")
  }
  
} else {
  cat("No data available for 48H forest plots.\n")
}
```

```{r da-lefse, message=FALSE, warning=FALSE}
# if the output file already exists, skip the analysis and load the results
if(file.exists("outputs/lefse_genus_results.csv")) {
  lefse_group <- readRDS("outputs/lefse_group_genus.rds")
  lefse_carb <- readRDS("outputs/lefse_carb_genus.rds")
  lefse_time <- readRDS("outputs/lefse_time_genus.rds")
  cat("Loaded LEfSe results from file.\n")
} else {
# LEfSe via microbiomeMarker; use subclass for within-subject pairing
set.seed(123)
lefse_group <- tryCatch({
  microbiomeMarker::run_lefse(
    ps_genus_filt,
    group = "group",
    norm = "CPM",
    taxa_rank = "Genus",
    wilcoxon_cutoff = 0.05,
    lda_cutoff = 2.0,
    bootstrap_n = 30
  )
}, error = function(e) NULL)

lefse_carb <- tryCatch({
  microbiomeMarker::run_lefse(
    ps_genus_filt,
    group = "carbohydrate",
    norm = "CPM",
    taxa_rank = "Genus",
    wilcoxon_cutoff = 0.05,
    lda_cutoff = 2.0,
    bootstrap_n = 30
  )
}, error = function(e) NULL)

lefse_time <- tryCatch({
  microbiomeMarker::run_lefse(
    ps_genus_filt,
    group = "timepoint",
    norm = "CPM",
    taxa_rank = "Genus",
    wilcoxon_cutoff = 0.05,
    lda_cutoff = 2.0,
    bootstrap_n = 30
  )
}, error = function(e) NULL)

saveRDS(lefse_group, file = "outputs/lefse_group_genus.rds")
saveRDS(lefse_carb,  file = "outputs/lefse_carb_genus.rds")
saveRDS(lefse_time,  file = "outputs/lefse_time_genus.rds")
}

if (!is.null(lefse_carb)) {
  p_lda <- microbiomeMarker::plot_ef_bar(lefse_carb)
  ggsave("outputs/lefse_lda_carb_genus.png", p_lda, width = 8, height = 6, dpi = 300)
  p_lda
}
```

### Notes on Modeling Choices

- Corncob: SubjectID is included as a blocking (fixed) factor since random effects are not supported in differentialTest. This adjusts for repeated measures when testing each effect by drop-one comparison.
- ANCOM-BC2: Uses `rand_formula = (1|SubjectID)` to include a subject-specific random intercept when available; falls back to ANCOM-BC v1 otherwise.
- LEfSe: Uses `subclass = SubjectID` to perform paired LEfSe appropriate for repeated measures.

```{r ancombc-forest-plots, message=FALSE, warning=FALSE, fig.width=14, fig.height=24}
# Forest plots for ANCOM-BC results (effect sizes with 95% CI)

make_pretty_term <- function(term_key) {
  # Map internal column suffixes to readable labels
  term_key <- as.character(term_key)
  term_key <- gsub("[()]+", "", term_key)
  case_when(
    term_key == "groupObese" ~ "Group: Obese vs Healthy",
    term_key == "timepoint48H" ~ "Timepoint: 48H vs reference",
    term_key == "carbohydrateRDC" ~ "Carbohydrate: RDC vs No Added Carb",
    term_key == "carbohydrateSDC" ~ "Carbohydrate: SDC vs No Added Carb",
    TRUE ~ term_key
  )
}

plot_forest_from_wide <- function(res_wide, term_key, top_n = 25, alpha = 0.05,
                                  title_prefix = "ANCOM-BC Genus Effects") {
  lfc_col <- paste0("lfc_", term_key)
  se_col  <- paste0("se_", term_key)
  q_col   <- paste0("q_", term_key)
  diff_col<- paste0("diff_", term_key)

  if (!all(c("taxon", lfc_col) %in% names(res_wide))) return(NULL)

  df <- res_wide %>%
    transmute(
      taxon,
      term = make_pretty_term(term_key),
      lfc = as.numeric(.data[[lfc_col]]),
      se  = ifelse(se_col %in% names(res_wide), as.numeric(.data[[se_col]]), NA_real_),
      q   = ifelse(q_col %in% names(res_wide), as.numeric(.data[[q_col]]), NA_real_),
      diff= ifelse(diff_col %in% names(res_wide), .data[[diff_col]], NA)
    ) %>%
    mutate(
      # Ensure se is numeric and handle any remaining non-numeric values
      se = ifelse(is.na(se) | !is.finite(se), NA_real_, as.numeric(se)),
      ci_l = ifelse(is.na(se), NA_real_, lfc - 1.96 * se), 
      ci_u = ifelse(is.na(se), NA_real_, lfc + 1.96 * se),
      sig = ifelse(!is.na(q) & q < alpha, TRUE, FALSE),
      direction = ifelse(lfc > 0, "Higher", "Lower")
    )

  # Keep significant taxa, or top by |lfc|
  df_sig <- df %>% filter(sig)
  if (nrow(df_sig) == 0) {
    df_sig <- df %>% slice_max(order_by = abs(lfc), n = min(top_n, nrow(df)))
  } else {
    df_sig <- df_sig %>% arrange(desc(abs(lfc))) %>% slice_head(n = min(top_n, nrow(df_sig)))
  }

  df_sig <- df_sig %>% mutate(taxon = factor(taxon, levels = rev(unique(taxon[order(lfc)]))))

  ggplot(df_sig, aes(x = lfc, y = taxon, color = direction)) +
    geom_vline(xintercept = 0, linetype = 2, color = "grey50") +
    geom_errorbarh(aes(xmin = ci_l, xmax = ci_u), height = 0.2, alpha = 0.9) +
    geom_point(size = 2) +
    ggsci::scale_color_lancet() +
    labs(x = "Log fold-change (bias-corrected)", y = NULL,
         title = paste(title_prefix, "-", make_pretty_term(term_key))) +
    theme_classic(base_size = 12) +
    theme(axis.text = element_text(color = "black"))
}

plot_forest_from_long <- function(res_long, term_key, top_n = 25, alpha = 0.05,
                                  title_prefix = "ANCOM-BC Genus Effects") {
  df <- res_long %>% filter(term == term_key) %>%
    transmute(
      taxon, 
      term = make_pretty_term(term_key), 
      lfc = as.numeric(beta), 
      se = as.numeric(se), 
      q = as.numeric(q), 
      diff = diff_abn
    ) %>%
    mutate(
      # Ensure se is numeric and handle any remaining non-numeric values
      se = ifelse(is.na(se) | !is.finite(se), NA_real_, as.numeric(se)),
      ci_l = ifelse(is.na(se), NA_real_, lfc - 1.96 * se), 
      ci_u = ifelse(is.na(se), NA_real_, lfc + 1.96 * se),
      sig = ifelse(!is.na(q) & q < alpha, TRUE, FALSE),
      direction = ifelse(lfc > 0, "Higher", "Lower")
    )

  df_sig <- df %>% filter(sig)
  if (nrow(df_sig) == 0) {
    df_sig <- df %>% slice_max(order_by = abs(lfc), n = min(top_n, nrow(df)))
  } else {
    df_sig <- df_sig %>% arrange(desc(abs(lfc))) %>% slice_head(n = min(top_n, nrow(df_sig)))
  }
  df_sig <- df_sig %>% mutate(taxon = factor(taxon, levels = rev(unique(taxon[order(lfc)]))))

  ggplot(df_sig, aes(x = lfc, y = taxon, color = direction)) +
    geom_vline(xintercept = 0, linetype = 2, color = "grey50") +
    geom_errorbarh(aes(xmin = ci_l, xmax = ci_u), height = 0.2, alpha = 0.9) +
    geom_point(size = 2) +
    ggsci::scale_color_lancet() +
    labs(x = "Log fold-change (bias-corrected)", y = NULL,
         title = paste(title_prefix, "-", make_pretty_term(term_key))) +
    theme_classic(base_size = 12) +
    theme(axis.text = element_text(color = "black"))
}

# Decide which plotting path to use based on res_tbl structure
is_wide <- any(grepl("^lfc_", names(res_tbl)))

terms_to_plot <- c("groupObese", "timepoint48H", "carbohydrateRDC", "carbohydrateSDC")
plots <- list()
for (tm in terms_to_plot) {
  p <- if (is_wide) plot_forest_from_wide(res_tbl, tm) else plot_forest_from_long(res_tbl, tm)
  if (!is.null(p)) plots[[tm]] <- p
}

# Save individual plots and show a combined grid of available ones
plot_list <- plots[!vapply(plots, is.null, logical(1))]
if (length(plot_list) > 0) {
  for (nm in names(plot_list)) {
    ggsave(file.path("outputs", paste0("ancombc_forest_", nm, ".png")), plot_list[[nm]],
           width = 8.5, height = 6.5, dpi = 300)
  }
  cowplot::plot_grid(plotlist = plot_list, ncol = 1)
}
```


```{r supervised-learning-analysis}
library(microbiomemarker)

set.seed(12345)
res <- run_sl(
    ps_genus_filt %>% subset_samples(carbohydrate%in%c("RDC", "SDC") & timepoint == "48H"),
    group = "carbohydrate",
    nfolds = 2,
    nrepeats = 1,
    taxa_rank = "Genus",
    top_n = 25,
    norm = "TSS",
    method = "LR",
)

```

```{r adapt-da-analysis}
library(ADAPT)

adapt_output <- adapt(input_data=ps_genus_filt, 
  cond.var = "carbohydrate", 
  base.cond = "No Added Carb", 
  time.var = "timepoint", 
  random.var = "subjectID")
  adapt_DAtaxa <- adapt_output@signal
  adapt_details <- adapt_output@details
  adapt_summary <- data.frame(Taxa = adapt_DAtaxa, ADAPT=adapt_details[adapt_DAtaxa,  "log10foldchange"] > 0)
  adapt_summary$ADAPT <- 2*(adapt_summary$ADAPT-0.5)

```

```{r maaslin2-da-analysis}
library(Maaslin2)

count_df <- data.frame(otu_table(ps_genus_filt))

maaslin2_output <-invisible(Maaslin2(input_data=count_df, input_metadata=metadata,
                                     output = file.path("outputs", 'maaslin2_output'),
                                     min_prevalence=0.05,
                                     max_significance = 0.05,
                                     fixed_effects=c("group", "carbohydrate", "timepoint"),
                                     random_effects=c("subjectID"),
                                     cores=10, plot_heatmap = TRUE,plot_scatter = TRUE))
maaslin2_result <- maaslin2_output$results
maaslin2_DAtaxa <- maaslin2_result$feature[maaslin2_result$qval < 0.05]
library(dplyr)
maaslin2_summary <- maaslin2_result %>% filter(qval < 0.05) %>%
  dplyr::select(feature, coef) %>% mutate(Taxa=feature, Maaslin2=coef > 0, .keep="none")
maaslin2_summary$Maaslin2 <- 2*(maaslin2_summary$Maaslin2 - 0.5)
```