---
title: "Microbiome Analysis v3 - ASV Finder"
author: "Abbott Carbohydrate Obesity Project"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    code_folding: hide
    theme: flatly
    highlight: tango
    number_sections: true
    fig_width: 10
    fig_height: 6
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 5.5,
  cache = FALSE
)
```

## 1. Analysis Roadmap

This v3 report refines the v2 workflow to deliver a cleaner, more cohesive analysis with harmonized graphics, consistent statistical modeling, and dynamic interpretation. The key updates include:

- Streamlined data processing with shared helper functions and reproducible checks.
- Consistent mixed-effects modeling that respects the repeated-measures design (`SubjectID` random intercepts).
- False discovery rate control (Benjamini–Hochberg) applied to all multi-comparison contexts.
- Publication-quality visual styling and unified color palettes for categorical variables.
- Interpretation blocks populated via inline code so narrative text always reflects current outputs.
- Expanded differential-abundance section spanning corncob, ANCOM-BC2, ALDEx2, LEfSe, MaAsLin2, ADAPT, random forest, and consensus summaries.
- Dedicated focus on 48 h responses, baseline clustering, and analyses with combined groups for improved power when appropriate.

## 2. System Check and Memory Optimization

```{r system-check}
# Check system resources and optimize for large datasets
cat("System Information:\n")
cat("R version:", R.version.string, "\n")
cat("Platform:", R.version$platform, "\n")
cat("Memory limit:", memory.limit(), "MB\n")

# Check available memory
if (.Platform$OS.type == "unix") {
  mem_info <- system("free -h", intern = TRUE)
  cat("System memory:\n")
  cat(paste(mem_info, collapse = "\n"), "\n")
}

# Set memory-efficient options
options(
  scipen = 999,  # Avoid scientific notation
  digits = 4,    # Limit decimal places
  max.print = 100,  # Limit printed output
  stringsAsFactors = FALSE
)

# Increase memory limit if possible
tryCatch({
  if (.Platform$OS.type == "windows") {
    memory.limit(size = 8000)  # 8GB
  }
}, error = function(e) {
  cat("Could not increase memory limit:", e$message, "\n")
})

cat("System check completed.\n")
```

## 3. Libraries, Options, and Helper Functions

```{r libraries}
required_packages <- c(
  "tidyverse", "readr", "janitor", "glue", "scales", "patchwork",
  "phyloseq", "microbiome", "vegan", "betapart", "ape", "Biostrings",
  "broom", "broom.mixed", "lme4", "lmerTest", "emmeans", "performance",
  "rstatix", "ggpubr", "ggtext", "cowplot", "ggsci", "scico",
  "here", "corncob", "ANCOMBC", "ALDEx2", "microbiomeMarker", "ADAPT",
  "Maaslin2", "gt", "cluster", "nlme", "randomForest", "ranger", "caret", "phyloseqGraphTest"
)

# Function to install missing packages
install_missing_packages <- function(packages) {
  missing_packages <- packages[!vapply(packages, requireNamespace, logical(1), quietly = TRUE)]
  
  if (length(missing_packages) > 0) {
    cat("Installing missing packages:", paste(missing_packages, collapse = ", "), "\n")
    
    # Bioconductor packages
    bioc_packages <- c("phyloseq", "microbiome", "Biostrings", "ANCOMBC", "ALDEx2", "microbiomeMarker", "Maaslin2", "phyloseqGraphTest", "ADAPT")
    bioc_to_install <- intersect(missing_packages, bioc_packages)
    
    # CRAN packages
    cran_to_install <- setdiff(missing_packages, bioc_packages)
    
    # Install Bioconductor packages
    if (length(bioc_to_install) > 0) {
      if (!requireNamespace("BiocManager", quietly = TRUE)) {
        install.packages("BiocManager")
      }
      BiocManager::install(bioc_to_install, update = FALSE, ask = FALSE)
    }
    
    # Install CRAN packages
    if (length(cran_to_install) > 0) {
      install.packages(cran_to_install, repos = "https://cloud.r-project.org/")
    }
    
    cat("Package installation completed.\n")
  }
}

# Install missing packages
install_missing_packages(required_packages)

invisible(lapply(required_packages, library, character.only = TRUE))

options(dplyr.summarise.inform = FALSE)
set.seed(12345)

# Output directory for derived artefacts
outputs_dir <- here::here("outputs")
dir.create(outputs_dir, showWarnings = FALSE, recursive = TRUE)
```

```{r theming}
# Harmonized aesthetic ---------------------------------------------------

palette_group <- setNames(scico::scico(2, palette = "berlin"), c("Healthy", "Obese"))
palette_carbohydrate <- setNames(c("#1f77b4", "#ff7f0e", "#2ca02c"), c("No Added Carb", "RDC", "SDC"))
palette_timepoint <- setNames(scico::scico(2, palette = "roma"), c("0H", "48H"))

palette_cluster <- setNames(scico::scico(6, palette = "batlow"), paste0("Cluster", seq_len(6)))

custom_theme <- function(base_size = 12, base_family = "sans") {
  theme_minimal(base_size = base_size, base_family = base_family) +
    theme(
      panel.grid.major = element_line(color = "grey85", linewidth = 0.4),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(color = "grey65", fill = NA, linewidth = 0.4),
      plot.title = element_text(face = "bold", size = base_size + 2),
      plot.subtitle = element_text(size = base_size, margin = ggplot2::margin(b = 6)),
      plot.caption = element_text(size = base_size - 2, color = "grey40"),
      strip.text = element_text(face = "bold", size = base_size - 1),
      legend.position = "right",
      legend.title = element_text(face = "bold"),
      legend.background = element_rect(fill = "white", color = NA)
    )
}

theme_set(custom_theme())

format_p <- function(p) {
  case_when(
    is.na(p) ~ "NA",
    p < 0.001 ~ "<0.001",
    TRUE ~ sprintf("%.3f", p)
  )
}

fdr_adjust <- function(p) stats::p.adjust(p, method = "BH")

add_fdr <- function(tbl, p_col = "p_value", new_col = "fdr") {
  tbl %>% dplyr::mutate(!!new_col := fdr_adjust(.data[[p_col]]))
}

anova_table <- function(model) {
  if (inherits(model, "lme")) {
    anova_df <- nlme::anova.lme(model, type = "marginal") %>%
      rownames_to_column("term") %>%
      as_tibble() %>%
      janitor::clean_names() %>%
      dplyr::rename(
        num_df = num_df,
        den_df = den_df,
        f_value = f_value,
        p_value = p_value
      )

    if (!"p_value" %in% names(anova_df)) {
      warning("anova.lme output did not contain p-values; returning empty tibble")
      return(tibble())
    }

    anova_df %>% add_fdr("p_value", "fdr")
  } else {
    anova(model, type = 3) %>%
      rownames_to_column("term") %>%
      as_tibble() %>%
      janitor::clean_names() %>%
      dplyr::rename(p_value = pr_greater_f_) %>%
      add_fdr("p_value", "fdr")
  }
}

emmeans_pairs <- function(model, specs, adjust = "BH", by = NULL) {
  emmeans::emmeans(model, specs = specs, by = by) %>%
    emmeans::contrast(method = "pairwise", adjust = adjust) %>%
    broom::tidy()
}
```

```{r util-functions}
# Helper utilities -------------------------------------------------------

collapse_taxa <- function(ps_obj, rank = "Genus", top_n = 15) {
  ps_rel <- microbiome::transform(ps_obj, "compositional")
  tax_rank <- tax_table(ps_rel)[, rank]
  tax_rank[is.na(tax_rank) | tax_rank == ""] <- paste0(rank, "_Unassigned")
  tax_table(ps_rel)[, rank] <- tax_rank

  ps_rank <- tax_glom(ps_rel, taxrank = rank, NArm = FALSE)
  abund <- taxa_sums(ps_rank)
  keep <- rank(abund, ties.method = "first") > (ntaxa(ps_rank) - top_n)
  ps_keep <- prune_taxa(keep, ps_rank)
  other <- prune_taxa(!keep, ps_rank)

  if (ntaxa(other) > 0) {
    other_sum <- tax_table(other)[, rank]
    tax_table(other)[, rank] <- "Other"
    ps_rank <- merge_phyloseq(ps_keep, other)
  } else {
    ps_rank <- ps_keep
  }

  ps_rank
}

plot_bar_compositional <- function(ps_obj, fill_var, facet_var = NULL, rank = "Genus", top_n = 15) {
  ps_rank <- collapse_taxa(ps_obj, rank = rank, top_n = top_n)
  df <- psmelt(ps_rank)

  fill_palette <- scico::scico(length(unique(df[[rank]])) + 1, palette = "batlow")

  ggplot(df, aes(x = .data[[fill_var]], y = Abundance, fill = .data[[rank]])) +
    geom_col(position = "stack") +
    scale_y_continuous(labels = scales::percent) +
    scale_fill_manual(values = fill_palette, guide = guide_legend(ncol = 1)) +
    labs(
      x = NULL,
      y = "Relative abundance",
      fill = rank
    ) +
    custom_theme() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    facet_wrap(vars(.data[[facet_var]]), scales = "free_x") %||%
    identity
}

format_emm_table <- function(tbl) {
  tbl %>%
    dplyr::mutate(
      estimate = round(estimate, 3),
      conf.low = round(conf.low, 3),
      conf.high = round(conf.high, 3),
      p_raw = p.value,
      fdr = p.adjust(p.value, method = "BH")
    ) %>%
    dplyr::mutate(
      p.value = format_p(p_raw),
      fdr = format_p(fdr)
    ) %>%
    dplyr::select(-p_raw)
}

inline_glue <- function(...) {
  glue::glue_data(.x = environment(), ...)
}

`%||%` <- function(x, y) if (!is.null(x)) x else y

cache_rds <- function(path, code) {
  if (file.exists(path)) {
    message("Loading cached result: ", path)
    readRDS(path)
  } else {
    dir.create(dirname(path), showWarnings = FALSE, recursive = TRUE)
    message("Caching result to: ", path)
    result <- code
    if (!is.null(result)) {
      saveRDS(result, path)
    }
    result
  }
}

pull_safe <- function(tbl, col) {
  if (is.null(tbl) || nrow(tbl) == 0 || !(col %in% names(tbl))) {
    return(character())
  }
  tbl[[col]]
}

first_match_col <- function(tbl, candidates) {
  candidates <- candidates[candidates %in% names(tbl)]
  if (length(candidates) == 0) return(NULL)
  candidates[[1]]
}
```

## 3. Data Import and Curation

```{r data-load}
metadata_raw <- readr::read_tsv("sample.metadata.tsv", show_col_types = FALSE) %>%
  janitor::clean_names()

metadata <- metadata_raw %>%
  separate(customer_label, into = c("subject_id", "replicate_code", "timepoint"), sep = "\\.", remove = FALSE) %>%
  dplyr::mutate(
    carbohydrate = case_when(
      str_detect(replicate_code, "R") ~ "RDC",
      str_detect(replicate_code, "S") ~ "SDC",
      TRUE ~ "No Added Carb"
    ),
    carbohydrate = factor(carbohydrate, levels = c("No Added Carb", "RDC", "SDC")),
    timepoint = factor(timepoint, levels = c("0H", "48H")),
    replicate = readr::parse_number(replicate_code) %>% replace_na(1)
  ) %>%
  dplyr::rename(SampleID = sample_id) %>%
  dplyr::select(SampleID, subject_id, timepoint, carbohydrate, replicate, customer_label, raw_seq_files)

healthy_lookup <- readr::read_tsv(
  file = here::here("zr24558.16S_250813.zymo", "1.Healthy.TimepointComparism..Bac16Sv34", "Sample_Information", "sample.metadata.csv"),
  show_col_types = FALSE
) %>%
  janitor::clean_names() %>%
  separate(customer_label, into = c("subject_id", "tmp", "tp"), sep = "\\.", remove = FALSE) %>%
  dplyr::distinct(subject_id) %>%
  dplyr::mutate(group = "Healthy")

metadata <- metadata %>%
  dplyr::mutate(
    group = if_else(subject_id %in% healthy_lookup$subject_id, "Healthy", "Obese"),
    group = factor(group, levels = c("Healthy", "Obese")),
    subject_id = factor(subject_id)
  )

metadata_summary <- metadata %>%
  dplyr::count(group, carbohydrate, timepoint, name = "n_samples")

subject_counts <- metadata %>%
  dplyr::distinct(subject_id, group) %>%
  dplyr::count(group, name = "n_subjects")

metadata_overview <- metadata_summary %>%
  dplyr::group_by(group) %>%
  dplyr::summarise(total = sum(n_samples), .groups = "drop") %>%
  dplyr::left_join(subject_counts, by = "group")
```

**Cohort coverage**

- Samples: `r scales::comma(nrow(metadata))`
- Subjects: `r scales::comma(n_distinct(metadata$subject_id))`
- Group allocation: `r glue::glue_collapse(glue::glue('{metadata_overview$group}: {metadata_overview$n_subjects} subjects, {metadata_overview$total} samples'), sep = '; ')`
- Carbohydrate distribution: `r glue::glue_collapse(levels(metadata$carbohydrate), sep = ', ')`

```{r metadata-table}
metadata_summary %>%
  dplyr::arrange(group, carbohydrate, timepoint) %>%
  gt::gt(rowname_col = "timepoint") %>%
  gt::tab_header(title = "Sample counts by group, carbohydrate, and timepoint") %>%
  gt::fmt_number(columns = "n_samples", decimals = 0)
```
```
### 3.1 Amplicon Sequence Variants and Taxonomy

```{r phyloseq-import}
asv_raw <- readr::read_tsv("dada2_asv_2/asv_table.tsv", show_col_types = FALSE)

asv_ids <- stringr::str_extract(asv_raw$ASV_ID, "ASV_\\d+")

otu_mat <- asv_raw %>%
  dplyr::select(-ASV_ID, -Sequence, -Length, -Total_Abundance, -Relative_Abundance) %>%
  as.data.frame()

colnames(otu_mat) <- stringr::str_remove(colnames(otu_mat), "_filtered$")
rownames(otu_mat) <- asv_ids

taxonomy_raw <- readr::read_tsv("dada2_asv_2/strobelca_taxonomy.tsv", show_col_types = FALSE)

taxonomy_mat <- taxonomy_raw %>%
  dplyr::mutate(ASV = stringr::str_extract(ASV_ID, "ASV_\\d+")) %>%
  dplyr::select(ASV, Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  dplyr::mutate(across(everything(), ~ replace_na(., "Unassigned"))) %>%
  column_to_rownames("ASV") %>%
  as.matrix()

stopifnot(all(rownames(otu_mat) %in% rownames(taxonomy_mat)))

metadata_ps <- metadata %>%
  dplyr::mutate(SampleID = as.character(SampleID)) %>%
  column_to_rownames("SampleID")

common_samples <- intersect(colnames(otu_mat), rownames(metadata_ps))
otu_mat <- otu_mat[, common_samples]
metadata_ps <- metadata_ps[common_samples, ]

ps <- phyloseq(
  otu_table(as.matrix(otu_mat), taxa_are_rows = TRUE),
  sample_data(metadata_ps),
  tax_table(taxonomy_mat[rownames(otu_mat), ])
)

ps <- prune_taxa(taxa_sums(ps) > 0, ps)
ps <- prune_samples(sample_sums(ps) > 0, ps)

sample_data(ps)$SubjectID <- sample_data(ps)$subject_id
sample_data(ps)$Group <- sample_data(ps)$group
sample_data(ps)$Carbohydrate <- sample_data(ps)$carbohydrate
sample_data(ps)$Timepoint <- sample_data(ps)$timepoint

ps_rel <- microbiome::transform(ps, "compositional")
```

```{r sequencing-depth}
seq_depth <- tibble(
  SampleID = sample_names(ps),
  reads = sample_sums(ps)
) %>%
  dplyr::left_join(as_tibble(metadata, .name_repair = "minimal"), by = "SampleID")

coverage_stats <- seq_depth %>%
  dplyr::summarise(
    min_reads = min(reads),
    median_reads = median(reads),
    max_reads = max(reads)
  )

coverage_stats
```

**Retained ASV reads** – minimum `r scales::comma(coverage_stats$min_reads)`, median `r scales::comma(coverage_stats$median_reads)`, maximum `r scales::comma(coverage_stats$max_reads)` reads per sample represented in the denoised ASV table after filtering and chimera removal.


## 4. Alpha Diversity

```{r alpha-diversity}
alpha_metrics <- estimate_richness(ps, measures = c("Observed", "Shannon", "Simpson")) %>%
  rownames_to_column("SampleID")

alpha_long <- alpha_metrics %>%
  pivot_longer(cols = -SampleID, names_to = "metric", values_to = "value") %>%
  dplyr::left_join(as_tibble(metadata, .name_repair = "minimal"), by = "SampleID") %>%
  dplyr::mutate(
    metric = factor(metric, levels = c("Shannon", "Observed", "Simpson"))
  )

prep_mixed_data <- function(data, extra_vars = character()) {
  keep_vars <- unique(c("value", "subject_id", extra_vars))
  keep_vars <- keep_vars[keep_vars %in% names(data)]

  data <- data %>%
    dplyr::select(dplyr::all_of(keep_vars)) %>%
    tidyr::drop_na()

  if (nrow(data) == 0) {
    return(NULL)
  }

  data <- droplevels(as.data.frame(data))

  categorical_vars <- intersect(c("subject_id", extra_vars, "baseline_cluster"), names(data))
  data[categorical_vars] <- lapply(data[categorical_vars], function(x) factor(x))

  if (any(vapply(data[categorical_vars], function(x) length(levels(x)) < 2, logical(1)))) {
    return(NULL)
  }

  data
}

safe_lme <- function(fixed, random, data, extra_vars = character()) {
  data_prepped <- prep_mixed_data(data, extra_vars)
  if (is.null(data_prepped)) {
    return(NULL)
  }

  tryCatch(
    nlme::lme(
      fixed = fixed,
      data = data_prepped,
      random = random,
      method = "REML",
      control = nlme::lmeControl(opt = "optim")
    ),
    error = function(e) {
      message("Mixed model failed: ", e$message)
      NULL
    }
  )
}

fit_alpha_model <- function(df) {
  safe_lme(
    fixed = value ~ group * carbohydrate + group * timepoint + carbohydrate * timepoint,
    random = ~1 | subject_id,
    data = df,
    extra_vars = c("group", "carbohydrate", "timepoint")
  )
}

fit_alpha_model_collapsed <- function(df) {
  safe_lme(
    fixed = value ~ carbohydrate * timepoint,
    random = ~1 | subject_id,
    data = df,
    extra_vars = c("carbohydrate", "timepoint")
  )
}

fit_alpha_model_48h <- function(df) {
  df_levels <- df %>%
    dplyr::filter(!is.na(value)) %>%
    droplevels()

  group_levels <- dplyr::n_distinct(df_levels$group)
  carb_levels <- dplyr::n_distinct(df_levels$carbohydrate)

  if (group_levels <= 1 && carb_levels <= 1) {
    return(NULL)
  }

  if (group_levels > 1 && carb_levels > 1) {
    fixed_formula <- value ~ group * carbohydrate
    extra_vars <- c("group", "carbohydrate")
  } else if (carb_levels > 1) {
    fixed_formula <- value ~ carbohydrate
    extra_vars <- c("carbohydrate")
  } else {
    fixed_formula <- value ~ group
    extra_vars <- c("group")
  }

  safe_lme(
    fixed = fixed_formula,
    random = ~1 | subject_id,
    data = df,
    extra_vars = extra_vars
  )
}

fit_cluster_model <- function(df) {
  safe_lme(
    fixed = value ~ baseline_cluster * carbohydrate,
    random = ~1 | subject_id,
    data = df,
    extra_vars = c("baseline_cluster", "carbohydrate")
  )
}

alpha_metrics_vec <- levels(alpha_long$metric)
alpha_results <- vector("list", length(alpha_metrics_vec))
names(alpha_results) <- alpha_metrics_vec

for (metric_name in alpha_metrics_vec) {
  df_metric <- alpha_long %>%
    dplyr::filter(metric == metric_name) %>%
    droplevels()

  model <- fit_alpha_model(df_metric)
  anova_tbl <- if (is.null(model)) tibble::tibble() else anova_table(model)

  alpha_results[[metric_name]] <- list(
    data = df_metric,
    model = model,
    anova = anova_tbl
  )
}

alpha_models <- tibble(
  metric = alpha_metrics_vec,
  data = lapply(alpha_metrics_vec, function(m) alpha_results[[m]]$data),
  model = lapply(alpha_metrics_vec, function(m) alpha_results[[m]]$model),
  anova = lapply(alpha_metrics_vec, function(m) alpha_results[[m]]$anova)
)

alpha_significance <- alpha_models %>%
  dplyr::mutate(
    sig_terms = purrr::map_chr(
      anova,
      ~ {
        if (is.null(.x) || nrow(.x) == 0) {
          return("Model unavailable for this metric")
        }
        .x %>%
          dplyr::filter(fdr < 0.1) %>%
          dplyr::mutate(term = stringr::str_replace_all(term, ":", " × ")) %>%
          dplyr::mutate(label = glue::glue("{term} (q = {format_p(fdr)})")) %>%
          dplyr::pull(label) %>%
          glue::glue_collapse(sep = "; ") %>%
          {
            ifelse(. == "", "No terms reach q < 0.10", .)
          }
      }
    )
  ) %>%
  dplyr::select(metric, sig_terms)

```

```{r alpha-pairwise-tests}
# First, let's check the data structure
cat("Checking alpha_long data structure:\n")
cat("Total observations:", nrow(alpha_long), "\n")
cat("Metrics:", paste(unique(alpha_long$metric), collapse = ", "), "\n")
cat("Groups:", paste(unique(alpha_long$group), collapse = ", "), "\n")
cat("Timepoints:", paste(unique(alpha_long$timepoint), collapse = ", "), "\n")
cat("Carbohydrate levels:", paste(unique(alpha_long$carbohydrate), collapse = ", "), "\n")

# Check sample sizes per group
data_summary <- alpha_long %>%
  dplyr::group_by(metric, group, timepoint, carbohydrate) %>%
  dplyr::summarise(n = n(), .groups = "drop") %>%
  dplyr::arrange(metric, group, timepoint, carbohydrate)

cat("\nSample sizes per group:\n")
print(data_summary)

# Perform pairwise Wilcoxon tests for each metric across group-timepoint combinations
alpha_pairwise_tests <- alpha_long %>%
  dplyr::group_by(metric, group, timepoint) %>%
  rstatix::wilcox_test(value ~ carbohydrate, p.adjust.method = "BH") %>%
  rstatix::add_significance() %>%
  dplyr::ungroup()

cat("\nPairwise test results:\n")
cat("Number of tests performed:", nrow(alpha_pairwise_tests), "\n")
cat("Columns in results:", paste(names(alpha_pairwise_tests), collapse = ", "), "\n")

# Calculate max values for positioning annotations
alpha_max_values <- alpha_long %>%
  dplyr::group_by(metric, group, timepoint) %>%
  dplyr::summarise(max_val = max(value, na.rm = TRUE), .groups = "drop")

# Filter out non-significant comparisons and add y-positions for significance annotations
alpha_pairwise_tests <- alpha_pairwise_tests %>%
  dplyr::filter(p.adj.signif != "ns") %>%  # Remove non-significant comparisons
  dplyr::left_join(alpha_max_values, by = c("metric", "group", "timepoint")) %>%
  dplyr::group_by(metric, group, timepoint) %>%
  dplyr::mutate(
    y.position = max_val * 1.15 + (dplyr::row_number() - 1) * (max_val * 0.08)
  ) %>%
  dplyr::ungroup()

cat("After filtering non-significant comparisons:\n")
cat("Number of significant comparisons:", nrow(alpha_pairwise_tests), "\n")

cat("Final test data structure:\n")
print(head(alpha_pairwise_tests))
```

```{r alpha-figures}
alpha_plot <- function(data, metric_name, test_data) {
  # Filter data for current metric
  plot_data <- data %>%
    dplyr::filter(metric == metric_name)
  
  # Filter test results for current metric and remove non-significant comparisons
  test_data_metric <- test_data %>%
    dplyr::filter(metric == metric_name) %>%
    dplyr::filter(p.adj.signif != "ns")
  
  # Create base plot
  p <- plot_data %>%
    ggplot(aes(x = carbohydrate, y = value, fill = carbohydrate)) +
    geom_boxplot(alpha = 0.8, outlier.alpha = 0.4, width = 0.7) +
    geom_jitter(aes(color = group), width = 0.12, size = 1.5, alpha = 0.7, show.legend = FALSE) +
    scale_fill_manual(values = palette_carbohydrate) +
    scale_color_manual(values = palette_group) +
    labs(x = NULL, y = metric_name, title = metric_name) +
    facet_grid(group ~ timepoint, scales = "free_y") +
    custom_theme() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  # Add significance annotations only if there are significant comparisons
  if (nrow(test_data_metric) > 0) {
    p <- p +
      ggpubr::stat_pvalue_manual(
        test_data_metric,
        label = "p.adj.signif",
        tip.length = 0.005,
        bracket.nudge.y = 0.01,
        step.increase = 0.03
      )
  }
  
  return(p)
}

# Create plots with significance annotations
alpha_plots <- purrr::map(levels(alpha_long$metric), ~ alpha_plot(alpha_long, .x, alpha_pairwise_tests))

alpha_plots[[2]]
alpha_plots[[1]]
alpha_plots[[3]]

#alpha_plots[[1]] / alpha_plots[[2]] / alpha_plots[[3]] + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

```{r alpha-interpretation, results='asis'}
alpha_significance %>%
  dplyr::mutate(text = glue::glue("**{metric}**: {sig_terms}")) %>%
  dplyr::pull(text) %>%
  glue::glue_collapse(sep = "\n") %>%
  cat()
```

### 4.1 Combined-Group Modeling

The limited group-level separation motivates a secondary model that collapses healthy and obese cohorts to evaluate carbohydrate effects with improved power.

```{r alpha-collapsed}
alpha_collapsed <- alpha_long %>%
  dplyr::mutate(group_collapsed = factor("Combined"))

alpha_models_collapsed <- alpha_collapsed %>%
  tidyr::nest(data = -metric) %>%
  dplyr::mutate(
    model = purrr::map(data, fit_alpha_model_collapsed),
    anova = purrr::map(model, ~ if (is.null(.x)) tibble::tibble() else anova_table(.x))
  )

alpha_models_collapsed %>%
  dplyr::select(metric, anova) %>%
  tidyr::unnest(anova) %>%
  gt::gt() %>%
  gt::fmt_number(columns = c(f_value), decimals = 2) %>%
  gt::fmt_number(columns = c(p_value, fdr), decimals = 3) %>%
  gt::cols_label(term = "Effect", num_df = "df num", den_df = "df den", f_value = "F", p_value = "p", fdr = "q")
```

```{r alpha-combined-plots}
# Create combined data where healthy and obese groups are merged
alpha_combined <- alpha_long %>%
  dplyr::mutate(group = "Combined") %>%
  dplyr::mutate(group = factor(group))

# Create pairwise tests for the combined data
alpha_combined_pairwise_tests <- alpha_combined %>%
  dplyr::group_by(metric, group, timepoint) %>%
  rstatix::wilcox_test(value ~ carbohydrate, p.adjust.method = "BH") %>%
  rstatix::add_significance() %>%
  rstatix::add_y_position(fun =  "max", scales = "free_y") %>%
  dplyr::ungroup()

# plot here but with the healthy and obese combined
alpha_combined_plots <- purrr::map(levels(alpha_long$metric), ~ alpha_plot(alpha_combined, .x, alpha_combined_pairwise_tests))

alpha_combined_plots[[1]]
alpha_combined_plots[[2]]
alpha_combined_plots[[3]]

```

```{r alpha-collapsed-interpretation, results='asis'}
alpha_collapsed_summary <- alpha_models_collapsed %>%
  dplyr::mutate(
    sig_terms = purrr::map_chr(
      anova,
      ~ {
        if (is.null(.x) || nrow(.x) == 0) {
          return("Model unavailable for this metric")
        }
        .x %>%
          dplyr::filter(fdr < 0.1) %>%
          dplyr::mutate(term = stringr::str_replace_all(term, ":", " × ")) %>%
          dplyr::mutate(label = glue::glue("{term} (q = {format_p(fdr)})")) %>%
          dplyr::pull(label) %>%
          glue::glue_collapse(sep = "; ") %>%
          {
            ifelse(. == "", "No carbohydrate × timepoint effects detected at q < 0.10", .)
          }
      }
    )
  )

alpha_collapsed_summary %>%
  dplyr::mutate(text = glue::glue("**{metric}** (groups combined): {sig_terms}")) %>%
  dplyr::pull(text) %>%
  glue::glue_collapse(sep = "\n") %>%
  cat()
```

### 4.2 48 h Focus

```{r alpha-48h}
alpha_48h <- alpha_long %>% dplyr::filter(timepoint == "48H")

alpha_48_model <- alpha_48h %>%
  tidyr::nest(data = -metric) %>%
  dplyr::mutate(
    model = purrr::map(data, fit_alpha_model_48h),
    anova = purrr::map(model, ~ if (is.null(.x)) tibble::tibble() else anova_table(.x)),
    contrasts = purrr::map2(
      model,
      data,
      ~ {
        if (is.null(.x)) return(tibble::tibble())

        df_levels <- .y %>%
          dplyr::filter(!is.na(value)) %>%
          droplevels()

        group_levels <- dplyr::n_distinct(df_levels$group)
        carb_levels <- dplyr::n_distinct(df_levels$carbohydrate)

        if (group_levels > 1 && carb_levels > 1) {
          emmeans::emmeans(.x, ~ carbohydrate | group, data = df_levels) %>%
            emmeans::contrast("pairwise", adjust = "BH") %>%
            broom::tidy()
        } else if (carb_levels > 1) {
          emmeans::emmeans(.x, specs = "carbohydrate", data = df_levels) %>%
            emmeans::contrast("pairwise", adjust = "BH") %>%
            broom::tidy()
        } else if (group_levels > 1) {
          emmeans::emmeans(.x, specs = "group", data = df_levels) %>%
            emmeans::contrast("pairwise", adjust = "BH") %>%
            broom::tidy()
        } else {
          tibble::tibble()
        }
      }
    )
  )

alpha_48_model$anova
```

```{r alpha-48h-interpretation, results='asis'}
alpha_48_text <- alpha_48_model %>%
  dplyr::mutate(
    sig_terms = purrr::map_chr(
      anova,
      ~ {
        if (is.null(.x) || nrow(.x) == 0) {
          return("Model unavailable for this metric")
        }
        .x %>%
          dplyr::filter(fdr < 0.1) %>%
          dplyr::mutate(term = stringr::str_replace_all(term, ":", " × ")) %>%
          dplyr::mutate(label = glue::glue("{term} (q = {format_p(fdr)})")) %>%
          dplyr::pull(label) %>%
          glue::glue_collapse(sep = "; ") %>%
          {
            ifelse(. == "", "No significant effects after FDR correction", .)
          }
      }
    )
  )

alpha_48_text %>%
  dplyr::mutate(text = glue::glue("**{metric}** (48 h): {sig_terms}")) %>%
  dplyr::pull(text) %>%
  glue::glue_collapse(sep = "\n") %>%
  cat()
```

```{r alpha-48h-wilcoxon}
# Perform pairwise Wilcoxon tests for each metric
alpha_48h_wilcoxon <- alpha_48h %>%
  group_by(metric) %>%
  wilcox_test(value ~ carbohydrate, p.adjust.method = "BH") %>%
  add_significance()

# Calculate max values for each metric to position annotations
max_values <- alpha_48h %>%
  group_by(metric) %>%
  summarise(max_val = max(value, na.rm = TRUE), .groups = "drop")

# Add manual y-positions based on data range for each metric
alpha_48h_wilcoxon <- alpha_48h_wilcoxon %>%
  left_join(max_values, by = "metric") %>%
  group_by(metric) %>%
  mutate(
    y.position = case_when(
      metric == "Shannon" ~ max_val * 1.1,
      metric == "Observed" ~ max_val * 1.15,
      metric == "Simpson" ~ max_val * 1.1
    ),
    y.position = y.position + (row_number() - 1) * (max_val * 0.1)
  ) %>%
  ungroup()

# Check the structure of the results
print("Columns in alpha_48h_wilcoxon:")
print(names(alpha_48h_wilcoxon))
print("First few rows:")
print(head(alpha_48h_wilcoxon))
```

```{r alpha-48h-plot}
alpha_48h %>%
  ggplot(aes(x = carbohydrate, y = value, fill = carbohydrate)) +
  #geom_violin(trim = FALSE, alpha = 0.8) +
  geom_boxplot(alpha = 0.9) +
  geom_jitter(width = 0.12, size = 1.5, alpha = 0.7, show.legend = FALSE) +
  stat_pvalue_manual(alpha_48h_wilcoxon, 
                     label = "p.adj.signif", 
                     tip.length = 0.01,
                     bracket.nudge.y = 0.05) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_fill_manual(values = palette_carbohydrate) +
  labs(x = NULL, y = "Diversity", title = "Alpha diversity at 48 h") +
  custom_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 5. Beta Diversity and Community Structure

```{r beta-prep}
make_pcoa_df <- function(ps_obj, distance_method) {
  ord <- ordinate(ps_obj, method = "PCoA", distance = distance_method)
  vecs <- as_tibble(ord$vectors[, 1:2], rownames = "SampleID")
  colnames(vecs)[2:3] <- c("PC1", "PC2")
  vecs <- vecs %>%
    dplyr::mutate(
      axis1_var = ord$values$Relative_eig[1],
      axis2_var = ord$values$Relative_eig[2]
    ) %>%
    dplyr::left_join(as_tibble(metadata, .name_repair = "minimal"), by = "SampleID")
  vecs$distance_method <- distance_method
  vecs
}

dist_methods <- c("bray", "jaccard")

pcoa_dfs <- purrr::map_dfr(dist_methods, ~ make_pcoa_df(ps_rel, .x))

```

```{r permanova}
permanova_results_raw <- purrr::map_dfr(
  dist_methods,
  ~ {
    dist_mat <- phyloseq::distance(ps_rel, method = .x)
    df <- as(sample_data(ps_rel), "data.frame")
    adonis_out <- vegan::adonis2(
      dist_mat ~ group * carbohydrate + timepoint,
      data = df,
      permutations = 999,
      strata = df$subject_id
    )
    adonis_tbl <- as.data.frame(adonis_out) %>% rownames_to_column('term') %>% janitor::clean_names()
    adonis_tbl$distance <- .x
    adonis_tbl
  }
)

permanova_p_col <- intersect(c("pr_greater_f_", "pr_greater_f", "pr_f", "p_value"), names(permanova_results_raw))
permanova_f_col <- intersect(c("f", "f_model", "pseudo_f"), names(permanova_results_raw))
permanova_r2_col <- intersect(c("r2", "r_squared", "r2_adj"), names(permanova_results_raw))

if (length(permanova_p_col) == 0 || length(permanova_f_col) == 0 || length(permanova_r2_col) == 0) {
  stop(glue::glue(
    "Unexpected PERMANOVA columns. Available: {glue::glue_collapse(names(permanova_results_raw), sep = ', ')}"
  ))
}

permanova_results <- permanova_results_raw %>%
  dplyr::mutate(
    f = .data[[permanova_f_col[1]]],
    r2 = .data[[permanova_r2_col[1]]],
    p_value = .data[[permanova_p_col[1]]],
    q_value = fdr_adjust(p_value)
  )

permanova_results %>%
  gt::gt() %>%
  gt::fmt_number(columns = c(f, r2), decimals = 3) %>%
  gt::fmt_number(columns = c(p_value, q_value), decimals = 3) %>%
  gt::cols_label(term = "Effect", f = "F", r2 = "R2", p_value = "p", q_value = "q") %>%
  gt::tab_header("PERMANOVA results (subject-level strata)")
```

```{r pcoa-plot}
pcoa_plot <- function(df, method_name) {
  df_subset <- df %>% dplyr::filter(distance_method == method_name)
  var1 <- scales::percent(round(df_subset$axis1_var[1], 3))
  var2 <- scales::percent(round(df_subset$axis2_var[1], 3))

  # Get PERMANOVA results for this distance method - use group:carbohydrate interaction
  permanova_subset <- permanova_results %>%
    dplyr::filter(distance == method_name, term == "group:carbohydrate")

  # Create annotation text and position
  if (nrow(permanova_subset) > 0) {
    r2_val <- round(permanova_subset$r2[1], 3)
    p_val <- round(permanova_subset$p_value[1], 3)
    annotation_text <- glue::glue("R² = {r2_val}, p = {p_val}")

    # Calculate per-facet label positions, guard against missing/flat ranges
    annotation_df <- df_subset %>%
      dplyr::filter(!is.na(PC1), !is.na(PC2)) %>%
      dplyr::group_by(timepoint) %>%
      dplyr::summarise(
        x_min = min(PC1),
        x_max = max(PC1),
        y_min = min(PC2),
        y_max = max(PC2),
        .groups = "drop"
      ) %>%
      dplyr::mutate(
        x_span = x_max - x_min,
        y_span = y_max - y_min,
        PC1 = ifelse(is.finite(x_span) & x_span > 0, x_min + 0.05 * x_span, x_min),
        PC2 = ifelse(is.finite(y_span) & y_span > 0, y_max - 0.05 * y_span, y_max),
        label = annotation_text
      ) %>%
      dplyr::select(timepoint, PC1, PC2, label)
  } else {
    annotation_df <- data.frame(
      PC1 = numeric(0),
      PC2 = numeric(0),
      label = character(0),
      timepoint = character(0)
    )
  }

  p <- ggplot(df_subset, aes(x = PC1, y = PC2, color = carbohydrate, shape = group)) +
    geom_point(size = 2.5, alpha = 0.85) +
    stat_ellipse(aes(group = interaction(carbohydrate, timepoint), color = carbohydrate),
      linewidth = 0.4, linetype = "dashed", show.legend = FALSE
    ) +
    facet_wrap(~timepoint) +
    scale_color_manual(values = palette_carbohydrate) +
    scale_shape_manual(values = c(16, 17)) +
    labs(
      x = glue::glue("PC1 ({var1})"),
      y = glue::glue("PC2 ({var2})"),
      title = glue::glue("PCoA – {stringr::str_to_title(method_name)} distance")
    ) +
    custom_theme()

  if (nrow(annotation_df) > 0) {
    p <- p + geom_label(
      data = annotation_df,
      aes(x = PC1, y = PC2, label = label),
      hjust = 0, vjust = 1, size = 3, fontface = "bold",
      inherit.aes = FALSE, fill = "white", alpha = 0.8
    )
  }

  p
}

(pcoa_plot(pcoa_dfs, "bray") | pcoa_plot(pcoa_dfs, "jaccard")) + plot_annotation(tag_levels = "A")
```

```{r dispersion-tests}
betadisp_results <- purrr::map_dfr(
  dist_methods,
  ~ {
    dist_mat <- phyloseq::distance(ps_rel, method = .x)
    df <- as(sample_data(ps_rel), "data.frame")
    bd <- vegan::betadisper(dist_mat, df$carbohydrate)
    perm <- vegan::permutest(bd, permutations = 999)

    # Extract values more robustly
    perm_df <- as.data.frame(perm$tab)
    f_row <- which(rownames(perm_df) %in% c("F", "Groups"))

    tibble(
      distance = .x,
      stat = perm_df[f_row[1], "F"],
      p_value = perm_df[f_row[1], "Pr(>F)"]
    )
  }
) %>%
  dplyr::mutate(q_value = fdr_adjust(p_value))

betadisp_results %>%
  gt::gt() %>%
  gt::fmt_number(columns = c(stat, p_value, q_value), decimals = 3) %>%
  gt::cols_label(stat = "F", p_value = "p", q_value = "q") %>%
  gt::tab_header("Homogeneity of dispersion (carbohydrate groups)")
```

### Interpretation

```{r beta-interpretation, results='asis'}
permanova_summary <- permanova_results %>%
  dplyr::mutate(q = fdr_adjust(p_value)) %>%
  dplyr::filter(term != "Residual" & q < 0.1) %>%
  dplyr::mutate(item = glue::glue("{stringr::str_to_title(distance)}: {term} (q = {format_p(q)})")) %>%
  dplyr::pull(item)

if (length(permanova_summary) == 0) {
  cat("No beta-diversity terms reach q < 0.10 after FDR adjustment. Patterns are primarily driven by subject-level variability.")
} else {
  cat(glue::glue_collapse(permanova_summary, sep = "\n"))
}
```

### 5.1 Beta Diversity at 48 h

```{r beta-48h}
ps_48h <- phyloseq::subset_samples(ps_rel, Timepoint == "48H")

pcoa_48h <- make_pcoa_df(ps_48h, "bray")

pcoa_48h %>%
  ggplot(aes(x = PC1, y = PC2, color = carbohydrate, shape = group)) +
  geom_point(size = 3, alpha = 0.9) +
  stat_ellipse(aes(group = carbohydrate, color = carbohydrate), linewidth = 0.8, linetype = "dashed", alpha = 0.8) +
  scale_color_manual(values = palette_carbohydrate) +
  scale_shape_manual(values = c(16, 17)) +
  labs(
    x = glue::glue("PC1 ({scales::percent(round(pcoa_48h$axis1_var[1], 3))})"),
    y = glue::glue("PC2 ({scales::percent(round(pcoa_48h$axis2_var[1], 3))})"),
    title = "PCoA – Bray-Curtis distance (48 h samples)"
  ) +
  custom_theme()
```

```{r beta-48h-permanova}
metadata_48h <- as(sample_data(ps_48h), "data.frame")
permanova_48h <- vegan::adonis2(
  phyloseq::distance(ps_48h, method = "bray") ~ group * carbohydrate,
  data = metadata_48h,
  permutations = 999,
  strata = metadata_48h$subject_id
)

permanova_48h_tbl <- as.data.frame(permanova_48h) %>% rownames_to_column('term') %>% janitor::clean_names()

# Find the correct column names for p-value, F-statistic, and R-squared
permanova_48h_p_col <- intersect(c("pr_greater_f_", "pr_greater_f", "pr_f", "p_value"), names(permanova_48h_tbl))
permanova_48h_f_col <- intersect(c("f", "f_model", "pseudo_f"), names(permanova_48h_tbl))
permanova_48h_r2_col <- intersect(c("r2", "r_squared", "r2_adj"), names(permanova_48h_tbl))

permanova_48h_tbl %>%
  dplyr::mutate(p_value = !!sym(permanova_48h_p_col), q = fdr_adjust(p_value)) %>%
  gt::gt() %>%
  gt::fmt_number(columns = c(!!sym(permanova_48h_f_col), !!sym(permanova_48h_r2_col)), decimals = 3) %>%
  gt::fmt_number(columns = c(p_value, q), decimals = 3) %>%
  gt::cols_label(term = "Effect", !!sym(permanova_48h_f_col) := "F", !!sym(permanova_48h_r2_col) := "R2", p_value = "p", q = "q")
```

## 6. Baseline Clustering and Trajectories

```{r cluster-baseline}
ps_0h <- phyloseq::subset_samples(ps_rel, Timepoint == "0H")
stopifnot(nsamples(ps_0h) > 0)

bray_0h <- phyloseq::distance(ps_0h, method = "bray")
dist_matrix <- as.matrix(bray_0h)

silhouette_scores <- map_dbl(2:6, ~ {
  pam_fit <- cluster::pam(bray_0h, k = .x)
  mean(cluster::silhouette(pam_fit)[, "sil_width"])
})

best_k <- which.max(silhouette_scores) + 1
pam_best <- cluster::pam(bray_0h, k = best_k)

cluster_assignments <- tibble(
  SampleID = sample_names(ps_0h),
  baseline_cluster = factor(paste0("Cluster", pam_best$clustering), levels = paste0("Cluster", seq_len(best_k)))
)

cluster_summary <- cluster_assignments %>%
  dplyr::left_join(as_tibble(metadata, .name_repair = "minimal"), by = "SampleID") %>%
  dplyr::count(baseline_cluster, carbohydrate, group, name = "n")

cluster_summary
```

```{r cluster-plot}
cluster_ord <- make_pcoa_df(ps_0h, "bray") %>%
  dplyr::left_join(cluster_assignments, by = "SampleID")

cluster_ord %>%
  ggplot(aes(x = PC1, y = PC2, color = baseline_cluster, shape = carbohydrate)) +
  geom_point(size = 3, alpha = 0.9) +
  stat_ellipse(aes(group = baseline_cluster), linewidth = 0.5) +
  scale_color_manual(values = palette_cluster) +
  labs(
    title = "Baseline (0 h) clusters",
    x = glue::glue("PC1 ({scales::percent(round(cluster_ord$axis1_var[1], 3))})"),
    y = glue::glue("PC2 ({scales::percent(round(cluster_ord$axis2_var[1], 3))})")
  ) +
  custom_theme()
```

```{r cluster-propagation}
metadata_with_cluster <- metadata %>%
  dplyr::left_join(cluster_assignments %>% dplyr::select(SampleID, baseline_cluster), by = "SampleID")

# propagate baseline cluster to all timepoints via subject_id
subject_clusters <- cluster_assignments %>%
  dplyr::left_join(metadata %>% dplyr::select(SampleID, subject_id), by = "SampleID") %>%
  dplyr::group_by(subject_id) %>%
  dplyr::summarise(baseline_cluster = dplyr::first(baseline_cluster), .groups = "drop")

metadata_with_cluster <- metadata %>%
  dplyr::left_join(subject_clusters, by = "subject_id")

ps_with_cluster <- ps_rel
sample_data(ps_with_cluster)$baseline_cluster <- metadata_with_cluster$baseline_cluster[match(sample_names(ps_with_cluster), metadata_with_cluster$SampleID)]
```

```{r cluster-48h}
# Only include subjects that have baseline cluster assignments
subjects_with_clusters <- subject_clusters$subject_id
ps_48h_cluster <- phyloseq::subset_samples(ps_with_cluster, Timepoint == "48H" & subject_id %in% subjects_with_clusters)

# Check how many samples we have after filtering
cat("48H samples with baseline cluster assignments:", nsamples(ps_48h_cluster), "\n")
cat("Total 48H samples:", sum(sample_data(ps_with_cluster)$Timepoint == "48H"), "\n")

# Show which subjects are missing baseline samples
all_48h_subjects <- as(sample_data(ps_with_cluster), "data.frame") %>% 
  dplyr::filter(Timepoint == "48H") %>% 
  dplyr::pull(subject_id) %>% 
  unique()

missing_baseline_subjects <- setdiff(all_48h_subjects, subjects_with_clusters)
if(length(missing_baseline_subjects) > 0) {
  cat("Subjects missing baseline (0H) samples:", paste(missing_baseline_subjects, collapse = ", "), "\n")
  cat("These subjects will be excluded from cluster-based 48H analysis.\n")
} else {
  cat("All 48H subjects have baseline cluster assignments.\n")
}

ord_48h_cluster <- make_pcoa_df(ps_48h_cluster, "bray") %>%
  dplyr::mutate(baseline_cluster = sample_data(ps_48h_cluster)$baseline_cluster[match(SampleID, sample_names(ps_48h_cluster))])

# Create dynamic color palette based on actual clusters present
actual_clusters <- levels(factor(ord_48h_cluster$baseline_cluster))
cluster_palette_dynamic <- setNames(scico::scico(length(actual_clusters), palette = "batlow"), actual_clusters)

ord_48h_cluster %>%
  ggplot(aes(x = PC1, y = PC2, color = baseline_cluster, shape = carbohydrate)) +
  geom_point(size = 3, alpha = 0.9) +
  stat_ellipse(aes(group = interaction(baseline_cluster, carbohydrate)), 
               linewidth = 0.5, linetype = "dotted", 
               level = 0.95, type = "norm") +
  scale_color_manual(values = cluster_palette_dynamic) +
  scale_shape_manual(values = c(16, 17, 18)) +
  labs(
    title = "48 h beta diversity colored by baseline cluster",
    x = glue::glue("PC1 ({scales::percent(round(ord_48h_cluster$axis1_var[1], 3))})"),
    y = glue::glue("PC2 ({scales::percent(round(ord_48h_cluster$axis2_var[1], 3))})")
  ) +
  custom_theme()
```
```{r cluster-48h-facet-plots}
ord_48h_cluster %>% ggplot(aes(PC1, PC2)) +
  geom_point(aes(color = carbohydrate), size = 3) +
  facet_wrap(~baseline_cluster)
```

```{r cluster-statistics}
cluster_response <- ord_48h_cluster %>%
  dplyr::filter(!is.na(baseline_cluster)) %>%
  dplyr::group_by(baseline_cluster, carbohydrate) %>%
  dplyr::summarise(
    mean_PC1 = mean(PC1),
    mean_PC2 = mean(PC2),
    n = n(),
    .groups = "drop"
  )

cluster_response
```

```{r cluster-anova}
ps_48h_counts <- phyloseq::subset_samples(ps, Timepoint == "48H")
alpha_48h_cluster <- estimate_richness(ps_48h_counts, measures = c("Observed", "Shannon", "Simpson")) %>%
  rownames_to_column("SampleID") %>%
  pivot_longer(-SampleID, names_to = "metric", values_to = "value") %>%
  dplyr::left_join(metadata_with_cluster, by = "SampleID") %>%
  dplyr::filter(!is.na(baseline_cluster))

cluster_models <- alpha_48h_cluster %>%
  tidyr::nest(data = -metric) %>%
  dplyr::mutate(
    model = purrr::map(data, fit_cluster_model),
    anova = purrr::map(model, ~ if (is.null(.x)) tibble::tibble() else anova_table(.x))
  )

cluster_models %>%
  dplyr::select(metric, anova) %>%
  tidyr::unnest(anova) %>%
  gt::gt() %>%
  gt::fmt_number(columns = c(f_value), decimals = 3) %>%
  gt::fmt_number(columns = c(p_value, fdr), decimals = 3) %>%
  gt::cols_label(term = "Effect", num_df = "df num", den_df = "df den", f_value = "F", p_value = "p", fdr = "q")
```

```{r cluster-interpretation, results='asis'}
cluster_text <- cluster_models %>%
  dplyr::mutate(summary = purrr::map_chr(
    anova,
    ~ {
      if (is.null(.x) || nrow(.x) == 0) {
        return("Model unavailable for this metric at 48 h")
      }
      .x %>%
        dplyr::filter(fdr < 0.1) %>%
        dplyr::mutate(label = glue::glue("{term} (q = {format_p(fdr)})")) %>%
        dplyr::pull(label) %>%
        glue::glue_collapse(sep = "; ") %>%
        {ifelse(. == "", "No significant cluster–carbohydrate interactions detected at 48 h", .)}
    }
  ))

cluster_text %>%
  dplyr::mutate(text = glue::glue("**{metric}**: {summary}")) %>%
  dplyr::pull(text) %>%
  glue::glue_collapse(sep = "\n") %>%
  cat()
```

## 7. Community Composition Profiles

```{r composition-overview}
ps_genus <- collapse_taxa(ps_with_cluster, rank = "Genus", top_n = 20)

ggplot(psmelt(microbiome::transform(ps_genus, "compositional")),
       aes(x = carbohydrate, y = Abundance, fill = Genus)) +
  geom_col(position = "fill") +
  facet_grid(group ~ timepoint) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = scico::scico(21, palette = "batlow")) +
  labs(
    x = NULL,
    y = "Relative abundance",
    title = "Genus-level composition by group, carbohydrate, and timepoint"
  ) +
  custom_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r composition-48h}
ps_48h_genus <- phyloseq::subset_samples(ps_genus, Timepoint == "48H")

ggplot(psmelt(microbiome::transform(ps_48h_genus, "compositional")),
       aes(x = carbohydrate, y = Abundance, fill = Genus)) +
  geom_col(position = "fill") +
  facet_wrap(~ group) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = scico::scico(21, palette = "batlow")) +
  labs(
    x = NULL,
    y = "Relative abundance",
    title = "Genus-level composition at 48 h"
  ) +
  custom_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r composition-cluster}
ps_cluster <- phyloseq::subset_samples(ps_genus, !is.na(sample_data(ps_genus)$baseline_cluster))

ggplot(psmelt(microbiome::transform(ps_cluster, "compositional")),
       aes(x = baseline_cluster, y = Abundance, fill = Genus)) +
  geom_col(position = "fill") +
  facet_wrap(~ carbohydrate + timepoint, ncol = 3) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = scico::scico(21, palette = "batlow")) +
  labs(
    x = "Baseline cluster",
    y = "Relative abundance",
    title = "Genus-level composition by baseline cluster"
  ) +
  custom_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Alluvial flow diagrams: Genus abundance changes from 0H to 48H

```{r alluvial-diagram, fig.width=14, fig.height=10}
# Install ggalluvial if not available
if (!requireNamespace("ggalluvial", quietly = TRUE)) {
  install.packages("ggalluvial")
}
library(ggalluvial)

# Prepare data for alluvial diagrams
ps_genus_rel <- transform_sample_counts(ps_genus, function(x) x / sum(x))

# Extract data
otu_mat <- as.matrix(otu_table(ps_genus_rel))
if (!taxa_are_rows(ps_genus_rel)) {
  otu_mat <- t(otu_mat)
}

tax_df <- as.data.frame(tax_table(ps_genus_rel))
meta_df <- data.frame(sample_data(ps_genus_rel))

# Create long format data
alluvial_data <- otu_mat %>%
  as.data.frame() %>%
  tibble::rownames_to_column("ASV") %>%
  tidyr::pivot_longer(-ASV, names_to = "SampleID", values_to = "Abundance") %>%
  dplyr::left_join(
    tax_df %>% tibble::rownames_to_column("ASV") %>% dplyr::select(ASV, Genus),
    by = "ASV"
  ) %>%
  dplyr::left_join(
    meta_df %>% tibble::rownames_to_column("SampleID") %>%
      dplyr::select(SampleID, Timepoint, Carbohydrate),
    by = "SampleID"
  ) %>%
  dplyr::group_by(Genus, Timepoint, Carbohydrate) %>%
  dplyr::summarise(mean_abundance = mean(Abundance), .groups = "drop") %>%
  dplyr::filter(mean_abundance > 0.01)  # Keep genera >1%

# Get top genera
top_genera <- alluvial_data %>%
  dplyr::group_by(Genus) %>%
  dplyr::summarise(max_abund = max(mean_abundance)) %>%
  dplyr::arrange(desc(max_abund)) %>%
  head(12) %>%
  pull(Genus)

# Filter to top genera and add "Other" category
alluvial_plot_data <- alluvial_data %>%
  dplyr::mutate(
    Genus_grouped = ifelse(Genus %in% top_genera, as.character(Genus), "Other")
  ) %>%
  dplyr::group_by(Genus_grouped, Timepoint, Carbohydrate) %>%
  dplyr::summarise(Abundance = sum(mean_abundance), .groups = "drop")

# Create color palette
genus_colors <- c(
  scico::scico(length(top_genera), palette = "batlow"),
  "Other" = "grey80"
)
names(genus_colors)[1:length(top_genera)] <- top_genera

# Create alluvial plots for each carbohydrate
plot_list <- list()

for (carb in c("No Added Carb", "RDC", "SDC")) {
  plot_data <- alluvial_plot_data %>%
    dplyr::filter(Carbohydrate == carb) %>%
    dplyr::mutate(
      Genus_grouped = factor(Genus_grouped,
                            levels = c(top_genera[top_genera %in% unique(Genus_grouped)], "Other")),
      # Create label columns that show only on respective sides
      label_left = ifelse(Timepoint == "0H", as.character(Genus_grouped), ""),
      label_right = ifelse(Timepoint == "48H", as.character(Genus_grouped), "")
    )

  p <- ggplot(plot_data,
              aes(x = Timepoint, y = Abundance,
                  alluvium = Genus_grouped, stratum = Genus_grouped, fill = Genus_grouped)) +
    geom_alluvium(alpha = 0.7, width = 0.3) +
    geom_stratum(width = 0.3, alpha = 0.9) +
    # Add labels to the left side (0H only)
    stat_stratum(geom = "text", aes(label = label_left),
                 size = 2.3, hjust = 1.3, fontface = "bold") +
    # Add labels to the right side (48H only)
    stat_stratum(geom = "text", aes(label = label_right),
                 size = 2.3, hjust = -0.5, fontface = "bold") +
    scale_fill_manual(values = genus_colors) +
    scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
    scale_x_discrete(expand = expansion(add = c(1.2, 1.2))) +
    labs(
      title = carb,
      x = "Timepoint",
      y = "Relative Abundance",
      fill = "Genus"
    ) +
    custom_theme() +
    theme(
      legend.position = "none",
      panel.grid.major.x = element_blank()
    )

  plot_list[[carb]] <- p
}

# Combine plots
library(patchwork)
combined_plot <- wrap_plots(plot_list, ncol = 3) +
  plot_annotation(
    title = "Genus Composition Changes from 0H to 48H by Carbohydrate Type",
    subtitle = "Flow diagram showing relative abundance changes (top 12 genera)",
    theme = theme(plot.title = element_text(size = 16, face = "bold"))
  )

print(combined_plot)
```

### Validation: Bar plots of genus abundances by timepoint and carbohydrate

```{r chord-validation, fig.width=14, fig.height=10}
# Extract genus-level data for validation
ps_genus_rel <- transform_sample_counts(ps_genus, function(x) x / sum(x))

# Get OTU table and metadata
otu_mat <- as.matrix(otu_table(ps_genus_rel))
if (!taxa_are_rows(ps_genus_rel)) {
  otu_mat <- t(otu_mat)
}
otu_df <- as.data.frame(t(otu_mat))

tax_df <- as.data.frame(tax_table(ps_genus_rel))
meta_df <- data.frame(sample_data(ps_genus_rel))

# Combine data
genus_long <- otu_df %>%
  tibble::rownames_to_column("SampleID") %>%
  tidyr::pivot_longer(-SampleID, names_to = "ASV", values_to = "Abundance") %>%
  dplyr::left_join(
    tax_df %>% tibble::rownames_to_column("ASV") %>% dplyr::select(ASV, Genus),
    by = "ASV"
  ) %>%
  dplyr::left_join(
    meta_df %>% tibble::rownames_to_column("SampleID") %>%
      dplyr::select(SampleID, Timepoint, Carbohydrate),
    by = "SampleID"
  )

# Calculate mean abundance per genus, timepoint, and carbohydrate
genus_summary <- genus_long %>%
  group_by(Genus, Timepoint, Carbohydrate) %>%
  summarise(
    mean_abundance = mean(Abundance),
    se_abundance = sd(Abundance) / sqrt(n()),
    .groups = "drop"
  ) %>%
  filter(mean_abundance > 0.01)  # Only keep genera >1% mean abundance

# Get top genera overall
top_genera <- genus_summary %>%
  group_by(Genus) %>%
  summarise(max_abundance = max(mean_abundance), .groups = "drop") %>%
  arrange(desc(max_abundance)) %>%
  head(15) %>%
  pull(Genus)

# Filter to top genera
genus_summary_top <- genus_summary %>%
  filter(Genus %in% top_genera)

# Create faceted bar plot
p_validation <- ggplot(genus_summary_top,
                       aes(x = Timepoint, y = mean_abundance, fill = Timepoint)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_errorbar(aes(ymin = mean_abundance - se_abundance,
                    ymax = mean_abundance + se_abundance),
                width = 0.2, position = position_dodge(0.7)) +
  facet_grid(Genus ~ Carbohydrate, scales = "free_y") +
  scale_fill_manual(values = c("0H" = "#E69F00", "48H" = "#56B4E9")) +
  labs(
    title = "Validation: Genus Abundance by Timepoint and Carbohydrate Type",
    subtitle = "Mean relative abundance ± SE for top 15 genera (>1% abundance)",
    x = "Timepoint",
    y = "Mean Relative Abundance",
    fill = "Timepoint"
  ) +
  custom_theme() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    strip.text.y = element_text(angle = 0, hjust = 0),
    panel.spacing = unit(0.5, "lines")
  )

print(p_validation)

# Print summary statistics for key changes
cat("\n### Summary of abundance changes from 0H to 48H:\n\n")

change_summary <- genus_summary_top %>%
  dplyr::select(Genus, Timepoint, Carbohydrate, mean_abundance) %>%
  pivot_wider(names_from = Timepoint, values_from = mean_abundance,
              names_prefix = "T") %>%
  mutate(
    change = T48H - T0H,
    pct_change = ((T48H - T0H) / T0H) * 100,
    direction = case_when(
      abs(change) < 0.005 ~ "No change",
      change > 0 ~ "Increase",
      change < 0 ~ "Decrease"
    )
  ) %>%
  arrange(Carbohydrate, desc(abs(change)))

# Print changes for each carbohydrate type
for (carb in c("No Added Carb", "RDC", "SDC")) {
  cat(paste0("\n", carb, ":\n"))
  carb_changes <- change_summary %>%
    filter(Carbohydrate == carb, abs(change) > 0.01) %>%
    arrange(desc(abs(change))) %>%
    head(10)

  if (nrow(carb_changes) > 0) {
    print(carb_changes %>%
            dplyr::select(Genus, T0H, T48H, change, pct_change, direction) %>%
            mutate(across(where(is.numeric), ~round(., 4))))
  } else {
    cat("  No major changes detected (>1% absolute change)\n")
  }
}
```

## 8. Differential Abundance and Predictive Modeling

### 8.1 Shared Setup

```{r da-setup}
ps_counts <- ps
sample_data(ps_counts)$subject_id <- factor(sample_data(ps_counts)$subject_id)

ps_genus_counts <- tax_glom(ps_counts, taxrank = "Genus", NArm = FALSE)
ps_genus_counts <- prune_taxa(taxa_sums(ps_genus_counts) > 0, ps_genus_counts)

metadata_counts <- as(sample_data(ps_genus_counts), "data.frame") %>%
  as_tibble(rownames = "SampleID")

prevalence_cutoff <- 0.05
min_prev <- ceiling(prevalence_cutoff * nsamples(ps_genus_counts))
keep_taxa <- taxa_sums(ps_genus_counts) > 0 & rowSums(otu_table(ps_genus_counts) > 0) >= min_prev
ps_genus_filtered <- prune_taxa(keep_taxa, ps_genus_counts)

ps_genus_48h <- subset_samples(ps_genus_filtered, Timepoint == "48H")
ps_genus_0h <- subset_samples(ps_genus_filtered, Timepoint == "0H")
```

### 8.2 Corncob (beta-binomial)

```{r corncob}
corncob_path <- file.path(outputs_dir, "corncob_genus_results.rds")

corncob_test <- cache_rds(
  corncob_path,
  {
    set.seed(12345)
    tryCatch({
      corncob::differentialTest(
        formula = ~ group + carbohydrate * timepoint + subject_id,
        phi.formula = ~ group + carbohydrate * timepoint + subject_id,
        formula_null = ~ group + timepoint + subject_id,
        phi.formula_null = ~ group + timepoint + subject_id,
        test = "Wald",
        data = ps_genus_filtered,
        fdr_cutoff = 0.1,
        B = 50
      )
    }, error = function(e) {
      message("Corncob failed: ", e$message)
      NULL
    })
  }
)

if (!is.null(corncob_test)) {
  if (!is.null(corncob_test$wald_results)) {
    corncob_results <- corncob_test$wald_results %>%
      dplyr::mutate(q_value = fdr_adjust(p_val)) %>%
      dplyr::arrange(q_value)
  } else if (!is.null(corncob_test$p) && !is.null(corncob_test$p_fdr)) {
    p_mat <- corncob_test$p
    q_mat <- corncob_test$p_fdr

    if (is.null(dim(p_mat))) {
      p_vals <- as.numeric(p_mat)
    } else {
      p_vals <- as.numeric(p_mat[, 1])
    }

    if (is.null(dim(q_mat))) {
      q_vals = as.numeric(q_mat)
    } else {
      q_vals = as.numeric(q_mat[, 1])
    }

    taxa_names <- rownames(p_mat)
    if (is.null(taxa_names)) {
      taxa_names <- names(p_mat)
    }
    if (is.null(taxa_names)) {
      taxa_names <- paste0("taxon_", seq_along(p_vals))
    }

    corncob_results <- tibble::tibble(
      taxon = taxa_names,
      p_val = p_vals,
      q_value = q_vals
    ) %>%
      dplyr::arrange(q_value)
  } else {
    corncob_results <- tibble()
  }
} else {
  corncob_results <- tibble()
}

corncob_results

plot(corncob_test)
```

### 8.2.1 Corncob (48H, RDC vs SDC only)

```{r corncob-48h-rdc-sdc}
# Filter data for 48H timepoint and only RDC and SDC groups
ps_genus_48h_rdc_sdc <- subset_samples(ps_genus_filtered, 
                                       Timepoint == "48H" & 
                                       carbohydrate %in% c("RDC", "SDC"))

# Check sample sizes
cat("48H RDC/SDC samples:", nsamples(ps_genus_48h_rdc_sdc), "\n")
cat("RDC samples:", sum(sample_data(ps_genus_48h_rdc_sdc)$carbohydrate == "RDC"), "\n")
cat("SDC samples:", sum(sample_data(ps_genus_48h_rdc_sdc)$carbohydrate == "SDC"), "\n")

corncob_48h_path <- file.path(outputs_dir, "corncob_genus_48h_rdc_sdc_results.rds")

corncob_48h_test <- cache_rds(
  corncob_48h_path,
  {
    set.seed(12345)
    tryCatch({
      corncob::differentialTest(
        formula = ~ group + carbohydrate + subject_id,
        phi.formula = ~ group + carbohydrate + subject_id,
        formula_null = ~ group + subject_id,
        phi.formula_null = ~ group + subject_id,
        test = "Wald",
        data = ps_genus_48h_rdc_sdc,
        fdr_cutoff = 0.1,
        B = 50
      )
    }, error = function(e) {
      message("Corncob 48H RDC/SDC failed: ", e$message)
      NULL
    })
  }
)

if (!is.null(corncob_48h_test)) {
  if (!is.null(corncob_48h_test$wald_results)) {
    corncob_48h_results <- corncob_48h_test$wald_results %>%
      dplyr::mutate(q_value = fdr_adjust(p_val)) %>%
      dplyr::arrange(q_value)
  } else if (!is.null(corncob_48h_test$p) && !is.null(corncob_48h_test$p_fdr)) {
    p_mat <- corncob_48h_test$p
    q_mat <- corncob_48h_test$p_fdr

    if (is.null(dim(p_mat))) {
      p_vals <- as.numeric(p_mat)
    } else {
      p_vals <- as.numeric(p_mat[, 1])
    }

    if (is.null(dim(q_mat))) {
      q_vals = as.numeric(q_mat)
    } else {
      q_vals = as.numeric(q_mat[, 1])
    }

    taxa_names <- rownames(p_mat)
    if (is.null(taxa_names)) {
      taxa_names <- names(p_mat)
    }
    if (is.null(taxa_names)) {
      taxa_names <- paste0("taxon_", seq_along(p_vals))
    }

    corncob_48h_results <- tibble::tibble(
      taxon = taxa_names,
      p_val = p_vals,
      q_value = q_vals
    ) %>%
      dplyr::arrange(q_value)
  } else {
    corncob_48h_results <- tibble()
  }
} else {
  corncob_48h_results <- tibble()
}

corncob_48h_results

# Show significant results (q < 0.1)
corncob_48h_significant <- corncob_48h_results %>%
  dplyr::filter(q_value < 0.1, !is.na(q_value)) %>%
  dplyr::arrange(q_value)

if (nrow(corncob_48h_significant) > 0) {
  cat("Significant taxa (q < 0.1):\n")
  print(corncob_48h_significant)
} else {
  cat("No significant taxa found at q < 0.1\n")
}

# Try to plot with error handling
if (!is.null(corncob_48h_test)) {
  tryCatch({
    plot(corncob_48h_test)
  }, error = function(e) {
    message("Plotting failed: ", e$message)
    message("This is likely due to dimension mismatches in the corncob results object")
  })
}
```

### 8.3 ANCOM-BC2

```{r ancombc2}
ancombc2_path <- file.path(outputs_dir, "ancombc2_genus_results.rds")

# Function to run ANCOMBC2 with multiple fallback strategies
run_ancombc2_safe <- function(ps_data, tax_level = "Genus") {
  
  # Strategy 1: Try with reduced complexity and conservative settings
  tryCatch({
    message("Attempting ANCOMBC2 with conservative settings...")
    ANCOMBC::ancombc2(
      data = ps_data,
      tax_level = tax_level,
      fix_formula = "group + carbohydrate * timepoint",
      rand_formula = "(1|subject_id)",
      p_adj_method = "BH",
      prv_cut = 0.15,  # Increased from 0.10 to reduce data size
      lib_cut = 0,
      s0_perc = 0.1,   # Increased from 0.05 for stability
      group = "carbohydrate",
      struc_zero = TRUE,  # Disabled to reduce complexity
      neg_lb = TRUE,      # Disabled to reduce complexity
      alpha = 0.05,
      global = TRUE,      # Disabled to reduce computation
      verbose = TRUE
    )
  }, error = function(e1) {
    message("Strategy 1 failed: ", e1$message)
    
    # Strategy 2: Try with simplified formula (no interaction)
    tryCatch({
      message("Attempting ANCOMBC2 with simplified formula...")
      ANCOMBC::ancombc2(
        data = ps_data,
        tax_level = tax_level,
        fix_formula = "group + carbohydrate + timepoint",
        rand_formula = "(1|subject_id)",
        p_adj_method = "BH",
        prv_cut = 0.20,  # Further increased
        lib_cut = 0,
        s0_perc = 0.1,
        group = "carbohydrate",
        struc_zero = TRUE,
        neg_lb = TRUE,
        alpha = 0.05,
        global = TRUE,
        verbose = TRUE
      )
    }, error = function(e2) {
      message("Strategy 2 failed: ", e2$message)
      
      # Strategy 3: Try without random effects
      tryCatch({
        message("Attempting ANCOMBC2 without random effects...")
        ANCOMBC::ancombc2(
          data = ps_data,
          tax_level = tax_level,
          fix_formula = "group + carbohydrate + timepoint",
          rand_formula = NULL,  # No random effects
          p_adj_method = "BH",
          prv_cut = 0.25,
          lib_cut = 0,
          s0_perc = 0.1,
          group = "carbohydrate",
          struc_zero = TRUE,
          neg_lb = TRUE,
          alpha = 0.05,
          global = TRUE,
          verbose = TRUE
        )
      }, error = function(e3) {
        message("Strategy 3 failed: ", e3$message)
        
        # Strategy 4: Try with even more conservative settings
        tryCatch({
          message("Attempting ANCOMBC2 with minimal settings...")
          ANCOMBC::ancombc2(
            data = ps_data,
            tax_level = tax_level,
            fix_formula = "carbohydrate",  # Only main effect
            rand_formula = NULL,
            p_adj_method = "BH",
            prv_cut = 0.30,
            lib_cut = 0,
            s0_perc = 0.1,
            group = "carbohydrate",
            struc_zero = TRUE,
            neg_lb = TRUE,
            alpha = 0.05,
            global = TRUE,
            verbose = TRUE
          )
        }, error = function(e4) {
          message("All ANCOMBC2 strategies failed. Last error: ", e4$message)
          return(NULL)
        })
      })
    })
  })
}

ancombc2_res <- cache_rds(
  ancombc2_path,
  {
    run_ancombc2_safe(ps_counts, "Genus")
  }
)

if (!is.null(ancombc2_res)) {
  ancombc2_tbl <- purrr::map_dfr(names(ancombc2_res$lfc), function(effect) {
    tibble(
      taxon = rownames(ancombc2_res$lfc[[effect]]),
      effect = effect,
      beta = ancombc2_res$lfc[[effect]][, 1],
      se = ancombc2_res$se[[effect]][, 1],
      W = ancombc2_res$W[[effect]][, 1],
      p_value = ancombc2_res$p_val[[effect]][, 1],
      q = ancombc2_res$q_val[[effect]][, 1],
      diff_abn = ancombc2_res$diff_abn[[effect]][, 1]
    )
  }) %>% dplyr::arrange(q)
} else {
  ancombc2_tbl <- tibble()
}

ancombc2_tbl %>%
  dplyr::filter(q < 0.1)

# If ANCOMBC2 failed, try alternative approach
if (nrow(ancombc2_tbl) == 0) {
  cat("ANCOM-BC2 analysis failed. Trying alternative approach with DESeq2...\n")
  
  # Alternative: Use DESeq2 for differential abundance analysis
  tryCatch({
    library(DESeq2)
    
    # Convert to DESeq2 format
    dds <- phyloseq_to_deseq2(ps_counts, ~ group + carbohydrate + timepoint)
    
    # Run DESeq2
    dds <- DESeq(dds, test = "Wald", fitType = "parametric")
    
    # Get results for carbohydrate effect
    res_carb <- results(dds, name = "carbohydrate_RDC_vs_No_Added_Carb", alpha = 0.05)
    res_carb <- res_carb[order(res_carb$padj), ]
    
    # Convert to data frame
    deseq2_tbl <- data.frame(
      taxon = rownames(res_carb),
      effect = "carbohydrate_RDC_vs_No_Added_Carb",
      beta = res_carb$log2FoldChange,
      se = res_carb$lfcSE,
      p_value = res_carb$pvalue,
      q = res_carb$padj,
      diff_abn = res_carb$padj < 0.05
    ) %>%
      filter(!is.na(q)) %>%
      dplyr::arrange(q)
    
    # Save results
    write_csv(deseq2_tbl, file.path(outputs_dir, "deseq2_genus_results.csv"))
    
    # Display summary
    cat("DESeq2 Results Summary (Alternative to ANCOM-BC2):\n")
    cat("Total taxa tested:", nrow(deseq2_tbl), "\n")
    cat("Significant taxa (q < 0.05):", sum(deseq2_tbl$q < 0.05, na.rm = TRUE), "\n")
    
    # Show top significant results
    deseq2_tbl %>%
      filter(q < 0.1) %>%
      head(10) %>%
      kable(caption = "Top 10 DESeq2 Significant Results (Alternative)")
    
  }, error = function(e) {
    cat("DESeq2 alternative also failed:", e$message, "\n")
    cat("Consider using simpler statistical approaches or reducing data complexity.\n")
  })
}
```

### 8.4 ANCOM-BC (48 h, RDC reference)

```{r ancombc48}
ps_48h_counts <- subset_samples(ps_counts, Timepoint == "48H")

ancombc48_path <- file.path(outputs_dir, "ancombc_48h_results.rds")

ancombc_48h <- cache_rds(
  ancombc48_path,
  {
    tryCatch({
      ANCOMBC::ancombc(
        phyloseq = ps_48h_counts,
        formula = "group + carbohydrate",
        p_adj_method = "BH",
        zero_cut = 0.90,
        lib_cut = 0,
        group = "carbohydrate",
        struc_zero = TRUE,
        neg_lb = TRUE,
        alpha = 0.05,
        global = TRUE
      )
    }, error = function(e) {
      message("ANCOM-BC 48h failed: ", e$message)
      NULL
    })
  }
)

if (!is.null(ancombc_48h)) {
  ancombc48_tbl <- purrr::map_dfr(names(ancombc_48h$beta), function(effect) {
    tibble(
      taxon = rownames(ancombc_48h$beta[[effect]]),
      effect = effect,
      beta = ancombc_48h$beta[[effect]][, 1],
      se = ancombc_48h$se[[effect]][, 1],
      W = ancombc_48h$W[[effect]][, 1],
      p = ancombc_48h$p_val[[effect]][, 1],
      q = ancombc_48h$q_val[[effect]][, 1]
    )
  }) %>%
    dplyr::filter(q < 0.1)
} else {
  ancombc48_tbl <- tibble()
}

ancombc48_tbl
```

### 8.5 ALDEx2 (48 h carbohydrate effect)

```{r aldex}
aldex_path <- file.path(outputs_dir, "aldex2_48h_results.rds")

aldex_tbl <- cache_rds(
  aldex_path,
  {
    counts_48h <- as(otu_table(ps_genus_48h), "matrix")
    conds_48h <- sample_data(ps_genus_48h)$carbohydrate

    aldex_clr <- ALDEx2::aldex.clr(counts_48h, conds_48h, mc.samples = 64, denom = "all", verbose = FALSE)
    aldex_kw <- ALDEx2::aldex.kw(aldex_clr)
    aldex_effect <- ALDEx2::aldex.effect(aldex_clr)

    aldex_kw %>%
      rownames_to_column("taxon") %>%
      dplyr::left_join(aldex_effect %>% rownames_to_column("taxon"), by = "taxon") %>%
      dplyr::mutate(q = fdr_adjust(kw.ep)) %>%
      dplyr::arrange(q)
  }
)

aldex_tbl %>% dplyr::filter(q < 0.1)
```

### 8.6 LEfSe (48 h, subclass = group)

```{r lefse}
lefse_path <- file.path(outputs_dir, "lefse_48h_results.rds")

lefse_res <- cache_rds(
  lefse_path,
  {
    tryCatch({
      microbiomeMarker::run_lefse(
        object = ps_genus_48h,
        class = "carbohydrate",
        subclass = "group",
        taxa_rank = "Genus",
        kw_cutoff = 0.05,
        lda_cutoff = 2
      )
    }, error = function(e) {
      message("LEfSe failed: ", e$message)
      NULL
    })
  }
)

lefse_tbl <- if (!is.null(lefse_res)) {
  microbiomeMarker::marker_table(lefse_res) %>%
    as_tibble() %>%
    dplyr::arrange(desc(lda))
} else {
  tibble()
}

lefse_tbl
```

### 8.7 MaAsLin2

```{r maaslin2}
maaslin_path <- file.path(outputs_dir, "maaslin2_genus_results.rds")

maaslin_counts <- as.data.frame(otu_table(ps_genus_filtered))
maaslin_meta <- as(sample_data(ps_genus_filtered), "data.frame")
maaslin_meta$SampleID <- rownames(maaslin_meta)

maaslin_fit <- cache_rds(
  maaslin_path,
  {
    Maaslin2::Maaslin2(
      input_data = maaslin_counts,
      input_metadata = maaslin_meta,
      output = file.path(outputs_dir, "maaslin2_genus"),
      min_prevalence = prevalence_cutoff,
      normalization = "TSS",
      transform = "LOG",
      analysis_method = "LM",
      fixed_effects = c("group", "carbohydrate", "timepoint"),
      random_effects = c("subject_id"),
      plot_heatmap = FALSE,
      plot_scatter = FALSE
    )
  }
)

maaslin_tbl <- maaslin_fit$results %>%
  as_tibble() %>%
  dplyr::mutate(q = fdr_adjust(qval)) %>%
  dplyr::arrange(q)

maaslin_tbl %>% dplyr::filter(q < 0.1)
```

### 8.8 ADAPT

```{r adapt}
adapt_path <- file.path(outputs_dir, "adapt_results.rds")

adapt_res <- cache_rds(
  adapt_path,
  {
    tryCatch({
      ADAPT::adapt(
        input_data = ps_genus_filtered,
        cond.var = "carbohydrate",
        base.cond = "No Added Carb",
        time.var = "timepoint",
        random.var = "subject_id"
      )
    }, error = function(e) {
      message("ADAPT failed: ", e$message)
      NULL
    })
  }
)

adapt_tbl <- if (!is.null(adapt_res)) {
  signals <- adapt_res@signal
  details <- adapt_res@details
  tibble(
    taxon = signals,
    log10_fc = details[signals, "log10foldchange"],
    direction = if_else(log10_fc > 0, "Enriched", "Depleted")
  )
} else {
  tibble()
}

adapt_tbl
```

### 8.9 Random Forest Classification (48 h carbohydrate)

```{r random-forest}
rf_path <- file.path(outputs_dir, "random_forest_markers.rds")

rf_markers <- cache_rds(
  rf_path,
  {
    set.seed(12345)
    rf_res <- tryCatch({
      microbiomeMarker::run_sl(
        object = ps_genus_48h,
        group = "carbohydrate",
        taxa_rank = "Genus",
        norm = "TSS",
        method = "RF",
        nfolds = 5,
        nrepeats = 1,
        top_n = 30
      )
    }, error = function(e) {
      message("Random forest failed: ", e$message)
      NULL
    })

    if (!is.null(rf_res)) {
      microbiomeMarker::marker_table(rf_res) %>%
        as_tibble() %>%
        dplyr::arrange(desc(importance))
    } else {
      NULL
    }
  }
)

rf_markers <- rf_markers %||% tibble()

rf_markers
```
```
### 8.10 Cross-Method Signal Concordance

```{r da-consensus}
da_lists <- list(
  Corncob = pull_safe(corncob_results %>% dplyr::filter(q_value < 0.1), "taxon"),
  ANCOMBC2 = {
    col <- first_match_col(ancombc2_tbl, c("taxa_id", "taxon"))
    if (is.null(col)) character() else pull_safe(ancombc2_tbl %>% dplyr::filter(q < 0.1), col)
  },
  ANCOMBC48 = pull_safe(ancombc48_tbl, "taxon"),
  ALDEx2 = pull_safe(aldex_tbl %>% dplyr::filter(q < 0.1), "taxon"),
  LEfSe = pull_safe(lefse_tbl, "taxa"),
  MaAsLin2 = pull_safe(maaslin_tbl %>% dplyr::filter(q < 0.1), "feature"),
  ADAPT = pull_safe(adapt_tbl, "taxon"),
  RandomForest = {
    col <- first_match_col(rf_markers, c("taxa", "feature", "marker"))
    if (is.null(col)) character() else pull_safe(rf_markers %>% slice_head(n = 15), col)
  }
)

da_long <- enframe(da_lists, name = "method", value = "taxa") %>%
  tidyr::unnest(taxa) %>%
  dplyr::distinct(method, taxa) %>%
  dplyr::count(taxa, method) %>%
  dplyr::mutate(present = 1)

consensus_matrix <- da_long %>%
  pivot_wider(names_from = method, values_from = present, values_fill = 0)

consensus_matrix %>% dplyr::arrange(desc(rowSums(across(where(is.numeric)))))
```

```{r forest-plot}
if (nrow(ancombc48_tbl) > 0) {
  ancombc48_tbl %>%
    dplyr::mutate(
      taxon = stringr::str_replace_all(taxon, "_", " "),
      effect = factor(effect)
    ) %>%
    ggplot(aes(x = beta, y = reorder(taxon, beta), xmin = beta - 1.96 * se, xmax = beta + 1.96 * se, color = effect)) +
    geom_point(size = 2.5) +
    geom_errorbar(width = 0.2) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    labs(
      x = "Log fold-change vs RDC",
      y = NULL,
      title = "48 h ANCOM-BC effect sizes"
    ) +
    custom_theme()
} else {
  ggplot() + annotate("text", x = 0, y = 0, label = "No ANCOM-BC signals at q < 0.10") +
    theme_void()
}
```
```
```{r da-interpretation, results='asis'}
consensus_summary <- consensus_matrix %>%
  dplyr::mutate(method_count = rowSums(across(where(is.numeric)))) %>%
  dplyr::arrange(desc(method_count))

if (nrow(consensus_summary) == 0) {
  cat("No overlapping differential taxa detected across the configured methods at q < 0.10.")
} else {
  top_taxa <- consensus_summary %>%
    slice_max(order_by = method_count, n = 10) %>%
    dplyr::mutate(text = glue::glue("{taxa}: detected by {method_count} methods")) %>%
    dplyr::pull(text)
  cat(glue::glue_collapse(top_taxa, sep = "\n"))
}
```
```
## 9. Summary and Next Steps

```{r final-summary, results='asis'}
alpha_highlights <- alpha_significance %>%
  dplyr::mutate(text = glue::glue("{metric}: {sig_terms}")) %>%
  dplyr::pull(text) %>%
  glue::glue_collapse(sep = "; ")

permanova_highlights <- permanova_results %>%
  dplyr::mutate(p_value = pr_greater_f_, q = fdr_adjust(p_value)) %>%
  dplyr::filter(term != "Residual", q < 0.1) %>%
  dplyr::mutate(text = glue::glue("{stringr::str_to_title(distance)} {term}")) %>%
  dplyr::pull(text) %>%
  glue::glue_collapse(sep = "; ")

if (permanova_highlights == "") permanova_highlights <- "No beta-diversity terms at q < 0.10"

consensus_highlights <- if (exists("consensus_summary") && nrow(consensus_summary) > 0) {
  consensus_summary %>%
    slice_max(order_by = method_count, n = 5) %>%
    dplyr::mutate(text = glue::glue("{taxa} ({method_count} methods)")) %>%
    dplyr::pull(text) %>%
    glue::glue_collapse(sep = "; ")
} else {
  "No shared differential taxa"
}

cat(glue::glue(
  "- Alpha diversity signals: {alpha_highlights}\n- Beta diversity drivers: {permanova_highlights}\n- Differential abundance consensus: {consensus_highlights}"
))
```

**Suggested extensions**

1. Validate differential findings with targeted modeling (e.g., mixed-effects regressions on log-relative abundances) for key taxa.
2. Incorporate functional pathway predictions (PICRUSt2, HUMAnN) to relate taxonomic shifts to metabolic potential.
3. Explore subject-level trajectories using longitudinal models (e.g., smoothing splines) to capture individual response heterogeneity.
```
