---
title: "Microbiome Analysis v3 - ASV Finder"
author: "Abbott Carbohydrate Obesity Project"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    code_folding: hide
    theme: flatly
    highlight: tango
    number_sections: true
    fig_width: 10
    fig_height: 6
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 5.5,
  cache = FALSE
)
```

## 1. Analysis Roadmap

This v3 report refines the v2 workflow to deliver a cleaner, more cohesive analysis with harmonized graphics, consistent statistical modeling, and dynamic interpretation. The key updates include:

- Streamlined data processing with shared helper functions and reproducible checks.
- Consistent mixed-effects modeling that respects the repeated-measures design (`SubjectID` random intercepts).
- False discovery rate control (Benjamini–Hochberg) applied to all multi-comparison contexts.
- Publication-quality visual styling and unified color palettes for categorical variables.
- Interpretation blocks populated via inline code so narrative text always reflects current outputs.
- Expanded differential-abundance section spanning corncob, ANCOM-BC2, ALDEx2, LEfSe, MaAsLin2, ADAPT, random forest, and consensus summaries.
- Dedicated focus on 48 h responses, baseline clustering, and analyses with combined groups for improved power when appropriate.

## 2. System Check and Memory Optimization

```{r system-check}
# Check system resources and optimize for large datasets
cat("System Information:\n")
cat("R version:", R.version.string, "\n")
cat("Platform:", R.version$platform, "\n")
cat("Memory limit:", memory.limit(), "MB\n")

# Check available memory
if (.Platform$OS.type == "unix") {
  mem_info <- system("free -h", intern = TRUE)
  cat("System memory:\n")
  cat(paste(mem_info, collapse = "\n"), "\n")
}

# Set memory-efficient options
options(
  scipen = 999,  # Avoid scientific notation
  digits = 4,    # Limit decimal places
  max.print = 100,  # Limit printed output
  stringsAsFactors = FALSE
)

# Increase memory limit if possible
tryCatch({
  if (.Platform$OS.type == "windows") {
    memory.limit(size = 8000)  # 8GB
  }
}, error = function(e) {
  cat("Could not increase memory limit:", e$message, "\n")
})

cat("System check completed.\n")
```

## 3. Libraries, Options, and Helper Functions

```{r libraries}
required_packages <- c(
  "tidyverse", "readr", "janitor", "glue", "scales", "patchwork",
  "phyloseq", "microbiome", "vegan", "betapart", "ape", "Biostrings",
  "broom", "broom.mixed", "lme4", "lmerTest", "emmeans", "performance",
  "rstatix", "ggpubr", "ggtext", "cowplot", "ggsci", "scico",
  "here", "corncob", "ANCOMBC", "ALDEx2", "microbiomeMarker", "ADAPT",
  "Maaslin2", "gt", "cluster", "nlme", "randomForest", "ranger", "caret", "phyloseqGraphTest"
)

# Function to install missing packages
install_missing_packages <- function(packages) {
  missing_packages <- packages[!vapply(packages, requireNamespace, logical(1), quietly = TRUE)]
  
  if (length(missing_packages) > 0) {
    cat("Installing missing packages:", paste(missing_packages, collapse = ", "), "\n")
    
    # Bioconductor packages
    bioc_packages <- c("phyloseq", "microbiome", "Biostrings", "ANCOMBC", "ALDEx2", "microbiomeMarker", "Maaslin2", "phyloseqGraphTest", "ADAPT")
    bioc_to_install <- intersect(missing_packages, bioc_packages)
    
    # CRAN packages
    cran_to_install <- setdiff(missing_packages, bioc_packages)
    
    # Install Bioconductor packages
    if (length(bioc_to_install) > 0) {
      if (!requireNamespace("BiocManager", quietly = TRUE)) {
        install.packages("BiocManager")
      }
      BiocManager::install(bioc_to_install, update = FALSE, ask = FALSE)
    }
    
    # Install CRAN packages
    if (length(cran_to_install) > 0) {
      install.packages(cran_to_install, repos = "https://cloud.r-project.org/")
    }
    
    cat("Package installation completed.\n")
  }
}

# Install missing packages
install_missing_packages(required_packages)

invisible(lapply(required_packages, library, character.only = TRUE))

options(dplyr.summarise.inform = FALSE)
set.seed(12345)

# Output directory for derived artefacts
outputs_dir <- here::here("outputs_v3")
dir.create(outputs_dir, showWarnings = FALSE, recursive = TRUE)
```

```{r theming}
# Harmonized aesthetic ---------------------------------------------------

palette_group <- setNames(scico::scico(2, palette = "berlin"), c("Healthy", "Obese"))
palette_carbohydrate <- setNames(c("#1f77b4", "#ff7f0e", "#2ca02c"), c("No Added Carb", "RDC", "SDC"))
palette_timepoint <- setNames(scico::scico(2, palette = "roma"), c("0H", "48H"))

palette_cluster <- setNames(scico::scico(6, palette = "batlow"), paste0("Cluster", seq_len(6)))

custom_theme <- function(base_size = 12, base_family = "sans") {
  theme_minimal(base_size = base_size, base_family = base_family) +
    theme(
      panel.grid.major = element_line(color = "grey85", linewidth = 0.4),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(color = "grey65", fill = NA, linewidth = 0.4),
      plot.title = element_text(face = "bold", size = base_size + 2),
      plot.subtitle = element_text(size = base_size, margin = ggplot2::margin(b = 6)),
      plot.caption = element_text(size = base_size - 2, color = "grey40"),
      strip.text = element_text(face = "bold", size = base_size - 1),
      legend.position = "right",
      legend.title = element_text(face = "bold"),
      legend.background = element_rect(fill = "white", color = NA)
    )
}

theme_set(custom_theme())

format_p <- function(p) {
  case_when(
    is.na(p) ~ "NA",
    p < 0.001 ~ "<0.001",
    TRUE ~ sprintf("%.3f", p)
  )
}

fdr_adjust <- function(p) stats::p.adjust(p, method = "BH")

add_fdr <- function(tbl, p_col = "p_value", new_col = "fdr") {
  tbl %>% dplyr::mutate(!!new_col := fdr_adjust(.data[[p_col]]))
}

anova_table <- function(model) {
  if (inherits(model, "lme")) {
    anova_df <- nlme::anova.lme(model, type = "marginal") %>%
      rownames_to_column("term") %>%
      as_tibble() %>%
      janitor::clean_names() %>%
      dplyr::rename(
        num_df = num_df,
        den_df = den_df,
        f_value = f_value,
        p_value = p_value
      )

    if (!"p_value" %in% names(anova_df)) {
      warning("anova.lme output did not contain p-values; returning empty tibble")
      return(tibble())
    }

    anova_df %>% add_fdr("p_value", "fdr")
  } else {
    anova(model, type = 3) %>%
      rownames_to_column("term") %>%
      as_tibble() %>%
      janitor::clean_names() %>%
      dplyr::rename(p_value = pr_greater_f_) %>%
      add_fdr("p_value", "fdr")
  }
}

emmeans_pairs <- function(model, specs, adjust = "BH", by = NULL) {
  emmeans::emmeans(model, specs = specs, by = by) %>%
    emmeans::contrast(method = "pairwise", adjust = adjust) %>%
    broom::tidy()
}
```

```{r util-functions}
# Helper utilities -------------------------------------------------------

collapse_taxa <- function(ps_obj, rank = "Genus", top_n = 15) {
  ps_rel <- microbiome::transform(ps_obj, "compositional")
  tax_rank <- tax_table(ps_rel)[, rank]
  tax_rank[is.na(tax_rank) | tax_rank == ""] <- paste0(rank, "_Unassigned")
  tax_table(ps_rel)[, rank] <- tax_rank

  ps_rank <- tax_glom(ps_rel, taxrank = rank, NArm = FALSE)
  abund <- taxa_sums(ps_rank)
  keep <- rank(abund, ties.method = "first") > (ntaxa(ps_rank) - top_n)
  ps_keep <- prune_taxa(keep, ps_rank)
  other <- prune_taxa(!keep, ps_rank)

  if (ntaxa(other) > 0) {
    other_sum <- tax_table(other)[, rank]
    tax_table(other)[, rank] <- "Other"
    ps_rank <- merge_phyloseq(ps_keep, other)
  } else {
    ps_rank <- ps_keep
  }

  ps_rank
}

plot_bar_compositional <- function(ps_obj, fill_var, facet_var = NULL, rank = "Genus", top_n = 15) {
  ps_rank <- collapse_taxa(ps_obj, rank = rank, top_n = top_n)
  df <- psmelt(ps_rank)

  fill_palette <- scico::scico(length(unique(df[[rank]])) + 1, palette = "batlow")

  ggplot(df, aes(x = .data[[fill_var]], y = Abundance, fill = .data[[rank]])) +
    geom_col(position = "stack") +
    scale_y_continuous(labels = scales::percent) +
    scale_fill_manual(values = fill_palette, guide = guide_legend(ncol = 1)) +
    labs(
      x = NULL,
      y = "Relative abundance",
      fill = rank
    ) +
    custom_theme() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    facet_wrap(vars(.data[[facet_var]]), scales = "free_x") %||%
    identity
}

format_emm_table <- function(tbl) {
  tbl %>%
    dplyr::mutate(
      estimate = round(estimate, 3),
      conf.low = round(conf.low, 3),
      conf.high = round(conf.high, 3),
      p_raw = p.value,
      fdr = p.adjust(p.value, method = "BH")
    ) %>%
    dplyr::mutate(
      p.value = format_p(p_raw),
      fdr = format_p(fdr)
    ) %>%
    dplyr::select(-p_raw)
}

inline_glue <- function(...) {
  glue::glue_data(.x = environment(), ...)
}

`%||%` <- function(x, y) if (!is.null(x)) x else y

cache_rds <- function(path, code) {
  if (file.exists(path)) {
    message("Loading cached result: ", path)
    readRDS(path)
  } else {
    dir.create(dirname(path), showWarnings = FALSE, recursive = TRUE)
    message("Caching result to: ", path)
    result <- code
    if (!is.null(result)) {
      saveRDS(result, path)
    }
    result
  }
}

pull_safe <- function(tbl, col) {
  if (is.null(tbl) || nrow(tbl) == 0 || !(col %in% names(tbl))) {
    return(character())
  }
  tbl[[col]]
}

first_match_col <- function(tbl, candidates) {
  candidates <- candidates[candidates %in% names(tbl)]
  if (length(candidates) == 0) return(NULL)
  candidates[[1]]
}
```

## 3. Data Import and Curation

```{r data-load}
metadata_raw <- readr::read_tsv("sample.metadata.tsv", show_col_types = FALSE) %>%
  janitor::clean_names()

metadata <- metadata_raw %>%
  separate(customer_label, into = c("subject_id", "replicate_code", "timepoint"), sep = "\\.", remove = FALSE) %>%
  dplyr::mutate(
    carbohydrate = case_when(
      str_detect(replicate_code, "R") ~ "RDC",
      str_detect(replicate_code, "S") ~ "SDC",
      TRUE ~ "No Added Carb"
    ),
    carbohydrate = factor(carbohydrate, levels = c("No Added Carb", "RDC", "SDC")),
    timepoint = factor(timepoint, levels = c("0H", "48H")),
    replicate = readr::parse_number(replicate_code) %>% replace_na(1)
  ) %>%
  dplyr::rename(SampleID = sample_id) %>%
  dplyr::select(SampleID, subject_id, timepoint, carbohydrate, replicate, customer_label, raw_seq_files)

healthy_lookup <- readr::read_tsv(
  file = here::here("zr24558.16S_250813.zymo", "1.Healthy.TimepointComparism..Bac16Sv34", "Sample_Information", "sample.metadata.csv"),
  show_col_types = FALSE
) %>%
  janitor::clean_names() %>%
  separate(customer_label, into = c("subject_id", "tmp", "tp"), sep = "\\.", remove = FALSE) %>%
  dplyr::distinct(subject_id) %>%
  dplyr::mutate(group = "Healthy")

metadata <- metadata %>%
  dplyr::mutate(
    group = if_else(subject_id %in% healthy_lookup$subject_id, "Healthy", "Obese"),
    group = factor(group, levels = c("Healthy", "Obese")),
    subject_id = factor(subject_id)
  )

metadata_summary <- metadata %>%
  dplyr::count(group, carbohydrate, timepoint, name = "n_samples")

subject_counts <- metadata %>%
  dplyr::distinct(subject_id, group) %>%
  dplyr::count(group, name = "n_subjects")

metadata_overview <- metadata_summary %>%
  dplyr::group_by(group) %>%
  dplyr::summarise(total = sum(n_samples), .groups = "drop") %>%
  dplyr::left_join(subject_counts, by = "group")
```

**Cohort coverage**

- Samples: `r scales::comma(nrow(metadata))`
- Subjects: `r scales::comma(n_distinct(metadata$subject_id))`
- Group allocation: `r glue::glue_collapse(glue::glue('{metadata_overview$group}: {metadata_overview$n_subjects} subjects, {metadata_overview$total} samples'), sep = '; ')`
- Carbohydrate distribution: `r glue::glue_collapse(levels(metadata$carbohydrate), sep = ', ')`

```{r metadata-table}
metadata_summary %>%
  dplyr::arrange(group, carbohydrate, timepoint) %>%
  gt::gt(rowname_col = "timepoint") %>%
  gt::tab_header(title = "Sample counts by group, carbohydrate, and timepoint") %>%
  gt::fmt_number(columns = "n_samples", decimals = 0)
```
### 3.1 Amplicon Sequence Variants and Taxonomy

```{r phyloseq-import}
asv_raw <- readr::read_tsv("dada2_asv_2/asv_table.tsv", show_col_types = FALSE)

asv_ids <- stringr::str_extract(asv_raw$ASV_ID, "ASV_\\d+")

otu_mat <- asv_raw %>%
  dplyr::select(-ASV_ID, -Sequence, -Length, -Total_Abundance, -Relative_Abundance) %>%
  as.data.frame()

colnames(otu_mat) <- stringr::str_remove(colnames(otu_mat), "_filtered$")
rownames(otu_mat) <- asv_ids

taxonomy_raw <- readr::read_tsv("dada2_asv_2/raxtax_taxonomy.tsv", show_col_types = FALSE)
rownames(taxonomy_raw) <- paste0("ASV_", 1:nrow(taxonomy_raw))

taxonomy_mat <- taxonomy_raw %>%
  dplyr::mutate(ASV = stringr::str_extract(ASV_ID, "ASV_\\d+")) %>%
  dplyr::select(ASV, Kingdom, Phylum, Class, Order, Family, Genus, Species) %>%
  dplyr::mutate(across(everything(), ~ replace_na(., "Unassigned"))) %>%
  column_to_rownames("ASV") %>%
  as.matrix()

stopifnot(all(rownames(otu_mat) %in% rownames(taxonomy_mat)))

metadata_ps <- metadata %>%
  dplyr::mutate(SampleID = as.character(SampleID)) %>%
  column_to_rownames("SampleID")

common_samples <- intersect(colnames(otu_mat), rownames(metadata_ps))
otu_mat <- otu_mat[, common_samples]
metadata_ps <- metadata_ps[common_samples, ]

ps <- phyloseq(
  otu_table(as.matrix(otu_mat), taxa_are_rows = TRUE),
  sample_data(metadata_ps),
  tax_table(taxonomy_mat[rownames(otu_mat), ])
)

ps <- prune_taxa(taxa_sums(ps) > 0, ps)
ps <- prune_samples(sample_sums(ps) > 0, ps)

sample_data(ps)$SubjectID <- sample_data(ps)$subject_id
sample_data(ps)$Group <- sample_data(ps)$group
sample_data(ps)$Carbohydrate <- sample_data(ps)$carbohydrate
sample_data(ps)$Timepoint <- sample_data(ps)$timepoint

ps_rel <- microbiome::transform(ps, "compositional")
```

```{r sequencing-depth}
seq_depth <- tibble(
  SampleID = sample_names(ps),
  reads = sample_sums(ps)
) %>%
  dplyr::left_join(as_tibble(metadata, .name_repair = "minimal"), by = "SampleID")

coverage_stats <- seq_depth %>%
  dplyr::summarise(
    min_reads = min(reads),
    median_reads = median(reads),
    max_reads = max(reads)
  )

coverage_stats
```

**Retained ASV reads** – minimum `r scales::comma(coverage_stats$min_reads)`, median `r scales::comma(coverage_stats$median_reads)`, maximum `r scales::comma(coverage_stats$max_reads)` reads per sample represented in the denoised ASV table after filtering and chimera removal.


## 4. Alpha Diversity

```{r alpha-diversity}
alpha_metrics <- estimate_richness(ps, measures = c("Observed", "Shannon", "Simpson")) %>%
  rownames_to_column("SampleID")

alpha_long <- alpha_metrics %>%
  pivot_longer(cols = -SampleID, names_to = "metric", values_to = "value") %>%
  dplyr::left_join(as_tibble(metadata, .name_repair = "minimal"), by = "SampleID") %>%
  dplyr::mutate(
    metric = factor(metric, levels = c("Shannon", "Observed", "Simpson"))
  )

prep_mixed_data <- function(data, extra_vars = character()) {
  keep_vars <- unique(c("value", "subject_id", extra_vars))
  keep_vars <- keep_vars[keep_vars %in% names(data)]

  data <- data %>%
    dplyr::select(dplyr::all_of(keep_vars)) %>%
    tidyr::drop_na()

  if (nrow(data) == 0) {
    return(NULL)
  }

  data <- droplevels(as.data.frame(data))

  categorical_vars <- intersect(c("subject_id", extra_vars, "baseline_cluster"), names(data))
  data[categorical_vars] <- lapply(data[categorical_vars], function(x) factor(x))

  if (any(vapply(data[categorical_vars], function(x) length(levels(x)) < 2, logical(1)))) {
    return(NULL)
  }

  data
}

safe_lme <- function(fixed, random, data, extra_vars = character()) {
  data_prepped <- prep_mixed_data(data, extra_vars)
  if (is.null(data_prepped)) {
    return(NULL)
  }

  tryCatch(
    nlme::lme(
      fixed = fixed,
      data = data_prepped,
      random = random,
      method = "REML",
      control = nlme::lmeControl(opt = "optim")
    ),
    error = function(e) {
      message("Mixed model failed: ", e$message)
      NULL
    }
  )
}

fit_alpha_model <- function(df) {
  safe_lme(
    fixed = value ~ group * carbohydrate + group * timepoint + carbohydrate * timepoint,
    random = ~1 | subject_id,
    data = df,
    extra_vars = c("group", "carbohydrate", "timepoint")
  )
}

fit_alpha_model_collapsed <- function(df) {
  safe_lme(
    fixed = value ~ carbohydrate * timepoint,
    random = ~1 | subject_id,
    data = df,
    extra_vars = c("carbohydrate", "timepoint")
  )
}

fit_alpha_model_48h <- function(df) {
  df_levels <- df %>%
    dplyr::filter(!is.na(value)) %>%
    droplevels()

  group_levels <- dplyr::n_distinct(df_levels$group)
  carb_levels <- dplyr::n_distinct(df_levels$carbohydrate)

  if (group_levels <= 1 && carb_levels <= 1) {
    return(NULL)
  }

  if (group_levels > 1 && carb_levels > 1) {
    fixed_formula <- value ~ group * carbohydrate
    extra_vars <- c("group", "carbohydrate")
  } else if (carb_levels > 1) {
    fixed_formula <- value ~ carbohydrate
    extra_vars <- c("carbohydrate")
  } else {
    fixed_formula <- value ~ group
    extra_vars <- c("group")
  }

  safe_lme(
    fixed = fixed_formula,
    random = ~1 | subject_id,
    data = df,
    extra_vars = extra_vars
  )
}

fit_cluster_model <- function(df) {
  safe_lme(
    fixed = value ~ baseline_cluster * carbohydrate,
    random = ~1 | subject_id,
    data = df,
    extra_vars = c("baseline_cluster", "carbohydrate")
  )
}

alpha_metrics_vec <- levels(alpha_long$metric)
alpha_results <- vector("list", length(alpha_metrics_vec))
names(alpha_results) <- alpha_metrics_vec

for (metric_name in alpha_metrics_vec) {
  df_metric <- alpha_long %>%
    dplyr::filter(metric == metric_name) %>%
    droplevels()

  model <- fit_alpha_model(df_metric)
  anova_tbl <- if (is.null(model)) tibble::tibble() else anova_table(model)

  alpha_results[[metric_name]] <- list(
    data = df_metric,
    model = model,
    anova = anova_tbl
  )
}

alpha_models <- tibble(
  metric = alpha_metrics_vec,
  data = lapply(alpha_metrics_vec, function(m) alpha_results[[m]]$data),
  model = lapply(alpha_metrics_vec, function(m) alpha_results[[m]]$model),
  anova = lapply(alpha_metrics_vec, function(m) alpha_results[[m]]$anova)
)

alpha_significance <- alpha_models %>%
  dplyr::mutate(
    sig_terms = purrr::map_chr(
      anova,
      ~ {
        if (is.null(.x) || nrow(.x) == 0) {
          return("Model unavailable for this metric")
        }
        .x %>%
          dplyr::filter(fdr < 0.1) %>%
          dplyr::mutate(term = stringr::str_replace_all(term, ":", " × ")) %>%
          dplyr::mutate(label = glue::glue("{term} (q = {format_p(fdr)})")) %>%
          dplyr::pull(label) %>%
          glue::glue_collapse(sep = "; ") %>%
          {
            ifelse(. == "", "No terms reach q < 0.10", .)
          }
      }
    )
  ) %>%
  dplyr::select(metric, sig_terms)

```

```{r alpha-pairwise-tests}
# First, let's check the data structure
cat("Checking alpha_long data structure:\n")
cat("Total observations:", nrow(alpha_long), "\n")
cat("Metrics:", paste(unique(alpha_long$metric), collapse = ", "), "\n")
cat("Groups:", paste(unique(alpha_long$group), collapse = ", "), "\n")
cat("Timepoints:", paste(unique(alpha_long$timepoint), collapse = ", "), "\n")
cat("Carbohydrate levels:", paste(unique(alpha_long$carbohydrate), collapse = ", "), "\n")

# Check sample sizes per group
data_summary <- alpha_long %>%
  dplyr::group_by(metric, group, timepoint, carbohydrate) %>%
  dplyr::summarise(n = n(), .groups = "drop") %>%
  dplyr::arrange(metric, group, timepoint, carbohydrate)

cat("\nSample sizes per group:\n")
print(data_summary)

# Perform pairwise Wilcoxon tests for each metric across group-timepoint combinations
alpha_pairwise_tests <- alpha_long %>%
  dplyr::group_by(metric, group, timepoint) %>%
  rstatix::wilcox_test(value ~ carbohydrate, p.adjust.method = "BH") %>%
  rstatix::add_significance() %>%
  dplyr::ungroup()

cat("\nPairwise test results:\n")
cat("Number of tests performed:", nrow(alpha_pairwise_tests), "\n")
cat("Columns in results:", paste(names(alpha_pairwise_tests), collapse = ", "), "\n")

# Calculate max values for positioning annotations
alpha_max_values <- alpha_long %>%
  dplyr::group_by(metric, group, timepoint) %>%
  dplyr::summarise(max_val = max(value, na.rm = TRUE), .groups = "drop")

# Filter out non-significant comparisons and add y-positions for significance annotations
alpha_pairwise_tests <- alpha_pairwise_tests %>%
  dplyr::filter(p.adj.signif != "ns") %>%  # Remove non-significant comparisons
  dplyr::left_join(alpha_max_values, by = c("metric", "group", "timepoint")) %>%
  dplyr::group_by(metric, group, timepoint) %>%
  dplyr::mutate(
    y.position = max_val * 1.15 + (dplyr::row_number() - 1) * (max_val * 0.08)
  ) %>%
  dplyr::ungroup()

cat("After filtering non-significant comparisons:\n")
cat("Number of significant comparisons:", nrow(alpha_pairwise_tests), "\n")

cat("Final test data structure:\n")
print(head(alpha_pairwise_tests))
```

```{r alpha-figures}
alpha_plot <- function(data, metric_name, test_data) {
  # Filter data for current metric
  plot_data <- data %>%
    dplyr::filter(metric == metric_name)
  
  # Filter test results for current metric and remove non-significant comparisons
  test_data_metric <- test_data %>%
    dplyr::filter(metric == metric_name) %>%
    dplyr::filter(p.adj.signif != "ns")
  
  # Create base plot
  p <- plot_data %>%
    ggplot(aes(x = carbohydrate, y = value, fill = carbohydrate)) +
    geom_boxplot(alpha = 0.8, outlier.alpha = 0.4, width = 0.7) +
    geom_jitter(aes(color = group), width = 0.12, size = 1.5, alpha = 0.7, show.legend = FALSE) +
    scale_fill_manual(values = palette_carbohydrate) +
    scale_color_manual(values = palette_group) +
    labs(x = NULL, y = metric_name, title = metric_name) +
    facet_grid(group ~ timepoint, scales = "free_y") +
    custom_theme() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  # Add significance annotations only if there are significant comparisons
  if (nrow(test_data_metric) > 0) {
    p <- p +
      ggpubr::stat_pvalue_manual(
        test_data_metric,
        label = "p.adj.signif",
        tip.length = 0.005,
        bracket.nudge.y = 0.01,
        step.increase = 0.03
      )
  }
  
  return(p)
}

# Create plots with significance annotations
alpha_plots <- purrr::map(levels(alpha_long$metric), ~ alpha_plot(alpha_long, .x, alpha_pairwise_tests))

alpha_plots[[2]]
alpha_plots[[1]]
alpha_plots[[3]]

#alpha_plots[[1]] / alpha_plots[[2]] / alpha_plots[[3]] + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

```{r alpha-interpretation, results='asis'}
alpha_significance %>%
  dplyr::mutate(text = glue::glue("**{metric}**: {sig_terms}")) %>%
  dplyr::pull(text) %>%
  glue::glue_collapse(sep = "\n") %>%
  cat()
```

### 4.1 Combined-Group Modeling

The limited group-level separation motivates a secondary model that collapses healthy and obese cohorts to evaluate carbohydrate effects with improved power.

```{r alpha-collapsed}
alpha_collapsed <- alpha_long %>%
  dplyr::mutate(group_collapsed = factor("Combined"))

alpha_models_collapsed <- alpha_collapsed %>%
  tidyr::nest(data = -metric) %>%
  dplyr::mutate(
    model = purrr::map(data, fit_alpha_model_collapsed),
    anova = purrr::map(model, ~ if (is.null(.x)) tibble::tibble() else anova_table(.x))
  )

alpha_models_collapsed %>%
  dplyr::select(metric, anova) %>%
  tidyr::unnest(anova) %>%
  gt::gt() %>%
  gt::fmt_number(columns = c(f_value), decimals = 2) %>%
  gt::fmt_number(columns = c(p_value, fdr), decimals = 3) %>%
  gt::cols_label(term = "Effect", num_df = "df num", den_df = "df den", f_value = "F", p_value = "p", fdr = "q")
```

```{r alpha-combined-plots}
# Create combined data where healthy and obese groups are merged
alpha_combined <- alpha_long %>%
  dplyr::mutate(group = "Combined") %>%
  dplyr::mutate(group = factor(group))

# Create pairwise tests for the combined data
alpha_combined_pairwise_tests <- alpha_combined %>%
  dplyr::group_by(metric, group, timepoint) %>%
  rstatix::wilcox_test(value ~ carbohydrate, p.adjust.method = "BH") %>%
  rstatix::add_significance() %>%
  rstatix::add_y_position(fun =  "max", scales = "free_y") %>%
  dplyr::ungroup()

# plot here but with the healthy and obese combined
alpha_combined_plots <- purrr::map(levels(alpha_long$metric), ~ alpha_plot(alpha_combined, .x, alpha_combined_pairwise_tests))

alpha_combined_plots[[1]]
alpha_combined_plots[[2]]
alpha_combined_plots[[3]]

```

```{r alpha-collapsed-interpretation, results='asis'}
alpha_collapsed_summary <- alpha_models_collapsed %>%
  dplyr::mutate(
    sig_terms = purrr::map_chr(
      anova,
      ~ {
        if (is.null(.x) || nrow(.x) == 0) {
          return("Model unavailable for this metric")
        }
        .x %>%
          dplyr::filter(fdr < 0.1) %>%
          dplyr::mutate(term = stringr::str_replace_all(term, ":", " × ")) %>%
          dplyr::mutate(label = glue::glue("{term} (q = {format_p(fdr)})")) %>%
          dplyr::pull(label) %>%
          glue::glue_collapse(sep = "; ") %>%
          {
            ifelse(. == "", "No carbohydrate × timepoint effects detected at q < 0.10", .)
          }
      }
    )
  )

alpha_collapsed_summary %>%
  dplyr::mutate(text = glue::glue("**{metric}** (groups combined): {sig_terms}")) %>%
  dplyr::pull(text) %>%
  glue::glue_collapse(sep = "\n") %>%
  cat()
```

### 4.2 48 h Focus

```{r alpha-48h}
alpha_48h <- alpha_long %>% dplyr::filter(timepoint == "48H")

alpha_48_model <- alpha_48h %>%
  tidyr::nest(data = -metric) %>%
  dplyr::mutate(
    model = purrr::map(data, fit_alpha_model_48h),
    anova = purrr::map(model, ~ if (is.null(.x)) tibble::tibble() else anova_table(.x)),
    contrasts = purrr::map2(
      model,
      data,
      ~ {
        if (is.null(.x)) return(tibble::tibble())

        df_levels <- .y %>%
          dplyr::filter(!is.na(value)) %>%
          droplevels()

        group_levels <- dplyr::n_distinct(df_levels$group)
        carb_levels <- dplyr::n_distinct(df_levels$carbohydrate)

        if (group_levels > 1 && carb_levels > 1) {
          emmeans::emmeans(.x, ~ carbohydrate | group, data = df_levels) %>%
            emmeans::contrast("pairwise", adjust = "BH") %>%
            broom::tidy()
        } else if (carb_levels > 1) {
          emmeans::emmeans(.x, specs = "carbohydrate", data = df_levels) %>%
            emmeans::contrast("pairwise", adjust = "BH") %>%
            broom::tidy()
        } else if (group_levels > 1) {
          emmeans::emmeans(.x, specs = "group", data = df_levels) %>%
            emmeans::contrast("pairwise", adjust = "BH") %>%
            broom::tidy()
        } else {
          tibble::tibble()
        }
      }
    )
  )

alpha_48_model$anova
```

```{r alpha-48h-interpretation, results='asis'}
alpha_48_text <- alpha_48_model %>%
  dplyr::mutate(
    sig_terms = purrr::map_chr(
      anova,
      ~ {
        if (is.null(.x) || nrow(.x) == 0) {
          return("Model unavailable for this metric")
        }
        .x %>%
          dplyr::filter(fdr < 0.1) %>%
          dplyr::mutate(term = stringr::str_replace_all(term, ":", " × ")) %>%
          dplyr::mutate(label = glue::glue("{term} (q = {format_p(fdr)})")) %>%
          dplyr::pull(label) %>%
          glue::glue_collapse(sep = "; ") %>%
          {
            ifelse(. == "", "No significant effects after FDR correction", .)
          }
      }
    )
  )

alpha_48_text %>%
  dplyr::mutate(text = glue::glue("**{metric}** (48 h): {sig_terms}")) %>%
  dplyr::pull(text) %>%
  glue::glue_collapse(sep = "\n") %>%
  cat()
```

```{r alpha-48h-wilcoxon}
# Perform pairwise Wilcoxon tests for each metric
alpha_48h_wilcoxon <- alpha_48h %>%
  group_by(metric) %>%
  wilcox_test(value ~ carbohydrate, p.adjust.method = "BH") %>%
  add_significance()

# Calculate max values for each metric to position annotations
max_values <- alpha_48h %>%
  group_by(metric) %>%
  summarise(max_val = max(value, na.rm = TRUE), .groups = "drop")

# Add manual y-positions based on data range for each metric
alpha_48h_wilcoxon <- alpha_48h_wilcoxon %>%
  left_join(max_values, by = "metric") %>%
  group_by(metric) %>%
  mutate(
    y.position = case_when(
      metric == "Shannon" ~ max_val * 1.1,
      metric == "Observed" ~ max_val * 1.15,
      metric == "Simpson" ~ max_val * 1.1
    ),
    y.position = y.position + (row_number() - 1) * (max_val * 0.1)
  ) %>%
  ungroup()

# Check the structure of the results
print("Columns in alpha_48h_wilcoxon:")
print(names(alpha_48h_wilcoxon))
print("First few rows:")
print(head(alpha_48h_wilcoxon))
```

```{r alpha-48h-plot}
alpha_48h %>%
  ggplot(aes(x = carbohydrate, y = value, fill = carbohydrate)) +
  #geom_violin(trim = FALSE, alpha = 0.8) +
  geom_boxplot(alpha = 0.9) +
  geom_jitter(width = 0.12, size = 1.5, alpha = 0.7, show.legend = FALSE) +
  stat_pvalue_manual(alpha_48h_wilcoxon, 
                     label = "p.adj.signif", 
                     tip.length = 0.01,
                     bracket.nudge.y = 0.05) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_fill_manual(values = palette_carbohydrate) +
  labs(x = NULL, y = "Diversity", title = "Alpha diversity at 48 h") +
  custom_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 5. Beta Diversity and Community Structure

```{r beta-prep}
make_pcoa_df <- function(ps_obj, distance_method) {
  ord <- ordinate(ps_obj, method = "PCoA", distance = distance_method)
  vecs <- as_tibble(ord$vectors[, 1:2], rownames = "SampleID")
  colnames(vecs)[2:3] <- c("PC1", "PC2")
  vecs <- vecs %>%
    dplyr::mutate(
      axis1_var = ord$values$Relative_eig[1],
      axis2_var = ord$values$Relative_eig[2]
    ) %>%
    dplyr::left_join(as_tibble(metadata, .name_repair = "minimal"), by = "SampleID")
  vecs$distance_method <- distance_method
  vecs
}

dist_methods <- c("bray", "jaccard")

pcoa_dfs <- purrr::map_dfr(dist_methods, ~ make_pcoa_df(ps_rel, .x))

```

```{r permanova}
permanova_results_raw <- purrr::map_dfr(
  dist_methods,
  ~ {
    dist_mat <- phyloseq::distance(ps_rel, method = .x)
    df <- as(sample_data(ps_rel), "data.frame")
    adonis_out <- vegan::adonis2(
      dist_mat ~ group * carbohydrate + timepoint,
      data = df,
      permutations = 999,
      strata = df$subject_id
    )
    adonis_tbl <- as.data.frame(adonis_out) %>% rownames_to_column('term') %>% janitor::clean_names()
    adonis_tbl$distance <- .x
    adonis_tbl
  }
)

permanova_p_col <- intersect(c("pr_greater_f_", "pr_greater_f", "pr_f", "p_value"), names(permanova_results_raw))
permanova_f_col <- intersect(c("f", "f_model", "pseudo_f"), names(permanova_results_raw))
permanova_r2_col <- intersect(c("r2", "r_squared", "r2_adj"), names(permanova_results_raw))

if (length(permanova_p_col) == 0 || length(permanova_f_col) == 0 || length(permanova_r2_col) == 0) {
  stop(glue::glue(
    "Unexpected PERMANOVA columns. Available: {glue::glue_collapse(names(permanova_results_raw), sep = ', ')}"
  ))
}

permanova_results <- permanova_results_raw %>%
  dplyr::mutate(
    f = .data[[permanova_f_col[1]]],
    r2 = .data[[permanova_r2_col[1]]],
    p_value = .data[[permanova_p_col[1]]],
    q_value = fdr_adjust(p_value)
  )

permanova_results %>%
  gt::gt() %>%
  gt::fmt_number(columns = c(f, r2), decimals = 3) %>%
  gt::fmt_number(columns = c(p_value, q_value), decimals = 3) %>%
  gt::cols_label(term = "Effect", f = "F", r2 = "R2", p_value = "p", q_value = "q") %>%
  gt::tab_header("PERMANOVA results (subject-level strata)")
```

```{r pcoa-plot, fig.width = 20, fig.height = 16}
pcoa_plot <- function(df, method_name) {
  df_subset <- df %>% dplyr::filter(distance_method == method_name)
  var1 <- scales::percent(round(df_subset$axis1_var[1], 3))
  var2 <- scales::percent(round(df_subset$axis2_var[1], 3))

  # Get PERMANOVA results for this distance method - use group:carbohydrate interaction
  permanova_subset <- permanova_results %>%
    dplyr::filter(distance == method_name, term == "group:carbohydrate")

  # Create annotation text and position
  if (nrow(permanova_subset) > 0) {
    r2_val <- round(permanova_subset$r2[1], 3)
    p_val <- round(permanova_subset$p_value[1], 3)
    annotation_text <- glue::glue("R² = {r2_val}, p = {p_val}")

    # Calculate per-facet label positions, guard against missing/flat ranges
    annotation_df <- df_subset %>%
      dplyr::filter(!is.na(PC1), !is.na(PC2)) %>%
      dplyr::group_by(timepoint) %>%
      dplyr::summarise(
        x_min = min(PC1),
        x_max = max(PC1),
        y_min = min(PC2),
        y_max = max(PC2),
        .groups = "drop"
      ) %>%
      dplyr::mutate(
        x_span = x_max - x_min,
        y_span = y_max - y_min,
        PC1 = ifelse(is.finite(x_span) & x_span > 0, x_min + 0.05 * x_span, x_min),
        PC2 = ifelse(is.finite(y_span) & y_span > 0, y_max - 0.05 * y_span, y_max),
        label = annotation_text
      ) %>%
      dplyr::select(timepoint, PC1, PC2, label)
  } else {
    annotation_df <- data.frame(
      PC1 = numeric(0),
      PC2 = numeric(0),
      label = character(0),
      timepoint = character(0)
    )
  }

  p <- ggplot(df_subset, aes(x = PC1, y = PC2, color = carbohydrate, shape = group)) +
    geom_point(size = 2.5, alpha = 0.85) +
    stat_ellipse(aes(group = interaction(carbohydrate, timepoint), color = carbohydrate),
      linewidth = 0.4, linetype = "dashed", show.legend = FALSE
    ) +
    facet_wrap(~timepoint) +
    scale_color_manual(values = palette_carbohydrate) +
    scale_shape_manual(values = c(16, 17)) +
    labs(
      x = glue::glue("PC1 ({var1})"),
      y = glue::glue("PC2 ({var2})"),
      title = glue::glue("PCoA – {stringr::str_to_title(method_name)} distance")
    ) +
    custom_theme()

  if (nrow(annotation_df) > 0) {
    p <- p + geom_label(
      data = annotation_df,
      aes(x = PC1, y = PC2, label = label),
      hjust = 0, vjust = 1, size = 3, fontface = "bold",
      inherit.aes = FALSE, fill = "white", alpha = 0.8
    )
  }

  p
}

(pcoa_plot(pcoa_dfs, "bray") | pcoa_plot(pcoa_dfs, "jaccard")) + plot_annotation(tag_levels = "A")
```

```{r dispersion-tests}
betadisp_results <- purrr::map_dfr(
  dist_methods,
  ~ {
    dist_mat <- phyloseq::distance(ps_rel, method = .x)
    df <- as(sample_data(ps_rel), "data.frame")
    bd <- vegan::betadisper(dist_mat, df$carbohydrate)
    perm <- vegan::permutest(bd, permutations = 999)

    # Extract values more robustly
    perm_df <- as.data.frame(perm$tab)
    f_row <- which(rownames(perm_df) %in% c("F", "Groups"))

    tibble(
      distance = .x,
      stat = perm_df[f_row[1], "F"],
      p_value = perm_df[f_row[1], "Pr(>F)"]
    )
  }
) %>%
  dplyr::mutate(q_value = fdr_adjust(p_value))

betadisp_results %>%
  gt::gt() %>%
  gt::fmt_number(columns = c(stat, p_value, q_value), decimals = 3) %>%
  gt::cols_label(stat = "F", p_value = "p", q_value = "q") %>%
  gt::tab_header("Homogeneity of dispersion (carbohydrate groups)")
```

### Interpretation

```{r beta-interpretation, results='asis'}
permanova_summary <- permanova_results %>%
  dplyr::mutate(q = fdr_adjust(p_value)) %>%
  dplyr::filter(term != "Residual" & q < 0.1) %>%
  dplyr::mutate(item = glue::glue("{stringr::str_to_title(distance)}: {term} (q = {format_p(q)})")) %>%
  dplyr::pull(item)

if (length(permanova_summary) == 0) {
  cat("No beta-diversity terms reach q < 0.10 after FDR adjustment. Patterns are primarily driven by subject-level variability.")
} else {
  cat(glue::glue_collapse(permanova_summary, sep = "\n"))
}
```

### 5.1 Beta Diversity at 48 h

```{r beta-48h}
ps_48h <- phyloseq::subset_samples(ps_rel, Timepoint == "48H")

pcoa_48h <- make_pcoa_df(ps_48h, "bray")

pcoa_48h %>%
  ggplot(aes(x = PC1, y = PC2, color = carbohydrate, shape = group)) +
  geom_point(size = 3, alpha = 0.9) +
  stat_ellipse(aes(group = carbohydrate, color = carbohydrate), linewidth = 0.8, linetype = "dashed", alpha = 0.8) +
  scale_color_manual(values = palette_carbohydrate) +
  scale_shape_manual(values = c(16, 17)) +
  labs(
    x = glue::glue("PC1 ({scales::percent(round(pcoa_48h$axis1_var[1], 3))})"),
    y = glue::glue("PC2 ({scales::percent(round(pcoa_48h$axis2_var[1], 3))})"),
    title = "PCoA – Bray-Curtis distance (48 h samples)"
  ) +
  custom_theme()
```

```{r beta-48h-permanova}
metadata_48h <- as(sample_data(ps_48h), "data.frame")
permanova_48h <- vegan::adonis2(
  phyloseq::distance(ps_48h, method = "bray") ~ group * carbohydrate,
  data = metadata_48h,
  permutations = 999,
  strata = metadata_48h$subject_id
)

permanova_48h_tbl <- as.data.frame(permanova_48h) %>% rownames_to_column('term') %>% janitor::clean_names()

# Find the correct column names for p-value, F-statistic, and R-squared
permanova_48h_p_col <- intersect(c("pr_greater_f_", "pr_greater_f", "pr_f", "p_value"), names(permanova_48h_tbl))
permanova_48h_f_col <- intersect(c("f", "f_model", "pseudo_f"), names(permanova_48h_tbl))
permanova_48h_r2_col <- intersect(c("r2", "r_squared", "r2_adj"), names(permanova_48h_tbl))

permanova_48h_tbl %>%
  dplyr::mutate(p_value = !!sym(permanova_48h_p_col), q = fdr_adjust(p_value)) %>%
  gt::gt() %>%
  gt::fmt_number(columns = c(!!sym(permanova_48h_f_col), !!sym(permanova_48h_r2_col)), decimals = 3) %>%
  gt::fmt_number(columns = c(p_value, q), decimals = 3) %>%
  gt::cols_label(term = "Effect", !!sym(permanova_48h_f_col) := "F", !!sym(permanova_48h_r2_col) := "R2", p_value = "p", q = "q")
```

## 6. Baseline Clustering and Trajectories

```{r cluster-baseline}
ps_0h <- phyloseq::subset_samples(ps_rel, Timepoint == "0H")
stopifnot(nsamples(ps_0h) > 0)

bray_0h <- phyloseq::distance(ps_0h, method = "bray")
dist_matrix <- as.matrix(bray_0h)

silhouette_scores <- map_dbl(2:6, ~ {
  pam_fit <- cluster::pam(bray_0h, k = .x)
  mean(cluster::silhouette(pam_fit)[, "sil_width"])
})

best_k <- which.max(silhouette_scores) + 1
pam_best <- cluster::pam(bray_0h, k = best_k)

cluster_assignments <- tibble(
  SampleID = sample_names(ps_0h),
  baseline_cluster = factor(paste0("Cluster", pam_best$clustering), levels = paste0("Cluster", seq_len(best_k)))
)

cluster_summary <- cluster_assignments %>%
  dplyr::left_join(as_tibble(metadata, .name_repair = "minimal"), by = "SampleID") %>%
  dplyr::count(baseline_cluster, carbohydrate, group, name = "n")

cluster_summary
```

```{r cluster-plot}
cluster_ord <- make_pcoa_df(ps_0h, "bray") %>%
  dplyr::left_join(cluster_assignments, by = "SampleID")

cluster_ord %>%
  ggplot(aes(x = PC1, y = PC2, color = baseline_cluster, shape = carbohydrate)) +
  geom_point(size = 3, alpha = 0.9) +
  stat_ellipse(aes(group = baseline_cluster), linewidth = 0.5) +
  scale_color_manual(values = palette_cluster) +
  labs(
    title = "Baseline (0 h) clusters",
    x = glue::glue("PC1 ({scales::percent(round(cluster_ord$axis1_var[1], 3))})"),
    y = glue::glue("PC2 ({scales::percent(round(cluster_ord$axis2_var[1], 3))})")
  ) +
  custom_theme()
```

```{r cluster-propagation}
metadata_with_cluster <- metadata %>%
  dplyr::left_join(cluster_assignments %>% dplyr::select(SampleID, baseline_cluster), by = "SampleID")

# propagate baseline cluster to all timepoints via subject_id
subject_clusters <- cluster_assignments %>%
  dplyr::left_join(metadata %>% dplyr::select(SampleID, subject_id), by = "SampleID") %>%
  dplyr::group_by(subject_id) %>%
  dplyr::summarise(baseline_cluster = dplyr::first(baseline_cluster), .groups = "drop")

metadata_with_cluster <- metadata %>%
  dplyr::left_join(subject_clusters, by = "subject_id")

ps_with_cluster <- ps_rel
sample_data(ps_with_cluster)$baseline_cluster <- metadata_with_cluster$baseline_cluster[match(sample_names(ps_with_cluster), metadata_with_cluster$SampleID)]
```

```{r cluster-48h}
# Only include subjects that have baseline cluster assignments
subjects_with_clusters <- subject_clusters$subject_id
ps_48h_cluster <- phyloseq::subset_samples(ps_with_cluster, Timepoint == "48H" & subject_id %in% subjects_with_clusters)

# Check how many samples we have after filtering
cat("48H samples with baseline cluster assignments:", nsamples(ps_48h_cluster), "\n")
cat("Total 48H samples:", sum(sample_data(ps_with_cluster)$Timepoint == "48H"), "\n")

# Show which subjects are missing baseline samples
all_48h_subjects <- as(sample_data(ps_with_cluster), "data.frame") %>% 
  dplyr::filter(Timepoint == "48H") %>% 
  dplyr::pull(subject_id) %>% 
  unique()

missing_baseline_subjects <- setdiff(all_48h_subjects, subjects_with_clusters)
if(length(missing_baseline_subjects) > 0) {
  cat("Subjects missing baseline (0H) samples:", paste(missing_baseline_subjects, collapse = ", "), "\n")
  cat("These subjects will be excluded from cluster-based 48H analysis.\n")
} else {
  cat("All 48H subjects have baseline cluster assignments.\n")
}

ord_48h_cluster <- make_pcoa_df(ps_48h_cluster, "bray") %>%
  dplyr::mutate(baseline_cluster = sample_data(ps_48h_cluster)$baseline_cluster[match(SampleID, sample_names(ps_48h_cluster))])

# Create dynamic color palette based on actual clusters present
actual_clusters <- levels(factor(ord_48h_cluster$baseline_cluster))
cluster_palette_dynamic <- setNames(scico::scico(length(actual_clusters), palette = "batlow"), actual_clusters)

ord_48h_cluster %>%
  ggplot(aes(x = PC1, y = PC2, color = baseline_cluster, shape = carbohydrate)) +
  geom_point(size = 3, alpha = 0.9) +
  stat_ellipse(aes(group = interaction(baseline_cluster, carbohydrate)), 
               linewidth = 0.5, linetype = "dotted", 
               level = 0.95, type = "norm") +
  scale_color_manual(values = cluster_palette_dynamic) +
  scale_shape_manual(values = c(16, 17, 18)) +
  labs(
    title = "48 h beta diversity colored by baseline cluster",
    x = glue::glue("PC1 ({scales::percent(round(ord_48h_cluster$axis1_var[1], 3))})"),
    y = glue::glue("PC2 ({scales::percent(round(ord_48h_cluster$axis2_var[1], 3))})")
  ) +
  custom_theme()
```
```{r cluster-48h-facet-plots}
ord_48h_cluster %>% ggplot(aes(PC1, PC2)) +
  geom_point(aes(color = carbohydrate), size = 3) +
  facet_wrap(~baseline_cluster)
```

```{r cluster-statistics}
cluster_response <- ord_48h_cluster %>%
  dplyr::filter(!is.na(baseline_cluster)) %>%
  dplyr::group_by(baseline_cluster, carbohydrate) %>%
  dplyr::summarise(
    mean_PC1 = mean(PC1),
    mean_PC2 = mean(PC2),
    n = n(),
    .groups = "drop"
  )

cluster_response
```

```{r cluster-anova}
ps_48h_counts <- phyloseq::subset_samples(ps, Timepoint == "48H")
alpha_48h_cluster <- estimate_richness(ps_48h_counts, measures = c("Observed", "Shannon", "Simpson")) %>%
  rownames_to_column("SampleID") %>%
  pivot_longer(-SampleID, names_to = "metric", values_to = "value") %>%
  dplyr::left_join(metadata_with_cluster, by = "SampleID") %>%
  dplyr::filter(!is.na(baseline_cluster))

cluster_models <- alpha_48h_cluster %>%
  tidyr::nest(data = -metric) %>%
  dplyr::mutate(
    model = purrr::map(data, fit_cluster_model),
    anova = purrr::map(model, ~ if (is.null(.x)) tibble::tibble() else anova_table(.x))
  )

cluster_models %>%
  dplyr::select(metric, anova) %>%
  tidyr::unnest(anova) %>%
  gt::gt() %>%
  gt::fmt_number(columns = c(f_value), decimals = 3) %>%
  gt::fmt_number(columns = c(p_value, fdr), decimals = 3) %>%
  gt::cols_label(term = "Effect", num_df = "df num", den_df = "df den", f_value = "F", p_value = "p", fdr = "q")
```

```{r cluster-interpretation, results='asis'}
cluster_text <- cluster_models %>%
  dplyr::mutate(summary = purrr::map_chr(
    anova,
    ~ {
      if (is.null(.x) || nrow(.x) == 0) {
        return("Model unavailable for this metric at 48 h")
      }
      .x %>%
        dplyr::filter(fdr < 0.1) %>%
        dplyr::mutate(label = glue::glue("{term} (q = {format_p(fdr)})")) %>%
        dplyr::pull(label) %>%
        glue::glue_collapse(sep = "; ") %>%
        {ifelse(. == "", "No significant cluster–carbohydrate interactions detected at 48 h", .)}
    }
  ))

cluster_text %>%
  dplyr::mutate(text = glue::glue("**{metric}**: {summary}")) %>%
  dplyr::pull(text) %>%
  glue::glue_collapse(sep = "\n") %>%
  cat()
```

## 7. Community Composition Profiles

```{r composition-overview}
ps_genus <- collapse_taxa(ps_with_cluster, rank = "Genus", top_n = 20)

ggplot(psmelt(microbiome::transform(ps_genus, "compositional")),
       aes(x = carbohydrate, y = Abundance, fill = Genus)) +
  geom_col(position = "fill") +
  facet_grid(group ~ timepoint) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = scico::scico(21, palette = "batlow")) +
  labs(
    x = NULL,
    y = "Relative abundance",
    title = "Genus-level composition by group, carbohydrate, and timepoint"
  ) +
  custom_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r composition-48h}
ps_48h_genus <- phyloseq::subset_samples(ps_genus, Timepoint == "48H")

ggplot(psmelt(microbiome::transform(ps_48h_genus, "compositional")),
       aes(x = carbohydrate, y = Abundance, fill = Genus)) +
  geom_col(position = "fill") +
  facet_wrap(~ group) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = scico::scico(21, palette = "batlow")) +
  labs(
    x = NULL,
    y = "Relative abundance",
    title = "Genus-level composition at 48 h"
  ) +
  custom_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r composition-cluster}
ps_cluster <- phyloseq::subset_samples(ps_genus, !is.na(sample_data(ps_genus)$baseline_cluster))

ggplot(psmelt(microbiome::transform(ps_cluster, "compositional")),
       aes(x = baseline_cluster, y = Abundance, fill = Genus)) +
  geom_col(position = "fill") +
  facet_wrap(~ carbohydrate + timepoint, ncol = 3) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = scico::scico(21, palette = "batlow")) +
  labs(
    x = "Baseline cluster",
    y = "Relative abundance",
    title = "Genus-level composition by baseline cluster"
  ) +
  custom_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Alluvial flow diagrams: Genus abundance changes from 0H to 48H

```{r alluvial-diagram, fig.width=14, fig.height=10}
# Install ggalluvial if not available
if (!requireNamespace("ggalluvial", quietly = TRUE)) {
  install.packages("ggalluvial")
}
library(ggalluvial)

# Prepare data for alluvial diagrams
ps_genus_rel <- transform_sample_counts(ps_genus, function(x) x / sum(x))

# Extract data
otu_mat <- as.matrix(otu_table(ps_genus_rel))
if (!taxa_are_rows(ps_genus_rel)) {
  otu_mat <- t(otu_mat)
}

tax_df <- as.data.frame(tax_table(ps_genus_rel))
meta_df <- data.frame(sample_data(ps_genus_rel))

# Create long format data
alluvial_data <- otu_mat %>%
  as.data.frame() %>%
  tibble::rownames_to_column("ASV") %>%
  tidyr::pivot_longer(-ASV, names_to = "SampleID", values_to = "Abundance") %>%
  dplyr::left_join(
    tax_df %>% tibble::rownames_to_column("ASV") %>% dplyr::select(ASV, Genus),
    by = "ASV"
  ) %>%
  dplyr::left_join(
    meta_df %>% tibble::rownames_to_column("SampleID") %>%
      dplyr::select(SampleID, Timepoint, Carbohydrate),
    by = "SampleID"
  ) %>%
  dplyr::group_by(Genus, Timepoint, Carbohydrate) %>%
  dplyr::summarise(mean_abundance = mean(Abundance), .groups = "drop") %>%
  dplyr::filter(mean_abundance > 0.01)  # Keep genera >1%

# Get top genera
top_genera <- alluvial_data %>%
  dplyr::group_by(Genus) %>%
  dplyr::summarise(max_abund = max(mean_abundance)) %>%
  dplyr::arrange(desc(max_abund)) %>%
  head(12) %>%
  pull(Genus)

# Filter to top genera and add "Other" category
alluvial_plot_data <- alluvial_data %>%
  dplyr::mutate(
    Genus_grouped = ifelse(Genus %in% top_genera, as.character(Genus), "Other")
  ) %>%
  dplyr::group_by(Genus_grouped, Timepoint, Carbohydrate) %>%
  dplyr::summarise(Abundance = sum(mean_abundance), .groups = "drop")

# Create color palette
genus_colors <- c(
  scico::scico(length(top_genera), palette = "batlow"),
  "Other" = "grey80"
)
names(genus_colors)[1:length(top_genera)] <- top_genera

# Create alluvial plots for each carbohydrate
plot_list <- list()

for (carb in c("No Added Carb", "RDC", "SDC")) {
  plot_data <- alluvial_plot_data %>%
    dplyr::filter(Carbohydrate == carb) %>%
    dplyr::mutate(
      Genus_grouped = factor(Genus_grouped,
                            levels = c(top_genera[top_genera %in% unique(Genus_grouped)], "Other")),
      # Create label columns that show only on respective sides
      label_left = ifelse(Timepoint == "0H", as.character(Genus_grouped), ""),
      label_right = ifelse(Timepoint == "48H", as.character(Genus_grouped), "")
    )

  p <- ggplot(plot_data,
              aes(x = Timepoint, y = Abundance,
                  alluvium = Genus_grouped, stratum = Genus_grouped, fill = Genus_grouped)) +
    geom_alluvium(alpha = 0.7, width = 0.3) +
    geom_stratum(width = 0.3, alpha = 0.9) +
    # Add labels to the left side (0H only)
    stat_stratum(geom = "text", aes(label = label_left),
                 size = 2.3, hjust = 1.3, fontface = "bold") +
    # Add labels to the right side (48H only)
    stat_stratum(geom = "text", aes(label = label_right),
                 size = 2.3, hjust = -0.5, fontface = "bold") +
    scale_fill_manual(values = genus_colors) +
    scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
    scale_x_discrete(expand = expansion(add = c(1.2, 1.2))) +
    labs(
      title = carb,
      x = "Timepoint",
      y = "Relative Abundance",
      fill = "Genus"
    ) +
    custom_theme() +
    theme(
      legend.position = "none",
      panel.grid.major.x = element_blank()
    )

  plot_list[[carb]] <- p
}

# Combine plots
library(patchwork)
combined_plot <- wrap_plots(plot_list, ncol = 3) +
  plot_annotation(
    title = "Genus Composition Changes from 0H to 48H by Carbohydrate Type",
    subtitle = "Flow diagram showing relative abundance changes (top 12 genera)",
    theme = theme(plot.title = element_text(size = 16, face = "bold"))
  )

print(combined_plot)
```

### Validation: Bar plots of genus abundances by timepoint and carbohydrate

```{r chord-validation, fig.width=14, fig.height=10}
# Extract genus-level data for validation
ps_genus_rel <- transform_sample_counts(ps_genus, function(x) x / sum(x))

# Get OTU table and metadata
otu_mat <- as.matrix(otu_table(ps_genus_rel))
if (!taxa_are_rows(ps_genus_rel)) {
  otu_mat <- t(otu_mat)
}
otu_df <- as.data.frame(t(otu_mat))

tax_df <- as.data.frame(tax_table(ps_genus_rel))
meta_df <- data.frame(sample_data(ps_genus_rel))

# Combine data
genus_long <- otu_df %>%
  tibble::rownames_to_column("SampleID") %>%
  tidyr::pivot_longer(-SampleID, names_to = "ASV", values_to = "Abundance") %>%
  dplyr::left_join(
    tax_df %>% tibble::rownames_to_column("ASV") %>% dplyr::select(ASV, Genus),
    by = "ASV"
  ) %>%
  dplyr::left_join(
    meta_df %>% tibble::rownames_to_column("SampleID") %>%
      dplyr::select(SampleID, Timepoint, Carbohydrate),
    by = "SampleID"
  )

# Calculate mean abundance per genus, timepoint, and carbohydrate
genus_summary <- genus_long %>%
  group_by(Genus, Timepoint, Carbohydrate) %>%
  summarise(
    mean_abundance = mean(Abundance),
    se_abundance = sd(Abundance) / sqrt(n()),
    .groups = "drop"
  ) %>%
  filter(mean_abundance > 0.01)  # Only keep genera >1% mean abundance

# Get top genera overall
top_genera <- genus_summary %>%
  group_by(Genus) %>%
  summarise(max_abundance = max(mean_abundance), .groups = "drop") %>%
  arrange(desc(max_abundance)) %>%
  head(15) %>%
  pull(Genus)

# Filter to top genera
genus_summary_top <- genus_summary %>%
  filter(Genus %in% top_genera)

# Create faceted bar plot
p_validation <- ggplot(genus_summary_top,
                       aes(x = Timepoint, y = mean_abundance, fill = Timepoint)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_errorbar(aes(ymin = mean_abundance - se_abundance,
                    ymax = mean_abundance + se_abundance),
                width = 0.2, position = position_dodge(0.7)) +
  facet_grid(Genus ~ Carbohydrate, scales = "free_y") +
  scale_fill_manual(values = c("0H" = "#E69F00", "48H" = "#56B4E9")) +
  labs(
    title = "Validation: Genus Abundance by Timepoint and Carbohydrate Type",
    subtitle = "Mean relative abundance ± SE for top 15 genera (>1% abundance)",
    x = "Timepoint",
    y = "Mean Relative Abundance",
    fill = "Timepoint"
  ) +
  custom_theme() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    strip.text.y = element_text(angle = 0, hjust = 0),
    panel.spacing = unit(0.5, "lines")
  )

print(p_validation)

# Print summary statistics for key changes
cat("\n### Summary of abundance changes from 0H to 48H:\n\n")

change_summary <- genus_summary_top %>%
  dplyr::select(Genus, Timepoint, Carbohydrate, mean_abundance) %>%
  pivot_wider(names_from = Timepoint, values_from = mean_abundance,
              names_prefix = "T") %>%
  mutate(
    change = T48H - T0H,
    pct_change = ((T48H - T0H) / T0H) * 100,
    direction = case_when(
      abs(change) < 0.005 ~ "No change",
      change > 0 ~ "Increase",
      change < 0 ~ "Decrease"
    )
  ) %>%
  arrange(Carbohydrate, desc(abs(change)))

# Print changes for each carbohydrate type
for (carb in c("No Added Carb", "RDC", "SDC")) {
  cat(paste0("\n", carb, ":\n"))
  carb_changes <- change_summary %>%
    filter(Carbohydrate == carb, abs(change) > 0.01) %>%
    arrange(desc(abs(change))) %>%
    head(10)

  if (nrow(carb_changes) > 0) {
    print(carb_changes %>%
            dplyr::select(Genus, T0H, T48H, change, pct_change, direction) %>%
            mutate(across(where(is.numeric), ~round(., 4))))
  } else {
    cat("  No major changes detected (>1% absolute change)\n")
  }
}
```

## 8. Differential Abundance and Predictive Modeling

### 8.1 Shared Setup

```{r da-setup}
ps_counts <- ps
sample_data(ps_counts)$subject_id <- factor(sample_data(ps_counts)$subject_id)

ps_genus_counts <- tax_glom(ps_counts, taxrank = "Genus", NArm = FALSE)
ps_genus_counts <- prune_taxa(taxa_sums(ps_genus_counts) > 0, ps_genus_counts)

metadata_counts <- as(sample_data(ps_genus_counts), "data.frame") %>%
  as_tibble(rownames = "SampleID")

prevalence_cutoff <- 0.005
min_prev <- ceiling(prevalence_cutoff * nsamples(ps_genus_counts))
keep_taxa <- taxa_sums(ps_genus_counts) > 0 & rowSums(otu_table(ps_genus_counts) > 0) >= min_prev
ps_genus_filtered <- prune_taxa(keep_taxa, ps_genus_counts)

ps_genus_48h <- subset_samples(ps_genus_filtered, Timepoint == "48H")
ps_genus_0h <- subset_samples(ps_genus_filtered, Timepoint == "0H")
```

### 8.2 Corncob (beta-binomial)

```{r corncob, fig.width = 20, fig.height = 10}
corncob_path <- file.path(outputs_dir, "corncob_genus_results.rds")

corncob_test <- cache_rds(
  corncob_path,
  {
    set.seed(12345)
    tryCatch({
      corncob::differentialTest(
        formula = ~ group + carbohydrate * timepoint + subject_id,
        phi.formula = ~ group + carbohydrate * timepoint + subject_id,
        formula_null = ~ group + timepoint + subject_id,
        phi.formula_null = ~ group + timepoint + subject_id,
        test = "Wald",
        data = ps_genus_filtered,
        fdr_cutoff = 0.1,
        B = 50
      )
    }, error = function(e) {
      message("Corncob failed: ", e$message)
      NULL
    })
  }
)

if (!is.null(corncob_test)) {
  if (!is.null(corncob_test$wald_results)) {
    # Get taxonomy information for genus-species conversion
    tax_table_df <- as.data.frame(tax_table(ps_genus_filtered))
    tax_table_df$feature <- rownames(tax_table_df)
    
    # Create taxonomic display names
    tax_table_df <- tax_table_df %>%
      dplyr::mutate(
        # Create genus_species display name
        genus_species = case_when(
          !is.na(Species) & Species != "Unassigned" & Species != "" ~ paste0(Genus, " ", Species),
          !is.na(Genus) & Genus != "Unassigned" & Genus != "" ~ paste0(Genus, " sp."),
          !is.na(Family) & Family != "Unassigned" & Family != "" ~ paste0(Family, " (family)"),
          !is.na(Order) & Order != "Unassigned" & Order != "" ~ paste0(Order, " (order)"),
          TRUE ~ feature  # Fallback to ASV ID if no good taxonomy
        )
      )
    
    corncob_results <- corncob_test$wald_results %>%
      dplyr::mutate(q_value = fdr_adjust(p_val)) %>%
      dplyr::left_join(tax_table_df %>% dplyr::select(feature, genus_species), by = c("taxon" = "feature")) %>%
      dplyr::mutate(
        # Use genus_species if available, otherwise fall back to taxon
        taxon_display = ifelse(!is.na(genus_species), genus_species, taxon)
      ) %>%
      dplyr::select(taxon, taxon_display, p_val, q_value) %>%
      dplyr::arrange(q_value)
  } else if (!is.null(corncob_test$p) && !is.null(corncob_test$p_fdr)) {
    p_mat <- corncob_test$p
    q_mat <- corncob_test$p_fdr

    if (is.null(dim(p_mat))) {
      p_vals <- as.numeric(p_mat)
    } else {
      p_vals <- as.numeric(p_mat[, 1])
    }

    if (is.null(dim(q_mat))) {
      q_vals = as.numeric(q_mat)
    } else {
      q_vals = as.numeric(q_mat[, 1])
    }

    taxa_names <- rownames(p_mat)
    if (is.null(taxa_names)) {
      taxa_names <- names(p_mat)
    }
    if (is.null(taxa_names)) {
      taxa_names <- paste0("taxon_", seq_along(p_vals))
    }

    # Get taxonomy information for genus-species conversion
    tax_table_df <- as.data.frame(tax_table(ps_genus_filtered))
    tax_table_df$feature <- rownames(tax_table_df)
    
    # Create taxonomic display names
    tax_table_df <- tax_table_df %>%
      dplyr::mutate(
        # Create genus_species display name
        genus_species = case_when(
          !is.na(Species) & Species != "Unassigned" & Species != "" ~ paste0(Genus, " ", Species),
          !is.na(Genus) & Genus != "Unassigned" & Genus != "" ~ paste0(Genus, " sp."),
          !is.na(Family) & Family != "Unassigned" & Family != "" ~ paste0(Family, " (family)"),
          !is.na(Order) & Order != "Unassigned" & Order != "" ~ paste0(Order, " (order)"),
          TRUE ~ feature  # Fallback to ASV ID if no good taxonomy
        )
      )
    
    corncob_results <- tibble::tibble(
      taxon = taxa_names,
      p_val = p_vals,
      q_value = q_vals
    ) %>%
      dplyr::left_join(tax_table_df %>% dplyr::select(feature, genus_species), by = c("taxon" = "feature")) %>%
      dplyr::mutate(
        # Use genus_species if available, otherwise fall back to taxon
        taxon_display = ifelse(!is.na(genus_species), genus_species, taxon)
      ) %>%
      dplyr::select(taxon, taxon_display, p_val, q_value) %>%
      dplyr::arrange(q_value)
  } else {
    corncob_results <- tibble()
  }
} else {
  corncob_results <- tibble()
}

corncob_results

plot(corncob_test)
```

### 8.2.1 Corncob (48H, RDC vs SDC only)

```{r corncob-48h-rdc-sdc}
# Filter data for 48H timepoint and only RDC and SDC groups
ps_genus_48h_rdc_sdc <- subset_samples(ps_genus_filtered, 
                                       Timepoint == "48H" & 
                                       carbohydrate %in% c("RDC", "SDC"))

# Check sample sizes
cat("48H RDC/SDC samples:", nsamples(ps_genus_48h_rdc_sdc), "\n")
cat("RDC samples:", sum(sample_data(ps_genus_48h_rdc_sdc)$carbohydrate == "RDC"), "\n")
cat("SDC samples:", sum(sample_data(ps_genus_48h_rdc_sdc)$carbohydrate == "SDC"), "\n")

corncob_48h_path <- file.path(outputs_dir, "corncob_genus_48h_rdc_sdc_results.rds")

corncob_48h_test <- cache_rds(
  corncob_48h_path,
  {
    set.seed(12345)
    tryCatch({
      corncob::differentialTest(
        formula = ~ carbohydrate + subject_id,
        phi.formula = ~ carbohydrate + subject_id,
        formula_null = ~ subject_id,
        phi.formula_null = ~ subject_id,
        test = "Wald",
        data = ps_genus_48h_rdc_sdc,
        fdr_cutoff = 0.1,
        B = 50
      )
    }, error = function(e) {
      message("Corncob 48h RDC vs SDC failed: ", e$message)
      NULL
    })
  }
)

if (!is.null(corncob_48h_test)) {
  if (!is.null(corncob_48h_test$wald_results)) {
    # Get taxonomy information for genus-species conversion
    tax_table_df_48h <- as.data.frame(tax_table(ps_genus_48h_rdc_sdc))
    tax_table_df_48h$feature <- rownames(tax_table_df_48h)
    
    # Create taxonomic display names
    tax_table_df_48h <- tax_table_df_48h %>%
      dplyr::mutate(
        # Create genus_species display name
        genus_species = case_when(
          !is.na(Species) & Species != "Unassigned" & Species != "" ~ paste0(Genus, " ", Species),
          !is.na(Genus) & Genus != "Unassigned" & Genus != "" ~ paste0(Genus, " sp."),
          !is.na(Family) & Family != "Unassigned" & Family != "" ~ paste0(Family, " (family)"),
          !is.na(Order) & Order != "Unassigned" & Order != "" ~ paste0(Order, " (order)"),
          TRUE ~ feature  # Fallback to ASV ID if no good taxonomy
        )
      )
    
    corncob_48h_results <- corncob_48h_test$wald_results %>%
      dplyr::mutate(q_value = fdr_adjust(p_val)) %>%
      dplyr::left_join(tax_table_df_48h %>% dplyr::select(feature, genus_species), by = c("taxon" = "feature")) %>%
      dplyr::mutate(
        # Use genus_species if available, otherwise fall back to taxon
        taxon_display = ifelse(!is.na(genus_species), genus_species, taxon)
      ) %>%
      dplyr::select(taxon, taxon_display, p_val, q_value) %>%
      dplyr::arrange(q_value)
  } else if (!is.null(corncob_48h_test$p) && !is.null(corncob_48h_test$p_fdr)) {
    p_mat_48h <- corncob_48h_test$p
    q_mat_48h <- corncob_48h_test$p_fdr

    if (is.null(dim(p_mat_48h))) {
      p_vals_48h <- as.numeric(p_mat_48h)
    } else {
      p_vals_48h <- as.numeric(p_mat_48h[, 1])
    }

    if (is.null(dim(q_mat_48h))) {
      q_vals_48h = as.numeric(q_mat_48h)
    } else {
      q_vals_48h = as.numeric(q_mat_48h[, 1])
    }

    taxa_names_48h <- rownames(p_mat_48h)
    if (is.null(taxa_names_48h)) {
      taxa_names_48h <- names(p_mat_48h)
    }
    if (is.null(taxa_names_48h)) {
      taxa_names_48h <- paste0("taxon_", seq_along(p_vals_48h))
    }

    # Get taxonomy information for genus-species conversion
    tax_table_df_48h <- as.data.frame(tax_table(ps_genus_48h_rdc_sdc))
    tax_table_df_48h$feature <- rownames(tax_table_df_48h)
    
    # Create taxonomic display names
    tax_table_df_48h <- tax_table_df_48h %>%
      dplyr::mutate(
        # Create genus_species display name
        genus_species = case_when(
          !is.na(Species) & Species != "Unassigned" & Species != "" ~ paste0(Genus, " ", Species),
          !is.na(Genus) & Genus != "Unassigned" & Genus != "" ~ paste0(Genus, " sp."),
          !is.na(Family) & Family != "Unassigned" & Family != "" ~ paste0(Family, " (family)"),
          !is.na(Order) & Order != "Unassigned" & Order != "" ~ paste0(Order, " (order)"),
          TRUE ~ feature  # Fallback to ASV ID if no good taxonomy
        )
      )
    
    corncob_48h_results <- tibble::tibble(
      taxon = taxa_names_48h,
      p_val = p_vals_48h,
      q_value = q_vals_48h
    ) %>%
      dplyr::left_join(tax_table_df_48h %>% dplyr::select(feature, genus_species), by = c("taxon" = "feature")) %>%
      dplyr::mutate(
        # Use genus_species if available, otherwise fall back to taxon
        taxon_display = ifelse(!is.na(genus_species), genus_species, taxon)
      ) %>%
      dplyr::select(taxon, taxon_display, p_val, q_value) %>%
      dplyr::arrange(q_value)
  } else {
    corncob_48h_results <- tibble()
  }
} else {
  corncob_48h_results <- tibble()
}

# Display results
cat("Corncob 48h RDC vs SDC Results:\n")
print(corncob_48h_results)

# Show significant results
if (nrow(corncob_48h_results) > 0) {
  sig_results_48h <- corncob_48h_results %>% dplyr::filter(q_value < 0.1)
  cat("\nSignificant results (q < 0.1):\n")
  print(sig_results_48h)
  
  if (nrow(sig_results_48h) > 0) {
    cat("\nNumber of significant taxa:", nrow(sig_results_48h), "\n")
  } else {
    cat("\nNo significant taxa found at q < 0.1\n")
  }
} else {
  cat("No results available\n")
}

# Plot results if available
if (!is.null(corncob_48h_test)) {
  plot(corncob_48h_test)
}
```

### 8.3 ANCOM-BC2

```{r ancombc2}
ancombc2_path <- file.path(outputs_dir, "ancombc2_genus_results.rds")

# Function to run ANCOMBC2 with multiple fallback strategies
run_ancombc2_safe <- function(ps_data, tax_level = "Genus") {
  
  # Strategy 1: Try with reduced complexity and conservative settings
  tryCatch({
    message("Attempting ANCOMBC2 with conservative settings...")
    ANCOMBC::ancombc2(
      data = ps_data,
      tax_level = tax_level,
      fix_formula = "group + carbohydrate * timepoint",
      rand_formula = "(1|subject_id)",
      p_adj_method = "BH",
      prv_cut = 0.15,  # Increased from 0.10 to reduce data size
      lib_cut = 0,
      s0_perc = 0.1,   # Increased from 0.05 for stability
      group = "carbohydrate",
      struc_zero = TRUE,  # Disabled to reduce complexity
      neg_lb = TRUE,      # Disabled to reduce complexity
      alpha = 0.05,
      global = TRUE,      # Disabled to reduce computation
      verbose = TRUE
    )
  }, error = function(e1) {
    message("Strategy 1 failed: ", e1$message)
    
    # Strategy 2: Try with simplified formula (no interaction)
    tryCatch({
      # Remove samples with too many zeros that cause estimation issues
      # Use a more lenient threshold to preserve more data
      sample_sums <- sample_sums(ps_counts)
      zero_prop <- apply(otu_table(ps_counts), 2, function(x) sum(x == 0) / length(x))
      problematic_samples <- names(zero_prop[zero_prop > 0.95])  # More lenient: 95% zeros
      
      if (length(problematic_samples) > 0) {
        message("Removing samples with >95% zeros: ", paste(problematic_samples, collapse = ", "))
        ps_counts_clean <- subset_samples(ps_counts, !sample_names(ps_counts) %in% problematic_samples)
      } else {
        ps_counts_clean <- ps_counts
      }
      
      # Check if we have enough samples left
      if (nsamples(ps_counts_clean) < 10) {
        stop("Insufficient samples remaining after filtering (", nsamples(ps_counts_clean), " samples)")
      }
      
      ANCOMBC::ancombc2(
        data = ps_counts_clean,
        tax_level = "Genus",
        fix_formula = "group + carbohydrate * timepoint",
        rand_formula = "(1|subject_id)",
        p_adj_method = "BH",
        prv_cut = 0.20,  # Further increased
        lib_cut = 0,
        s0_perc = 0.1,
        group = "carbohydrate",
        struc_zero = TRUE,
        neg_lb = TRUE,
        alpha = 0.05,
        global = TRUE,
        verbose = TRUE
      )
    }, error = function(e2) {
      message("Strategy 2 failed: ", e2$message)
      
      # Strategy 3: Try without random effects
      tryCatch({
        message("Attempting ANCOMBC2 without random effects...")
        ANCOMBC::ancombc2(
          data = ps_data,
          tax_level = tax_level,
          fix_formula = "group + carbohydrate + timepoint",
          rand_formula = NULL,  # No random effects
          p_adj_method = "BH",
          prv_cut = 0.25,
          lib_cut = 0,
          s0_perc = 0.1,
          group = "carbohydrate",
          struc_zero = TRUE,
          neg_lb = TRUE,
          alpha = 0.05,
          global = TRUE,
          verbose = TRUE
        )
      }, error = function(e3) {
        message("Strategy 3 failed: ", e3$message)
        
        # Strategy 4: Try with even more conservative settings
        tryCatch({
          message("Attempting ANCOMBC2 with minimal settings...")
          ANCOMBC::ancombc2(
            data = ps_data,
            tax_level = tax_level,
            fix_formula = "carbohydrate",  # Only main effect
            rand_formula = NULL,
            p_adj_method = "BH",
            prv_cut = 0.30,
            lib_cut = 0,
            s0_perc = 0.1,
            group = "carbohydrate",
            struc_zero = TRUE,
            neg_lb = TRUE,
            alpha = 0.05,
            global = TRUE,
            verbose = TRUE
          )
        }, error = function(e4) {
          message("All ANCOMBC2 strategies failed. Last error: ", e4$message)
          return(NULL)
        })
      })
    })
  })
}

ancombc2_res <- cache_rds(
  ancombc2_path,
  {
    run_ancombc2_safe(ps_counts, "Genus")
  }
)

if (!is.null(ancombc2_res)) {
  # Extract results from the res component
  res_data <- ancombc2_res$res
  
  # Get the effect names (excluding Intercept for main effects)
  effect_names <- c("groupObese", "carbohydrateRDC", "carbohydrateSDC", "timepoint48H", 
                    "carbohydrateRDC:timepoint48H", "carbohydrateSDC:timepoint48H")
  
  # Create a data frame from the results for each effect
  ancombc2_tbl <- purrr::map_dfr(effect_names, function(effect) {
    # Check if columns exist for this effect
    lfc_col <- paste0("lfc_", effect)
    se_col <- paste0("se_", effect)
    W_col <- paste0("W_", effect)
    p_col <- paste0("p_", effect)
    q_col <- paste0("q_", effect)
    diff_col <- paste0("diff_", effect)
    
    if (all(c(lfc_col, se_col, W_col, p_col, q_col, diff_col) %in% colnames(res_data))) {
      tibble(
        taxon = res_data$taxon,
        effect = effect,
        beta = res_data[[lfc_col]],
        se = res_data[[se_col]],
        W = res_data[[W_col]],
        p_value = res_data[[p_col]],
        q_value = res_data[[q_col]],
        diff_abn = res_data[[diff_col]]
      )
    } else {
      # Return empty tibble if columns don't exist
      tibble()
    }
  }) %>% dplyr::arrange(q_value)
} else {
  ancombc2_tbl <- tibble()
}

if (nrow(ancombc2_tbl) > 0 && "q_value" %in% names(ancombc2_tbl)) {
  ancombc2_tbl %>%
    dplyr::filter(q_value < 0.1)
} else {
  cat("No ANCOM-BC2 results available - analysis may have failed\n")
  tibble()
}
```

#### Forest Plot of ANCOM-BC2 Results

```{r ancombc2_forest_plot, fig.width = 20, fig.height = 10}
# Prepare data for forest plot
if (nrow(ancombc2_tbl) > 0 && "q_value" %in% names(ancombc2_tbl)) {
  forest_data <- ancombc2_tbl %>%
    dplyr::filter(q_value < 0.1) %>%
  dplyr::mutate(
    # Calculate confidence intervals
    ci_lower = beta - 1.96 * se,
    ci_upper = beta + 1.96 * se,
    # Create effect labels
    effect_label = case_when(
      effect == "groupObese" ~ "Obese vs Normal",
      effect == "carbohydrateRDC" ~ "RDC vs No Added Carb",
      effect == "carbohydrateSDC" ~ "SDC vs No Added Carb", 
      effect == "timepoint48H" ~ "48H vs Baseline",
      effect == "carbohydrateRDC:timepoint48H" ~ "RDC × 48H Interaction",
      effect == "carbohydrateSDC:timepoint48H" ~ "SDC × 48H Interaction",
      TRUE ~ effect
    ),
    # Create taxon labels with significance
    taxon_label = paste0(taxon, " (q=", round(q_value, 3), ")"),
    # Color by significance level
    significance = case_when(
      q_value < 0.001 ~ "q < 0.001",
      q_value < 0.01 ~ "q < 0.01", 
      q_value < 0.05 ~ "q < 0.05",
      q_value < 0.1 ~ "q < 0.1",
      TRUE ~ "q ≥ 0.1"
    )
  ) %>%
  # Order by effect size (absolute value)
  dplyr::arrange(desc(abs(beta)))

# Create forest plot
forest_plot <- ggplot(forest_data, aes(x = beta, y = reorder(taxon_label, beta))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(aes(color = significance), size = 2) +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper, color = significance), 
                 height = 0.2) +
  facet_wrap(~ effect_label, scales = "free_x", ncol = 2) +
  scale_color_manual(
    values = c("q < 0.001" = "#d73027", 
               "q < 0.01" = "#f46d43",
               "q < 0.05" = "#fdae61", 
               "q < 0.1" = "#fee08b"),
    name = "FDR-adjusted\np-value"
  ) +
  labs(
    title = "ANCOM-BC2 Results: Forest Plot of Significant Taxa",
    subtitle = "Effect sizes (log-fold changes) with 95% confidence intervals",
    x = "Log-fold change (β)",
    y = "Taxon"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 9),
    strip.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

  print(forest_plot)

  # Summary statistics
  cat("Summary of ANCOM-BC2 Results:\n")
  cat("Total significant taxa (q < 0.1):", nrow(forest_data), "\n")
  cat("By effect:\n")
  print(table(forest_data$effect_label))
} else {
  cat("No ANCOM-BC2 results available for forest plot - analysis may have failed\n")
}
```

#### Side-by-Side Comparison: RDC vs No Carb and SDC vs No Carb

```{r ancombc2_carb_comparison, fig.width = 20, fig.height = 10}
# First, create comparison table to identify taxa with opposite effects
if (nrow(ancombc2_tbl) > 0 && "effect" %in% names(ancombc2_tbl)) {
  comparison_table <- ancombc2_tbl %>%
    dplyr::filter(effect %in% c("carbohydrateRDC", "carbohydrateSDC")) %>%
  dplyr::select(taxon, effect, beta, se, p_value, q_value, diff_abn) %>%
  tidyr::pivot_wider(
    names_from = effect,
    values_from = c(beta, se, p_value, q_value, diff_abn),
    names_sep = "_"
  ) %>%
  dplyr::mutate(
    # Calculate difference between RDC and SDC effects
    effect_difference = beta_carbohydrateRDC - beta_carbohydrateSDC,
    # Determine which treatment has stronger effect
    stronger_effect = case_when(
      abs(beta_carbohydrateRDC) > abs(beta_carbohydrateSDC) ~ "RDC",
      abs(beta_carbohydrateSDC) > abs(beta_carbohydrateRDC) ~ "SDC",
      TRUE ~ "Equal"
    ),
    # Check if effects are opposite (different signs)
    opposite_effects = sign(beta_carbohydrateRDC) != sign(beta_carbohydrateSDC),
    # Check if either effect is significant
    either_significant = (q_value_carbohydrateRDC < 0.1) | (q_value_carbohydrateSDC < 0.1)
  )

# Filter taxa: significant (q <= 0.1) OR opposite effects
filtered_taxa <- comparison_table %>%
  dplyr::filter(either_significant | opposite_effects) %>%
  dplyr::pull(taxon)

# Filter for carbohydrate main effects only, keeping only filtered taxa
carb_data <- ancombc2_tbl %>%
  dplyr::filter(effect %in% c("carbohydrateRDC", "carbohydrateSDC") & 
                taxon %in% filtered_taxa) %>%
  dplyr::mutate(
    # Calculate confidence intervals
    ci_lower = beta - 1.96 * se,
    ci_upper = beta + 1.96 * se,
    # Create treatment labels
    treatment = case_when(
      effect == "carbohydrateRDC" ~ "RDC vs No Added Carb",
      effect == "carbohydrateSDC" ~ "SDC vs No Added Carb",
      TRUE ~ effect
    ),
    # Create taxon labels with significance
    taxon_label = paste0(taxon, " (q=", round(q_value, 3), ")"),
    # Color by significance level
    significance = case_when(
      q_value < 0.001 ~ "q < 0.001",
      q_value < 0.01 ~ "q < 0.01", 
      q_value < 0.05 ~ "q < 0.05",
      q_value < 0.1 ~ "q < 0.1",
      TRUE ~ "q ≥ 0.1"
    )
  ) %>%
  # Order by effect size (absolute value) for RDC first, then SDC
  dplyr::arrange(treatment, desc(abs(beta)))

# Create side-by-side forest plot with colored error bars
carb_plot <- ggplot(carb_data, aes(x = beta, y = reorder(taxon_label, beta))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper, color = significance), 
                 height = 0.15, size = 1) +
  geom_point(aes(color = significance), size = 3, fill = "white", shape = 21, stroke = 1.5) +
  facet_wrap(~ treatment, scales = "free_x", ncol = 2) +
  scale_color_manual(
    values = c("q < 0.001" = "#d73027", 
               "q < 0.01" = "#f46d43",
               "q < 0.05" = "#fdae61", 
               "q < 0.1" = "#fee08b",
               "q ≥ 0.1" = "#d9d9d9"),
    name = "FDR-adjusted\np-value"
  ) +
  labs(
    title = "Carbohydrate Effects: RDC vs SDC Comparison",
    subtitle = "Effect sizes (log-fold changes) with 95% confidence intervals\n(Showing significant taxa q ≤ 0.1 or taxa with opposite effects)",
    x = "Log-fold change (β)",
    y = "Taxon"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 9),
    axis.text.x = element_text(size = 10),
    strip.text = element_text(size = 11, face = "bold"),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "lightgray", color = "black")
  )

print(carb_plot)

# Filter comparison table to show only taxa in the plot
filtered_comparison <- comparison_table %>%
  dplyr::filter(taxon %in% filtered_taxa) %>%
  dplyr::arrange(desc(abs(effect_difference)))

cat("\nComparison Table: RDC vs SDC Effects (Filtered)\n")
cat("================================================\n")
print(filtered_comparison %>% 
  dplyr::select(taxon, 
                beta_carbohydrateRDC, q_value_carbohydrateRDC,
                beta_carbohydrateSDC, q_value_carbohydrateSDC,
                effect_difference, stronger_effect, opposite_effects) %>%
  dplyr::slice_head(n = 20))

# Summary statistics
cat("\nSummary of Carbohydrate Effects (Filtered Data):\n")
cat("Total taxa shown:", length(filtered_taxa), "\n")
cat("RDC significant taxa (q < 0.1):", sum(carb_data$effect == "carbohydrateRDC" & carb_data$q_value < 0.1), "\n")
cat("SDC significant taxa (q < 0.1):", sum(carb_data$effect == "carbohydrateSDC" & carb_data$q_value < 0.1), "\n")
cat("Taxa with opposite effects:", sum(filtered_comparison$opposite_effects, na.rm = TRUE), "\n")
cat("Taxa with stronger RDC effect:", sum(filtered_comparison$stronger_effect == "RDC", na.rm = TRUE), "\n")
cat("Taxa with stronger SDC effect:", sum(filtered_comparison$stronger_effect == "SDC", na.rm = TRUE), "\n")

# Show taxa with opposite effects
opposite_taxa <- filtered_comparison %>%
  dplyr::filter(opposite_effects) %>%
  dplyr::select(taxon, beta_carbohydrateRDC, beta_carbohydrateSDC, effect_difference)

  if(nrow(opposite_taxa) > 0) {
    cat("\nTaxa with Opposite Effects (RDC vs SDC):\n")
    cat("========================================\n")
    print(opposite_taxa)
  }
} else {
  cat("No ANCOM-BC2 comparison data available - analysis may have failed\n")
}
```


### 8.4 ANCOM-BC (48 h, RDC reference)

```{r ancombc48}
ps_48h_counts <- subset_samples(ps_counts, Timepoint == "48H")

ancombc48_path <- file.path(outputs_dir, "ancombc_48h_results.rds")

ancombc_48h <- cache_rds(
  ancombc48_path,
  {
    tryCatch({
      ANCOMBC::ancombc2(
        data = ps_48h_counts,
        tax_level = "Genus",
        fix_formula = "group + carbohydrate",
        rand_formula = "(1|subject_id)",
        p_adj_method = "BH",
        prv_cut = 0.10,
        lib_cut = 0,
        group = "carbohydrate",
        struc_zero = TRUE,
        neg_lb = TRUE,
        alpha = 0.05,
        global = TRUE
      )
    }, error = function(e) {
      message("ANCOM-BC 48h failed: ", e$message)
      NULL
    })
  }
)

if (!is.null(ancombc_48h)) {
  # Extract results from the res component (ANCOM-BC2 structure)
  res_data_48h <- ancombc_48h$res
  
  # Get the effect names (excluding Intercept for main effects)
  effect_names_48h <- c("groupObese", "carbohydrateRDC", "carbohydrateSDC")
  
  # Create a data frame from the results for each effect
  ancombc48_tbl <- purrr::map_dfr(effect_names_48h, function(effect) {
    # Check if columns exist for this effect
    lfc_col <- paste0("lfc_", effect)
    se_col <- paste0("se_", effect)
    W_col <- paste0("W_", effect)
    p_col <- paste0("p_", effect)
    q_col <- paste0("q_", effect)
    diff_col <- paste0("diff_", effect)
    
    if (all(c(lfc_col, se_col, W_col, p_col, q_col, diff_col) %in% colnames(res_data_48h))) {
      tibble(
        taxon = res_data_48h$taxon,
        effect = effect,
        beta = res_data_48h[[lfc_col]],
        se = res_data_48h[[se_col]],
        W = res_data_48h[[W_col]],
        p_value = res_data_48h[[p_col]],
        q_value = res_data_48h[[q_col]],
        diff_abn = res_data_48h[[diff_col]]
      )
    } else {
      # Return empty tibble if columns don't exist
      tibble()
    }
  }) %>% dplyr::filter(q_value < 0.1)
} else {
  ancombc48_tbl <- tibble()
}

ancombc48_tbl
```

#### Forest Plot of ANCOM-BC2 48h Results

```{r ancombc48_forest_plot}
# Prepare data for forest plot (show all data initially)
forest_data_48h <- ancombc48_tbl %>%
  dplyr::mutate(
    # Calculate confidence intervals
    ci_lower = beta - 1.96 * se,
    ci_upper = beta + 1.96 * se,
    # Create effect labels
    effect_label = case_when(
      effect == "groupObese" ~ "Obese vs Normal",
      effect == "carbohydrateRDC" ~ "RDC vs No Added Carb",
      effect == "carbohydrateSDC" ~ "SDC vs No Added Carb",
      TRUE ~ effect
    ),
    # Create taxon labels with significance
    taxon_label = paste0(taxon, " (q=", round(q_value, 3), ")"),
    # Color by significance level
    significance = case_when(
      q_value < 0.001 ~ "q < 0.001",
      q_value < 0.01 ~ "q < 0.01", 
      q_value < 0.05 ~ "q < 0.05",
      q_value < 0.1 ~ "q < 0.1",
      TRUE ~ "q ≥ 0.1"
    )
  ) %>%
  # Order by effect size (absolute value)
  dplyr::arrange(desc(abs(beta)))

# Create forest plot for 48h results
forest_plot_48h <- ggplot(forest_data_48h, aes(x = beta, y = reorder(taxon_label, beta))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper, color = significance), 
                 height = 0.15, size = 1) +
  geom_point(aes(color = significance), size = 3, fill = "white", shape = 21, stroke = 1.5) +
  facet_wrap(~ effect_label, scales = "free_x", ncol = 3) +
  scale_color_manual(
    values = c("q < 0.001" = "#d73027", 
               "q < 0.01" = "#f46d43",
               "q < 0.05" = "#fdae61", 
               "q < 0.1" = "#fee08b",
               "q ≥ 0.1" = "#d9d9d9"),
    name = "FDR-adjusted\np-value"
  ) +
  labs(
    title = "ANCOM-BC2 Results: 48h Timepoint Analysis",
    subtitle = "Effect sizes (log-fold changes) with 95% confidence intervals\n(48h timepoint only)",
    x = "Log-fold change (β)",
    y = "Taxon"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 9),
    axis.text.x = element_text(size = 10),
    strip.text = element_text(size = 11, face = "bold"),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "lightblue", color = "black")
  )

print(forest_plot_48h)

# Summary statistics for 48h analysis
cat("Summary of ANCOM-BC2 48h Results:\n")
cat("Total taxa shown:", nrow(forest_data_48h), "\n")
cat("By effect:\n")
print(table(forest_data_48h$effect_label))

# Show top effects by magnitude
cat("\nTop 10 Effects by Magnitude (48h):\n")
print(forest_data_48h %>%
  dplyr::select(taxon, effect_label, beta, q_value) %>%
  dplyr::arrange(desc(abs(beta))) %>%
  dplyr::slice_head(n = 10))
```

### 8.4.1 ANCOM-BC2: RDC vs SDC at 48h

```{r ancombc2_rdc_vs_sdc_48h}
# Subset to 48h timepoint only
ps_48h_counts <- subset_samples(ps_counts, Timepoint == "48H")

# Further subset to only RDC and SDC groups (exclude No Added Carb)
ps_48h_rdc_sdc <- subset_samples(ps_48h_counts, carbohydrate %in% c("RDC", "SDC"))

ancombc2_rdc_sdc_path <- file.path(outputs_dir, "ancombc2_rdc_vs_sdc_48h_results.rds")

ancombc2_rdc_sdc <- cache_rds(
  ancombc2_rdc_sdc_path,
  {
    tryCatch({
      # Remove samples with too many zeros that cause estimation issues
      # Use a more lenient threshold since 48h data may have more zeros
      sample_sums <- sample_sums(ps_48h_rdc_sdc)
      zero_prop <- apply(otu_table(ps_48h_rdc_sdc), 2, function(x) sum(x == 0) / length(x))
      problematic_samples <- names(zero_prop[zero_prop > 0.95])  # More lenient: 95% zeros
      
      if (length(problematic_samples) > 0) {
        message("Removing samples with >95% zeros: ", paste(problematic_samples, collapse = ", "))
        ps_48h_rdc_sdc_clean <- subset_samples(ps_48h_rdc_sdc, !sample_names(ps_48h_rdc_sdc) %in% problematic_samples)
        
        # Check if we still have enough samples
        if (nsamples(ps_48h_rdc_sdc_clean) < 10) {
          message("Too few samples after filtering, using original data")
          ps_48h_rdc_sdc_clean <- ps_48h_rdc_sdc
        }
      } else {
        ps_48h_rdc_sdc_clean <- ps_48h_rdc_sdc
      }
      
      ANCOMBC::ancombc2(
        data = ps_48h_rdc_sdc_clean,
        tax_level = "Genus",
        fix_formula = "group + carbohydrate",
        rand_formula = "(1|subject_id)",
        p_adj_method = "BH",
        prv_cut = 0.10,
        lib_cut = 0,
        s0_perc = 0.05,
        group = "carbohydrate",
        struc_zero = TRUE,
        neg_lb = TRUE,
        alpha = 0.05,
        global = TRUE
      )
    }, error = function(e) {
      message("ANCOM-BC2 RDC vs SDC 48h failed: ", e$message)
      NULL
    })
  }
)

if (!is.null(ancombc2_rdc_sdc)) {
  # Extract results from the res component
  res_data_rdc_sdc <- ancombc2_rdc_sdc$res
  
  # Get the effect names (focusing on carbohydrate comparison)
  effect_names_rdc_sdc <- c("groupObese", "carbohydrateSDC")
  
  # Create a data frame from the results for each effect
  ancombc2_rdc_sdc_tbl <- purrr::map_dfr(effect_names_rdc_sdc, function(effect) {
    # Check if columns exist for this effect
    lfc_col <- paste0("lfc_", effect)
    se_col <- paste0("se_", effect)
    W_col <- paste0("W_", effect)
    p_col <- paste0("p_", effect)
    q_col <- paste0("q_", effect)
    diff_col <- paste0("diff_", effect)
    
    if (all(c(lfc_col, se_col, W_col, p_col, q_col, diff_col) %in% colnames(res_data_rdc_sdc))) {
      tibble(
        taxon = res_data_rdc_sdc$taxon,
        effect = effect,
        beta = res_data_rdc_sdc[[lfc_col]],
        se = res_data_rdc_sdc[[se_col]],
        W = res_data_rdc_sdc[[W_col]],
        p_value = res_data_rdc_sdc[[p_col]],
        q_value = res_data_rdc_sdc[[q_col]],
        diff_abn = res_data_rdc_sdc[[diff_col]]
      )
    } else {
      # Return empty tibble if columns don't exist
      tibble()
    }
  }) %>% dplyr::arrange(q_value)
} else {
  ancombc2_rdc_sdc_tbl <- tibble()
}

if (nrow(ancombc2_rdc_sdc_tbl) > 0) {
  ancombc2_rdc_sdc_tbl %>%
    dplyr::filter(q_value < 0.1)
} else {
  cat("No results available - analysis may have failed due to insufficient data\n")
  tibble()
}
```

#### Forest Plot: RDC vs SDC at 48h

```{r ancombc2_rdc_sdc_forest_plot}
# Prepare data for forest plot
if (nrow(ancombc2_rdc_sdc_tbl) > 0) {
  forest_data_rdc_sdc <- ancombc2_rdc_sdc_tbl %>%
    dplyr::filter(q_value < 0.1) %>%
  dplyr::mutate(
    # Calculate confidence intervals
    ci_lower = beta - 1.96 * se,
    ci_upper = beta + 1.96 * se,
    # Create effect labels
    effect_label = case_when(
      effect == "groupObese" ~ "Obese vs Normal (48h)",
      effect == "carbohydrateSDC" ~ "SDC vs RDC (48h)",
      TRUE ~ effect
    ),
    # Create taxon labels with significance
    taxon_label = paste0(taxon, " (q=", round(q_value, 3), ")"),
    # Color by significance level
    significance = case_when(
      q_value < 0.001 ~ "q < 0.001",
      q_value < 0.01 ~ "q < 0.01", 
      q_value < 0.05 ~ "q < 0.05",
      q_value < 0.1 ~ "q < 0.1",
      TRUE ~ "q ≥ 0.1"
    )
  ) %>%
  # Order by effect size (absolute value)
  dplyr::arrange(desc(abs(beta)))
} else {
  forest_data_rdc_sdc <- tibble()
}

# Create forest plot for RDC vs SDC comparison
if (nrow(forest_data_rdc_sdc) > 0) {
forest_plot_rdc_sdc <- ggplot(forest_data_rdc_sdc, aes(x = beta, y = reorder(taxon_label, beta))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper, color = significance), 
                 height = 0.15, size = 1) +
  geom_point(aes(color = significance), size = 3, fill = "white", shape = 21, stroke = 1.5) +
  facet_wrap(~ effect_label, scales = "free_x", ncol = 2) +
  scale_color_manual(
    values = c("q < 0.001" = "#d73027", 
               "q < 0.01" = "#f46d43",
               "q < 0.05" = "#fdae61", 
               "q < 0.1" = "#fee08b",
               "q ≥ 0.1" = "#d9d9d9"),
    name = "FDR-adjusted\np-value"
  ) +
  labs(
    title = "ANCOM-BC2: RDC vs SDC Comparison at 48h",
    subtitle = "Effect sizes (log-fold changes) with 95% confidence intervals\n(Direct comparison between RDC and SDC at 48h timepoint)",
    x = "Log-fold change (β)",
    y = "Taxon"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 9),
    axis.text.x = element_text(size = 10),
    strip.text = element_text(size = 11, face = "bold"),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "lightgreen", color = "black")
  )

print(forest_plot_rdc_sdc)
} else {
  cat("No data available for forest plot - analysis may have failed\n")
}

# Summary statistics for RDC vs SDC comparison
if (nrow(forest_data_rdc_sdc) > 0) {
cat("Summary of RDC vs SDC 48h Comparison:\n")
cat("Total significant taxa (q < 0.1):", nrow(forest_data_rdc_sdc), "\n")
cat("By effect:\n")
print(table(forest_data_rdc_sdc$effect_label))

# Show top effects by magnitude
cat("\nTop 10 Effects by Magnitude (RDC vs SDC 48h):\n")
print(forest_data_rdc_sdc %>%
  dplyr::select(taxon, effect_label, beta, q_value) %>%
  dplyr::arrange(desc(abs(beta))) %>%
  dplyr::slice_head(n = 10))

# Focus on carbohydrate effect specifically
carb_effect_rdc_sdc <- forest_data_rdc_sdc %>%
  dplyr::filter(effect == "carbohydrateSDC")

if(nrow(carb_effect_rdc_sdc) > 0) {
  cat("\nCarbohydrate Effect (SDC vs RDC at 48h):\n")
  cat("==========================================\n")
  print(carb_effect_rdc_sdc %>%
    dplyr::select(taxon, beta, q_value) %>%
    dplyr::arrange(desc(abs(beta))))
}
} else {
  cat("No summary statistics available - analysis may have failed\n")
}
```

### 8.4.2 ANCOM-BC2: Group × Carbohydrate Interaction at 48h

```{r ancombc2_group_carb_interaction_48h}
# Subset to 48h timepoint only
ps_48h_counts <- subset_samples(ps_counts, Timepoint == "48H")

# Further subset to only RDC and SDC groups (exclude No Added Carb)
ps_48h_rdc_sdc <- subset_samples(ps_48h_counts, carbohydrate %in% c("RDC", "SDC"))

ancombc2_interaction_path <- file.path(outputs_dir, "ancombc2_group_carb_interaction_48h_results.rds")

ancombc2_interaction <- cache_rds(
  ancombc2_interaction_path,
  {
    tryCatch({
      # Remove samples with too many zeros that cause estimation issues
      # Use a more lenient threshold since 48h data may have more zeros
      sample_sums <- sample_sums(ps_48h_rdc_sdc)
      zero_prop <- apply(otu_table(ps_48h_rdc_sdc), 2, function(x) sum(x == 0) / length(x))
      problematic_samples <- names(zero_prop[zero_prop > 0.95])  # More lenient: 95% zeros
      
      if (length(problematic_samples) > 0) {
        message("Removing samples with >95% zeros: ", paste(problematic_samples, collapse = ", "))
        ps_48h_rdc_sdc_clean <- subset_samples(ps_48h_rdc_sdc, !sample_names(ps_48h_rdc_sdc) %in% problematic_samples)
        
        # Check if we still have enough samples
        if (nsamples(ps_48h_rdc_sdc_clean) < 10) {
          message("Too few samples after filtering, using original data")
          ps_48h_rdc_sdc_clean <- ps_48h_rdc_sdc
        }
      } else {
        ps_48h_rdc_sdc_clean <- ps_48h_rdc_sdc
      }
      
      ANCOMBC::ancombc2(
        data = ps_48h_rdc_sdc_clean,
        tax_level = "Genus",
        fix_formula = "group * carbohydrate",  # Include interaction term
        rand_formula = "(1|subject_id)",
        p_adj_method = "BH",
        prv_cut = 0.10,
        lib_cut = 0,
        s0_perc = 0.05,
        group = "carbohydrate",
        struc_zero = TRUE,
        neg_lb = TRUE,
        alpha = 0.05,
        global = TRUE
      )
    }, error = function(e) {
      message("ANCOM-BC2 Group × Carbohydrate Interaction 48h failed: ", e$message)
      NULL
    })
  }
)

if (!is.null(ancombc2_interaction)) {
  # Extract results from the res component
  res_data_interaction <- ancombc2_interaction$res
  
  # Get the effect names (including interaction term)
  effect_names_interaction <- c("groupObese", "carbohydrateSDC", "groupObese:carbohydrateSDC")
  
  # Create a data frame from the results for each effect
  ancombc2_interaction_tbl <- purrr::map_dfr(effect_names_interaction, function(effect) {
    # Check if columns exist for this effect
    lfc_col <- paste0("lfc_", effect)
    se_col <- paste0("se_", effect)
    W_col <- paste0("W_", effect)
    p_col <- paste0("p_", effect)
    q_col <- paste0("q_", effect)
    diff_col <- paste0("diff_", effect)
    
    if (all(c(lfc_col, se_col, W_col, p_col, q_col, diff_col) %in% colnames(res_data_interaction))) {
      tibble(
        taxon = res_data_interaction$taxon,
        effect = effect,
        beta = res_data_interaction[[lfc_col]],
        se = res_data_interaction[[se_col]],
        W = res_data_interaction[[W_col]],
        p_value = res_data_interaction[[p_col]],
        q_value = res_data_interaction[[q_col]],
        diff_abn = res_data_interaction[[diff_col]]
      )
    } else {
      # Return empty tibble if columns don't exist
      tibble()
    }
  }) %>% dplyr::arrange(q_value)
} else {
  ancombc2_interaction_tbl <- tibble()
}

if (nrow(ancombc2_interaction_tbl) > 0) {
  ancombc2_interaction_tbl %>%
    dplyr::filter(q_value < 0.1)
} else {
  cat("No results available - analysis may have failed due to insufficient data\n")
  tibble()
}
```

#### Forest Plot: Group × Carbohydrate Interaction at 48h

```{r ancombc2_interaction_forest_plot}
# Prepare data for forest plot
if (nrow(ancombc2_interaction_tbl) > 0) {
  forest_data_interaction <- ancombc2_interaction_tbl %>%
    dplyr::filter(q_value < 0.1) %>%
    dplyr::mutate(
      # Calculate confidence intervals
      ci_lower = beta - 1.96 * se,
      ci_upper = beta + 1.96 * se,
      # Create effect labels
      effect_label = case_when(
        effect == "groupObese" ~ "Obese vs Normal (48h)",
        effect == "carbohydrateSDC" ~ "SDC vs RDC (48h)",
        effect == "groupObese:carbohydrateSDC" ~ "Group × Carbohydrate Interaction (48h)",
        TRUE ~ effect
      ),
      # Create taxon labels with significance
      taxon_label = paste0(taxon, " (q=", round(q_value, 3), ")"),
      # Color by significance level
      significance = case_when(
        q_value < 0.001 ~ "q < 0.001",
        q_value < 0.01 ~ "q < 0.01", 
        q_value < 0.05 ~ "q < 0.05",
        q_value < 0.1 ~ "q < 0.1",
        TRUE ~ "q ≥ 0.1"
      )
    ) %>%
    # Order by effect size (absolute value)
    dplyr::arrange(desc(abs(beta)))
} else {
  forest_data_interaction <- tibble()
}

# Create forest plot for interaction analysis
if (nrow(forest_data_interaction) > 0) {
  forest_plot_interaction <- ggplot(forest_data_interaction, aes(x = beta, y = reorder(taxon_label, beta))) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    geom_errorbarh(aes(xmin = ci_lower, xmax = ci_upper, color = significance), 
                   height = 0.15, size = 1) +
    geom_point(aes(color = significance), size = 3, fill = "white", shape = 21, stroke = 1.5) +
    facet_wrap(~ effect_label, scales = "free_x", ncol = 3) +
    scale_color_manual(
      values = c("q < 0.001" = "#d73027", 
                 "q < 0.01" = "#f46d43",
                 "q < 0.05" = "#fdae61", 
                 "q < 0.1" = "#fee08b",
                 "q ≥ 0.1" = "#d9d9d9"),
      name = "FDR-adjusted\np-value"
    ) +
    labs(
      title = "ANCOM-BC2: Group × Carbohydrate Interaction at 48h",
      subtitle = "Effect sizes (log-fold changes) with 95% confidence intervals\n(Testing if carbohydrate effects differ between obese and normal weight individuals)",
      x = "Log-fold change (β)",
      y = "Taxon"
    ) +
    theme_minimal() +
    theme(
      axis.text.y = element_text(size = 9),
      axis.text.x = element_text(size = 10),
      strip.text = element_text(size = 11, face = "bold"),
      plot.title = element_text(size = 14, face = "bold"),
      plot.subtitle = element_text(size = 11),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      strip.background = element_rect(fill = "lightcoral", color = "black")
    )

  print(forest_plot_interaction)
} else {
  cat("No data available for forest plot - analysis may have failed\n")
}

# Summary statistics for interaction analysis
if (nrow(forest_data_interaction) > 0) {
  cat("Summary of Group × Carbohydrate Interaction 48h:\n")
  cat("Total significant taxa (q < 0.1):", nrow(forest_data_interaction), "\n")
  cat("By effect:\n")
  print(table(forest_data_interaction$effect_label))

  # Show top effects by magnitude
  cat("\nTop 10 Effects by Magnitude (Group × Carbohydrate Interaction 48h):\n")
  print(forest_data_interaction %>%
    dplyr::select(taxon, effect_label, beta, q_value) %>%
    dplyr::arrange(desc(abs(beta))) %>%
    dplyr::slice_head(n = 10))

  # Focus on interaction effect specifically
  interaction_effect <- forest_data_interaction %>%
    dplyr::filter(effect == "groupObese:carbohydrateSDC")

  if(nrow(interaction_effect) > 0) {
    cat("\nGroup × Carbohydrate Interaction Effect (48h):\n")
    cat("===============================================\n")
    print(interaction_effect %>%
      dplyr::select(taxon, beta, q_value) %>%
      dplyr::arrange(desc(abs(beta))))
    
    cat("\nInterpretation of Interaction Effects:\n")
    cat("- Positive β: Carbohydrate effect is stronger in obese individuals\n")
    cat("- Negative β: Carbohydrate effect is stronger in normal weight individuals\n")
  } else {
    cat("\nNo significant interaction effects found (q < 0.1)\n")
  }
} else {
  cat("No summary statistics available - analysis may have failed\n")
}
```



### 8.5 ALDEx2 (48 h carbohydrate effect)

```{r aldex}
aldex_path <- file.path(outputs_dir, "aldex2_48h_results.rds")

aldex_tbl <- cache_rds(
  aldex_path,
  {
    tryCatch({
      # Subset to 48h timepoint and only RDC vs SDC (two-group comparison)
      ps_48h_rdc_sdc <- subset_samples(ps_genus_48h, 
                                       Timepoint == "48H" & 
                                       carbohydrate %in% c("RDC", "SDC"))
      
      counts_48h <- as(otu_table(ps_48h_rdc_sdc), "matrix")
      conds_48h <- as.character(sample_data(ps_48h_rdc_sdc)$carbohydrate)  # Ensure character type

      aldex_clr <- ALDEx2::aldex.clr(counts_48h, conds_48h, mc.samples = 64, denom = "all", verbose = FALSE)
      aldex_tt <- ALDEx2::aldex.ttest(aldex_clr)  # Use t-test for two-group comparison
      aldex_effect <- ALDEx2::aldex.effect(aldex_clr)

      aldex_tt %>%
        rownames_to_column("taxon") %>%
        dplyr::left_join(aldex_effect %>% rownames_to_column("taxon"), by = "taxon") %>%
        dplyr::mutate(
          # Ensure numeric types for p-values
          we.ep = as.numeric(we.ep),
          we.eBH = as.numeric(we.eBH),
          q = we.eBH  # Use Benjamini-Hochberg corrected p-values from ALDEx2
        ) %>%
        dplyr::arrange(q)
    }, error = function(e) {
      message("ALDEx2 failed: ", e$message)
      tibble()
    })
  }
)

if (nrow(aldex_tbl) > 0) {
  aldex_tbl %>% dplyr::filter(q < 0.1)
} else {
  cat("No ALDEx2 results available - analysis may have failed\n")
  tibble()
}
```

#### Forest Plot: ALDEx2 RDC vs SDC at 48h

```{r aldex_forest_plot}
# Prepare data for forest plot
if (nrow(aldex_tbl) > 0) {
  # Get taxonomy information for ASV to genus/species conversion
  tax_table_df <- as.data.frame(tax_table(ps_genus_48h))
  
  forest_data_aldex <- aldex_tbl %>%
    dplyr::filter(q < 0.1) %>%
    dplyr::mutate(
      # Convert ASV to genus/species format
      taxon_display = case_when(
        # Check if we have genus and species information
        !is.na(tax_table_df[taxon, "Genus"]) & !is.na(tax_table_df[taxon, "Species"]) & 
        tax_table_df[taxon, "Species"] != "uncultured" & tax_table_df[taxon, "Species"] != "" ~
          paste0(tax_table_df[taxon, "Genus"], " ", tax_table_df[taxon, "Species"]),
        # If no species, use genus
        !is.na(tax_table_df[taxon, "Genus"]) & tax_table_df[taxon, "Genus"] != "uncultured" & 
        tax_table_df[taxon, "Genus"] != "" ~ tax_table_df[taxon, "Genus"],
        # Fallback to ASV if no genus info
        TRUE ~ taxon
      ),
      # For ALDEx2, we'll use the effect size without confidence intervals
      # since ALDEx2 doesn't provide standard errors in the traditional sense
      # The significance is already captured in the q-value
      # Create taxon labels with significance
      taxon_label = paste0(taxon_display, " (q=", round(q, 3), ")"),
      # Color by significance level
      significance = case_when(
        q < 0.001 ~ "q < 0.001",
        q < 0.01 ~ "q < 0.01", 
        q < 0.05 ~ "q < 0.05",
        q < 0.1 ~ "q < 0.1",
        TRUE ~ "q ≥ 0.1"
      )
    ) %>%
    # Order by effect size (absolute value)
    dplyr::arrange(desc(abs(effect)))
} else {
  forest_data_aldex <- tibble()
}

# Create forest plot for ALDEx2 results
if (nrow(forest_data_aldex) > 0) {
  forest_plot_aldex <- ggplot(forest_data_aldex, aes(x = effect, y = reorder(taxon_label, effect))) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    geom_point(aes(color = significance), size = 4, fill = "white", shape = 21, stroke = 1.5) +
    scale_color_manual(
      values = c("q < 0.001" = "#d73027", 
                 "q < 0.01" = "#f46d43",
                 "q < 0.05" = "#fdae61", 
                 "q < 0.1" = "#fee08b",
                 "q ≥ 0.1" = "#d9d9d9"),
      name = "FDR-adjusted\np-value"
    ) +
    labs(
      title = "ALDEx2: RDC vs SDC Comparison at 48h",
      subtitle = "Effect sizes (Cohen's d) - significance indicated by color\n(Direct comparison between RDC and SDC at 48h timepoint)",
      x = "Effect size (Cohen's d)",
      y = "Taxon"
    ) +
    theme_minimal() +
    theme(
      axis.text.y = element_text(size = 9),
      axis.text.x = element_text(size = 10),
      plot.title = element_text(size = 14, face = "bold"),
      plot.subtitle = element_text(size = 11),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )

  print(forest_plot_aldex)
} else {
  cat("No data available for ALDEx2 forest plot - analysis may have failed\n")
}

# Summary statistics for ALDEx2 analysis
if (nrow(forest_data_aldex) > 0) {
  cat("Summary of ALDEx2 RDC vs SDC 48h Comparison:\n")
  cat("Total significant taxa (q < 0.1):", nrow(forest_data_aldex), "\n")
  
  # Show top effects by magnitude
  cat("\nTop 10 Effects by Magnitude (ALDEx2 RDC vs SDC 48h):\n")
  print(forest_data_aldex %>%
    dplyr::select(taxon_display, effect, q, we.ep) %>%
    dplyr::arrange(desc(abs(effect))) %>%
    dplyr::slice_head(n = 10))
  
  # Interpretation
  cat("\nInterpretation of ALDEx2 Results:\n")
  cat("- Positive effect: Higher abundance in SDC compared to RDC\n")
  cat("- Negative effect: Higher abundance in RDC compared to SDC\n")
  cat("- Effect size magnitude: |effect| > 0.5 is considered moderate, |effect| > 0.8 is large\n")
} else {
  cat("No summary statistics available for ALDEx2 - analysis may have failed\n")
}
```

### 8.6 LEfSe (48H, RDC vs SDC)

```{r lefse}
lefse_path <- file.path(outputs_dir, "lefse_48h_rdc_vs_sdc_results.rds")

# Subset to only RDC and SDC at 48H
ps_genus_48h_rdc_sdc <- subset_samples(ps_genus_48h, 
                                      carbohydrate %in% c("RDC", "SDC"))

cat("48H RDC vs SDC samples for LEfSe:", nsamples(ps_genus_48h_rdc_sdc), "\n")
cat("Sample distribution by carbohydrate:\n")
print(table(sample_data(ps_genus_48h_rdc_sdc)$carbohydrate))

# Check available sample variables
cat("Available sample variables in ps_genus_48h_rdc_sdc:\n")
print(sample_variables(ps_genus_48h_rdc_sdc))

lefse_res <- cache_rds(
  lefse_path,
  {
    tryCatch({
      microbiomeMarker::run_lefse(
        ps = ps_genus_48h_rdc_sdc,
        group = "carbohydrate",
        subgroup = NULL,  # No subgroup, just RDC vs SDC comparison
        taxa_rank = "Genus",
        kw_cutoff = 0.05,
        lda_cutoff = 2
      )
    }, error = function(e) {
      message("LEfSe failed: ", e$message)
      NULL
    })
  }
)

lefse_tbl <- if (!is.null(lefse_res)) {
  microbiomeMarker::marker_table(lefse_res) %>%
    as_tibble() %>%
    dplyr::arrange(desc(ef_lda))
} else {
  tibble()
}

lefse_tbl
```

### 8.6.1 LEfSe Visualization

```{r lefse_viz}
if (nrow(lefse_tbl) > 0) {
  # Prepare data for visualization
  lefse_viz_data <- lefse_tbl %>%
    dplyr::mutate(
      # Create direction for bar colors (positive for SDC, negative for RDC)
      direction = ifelse(enrich_group == "SDC", 1, -1),
      # Create display labels with genus names and effect sizes
      label = paste0(feature, " (LDA=", round(ef_lda, 2), ")"),
      # Significance levels
      significance = case_when(
        padj < 0.001 ~ "***",
        padj < 0.01 ~ "**", 
        padj < 0.05 ~ "*",
        TRUE ~ ""
      ),
      # Combined label with significance
      display_label = paste0(label, significance)
    ) %>%
    # Limit to top 15 most significant results for readability
    dplyr::slice_head(n = 15) %>%
    # Reorder by effect size
    dplyr::arrange(ef_lda)
  
  # Create horizontal bar chart
  p_lefse <- ggplot(lefse_viz_data, aes(x = ef_lda * direction, y = reorder(display_label, ef_lda))) +
    geom_col(aes(fill = enrich_group), alpha = 0.8) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    scale_fill_manual(
      values = c("RDC" = "#2E86AB", "SDC" = "#A23B72"),
      name = "Enriched in",
      labels = c("RDC" = "RDC", "SDC" = "SDC")
    ) +
    labs(
      title = "LEfSe Results: 48H RDC vs SDC Comparison",
      subtitle = "Linear Discriminant Analysis (LDA) Effect Sizes",
      x = "LDA Score (Effect Size)",
      y = "Taxa (Genus level)",
      caption = "*** p<0.001, ** p<0.01, * p<0.05"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 14, face = "bold"),
      plot.subtitle = element_text(size = 12, color = "gray60"),
      axis.text.y = element_text(size = 10),
      axis.text.x = element_text(size = 10),
      axis.title = element_text(size = 11, face = "bold"),
      legend.position = "bottom",
      legend.title = element_text(face = "bold"),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      plot.caption = element_text(size = 9, color = "gray50")
    )
  
  print(p_lefse)
  
  # Save the plot
  ggsave(file.path(outputs_dir, "lefse_48h_rdc_vs_sdc_barplot.png"), 
         p_lefse, width = 12, height = 8, dpi = 300)
  
  # Summary statistics
  cat("\nLEfSe Summary (48H RDC vs SDC):\n")
  cat("- Total significant taxa:", nrow(lefse_tbl), "\n")
  cat("- Taxa enriched in RDC:", sum(lefse_tbl$enrich_group == "RDC"), "\n")
  cat("- Taxa enriched in SDC:", sum(lefse_tbl$enrich_group == "SDC"), "\n")
  cat("- Mean LDA score:", round(mean(lefse_tbl$ef_lda), 2), "\n")
  cat("- Max LDA score:", round(max(lefse_tbl$ef_lda), 2), "\n")
  
} else {
  cat("No significant LEfSe results found for 48H RDC vs SDC comparison.\n")
}
```

### 8.7 MaAsLin2

```{r maaslin2}
maaslin_path <- file.path(outputs_dir, "maaslin2_genus_results.rds")

maaslin_counts <- as.data.frame(otu_table(ps_genus_filtered))
maaslin_meta <- as(sample_data(ps_genus_filtered), "data.frame")
maaslin_meta$SampleID <- rownames(maaslin_meta)

maaslin_fit <- cache_rds(
  maaslin_path,
  {
    Maaslin2::Maaslin2(
      input_data = maaslin_counts,
      input_metadata = maaslin_meta,
      output = file.path(outputs_dir, "maaslin2_genus"),
      min_prevalence = prevalence_cutoff,
      normalization = "TSS",
      transform = "LOG",
      analysis_method = "LM",
      fixed_effects = c("group", "carbohydrate", "timepoint"),
      random_effects = c("subject_id"),
      plot_heatmap = FALSE,
      plot_scatter = FALSE
    )
  }
)

maaslin_tbl <- maaslin_fit$results %>%
  as_tibble() %>%
  dplyr::mutate(q = fdr_adjust(qval)) %>%
  dplyr::arrange(q)

maaslin_tbl %>% dplyr::filter(q < 0.1)
```

### 8.7.1 MaAsLin2 Visualization

```{r maaslin_viz, fig.width = 20, fig.height = 24}
if (nrow(maaslin_tbl) > 0) {
  # Get taxonomy information for ASV to genus/species conversion
  tax_table_df <- as.data.frame(tax_table(ps_genus_filtered))
  tax_table_df$feature <- rownames(tax_table_df)
  
  # Create taxonomic display names
  tax_table_df <- tax_table_df %>%
    dplyr::mutate(
      # Create genus_species display name
      genus_species = case_when(
        !is.na(Species) & Species != "Unassigned" & Species != "" ~ paste0(Genus, " ", Species),
        !is.na(Genus) & Genus != "Unassigned" & Genus != "" ~ paste0(Genus, " sp."),
        !is.na(Family) & Family != "Unassigned" & Family != "" ~ paste0(Family, " (family)"),
        !is.na(Order) & Order != "Unassigned" & Order != "" ~ paste0(Order, " (order)"),
        TRUE ~ feature  # Fallback to ASV ID if no good taxonomy
      )
    )
  
  # Filter for significant results and add taxonomy
  maaslin_sig <- maaslin_tbl %>% 
    dplyr::filter(q < 0.1) %>%
    dplyr::left_join(tax_table_df %>% dplyr::select(feature, genus_species), by = "feature") %>%
    dplyr::mutate(
      # Create significance levels
      significance = case_when(
        q < 0.001 ~ "***",
        q < 0.01 ~ "**", 
        q < 0.05 ~ "*",
        TRUE ~ ""
      ),
      # Create display labels with taxonomic names
      display_label = paste0(genus_species, " (", metadata, "=", value, ")"),
      # Create combined label with significance
      full_label = paste0(display_label, significance),
      # Create metadata-value combination for grouping
      metadata_value = paste0(metadata, "=", value)
    )
  
  if (nrow(maaslin_sig) > 0) {
    # Plot 1: Coefficient effect sizes by metadata variable
    p1 <- ggplot(maaslin_sig, aes(x = coef, y = reorder(full_label, coef))) +
      geom_col(aes(fill = metadata_value), alpha = 0.8) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
      scale_fill_viridis_d(name = "Variable", option = "plasma") +
      labs(
        title = "MaAsLin2 Results: Significant Associations",
        subtitle = "Coefficient Effect Sizes by Variable",
        x = "Coefficient (Effect Size)",
        y = "Features",
        caption = "*** q<0.001, ** q<0.01, * q<0.05"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12, color = "gray60"),
        axis.text.y = element_text(size = 9),
        axis.text.x = element_text(size = 10),
        axis.title = element_text(size = 11, face = "bold"),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        plot.caption = element_text(size = 9, color = "gray50")
      )
    
    print(p1)
    
    # Plot 2: Volcano plot (effect size vs significance)
    # Add taxonomy to full maaslin_tbl for volcano plot
    maaslin_tbl_with_tax <- maaslin_tbl %>%
      dplyr::left_join(tax_table_df %>% dplyr::select(feature, genus_species), by = "feature")
    
    # Create data for labeling significant points
    sig_points <- maaslin_tbl_with_tax %>% 
      dplyr::filter(q < 0.1) %>%
      dplyr::mutate(
        # Create genus name for labeling (extract just the genus part)
        genus_label = case_when(
          !is.na(genus_species) & grepl(" ", genus_species) ~ 
            stringr::str_extract(genus_species, "^[^ ]+"),  # Extract first word (genus)
          !is.na(genus_species) ~ genus_species,
          TRUE ~ feature
        )
      )
    
    p2 <- ggplot(maaslin_tbl_with_tax, aes(x = coef, y = -log10(q))) +
      geom_point(aes(color = q < 0.1), alpha = 0.6, size = 2) +
      geom_hline(yintercept = -log10(0.1), linetype = "dashed", color = "red", alpha = 0.7) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
      # Add text labels for significant points
      ggrepel::geom_text_repel(
        data = sig_points,
        aes(label = genus_label),
        size = 3,
        color = "#E31A1C",
        fontface = "bold",
        max.overlaps = 20,
        min.segment.length = 0.1,
        box.padding = 0.3,
        point.padding = 0.2,
        segment.color = "gray50",
        segment.alpha = 0.6
      ) +
      scale_color_manual(
        values = c("FALSE" = "gray70", "TRUE" = "#E31A1C"),
        name = "Significant",
        labels = c("FALSE" = "q ≥ 0.1", "TRUE" = "q < 0.1")
      ) +
      labs(
        title = "MaAsLin2 Volcano Plot",
        subtitle = "Effect Size vs Statistical Significance",
        x = "Coefficient (Effect Size)",
        y = "-log10(q-value)",
        caption = "Red dashed line: q = 0.1 threshold; Labels show significant taxa"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12, color = "gray60"),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 11, face = "bold"),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"),
        plot.caption = element_text(size = 9, color = "gray50")
      )
    
    print(p2)
    
    # Plot 3: Top effects by metadata variable
    top_effects <- maaslin_sig %>%
      dplyr::group_by(metadata) %>%
      dplyr::slice_max(abs(coef), n = 10) %>%
      dplyr::ungroup()
    
    p3 <- ggplot(top_effects, aes(x = coef, y = reorder(genus_species, coef))) +
      geom_col(aes(fill = metadata), alpha = 0.8) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
      scale_fill_manual(
        values = c("group" = "#1f77b4", "carbohydrate" = "#ff7f0e", "timepoint" = "#2ca02c"),
        name = "Variable"
      ) +
      labs(
        title = "MaAsLin2: Top Effects by Variable",
        subtitle = "Top 10 largest effects per metadata variable",
        x = "Coefficient (Effect Size)",
        y = "Features"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12, color = "gray60"),
        axis.text.y = element_text(size = 9),
        axis.text.x = element_text(size = 10),
        axis.title = element_text(size = 11, face = "bold"),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank()
      ) +
      facet_wrap(~metadata, scales = "free_y", ncol = 1)
    
    print(p3)
    
    # Save plots
    ggsave(file.path(outputs_dir, "maaslin2_effects_barplot.png"), 
           p1, width = 12, height = 10, dpi = 300)
    ggsave(file.path(outputs_dir, "maaslin2_volcano_plot.png"), 
           p2, width = 10, height = 8, dpi = 300)
    ggsave(file.path(outputs_dir, "maaslin2_top_effects.png"), 
           p3, width = 12, height = 12, dpi = 300)
    
    # Summary statistics
    cat("\nMaAsLin2 Summary:\n")
    cat("- Total features tested:", nrow(maaslin_tbl), "\n")
    cat("- Significant associations (q < 0.1):", nrow(maaslin_sig), "\n")
    cat("- Significant by variable:\n")
    print(table(maaslin_sig$metadata))
    cat("- Mean absolute coefficient:", round(mean(abs(maaslin_sig$coef)), 3), "\n")
    cat("- Max absolute coefficient:", round(max(abs(maaslin_sig$coef)), 3), "\n")
    
  } else {
    cat("No significant MaAsLin2 results found (q < 0.1).\n")
  }
  
} else {
  cat("No MaAsLin2 results available.\n")
}
```

### 8.7.2 MaAsLin2 (48H, RDC vs SDC)

```{r maaslin2_48h_rdc_sdc, fig.width = 20, fig.height = 16}
maaslin_48h_path <- file.path(outputs_dir, "maaslin2_48h_rdc_vs_sdc_results.rds")

# Subset to 48H timepoint and only RDC vs SDC
ps_genus_48h_rdc_sdc_maaslin <- subset_samples(ps_genus_filtered, 
                                              Timepoint == "48H" & 
                                              carbohydrate %in% c("RDC", "SDC"))

cat("48H RDC vs SDC samples for MaAsLin2:", nsamples(ps_genus_48h_rdc_sdc_maaslin), "\n")
cat("Sample distribution by carbohydrate:\n")
print(table(sample_data(ps_genus_48h_rdc_sdc_maaslin)$carbohydrate))

# Prepare data for MaAsLin2
maaslin_48h_counts <- as.data.frame(otu_table(ps_genus_48h_rdc_sdc_maaslin))
maaslin_48h_meta <- as(sample_data(ps_genus_48h_rdc_sdc_maaslin), "data.frame")
maaslin_48h_meta$SampleID <- rownames(maaslin_48h_meta)

# Run MaAsLin2 analysis
maaslin_48h_fit <- cache_rds(
  maaslin_48h_path,
  {
    Maaslin2::Maaslin2(
      input_data = maaslin_48h_counts,
      input_metadata = maaslin_48h_meta,
      output = file.path(outputs_dir, "maaslin2_48h_rdc_vs_sdc"),
      min_prevalence = prevalence_cutoff,
      normalization = "TSS",
      transform = "LOG",
      analysis_method = "LM",
      fixed_effects = c("carbohydrate"),
      random_effects = c("subject_id"),
      plot_heatmap = FALSE,
      plot_scatter = FALSE
    )
  }
)

maaslin_48h_tbl <- maaslin_48h_fit$results %>%
  as_tibble() %>%
  dplyr::mutate(q = fdr_adjust(qval)) %>%
  dplyr::arrange(q)

maaslin_48h_tbl %>% dplyr::filter(q < 0.1)
```

### 8.7.3 MaAsLin2 48H RDC vs SDC Visualization

```{r maaslin_48h_viz, fig.width = 20, fig.height = 16}
if (nrow(maaslin_48h_tbl) > 0) {
  # Get taxonomy information for ASV to genus/species conversion
  tax_table_df_48h <- as.data.frame(tax_table(ps_genus_48h_rdc_sdc_maaslin))
  tax_table_df_48h$feature <- rownames(tax_table_df_48h)
  
  # Create taxonomic display names
  tax_table_df_48h <- tax_table_df_48h %>%
    dplyr::mutate(
      # Create genus_species display name
      genus_species = case_when(
        !is.na(Species) & Species != "Unassigned" & Species != "" ~ paste0(Genus, " ", Species),
        !is.na(Genus) & Genus != "Unassigned" & Genus != "" ~ paste0(Genus, " sp."),
        !is.na(Family) & Family != "Unassigned" & Family != "" ~ paste0(Family, " (family)"),
        !is.na(Order) & Order != "Unassigned" & Order != "" ~ paste0(Order, " (order)"),
        TRUE ~ feature  # Fallback to ASV ID if no good taxonomy
      )
    )
  
  # Filter for significant results and add taxonomy
  maaslin_48h_sig <- maaslin_48h_tbl %>% 
    dplyr::filter(q < 0.1) %>%
    dplyr::left_join(tax_table_df_48h %>% dplyr::select(feature, genus_species), by = "feature") %>%
    dplyr::mutate(
      # Create significance levels
      significance = case_when(
        q < 0.001 ~ "***",
        q < 0.01 ~ "**", 
        q < 0.05 ~ "*",
        TRUE ~ ""
      ),
      # Create display labels with taxonomic names
      display_label = paste0(genus_species, " (", metadata, "=", value, ")"),
      # Create combined label with significance
      full_label = paste0(display_label, significance),
      # Create direction for bar colors (positive for SDC, negative for RDC)
      direction = ifelse(value == "SDC", 1, -1)
    )
  
  if (nrow(maaslin_48h_sig) > 0) {
    # Plot 1: Horizontal bar chart of significant associations
    p1_48h <- ggplot(maaslin_48h_sig, aes(x = coef * direction, y = reorder(full_label, coef))) +
      geom_col(aes(fill = value), alpha = 0.8) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
      scale_fill_manual(
        values = c("RDC" = "#2E86AB", "SDC" = "#A23B72"),
        name = "Enriched in",
        labels = c("RDC" = "RDC", "SDC" = "SDC")
      ) +
      labs(
        title = "MaAsLin2 Results: 48H RDC vs SDC Comparison",
        subtitle = "Coefficient Effect Sizes for Significant Associations",
        x = "Coefficient (Effect Size)",
        y = "Taxa (Genus level)",
        caption = "*** q<0.001, ** q<0.01, * q<0.05"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12, color = "gray60"),
        axis.text.y = element_text(size = 10),
        axis.text.x = element_text(size = 10),
        axis.title = element_text(size = 11, face = "bold"),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        plot.caption = element_text(size = 9, color = "gray50")
      )
    
    print(p1_48h)
    
    # Plot 2: Volcano plot for 48H RDC vs SDC
    maaslin_48h_tbl_with_tax <- maaslin_48h_tbl %>%
      dplyr::left_join(tax_table_df_48h %>% dplyr::select(feature, genus_species), by = "feature")
    
    p2_48h <- ggplot(maaslin_48h_tbl_with_tax, aes(x = coef, y = -log10(q))) +
      geom_point(aes(color = q < 0.1), alpha = 0.6, size = 2) +
      geom_hline(yintercept = -log10(0.1), linetype = "dashed", color = "red", alpha = 0.7) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
      scale_color_manual(
        values = c("FALSE" = "gray70", "TRUE" = "#E31A1C"),
        name = "Significant",
        labels = c("FALSE" = "q ≥ 0.1", "TRUE" = "q < 0.1")
      ) +
      labs(
        title = "MaAsLin2 Volcano Plot: 48H RDC vs SDC",
        subtitle = "Effect Size vs Statistical Significance",
        x = "Coefficient (Effect Size)",
        y = "-log10(q-value)",
        caption = "Red dashed line: q = 0.1 threshold"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12, color = "gray60"),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 11, face = "bold"),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"),
        plot.caption = element_text(size = 9, color = "gray50")
      )
    
    print(p2_48h)
    
    # Plot 3: Top effects by magnitude
    top_effects_48h <- maaslin_48h_sig %>%
      dplyr::slice_max(abs(coef), n = 15) %>%
      dplyr::arrange(coef)
    
    p3_48h <- ggplot(top_effects_48h, aes(x = coef * direction, y = reorder(genus_species, coef))) +
      geom_col(aes(fill = value), alpha = 0.8) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
      scale_fill_manual(
        values = c("RDC" = "#2E86AB", "SDC" = "#A23B72"),
        name = "Enriched in"
      ) +
      labs(
        title = "MaAsLin2: Top 15 Effects (48H RDC vs SDC)",
        subtitle = "Largest effect sizes by magnitude",
        x = "Coefficient (Effect Size)",
        y = "Taxa (Genus level)"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12, color = "gray60"),
        axis.text.y = element_text(size = 10),
        axis.text.x = element_text(size = 10),
        axis.title = element_text(size = 11, face = "bold"),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank()
      )
    
    print(p3_48h)
    
    # Save plots
    ggsave(file.path(outputs_dir, "maaslin2_48h_rdc_vs_sdc_barplot.png"), 
           p1_48h, width = 12, height = 10, dpi = 300)
    ggsave(file.path(outputs_dir, "maaslin2_48h_rdc_vs_sdc_volcano.png"), 
           p2_48h, width = 10, height = 8, dpi = 300)
    ggsave(file.path(outputs_dir, "maaslin2_48h_rdc_vs_sdc_top_effects.png"), 
           p3_48h, width = 12, height = 10, dpi = 300)
    
    # Summary statistics
    cat("\nMaAsLin2 48H RDC vs SDC Summary:\n")
    cat("- Total features tested:", nrow(maaslin_48h_tbl), "\n")
    cat("- Significant associations (q < 0.1):", nrow(maaslin_48h_sig), "\n")
    cat("- Taxa enriched in RDC:", sum(maaslin_48h_sig$value == "RDC"), "\n")
    cat("- Taxa enriched in SDC:", sum(maaslin_48h_sig$value == "SDC"), "\n")
    cat("- Mean absolute coefficient:", round(mean(abs(maaslin_48h_sig$coef)), 3), "\n")
    cat("- Max absolute coefficient:", round(max(abs(maaslin_48h_sig$coef)), 3), "\n")
    
  } else {
    cat("No significant MaAsLin2 results found for 48H RDC vs SDC comparison (q < 0.1).\n")
  }
  
} else {
  cat("No MaAsLin2 48H RDC vs SDC results available.\n")
}
```

### 8.7.4 MaAsLin2 (Both Timepoints, RDC vs SDC)

```{r maaslin2_both_timepoints_rdc_sdc}
maaslin_both_path <- file.path(outputs_dir, "maaslin2_both_timepoints_rdc_vs_sdc_results.rds")

# Subset to both timepoints and only RDC vs SDC
ps_genus_both_rdc_sdc_maaslin <- subset_samples(ps_genus_filtered, 
                                               carbohydrate %in% c("RDC", "SDC"))

cat("Both timepoints RDC vs SDC samples for MaAsLin2:", nsamples(ps_genus_both_rdc_sdc_maaslin), "\n")
cat("Sample distribution by carbohydrate:\n")
print(table(sample_data(ps_genus_both_rdc_sdc_maaslin)$carbohydrate))
cat("Sample distribution by timepoint:\n")
print(table(sample_data(ps_genus_both_rdc_sdc_maaslin)$Timepoint))

# Prepare data for MaAsLin2
maaslin_both_counts <- as.data.frame(otu_table(ps_genus_both_rdc_sdc_maaslin))
maaslin_both_meta <- as(sample_data(ps_genus_both_rdc_sdc_maaslin), "data.frame")
maaslin_both_meta$SampleID <- rownames(maaslin_both_meta)

# Run MaAsLin2 analysis with timepoint as covariate
maaslin_both_fit <- cache_rds(
  maaslin_both_path,
  {
    Maaslin2::Maaslin2(
      input_data = maaslin_both_counts,
      input_metadata = maaslin_both_meta,
      output = file.path(outputs_dir, "maaslin2_both_timepoints_rdc_vs_sdc"),
      min_prevalence = prevalence_cutoff,
      normalization = "TSS",
      transform = "LOG",
      analysis_method = "LM",
      fixed_effects = c("carbohydrate", "timepoint"),
      random_effects = c("subject_id"),
      plot_heatmap = FALSE,
      plot_scatter = FALSE
    )
  }
)

maaslin_both_tbl <- maaslin_both_fit$results %>%
  as_tibble() %>%
  dplyr::mutate(q = fdr_adjust(qval)) %>%
  dplyr::arrange(q)

# Show results for carbohydrate effect specifically
maaslin_both_carb <- maaslin_both_tbl %>% 
  dplyr::filter(metadata == "carbohydrate") %>%
  dplyr::filter(q < 0.1)

maaslin_both_carb
```

### 8.7.5 MaAsLin2 Both Timepoints RDC vs SDC Visualization

```{r maaslin_both_viz, fig.width = 20, fig.height = 16}
# Load required library for text repelling
library(ggrepel)

if (nrow(maaslin_both_tbl) > 0) {
  # Get taxonomy information for ASV to genus/species conversion
  tax_table_df_both <- as.data.frame(tax_table(ps_genus_both_rdc_sdc_maaslin))
  tax_table_df_both$feature <- rownames(tax_table_df_both)
  
  # Create taxonomic display names
  tax_table_df_both <- tax_table_df_both %>%
    dplyr::mutate(
      # Create genus_species display name
      genus_species = case_when(
        !is.na(Species) & Species != "Unassigned" & Species != "" ~ paste0(Genus, " ", Species),
        !is.na(Genus) & Genus != "Unassigned" & Genus != "" ~ paste0(Genus, " sp."),
        !is.na(Family) & Family != "Unassigned" & Family != "" ~ paste0(Family, " (family)"),
        !is.na(Order) & Order != "Unassigned" & Order != "" ~ paste0(Order, " (order)"),
        TRUE ~ feature  # Fallback to ASV ID if no good taxonomy
      )
    )
  
  # Filter for significant carbohydrate effects and add taxonomy
  maaslin_both_sig <- maaslin_both_tbl %>% 
    dplyr::filter(metadata == "carbohydrate") %>%
    dplyr::filter(q < 0.1) %>%
    dplyr::left_join(tax_table_df_both %>% dplyr::select(feature, genus_species), by = "feature") %>%
    dplyr::mutate(
      # Create significance levels
      significance = case_when(
        q < 0.001 ~ "***",
        q < 0.01 ~ "**", 
        q < 0.05 ~ "*",
        TRUE ~ ""
      ),
      # Create display labels with taxonomic names
      display_label = paste0(genus_species, " (", metadata, "=", value, ")"),
      # Create combined label with significance
      full_label = paste0(display_label, significance),
      # Create direction for bar colors (positive for SDC, negative for RDC)
      direction = ifelse(value == "SDC", 1, -1)
    )
  
  if (nrow(maaslin_both_sig) > 0) {
    # Plot 1: Horizontal bar chart of significant carbohydrate associations
    p1_both <- ggplot(maaslin_both_sig, aes(x = coef * direction, y = reorder(full_label, coef))) +
      geom_col(aes(fill = value), alpha = 0.8) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
      scale_fill_manual(
        values = c("RDC" = "#2E86AB", "SDC" = "#A23B72"),
        name = "Enriched in",
        labels = c("RDC" = "RDC", "SDC" = "SDC")
      ) +
      labs(
        title = "MaAsLin2 Results: RDC vs SDC (Both Timepoints)",
        subtitle = "Carbohydrate Effect Across 0H and 48H Timepoints",
        x = "Coefficient (Effect Size)",
        y = "Taxa (Genus level)",
        caption = "*** q<0.001, ** q<0.01, * q<0.05"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12, color = "gray60"),
        axis.text.y = element_text(size = 10),
        axis.text.x = element_text(size = 10),
        axis.title = element_text(size = 11, face = "bold"),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        plot.caption = element_text(size = 9, color = "gray50")
      )
    
    print(p1_both)
    
    # Plot 2: Volcano plot for carbohydrate effects
    maaslin_both_carb_with_tax <- maaslin_both_tbl %>%
      dplyr::filter(metadata == "carbohydrate") %>%
      dplyr::left_join(tax_table_df_both %>% dplyr::select(feature, genus_species), by = "feature")
    
    p2_both <- ggplot(maaslin_both_carb_with_tax, aes(x = coef, y = -log10(q))) +
      geom_point(aes(color = q < 0.1), alpha = 0.6, size = 2) +
      geom_text_repel(
        data = maaslin_both_carb_with_tax %>% filter(q < 0.1),
        aes(label = genus_species),
        size = 3,
        max.overlaps = 15,
        box.padding = 0.5,
        point.padding = 0.3,
        segment.color = "gray50",
        segment.size = 0.3,
        force = 2
      ) +
      geom_hline(yintercept = -log10(0.1), linetype = "dashed", color = "red", alpha = 0.7) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
      scale_color_manual(
        values = c("FALSE" = "gray70", "TRUE" = "#E31A1C"),
        name = "Significant",
        labels = c("FALSE" = "q ≥ 0.1", "TRUE" = "q < 0.1")
      ) +
      labs(
        title = "MaAsLin2 Volcano Plot: RDC vs SDC (Both Timepoints)",
        subtitle = "Carbohydrate Effect Size vs Statistical Significance",
        x = "Coefficient (Effect Size)",
        y = "-log10(q-value)",
        caption = "Red dashed line: q = 0.1 threshold"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12, color = "gray60"),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 11, face = "bold"),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"),
        plot.caption = element_text(size = 9, color = "gray50")
      )
    
    print(p2_both)
    
    # Plot 3: Top effects by magnitude
    top_effects_both <- maaslin_both_sig %>%
      dplyr::slice_max(abs(coef), n = 15) %>%
      dplyr::arrange(coef)
    
    p3_both <- ggplot(top_effects_both, aes(x = coef * direction, y = reorder(genus_species, coef))) +
      geom_col(aes(fill = value), alpha = 0.8) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
      scale_fill_manual(
        values = c("RDC" = "#2E86AB", "SDC" = "#A23B72"),
        name = "Enriched in"
      ) +
      labs(
        title = "MaAsLin2: Top 15 Carbohydrate Effects (Both Timepoints)",
        subtitle = "Largest effect sizes by magnitude across 0H and 48H",
        x = "Coefficient (Effect Size)",
        y = "Taxa (Genus level)"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12, color = "gray60"),
        axis.text.y = element_text(size = 10),
        axis.text.x = element_text(size = 10),
        axis.title = element_text(size = 11, face = "bold"),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank()
      )
    
    print(p3_both)
    
    # Plot 4: Comparison of timepoint effects
    timepoint_effects <- maaslin_both_tbl %>%
      dplyr::filter(metadata == "timepoint") %>%
      dplyr::filter(q < 0.1) %>%
      dplyr::left_join(tax_table_df_both %>% dplyr::select(feature, genus_species), by = "feature") %>%
      dplyr::mutate(
        significance = case_when(
          q < 0.001 ~ "***",
          q < 0.01 ~ "**", 
          q < 0.05 ~ "*",
          TRUE ~ ""
        ),
        display_label = paste0(genus_species, " (timepoint=", value, ")"),
        full_label = paste0(display_label, significance)
      )
    
    if (nrow(timepoint_effects) > 0) {
      p4_both <- ggplot(timepoint_effects, aes(x = coef, y = reorder(full_label, coef))) +
        geom_col(aes(fill = value), alpha = 0.8) +
        geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
        scale_fill_manual(
          values = c("0H" = "#FF6B6B", "48H" = "#4ECDC4"),
          name = "Timepoint"
        ) +
        labs(
          title = "MaAsLin2: Timepoint Effects (RDC vs SDC Analysis)",
          subtitle = "Taxa that change significantly between 0H and 48H",
          x = "Coefficient (Effect Size)",
          y = "Taxa (Genus level)",
          caption = "*** q<0.001, ** q<0.01, * q<0.05"
        ) +
        theme_minimal() +
        theme(
          plot.title = element_text(size = 14, face = "bold"),
          plot.subtitle = element_text(size = 12, color = "gray60"),
          axis.text.y = element_text(size = 10),
          axis.text.x = element_text(size = 10),
          axis.title = element_text(size = 11, face = "bold"),
          legend.position = "bottom",
          legend.title = element_text(face = "bold"),
          panel.grid.major.y = element_blank(),
          panel.grid.minor = element_blank(),
          plot.caption = element_text(size = 9, color = "gray50")
        )
      
      print(p4_both)
      
      # Save timepoint effects plot
      ggsave(file.path(outputs_dir, "maaslin2_both_timepoints_timepoint_effects.png"), 
             p4_both, width = 12, height = 10, dpi = 300)
    }
    
    # Save plots
    ggsave(file.path(outputs_dir, "maaslin2_both_timepoints_rdc_vs_sdc_barplot.png"), 
           p1_both, width = 12, height = 10, dpi = 300)
    ggsave(file.path(outputs_dir, "maaslin2_both_timepoints_rdc_vs_sdc_volcano.png"), 
           p2_both, width = 10, height = 8, dpi = 300)
    ggsave(file.path(outputs_dir, "maaslin2_both_timepoints_rdc_vs_sdc_top_effects.png"), 
           p3_both, width = 12, height = 10, dpi = 300)
    
    # Summary statistics
    cat("\nMaAsLin2 Both Timepoints RDC vs SDC Summary:\n")
    cat("- Total features tested:", nrow(maaslin_both_tbl), "\n")
    cat("- Significant carbohydrate associations (q < 0.1):", nrow(maaslin_both_sig), "\n")
    cat("- Taxa enriched in RDC:", sum(maaslin_both_sig$value == "RDC"), "\n")
    cat("- Taxa enriched in SDC:", sum(maaslin_both_sig$value == "SDC"), "\n")
    cat("- Mean absolute coefficient:", round(mean(abs(maaslin_both_sig$coef)), 3), "\n")
    cat("- Max absolute coefficient:", round(max(abs(maaslin_both_sig$coef)), 3), "\n")
    
    if (nrow(timepoint_effects) > 0) {
      cat("- Significant timepoint effects (q < 0.1):", nrow(timepoint_effects), "\n")
      cat("- Taxa enriched at 0H:", sum(timepoint_effects$value == "0H"), "\n")
      cat("- Taxa enriched at 48H:", sum(timepoint_effects$value == "48H"), "\n")
    }
    
  } else {
    cat("No significant MaAsLin2 carbohydrate effects found for RDC vs SDC comparison across both timepoints (q < 0.1).\n")
  }
  
} else {
  cat("No MaAsLin2 both timepoints RDC vs SDC results available.\n")
}
```

### 8.8 ADAPT

```{r adapt}
adapt_path <- file.path(outputs_dir, "adapt_results.rds")

adapt_res <- cache_rds(
  adapt_path,
  {
    tryCatch({
      ADAPT::adapt(
        input_data = ps_genus_filtered,
        cond.var = "carbohydrate",
        base.cond = "No Added Carb",
        adj.var = "timepoint"
      )
    }, error = function(e) {
      message("ADAPT failed: ", e$message)
      NULL
    })
  }
)

adapt_tbl <- if (!is.null(adapt_res)) {
  signals <- adapt_res@signal
  details <- adapt_res@details
  tibble(
    taxon = signals,
    log10_fc = details[signals, "log10foldchange"],
    direction = if_else(log10_fc > 0, "Enriched", "Depleted")
  )
} else {
  tibble()
}

adapt_tbl
```

### 8.9 Random Forest Classification (48 h carbohydrate)

```{r random-forest}
rf_path <- file.path(outputs_dir, "random_forest_markers.rds")

rf_markers <- cache_rds(
  rf_path,
  {
    set.seed(12345)
    rf_res <- tryCatch({
      # Subset to only RDC and SDC at 48H for two-group comparison
      ps_genus_48h_rdc_sdc <- subset_samples(ps_genus_48h, 
                                           carbohydrate %in% c("RDC", "SDC"))
      
      cat("48H RDC vs SDC samples for Random Forest:", nsamples(ps_genus_48h_rdc_sdc), "\n")
      cat("Sample distribution by carbohydrate:\n")
      print(table(sample_data(ps_genus_48h_rdc_sdc)$carbohydrate))
      
      microbiomeMarker::run_sl(
        ps_genus_48h_rdc_sdc,
        group = "carbohydrate",
        taxa_rank = "Genus",
        norm = "TSS",
        method = "RF",
        nfolds = 5,
        nrepeats = 1,
        top_n = 30
      )
    }, error = function(e) {
      message("Random forest failed: ", e$message)
      NULL
    })

    if (!is.null(rf_res)) {
      microbiomeMarker::marker_table(rf_res) %>%
        as_tibble() %>%
        dplyr::arrange(desc(importance))
    } else {
      NULL
    }
  }
)

rf_markers <- rf_markers %||% tibble()

rf_markers
```

### 8.10 Cross-Method Signal Concordance

```{r da-consensus}
# Get taxonomy information for ASV to genus conversion
tax_table_df <- as.data.frame(tax_table(ps_genus_48h))

# Function to convert ASV to genus name
asv_to_genus <- function(asv_ids) {
  if (length(asv_ids) == 0) return(character())
  
  genus_names <- sapply(asv_ids, function(asv_id) {
    if (asv_id %in% rownames(tax_table_df)) {
      genus <- tax_table_df[asv_id, "Genus"]
      if (!is.na(genus) && genus != "uncultured" && genus != "") {
        return(genus)
      } else {
        return(asv_id)  # Fallback to ASV if no genus info
      }
    } else {
      return(asv_id)  # Fallback to ASV if not found
    }
  })
  return(unname(genus_names))
}

da_lists <- list(
  Corncob = {
    if (nrow(corncob_results) > 0 && "q_value" %in% names(corncob_results)) {
      asv_ids <- pull_safe(corncob_results %>% dplyr::filter(q_value < 0.1), "taxon")
      asv_to_genus(asv_ids)
    } else {
      character()
    }
  },
  ANCOMBC2 = {
    if (nrow(ancombc2_tbl) > 0) {
      col <- first_match_col(ancombc2_tbl, c("taxa_id", "taxon"))
      if (!is.null(col) && "q" %in% names(ancombc2_tbl)) {
        asv_ids <- pull_safe(ancombc2_tbl %>% dplyr::filter(q < 0.1), col)
      } else {
        asv_ids <- pull_safe(ancombc2_tbl, col)
      }
      asv_to_genus(asv_ids)
    } else {
      character()
    }
  },
  ANCOMBC48 = {
    if (nrow(ancombc48_tbl) > 0 && "q_value" %in% names(ancombc48_tbl)) {
      asv_ids <- pull_safe(ancombc48_tbl %>% dplyr::filter(q_value < 0.1), "taxon")
      asv_to_genus(asv_ids)
    } else {
      character()
    }
  },
  ALDEx2 = {
    if (nrow(aldex_tbl) > 0 && "q" %in% names(aldex_tbl)) {
      asv_ids <- pull_safe(aldex_tbl %>% dplyr::filter(q < 0.1), "taxon")
      asv_to_genus(asv_ids)
    } else {
      character()
    }
  },
  LEfSe = {
    asv_ids <- pull_safe(lefse_tbl, "feature")
    asv_to_genus(asv_ids)
  },
  MaAsLin2 = {
    if (nrow(maaslin_tbl) > 0 && "q" %in% names(maaslin_tbl)) {
      asv_ids <- pull_safe(maaslin_tbl %>% dplyr::filter(q < 0.1), "feature")
      asv_to_genus(asv_ids)
    } else {
      character()
    }
  },
  ADAPT = {
    asv_ids <- pull_safe(adapt_tbl, "taxon")
    asv_to_genus(asv_ids)
  },
  RandomForest = {
    if (nrow(rf_markers) > 0) {
      col <- first_match_col(rf_markers, c("feature", "taxa", "marker"))
      if (!is.null(col)) {
        asv_ids <- pull_safe(rf_markers %>% slice_head(n = 15), col)
        asv_to_genus(asv_ids)
      } else {
        character()
      }
    } else {
      character()
    }
  }
)

da_long <- enframe(da_lists, name = "method", value = "taxa") %>%
  tidyr::unnest(taxa) %>%
  dplyr::distinct(method, taxa) %>%
  dplyr::count(taxa, method) %>%
  dplyr::mutate(present = 1)

consensus_matrix <- da_long %>%
  pivot_wider(names_from = method, values_from = present, values_fill = 0)

consensus_matrix %>% dplyr::arrange(desc(rowSums(across(where(is.numeric)))))
```

```{r da-interpretation, results='asis'}
consensus_summary <- consensus_matrix %>%
  dplyr::mutate(method_count = rowSums(across(where(is.numeric)))) %>%
  dplyr::arrange(desc(method_count))

if (nrow(consensus_summary) == 0) {
  cat("No overlapping differential taxa detected across the configured methods at q < 0.10.")
} else {
  top_taxa <- consensus_summary %>%
    slice_max(order_by = method_count, n = 10) %>%
    dplyr::mutate(text = glue::glue("{taxa}: detected by {method_count} methods")) %>%
    dplyr::pull(text)
  cat(glue::glue_collapse(top_taxa, sep = "\n"))
}
```
```
## 9. Summary and Next Steps

```{r final-summary, results='asis'}
alpha_highlights <- alpha_significance %>%
  dplyr::mutate(text = glue::glue("{metric}: {sig_terms}")) %>%
  dplyr::pull(text) %>%
  glue::glue_collapse(sep = "; ")

permanova_highlights <- permanova_results %>%
  dplyr::filter(term != "Residual", q_value < 0.1) %>%
  dplyr::mutate(text = glue::glue("{stringr::str_to_title(distance)} {term}")) %>%
  dplyr::pull(text) %>%
  glue::glue_collapse(sep = "; ")

if (permanova_highlights == "") permanova_highlights <- "No beta-diversity terms at q < 0.10"

consensus_highlights <- if (exists("consensus_summary") && nrow(consensus_summary) > 0) {
  consensus_summary %>%
    slice_max(order_by = method_count, n = 5) %>%
    dplyr::mutate(text = glue::glue("{taxa} ({method_count} methods)")) %>%
    dplyr::pull(text) %>%
    glue::glue_collapse(sep = "; ")
} else {
  "No shared differential taxa"
}

cat(glue::glue(
  "- Alpha diversity signals: {alpha_highlights}\n- Beta diversity drivers: {permanova_highlights}\n- Differential abundance consensus: {consensus_highlights}"
))
```

**Suggested extensions**

1. Validate differential findings with targeted modeling (e.g., mixed-effects regressions on log-relative abundances) for key taxa.
2. Incorporate functional pathway predictions (PICRUSt2, HUMAnN) to relate taxonomic shifts to metabolic potential.
3. Explore subject-level trajectories using longitudinal models (e.g., smoothing splines) to capture individual response heterogeneity.
```
